---
title: "R Fundamentals"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Installation and Setup

## Installing R

R can be downloaded from [CRAN](https://cran.r-project.org/). Choose the appropriate version for your operating system:

- **Windows**: Download the `.exe` installer
- **macOS**: Download the `.pkg` installer
- **Linux**: Use package manager or compile from source

## Installing RStudio

RStudio is the most popular IDE for R. Download from [RStudio](https://www.rstudio.com/products/rstudio/download/).

## RStudio Interface

The RStudio interface consists of four main panes:

1. **Source Editor**: Write and edit R scripts
2. **Console**: Execute R commands interactively
3. **Environment/History**: View objects and command history
4. **Files/Plots/Packages/Help**: Navigate files, view plots, manage packages

# 2. Basic Syntax

## Comments

```{r}
# This is a single-line comment

# R does not have multi-line comments
# Use multiple single-line comments instead
```

## Basic Arithmetic Operations

```{r}
# Addition
5 + 3

# Subtraction
10 - 4

# Multiplication
6 * 7

# Division
20 / 4

# Exponentiation
2^3

# Modulo (remainder)
17 %% 5

# Integer division
17 %/% 5
```

## Comparison Operators

```{r}
# Equal to
5 == 5

# Not equal to
5 != 3

# Greater than
10 > 5

# Less than
3 < 7

# Greater than or equal to
5 >= 5

# Less than or equal to
4 <= 6
```

## Logical Operators

```{r}
# AND
TRUE & FALSE

# OR
TRUE | FALSE

# NOT
!TRUE

# Element-wise AND
c(TRUE, FALSE, TRUE) & c(TRUE, TRUE, FALSE)

# Element-wise OR
c(TRUE, FALSE, TRUE) | c(TRUE, TRUE, FALSE)

# Short-circuit AND (only evaluates second if first is TRUE)
TRUE && FALSE

# Short-circuit OR (only evaluates second if first is FALSE)
TRUE || FALSE
```

# 3. Variables and Assignment

## Variable Assignment

```{r}
# Using <- (preferred in R)
x <- 10

# Using = (also works, but <- is conventional)
y = 20

# Using -> (right assignment, less common)
30 -> z

# Print values
x
y
z
```

## Variable Naming Rules

```{r}
# Valid variable names
my_variable <- 5
myVariable <- 10
my.variable <- 15
MyVariable2 <- 20

# Invalid variable names (will cause errors)
# 2myVariable <- 25  # Cannot start with a number
# my-variable <- 30  # Hyphens not allowed
# _myVariable <- 35  # Cannot start with underscore
```

## Variable Types

R automatically determines variable types:

```{r}
# Numeric (double)
num_var <- 42.5
class(num_var)

# Integer (must specify with L)
int_var <- 42L
class(int_var)

# Character
char_var <- "Hello, R!"
class(char_var)

# Logical
logical_var <- TRUE
class(logical_var)
```

# 4. Control Structures

## If-Else Statements

```{r}
# Simple if statement
x <- 10
if (x > 5) {
  print("x is greater than 5")
}

# If-else statement
x <- 3
if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is less than or equal to 5")
}

# If-else if-else chain
score <- 75
if (score >= 90) {
  grade <- "A"
} else if (score >= 80) {
  grade <- "B"
} else if (score >= 70) {
  grade <- "C"
} else if (score >= 60) {
  grade <- "D"
} else {
  grade <- "F"
}
print(paste("Grade:", grade))

# Vectorized ifelse (more efficient)
scores <- c(95, 82, 68, 55, 91)
grades <- ifelse(scores >= 90, "A",
          ifelse(scores >= 80, "B",
          ifelse(scores >= 70, "C",
          ifelse(scores >= 60, "D", "F"))))
grades
```

## Switch Statement

```{r}
# Switch with numeric index
x <- 2
result <- switch(x,
  "first",
  "second",
  "third",
  "default"
)
result

# Switch with character matching
day <- "Monday"
activity <- switch(day,
  "Monday" = "Start of work week",
  "Friday" = "End of work week",
  "Saturday" = "Weekend!",
  "Sunday" = "Weekend!",
  "Midweek"
)
activity
```

## For Loops

```{r}
# Simple for loop
for (i in 1:5) {
  print(i)
}

# Loop over vector
fruits <- c("apple", "banana", "cherry")
for (fruit in fruits) {
  print(paste("I like", fruit))
}

# Nested loops
for (i in 1:3) {
  for (j in 1:3) {
    print(paste(i, "*", j, "=", i*j))
  }
}

# Using seq_along (best practice for vectors)
for (i in seq_along(fruits)) {
  print(paste(i, ":", fruits[i]))
}
```

## While Loops

```{r}
# Simple while loop
count <- 1
while (count <= 5) {
  print(count)
  count <- count + 1
}

# While loop with break
x <- 1
while (TRUE) {
  print(x)
  x <- x + 1
  if (x > 5) {
    break
  }
}
```

## Repeat Loops

```{r}
# Repeat loop (infinite loop with break)
x <- 1
repeat {
  print(x)
  x <- x + 1
  if (x > 5) {
    break
  }
}
```

## Break and Next

```{r}
# Using break to exit loop early
for (i in 1:10) {
  if (i > 5) {
    break
  }
  print(i)
}

# Using next to skip iteration
for (i in 1:10) {
  if (i %% 2 == 0) {
    next  # Skip even numbers
  }
  print(i)
}
```

# 5. Functions

## Creating Basic Functions

```{r}
# Simple function
square <- function(x) {
  return(x^2)
}
square(5)

# Function without explicit return (last expression is returned)
cube <- function(x) {
  x^3
}
cube(4)

# Function with multiple arguments
add <- function(a, b) {
  a + b
}
add(3, 7)

# Function with default arguments
greet <- function(name, greeting = "Hello") {
  paste(greeting, name, "!")
}
greet("Alice")
greet("Bob", "Hi")
```

## Functions with Multiple Returns

```{r}
# Return multiple values using a list
calculate_stats <- function(x) {
  list(
    mean = mean(x),
    median = median(x),
    sd = sd(x),
    min = min(x),
    max = max(x)
  )
}

data <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
stats <- calculate_stats(data)
stats
```

## Variable Arguments

```{r}
# Using ... for variable number of arguments
sum_all <- function(...) {
  args <- list(...)
  sum(unlist(args))
}
sum_all(1, 2, 3, 4, 5)

# Combining ... with named arguments
print_with_separator <- function(sep = "-", ...) {
  args <- list(...)
  paste(unlist(args), collapse = sep)
}
print_with_separator(" | ", "apple", "banana", "cherry")
```

## Anonymous Functions

```{r}
# Traditional anonymous function
sapply(1:5, function(x) x^2)

# R 4.1+ shorthand syntax
sapply(1:5, \(x) x^2)
```

## Function Scope

```{r}
# Local vs global scope
x <- 10  # Global variable

test_scope <- function() {
  x <- 5  # Local variable
  print(paste("Inside function:", x))
}

test_scope()
print(paste("Outside function:", x))

# Accessing global variable with <<-
counter <- 0
increment <- function() {
  counter <<- counter + 1
  counter
}
increment()
increment()
counter
```

# 6. Package Management

## Installing Packages

```{r eval=FALSE}
# Install from CRAN
install.packages("ggplot2")

# Install multiple packages
install.packages(c("dplyr", "tidyr", "readr"))

# Install from Bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("DESeq2")

# Install from GitHub
# install.packages("devtools")
devtools::install_github("username/packagename")
```

## Loading Packages

```{r eval=FALSE}
# Load package
library(ggplot2)

# Load package without startup messages
suppressPackageStartupMessages(library(dplyr))

# Require package (returns FALSE if not available)
if (require("ggplot2")) {
  print("Package loaded successfully")
} else {
  install.packages("ggplot2")
  library(ggplot2)
}

# Use function without loading entire package
dplyr::filter(mtcars, mpg > 20)
```

## Managing Packages

```{r eval=FALSE}
# List installed packages
installed.packages()

# Update packages
update.packages()

# Update specific package
update.packages("ggplot2")

# Remove package
remove.packages("packagename")

# Check package version
packageVersion("ggplot2")

# View package help
help(package = "ggplot2")
```

## Common Essential Packages

```{r eval=FALSE}
# Tidyverse (collection of packages)
install.packages("tidyverse")
# Includes: ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats

# Data manipulation
install.packages("data.table")

# Bioinformatics
BiocManager::install(c("Biostrings", "GenomicRanges", "DESeq2"))

# Statistical analysis
install.packages(c("lme4", "survival", "MASS"))

# Machine learning
install.packages(c("caret", "randomForest", "xgboost"))

# Visualization
install.packages(c("plotly", "pheatmap", "ComplexHeatmap"))
```

# 7. Getting Help

## Help Functions

```{r eval=FALSE}
# Get help for a function
help(mean)
?mean

# Search for help
help.search("linear model")
??lm

# Get examples
example(mean)

# View function arguments
args(lm)

# View vignettes
vignette()
vignette("dplyr")

# Access package documentation
help(package = "ggplot2")
```

## Useful R Functions for Exploration

```{r}
# Get structure of object
str(mtcars)

# Summary statistics
summary(mtcars)

# View first few rows
head(mtcars)

# View last few rows
tail(mtcars)

# Dimensions
dim(mtcars)

# Column names
names(mtcars)

# Row names
rownames(mtcars)

# Check class/type
class(mtcars)
typeof(mtcars$mpg)
```

# 8. Working Directory

## Managing Working Directory

```{r eval=FALSE}
# Get current working directory
getwd()

# Set working directory
setwd("/path/to/directory")

# List files in current directory
list.files()

# List files with pattern
list.files(pattern = "\\.csv$")

# Check if file exists
file.exists("data.csv")

# Create directory
dir.create("new_folder")
```

# 9. Best Practices

## Code Style

```{r eval=FALSE}
# Use meaningful variable names
patient_age <- 45  # Good
x <- 45  # Bad

# Use <- for assignment (not =)
result <- compute_value()  # Good
result = compute_value()  # Works but not conventional

# Use spacing for readability
mean(c(1, 2, 3, 4, 5))  # Good
mean(c(1,2,3,4,5))  # Less readable

# Indent code blocks
if (condition) {
  do_something()
  do_another_thing()
}

# Comment your code
# Calculate mean expression across samples
mean_expression <- rowMeans(expression_matrix)
```

## Common Pitfalls

```{r}
# Be careful with == for floating point
0.1 + 0.2 == 0.3  # FALSE due to floating point precision
all.equal(0.1 + 0.2, 0.3)  # TRUE (with tolerance)

# Use identical() for exact comparison
identical(0.1 + 0.2, 0.3)  # FALSE

# Recycling in vectors
c(1, 2, 3, 4) + c(1, 2)  # Second vector is recycled: 1,2,1,2

# NA vs NULL
x <- c(1, 2, NA, 4)
y <- c(1, 2, NULL, 4)
length(x)  # 4
length(y)  # 3 (NULL is ignored)
```

# 10. Session Information

Always include session information in your reports for reproducibility:

```{r}
sessionInfo()
```

# Summary

This document covered R fundamentals including:

- Installation and setup
- Basic syntax and operators
- Variables and data types
- Control structures (if/else, loops)
- Functions and scope
- Package management
- Getting help
- Best practices

**Next Steps**: Proceed to Data Structures to learn about vectors, lists, data frames, and more complex R objects.

# References

- [R for Data Science](https://r4ds.had.co.nz/)
- [Advanced R](https://adv-r.hadley.nz/)
- [R Documentation](https://www.r-project.org/other-docs.html)
- [CRAN Task Views](https://cran.r-project.org/web/views/)
