---
title: "Data Structures in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview of R Data Structures

R has several fundamental data structures that differ in:

1. **Dimensionality**: 1D, 2D, or nD
2. **Homogeneity**: All elements same type (atomic) vs. mixed types (heterogeneous)

| Dimension | Homogeneous | Heterogeneous |
|-----------|-------------|---------------|
| 1D        | Atomic Vector | List |
| 2D        | Matrix      | Data Frame |
| nD        | Array       | - |

# 1. Vectors

Vectors are the most basic data structure in R. Everything in R is built from vectors.

## Atomic Vectors

There are six types of atomic vectors: logical, integer, double, character, complex, and raw.

### Numeric Vectors (Double)

```{r}
# Create numeric vector
x <- c(1.5, 2.7, 3.9, 4.2)
x

# Check type
typeof(x)
class(x)

# Check if numeric
is.numeric(x)
is.double(x)

# Vector arithmetic
x * 2
x + 10
sqrt(x)
```

### Integer Vectors

```{r}
# Create integer vector (must use L suffix)
y <- c(1L, 2L, 3L, 4L)
y

typeof(y)
is.integer(y)

# Coercion from numeric to integer
z <- as.integer(c(1.5, 2.7, 3.9))
z
```

### Character Vectors

```{r}
# Create character vector
genes <- c("TP53", "BRCA1", "EGFR", "KRAS")
genes

typeof(genes)
is.character(genes)

# String operations
nchar(genes)  # Number of characters
toupper(genes)
tolower(genes)
paste("Gene:", genes)
```

### Logical Vectors

```{r}
# Create logical vector
flags <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
flags

typeof(flags)
is.logical(flags)

# Logical operations
!flags  # NOT
flags & c(TRUE, TRUE, FALSE, FALSE, TRUE)  # AND
flags | c(FALSE, FALSE, FALSE, FALSE, FALSE)  # OR

# Sum and mean of logical vectors (TRUE = 1, FALSE = 0)
sum(flags)  # Count TRUEs
mean(flags)  # Proportion of TRUEs
```

## Vector Creation Methods

```{r}
# Using c() - combine
v1 <- c(1, 2, 3, 4, 5)

# Using : operator for sequences
v2 <- 1:10
v2

# Using seq() for more control
v3 <- seq(from = 0, to = 1, by = 0.1)
v3

v4 <- seq(from = 0, to = 1, length.out = 11)
v4

# Using rep() for repetition
v5 <- rep(1, times = 5)
v5

v6 <- rep(c(1, 2, 3), times = 3)
v6

v7 <- rep(c(1, 2, 3), each = 3)
v7

# Vector of specific length
v8 <- vector("numeric", length = 5)
v8

# Random vectors
set.seed(123)
rnorm(5)  # Normal distribution
runif(5)  # Uniform distribution
sample(1:10, 5)  # Random sample
```

## Vector Properties

```{r}
x <- c(10, 20, 30, 40, 50)

# Length
length(x)

# Type and class
typeof(x)
class(x)

# Structure
str(x)

# Names
names(x) <- c("a", "b", "c", "d", "e")
x
names(x)
```

## Vector Indexing and Subsetting

```{r}
x <- c(10, 20, 30, 40, 50)

# By position (1-indexed)
x[1]
x[3]
x[c(1, 3, 5)]

# By negative indices (exclude)
x[-1]
x[-c(1, 3)]

# By logical vector
x[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
x[x > 25]

# By names
names(x) <- c("a", "b", "c", "d", "e")
x["a"]
x[c("a", "c", "e")]

# Modifying elements
x[1] <- 100
x

x[x < 30] <- 0
x
```

## Vector Coercion

When you combine different types, R coerces to the most flexible type:
logical → integer → double → character

```{r}
# Automatic coercion
c(1, 2, "3")  # Becomes character
c(TRUE, FALSE, 1)  # Becomes numeric
c(1L, 2.5)  # Becomes double

# Explicit coercion
as.numeric(c("1", "2", "3"))
as.character(c(1, 2, 3))
as.logical(c(0, 1, 2, 0))  # 0 = FALSE, non-zero = TRUE
```

# 2. Matrices

Matrices are 2D homogeneous data structures.

## Creating Matrices

```{r}
# From vector with dimensions
m1 <- matrix(1:12, nrow = 3, ncol = 4)
m1

# By rows instead of columns
m2 <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
m2

# From combining vectors
v1 <- c(1, 2, 3)
v2 <- c(4, 5, 6)
v3 <- c(7, 8, 9)

m3 <- rbind(v1, v2, v3)  # Row bind
m3

m4 <- cbind(v1, v2, v3)  # Column bind
m4

# Identity matrix
diag(5)

# Matrix of specific values
matrix(0, nrow = 3, ncol = 4)
```

## Matrix Properties

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)

# Dimensions
dim(m)
nrow(m)
ncol(m)

# Row and column names
rownames(m) <- c("row1", "row2", "row3")
colnames(m) <- c("col1", "col2", "col3", "col4")
m

# Check if matrix
is.matrix(m)
class(m)
```

## Matrix Indexing

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)

# Single element
m[2, 3]

# Entire row
m[2, ]

# Entire column
m[, 3]

# Multiple rows/columns
m[c(1, 3), c(2, 4)]

# Logical indexing
m[m > 6]

# With names
rownames(m) <- c("A", "B", "C")
colnames(m) <- c("W", "X", "Y", "Z")
m["B", "Y"]
```

## Matrix Operations

```{r}
m1 <- matrix(1:4, nrow = 2)
m2 <- matrix(5:8, nrow = 2)

# Element-wise operations
m1 + m2
m1 * m2  # Element-wise multiplication
m1 / m2

# Matrix multiplication
m1 %*% m2

# Transpose
t(m1)

# Determinant
det(m1)

# Inverse
solve(m1)

# Eigenvalues and eigenvectors
eigen(m1)

# Row and column operations
colSums(m1)
rowSums(m1)
colMeans(m1)
rowMeans(m1)

# Apply function to rows or columns
apply(m1, 1, sum)  # Apply to rows (1)
apply(m1, 2, sum)  # Apply to columns (2)
```

# 3. Arrays

Arrays are nD extensions of matrices.

## Creating Arrays

```{r}
# 3D array
arr <- array(1:24, dim = c(3, 4, 2))
arr

# Check dimensions
dim(arr)

# Array indexing
arr[2, 3, 1]
arr[, , 1]  # First "slice"
arr[, , 2]  # Second "slice"
```

# 4. Lists

Lists can contain elements of different types and structures.

## Creating Lists

```{r}
# Simple list
my_list <- list(
  numbers = 1:5,
  characters = c("a", "b", "c"),
  logical = c(TRUE, FALSE),
  matrix = matrix(1:6, nrow = 2)
)
my_list

# List structure
str(my_list)

# Unnamed list
list(1:5, "a", TRUE)
```

## List Indexing

```{r}
# Single bracket [ ] returns a list
my_list[1]
class(my_list[1])

# Double bracket [[ ]] returns the element itself
my_list[[1]]
class(my_list[[1]])

# Dollar sign $ for named elements
my_list$numbers
my_list$matrix

# Multiple elements
my_list[c(1, 3)]

# By name
my_list["numbers"]
```

## List Operations

```{r}
# Add element
my_list$new_element <- "added"
my_list

# Remove element
my_list$new_element <- NULL
my_list

# Combine lists
list1 <- list(a = 1, b = 2)
list2 <- list(c = 3, d = 4)
c(list1, list2)

# Length
length(my_list)

# Names
names(my_list)
```

## Nested Lists

```{r}
# Lists within lists
nested <- list(
  level1 = list(
    level2a = c(1, 2, 3),
    level2b = list(
      level3 = "deep"
    )
  )
)

# Access nested elements
nested[[1]][[1]]
nested[[1]][[2]][[1]]
nested$level1$level2b$level3
```

# 5. Data Frames

Data frames are the most common data structure for statistical analysis. They are like matrices but can hold different types in each column.

## Creating Data Frames

```{r}
# From vectors
df <- data.frame(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "David", "Eve"),
  age = c(25, 30, 35, 28, 32),
  score = c(85.5, 92.0, 78.5, 88.0, 95.5),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE)
)
df

# From matrix
mat <- matrix(1:12, nrow = 4)
df_from_mat <- as.data.frame(mat)
df_from_mat

# From list
list_data <- list(
  x = 1:3,
  y = c("a", "b", "c")
)
df_from_list <- as.data.frame(list_data)
df_from_list
```

## Data Frame Properties

```{r}
# Dimensions
dim(df)
nrow(df)
ncol(df)

# Structure
str(df)

# Summary
summary(df)

# Names
names(df)
colnames(df)
rownames(df)

# First and last rows
head(df)
tail(df)
head(df, 3)
```

## Data Frame Indexing

```{r}
# By position
df[1, 2]  # Row 1, column 2
df[1, ]   # Row 1, all columns
df[, 2]   # All rows, column 2

# By name
df$name
df[, "name"]
df["name"]

# Multiple columns
df[, c("name", "age")]
df[c("name", "age")]

# Logical indexing
df[df$age > 30, ]
df[df$score > 85 & df$age < 35, ]

# Subset function
subset(df, age > 30)
subset(df, age > 30, select = c(name, score))
```

## Modifying Data Frames

```{r}
# Add column
df$grade <- ifelse(df$score >= 90, "A",
            ifelse(df$score >= 80, "B", "C"))
df

# Add row
new_row <- data.frame(id = 6, name = "Frank", age = 27,
                      score = 82, passed = TRUE, grade = "B")
df <- rbind(df, new_row)
df

# Remove column
df$grade <- NULL
df

# Rename columns
names(df)[names(df) == "score"] <- "test_score"
df

# Reorder columns
df <- df[, c("id", "name", "age", "test_score", "passed")]
df
```

## Data Frame Operations

```{r}
# Sort by column
df[order(df$age), ]
df[order(-df$test_score), ]  # Descending

# Unique values
unique(df$age)

# Count occurrences
table(df$age)

# Aggregate
aggregate(test_score ~ age, data = df, FUN = mean)
```

# 6. Factors

Factors are used for categorical data.

## Creating Factors

```{r}
# From character vector
blood_type <- c("A", "B", "O", "AB", "A", "O", "B", "A")
blood_factor <- factor(blood_type)
blood_factor

# Check class
class(blood_factor)
is.factor(blood_factor)

# View levels
levels(blood_factor)

# View underlying integer codes
as.numeric(blood_factor)
```

## Ordered Factors

```{r}
# For ordinal data
grade <- c("Good", "Excellent", "Poor", "Good", "Poor", "Excellent")
grade_factor <- factor(grade,
                       levels = c("Poor", "Good", "Excellent"),
                       ordered = TRUE)
grade_factor

# Comparison works with ordered factors
grade_factor[1] > grade_factor[3]
```

## Factor Operations

```{r}
# Change levels
blood_factor <- factor(blood_type)
levels(blood_factor) <- c("Type_A", "Type_AB", "Type_B", "Type_O")
blood_factor

# Add levels
blood_factor <- factor(blood_type, levels = c("A", "B", "AB", "O", "AB-"))
blood_factor

# Drop unused levels
blood_factor <- droplevels(blood_factor)
levels(blood_factor)

# Combine factors
f1 <- factor(c("a", "b", "c"))
f2 <- factor(c("d", "e", "f"))
factor(c(as.character(f1), as.character(f2)))

# Count factor levels
table(blood_factor)
```

## Converting Between Types

```{r}
# Factor to character
as.character(blood_factor)

# Factor to numeric (be careful!)
f <- factor(c("1", "2", "3"))
as.numeric(f)  # Returns internal codes (not what you want)
as.numeric(as.character(f))  # Correct way

# Character to factor
chars <- c("low", "medium", "high")
factor(chars, levels = c("low", "medium", "high"))
```

# 7. Tibbles (Modern Data Frames)

Tibbles are a modern reimagining of data frames from the tidyverse.

```{r}
library(tibble)

# Create tibble
tbl <- tibble(
  x = 1:5,
  y = x^2,
  z = letters[1:5]
)
tbl

# From data frame
as_tibble(df)

# Differences from data.frame:
# 1. Better printing (shows only first 10 rows)
# 2. Never converts strings to factors
# 3. Never changes column names
# 4. Never creates row names
# 5. Stricter subsetting (always returns tibble)
```

# 8. Special Values

## NA (Not Available)

```{r}
# NA represents missing values
x <- c(1, 2, NA, 4, 5)
x

# Check for NA
is.na(x)

# Operations with NA
mean(x)  # Returns NA
mean(x, na.rm = TRUE)  # Remove NA

# Count NAs
sum(is.na(x))

# Remove NAs
na.omit(x)
x[!is.na(x)]
```

## NULL

```{r}
# NULL represents absence of a value
x <- NULL
x

# Check for NULL
is.null(x)

# Length of NULL is 0
length(x)

# NULL in lists
my_list <- list(a = 1, b = NULL, c = 3)
my_list
length(my_list)  # Still counts NULL element
```

## NaN (Not a Number)

```{r}
# NaN from undefined mathematical operations
0/0
Inf - Inf

# Check for NaN
is.nan(0/0)

# NaN is also NA
is.na(NaN)
```

## Inf (Infinity)

```{r}
# Infinity
1/0
-1/0

# Check for infinity
is.infinite(1/0)
is.finite(1/0)
```

# 9. Type Checking Summary

```{r}
# Create example objects
vec <- c(1, 2, 3)
mat <- matrix(1:6, nrow = 2)
arr <- array(1:8, dim = c(2, 2, 2))
lst <- list(a = 1, b = 2)
df <- data.frame(x = 1:3, y = 4:6)
fct <- factor(c("a", "b", "c"))

# Type checking functions
is.vector(vec)
is.matrix(mat)
is.array(arr)
is.list(lst)
is.data.frame(df)
is.factor(fct)

# Note: matrices and arrays are also vectors
is.vector(mat)  # FALSE (has dim attribute)
is.array(mat)   # TRUE (matrix is 2D array)
```

# 10. Conversion Between Structures

```{r}
# Vector to matrix
v <- 1:6
matrix(v, nrow = 2)

# Matrix to data frame
m <- matrix(1:6, nrow = 2)
as.data.frame(m)

# List to data frame (equal length components)
l <- list(x = 1:3, y = 4:6)
as.data.frame(l)

# Data frame to matrix (numeric only)
df_num <- data.frame(a = 1:3, b = 4:6)
as.matrix(df_num)

# Data frame to list
as.list(df_num)

# Vector to list
as.list(1:3)
```

# Summary

This document covered all major R data structures:

- **Vectors**: Atomic (numeric, integer, character, logical) - 1D homogeneous
- **Matrices**: 2D homogeneous numeric arrays
- **Arrays**: nD homogeneous structures
- **Lists**: Heterogeneous collections
- **Data Frames**: Tabular data with mixed types
- **Factors**: Categorical data
- **Tibbles**: Modern data frames
- **Special Values**: NA, NULL, NaN, Inf

**Key Takeaways**:
- Choose the right structure for your data
- Understand indexing for each type
- Be aware of type coercion
- Use factors for categorical data
- Prefer tibbles for modern R workflows

**Next Steps**: Proceed to Data Import/Export to learn how to read and write different file formats.

# References

- [R for Data Science - Data Structures](https://r4ds.had.co.nz/)
- [Advanced R - Data Structures](https://adv-r.hadley.nz/vectors-chap.html)
- [R Documentation](https://www.rdocumentation.org/)
