---
title: "Data Import and Export in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Text Files (CSV, TSV, Delimited)

## Base R Functions

### read.table() Family

```{r}
# Read CSV (comma-separated)
data <- read.csv("data.csv")
data <- read.csv("data.csv", header = TRUE, stringsAsFactors = FALSE)

# Read TSV (tab-separated)
data <- read.delim("data.txt")
data <- read.table("data.txt", header = TRUE, sep = "\t")

# Custom delimiter
data <- read.table("data.txt", header = TRUE, sep = ";")

# Skip lines
data <- read.csv("data.csv", skip = 2)

# Specify column classes
data <- read.csv("data.csv", colClasses = c("character", "numeric", "factor"))

# Read only specific rows
data <- read.csv("data.csv", nrows = 100)

# Handle missing values
data <- read.csv("data.csv", na.strings = c("NA", "N/A", "", "missing"))
```

### write.table() Family

```{r}
# Write CSV
write.csv(data, "output.csv", row.names = FALSE)

# Write TSV
write.table(data, "output.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# Custom delimiter
write.table(data, "output.txt", sep = ";", row.names = FALSE)

# Control quoting
write.csv(data, "output.csv", row.names = FALSE, quote = FALSE)
```

## readr Package (Tidyverse)

Faster and more consistent than base R.

```{r}
library(readr)

# Read CSV
data <- read_csv("data.csv")

# Read TSV
data <- read_tsv("data.txt")

# Read with custom delimiter
data <- read_delim("data.txt", delim = ";")

# Specify column types
data <- read_csv("data.csv", col_types = cols(
  id = col_integer(),
  name = col_character(),
  value = col_double(),
  date = col_date(format = "%Y-%m-%d")
))

# Read from URL
data <- read_csv("https://example.com/data.csv")

# Read compressed files
data <- read_csv("data.csv.gz")

# Write files
write_csv(data, "output.csv")
write_tsv(data, "output.txt")
write_delim(data, "output.txt", delim = ";")

# Write compressed
write_csv(data, "output.csv.gz")
```

## data.table::fread() (Fastest)

```{r}
library(data.table)

# Read with fread (very fast for large files)
data <- fread("data.csv")

# Custom options
data <- fread("data.csv", sep = ",", header = TRUE)

# Select specific columns
data <- fread("data.csv", select = c("col1", "col2", "col5"))

# Drop columns
data <- fread("data.csv", drop = c("col3", "col4"))

# Write with fwrite
fwrite(data, "output.csv")
fwrite(data, "output.csv", sep = "\t")
```

# 2. Excel Files

## readxl Package (Read Only)

```{r}
library(readxl)

# Read Excel file (first sheet)
data <- read_excel("data.xlsx")

# Specify sheet by name or number
data <- read_excel("data.xlsx", sheet = "Sheet2")
data <- read_excel("data.xlsx", sheet = 2)

# Specify range
data <- read_excel("data.xlsx", range = "A1:D100")
data <- read_excel("data.xlsx", range = "Sheet1!B2:F50")

# Skip rows
data <- read_excel("data.xlsx", skip = 3)

# Specify column types
data <- read_excel("data.xlsx", col_types = c("text", "numeric", "date"))

# List sheets
excel_sheets("data.xlsx")

# Read all sheets
all_sheets <- lapply(excel_sheets("data.xlsx"),
                     function(x) read_excel("data.xlsx", sheet = x))
names(all_sheets) <- excel_sheets("data.xlsx")
```

## openxlsx Package (Read and Write)

```{r}
library(openxlsx)

# Read Excel
data <- read.xlsx("data.xlsx", sheet = 1)

# Write Excel
write.xlsx(data, "output.xlsx")

# Write multiple sheets
list_of_datasets <- list("Sheet1" = df1, "Sheet2" = df2, "Sheet3" = df3)
write.xlsx(list_of_datasets, "output.xlsx")

# Create workbook with formatting
wb <- createWorkbook()
addWorksheet(wb, "Data")
writeData(wb, "Data", data)

# Add styling
headerStyle <- createStyle(fontColour = "#ffffff", fgFill = "#4F81BD",
                           halign = "center", valign = "center",
                           textDecoration = "bold", border = "TopBottomLeftRight")
addStyle(wb, sheet = "Data", headerStyle, rows = 1, cols = 1:ncol(data), gridExpand = TRUE)

# Save workbook
saveWorkbook(wb, "formatted_output.xlsx", overwrite = TRUE)
```

## writexl Package (Simple Writing)

```{r}
library(writexl)

# Write single sheet
write_xlsx(data, "output.xlsx")

# Write multiple sheets
write_xlsx(list(Sheet1 = df1, Sheet2 = df2), "output.xlsx")
```

# 3. Statistical Software Formats

## haven Package (SPSS, SAS, Stata)

```{r}
library(haven)

# Read SPSS
data <- read_spss("data.sav")
data <- read_sav("data.sav")

# Read SAS
data <- read_sas("data.sas7bdat")

# Read Stata
data <- read_dta("data.dta")

# Write SPSS
write_sav(data, "output.sav")

# Write SAS
write_sas(data, "output.sas7bdat")

# Write Stata
write_dta(data, "output.dta")
```

# 4. JSON Files

## jsonlite Package

```{r}
library(jsonlite)

# Read JSON
data <- fromJSON("data.json")

# Read from URL
data <- fromJSON("https://api.example.com/data")

# Pretty print
data <- fromJSON("data.json", simplifyDataFrame = TRUE)

# Write JSON
write_json(data, "output.json")

# Pretty formatted JSON
write_json(data, "output.json", pretty = TRUE)

# Convert to/from JSON strings
json_string <- toJSON(data, pretty = TRUE)
data_back <- fromJSON(json_string)
```

# 5. XML Files

## xml2 Package

```{r}
library(xml2)

# Read XML
xml_data <- read_xml("data.xml")

# Parse XML structure
xml_structure(xml_data)

# Extract elements
nodes <- xml_find_all(xml_data, "//record")
values <- xml_text(xml_find_all(xml_data, "//field[@name='value']"))

# Extract attributes
xml_attr(nodes, "id")

# Convert to list
xml_list <- as_list(xml_data)

# Write XML
write_xml(xml_data, "output.xml")
```

# 6. Database Connections

## DBI and Database-Specific Packages

### SQLite

```{r}
library(DBI)
library(RSQLite)

# Connect to database
con <- dbConnect(RSQLite::SQLite(), "database.sqlite")

# List tables
dbListTables(con)

# Read entire table
data <- dbReadTable(con, "table_name")

# Query database
data <- dbGetQuery(con, "SELECT * FROM table_name WHERE value > 100")

# Write data to database
dbWriteTable(con, "new_table", data, overwrite = TRUE)

# Parameterized queries (safe from SQL injection)
query <- "SELECT * FROM table_name WHERE id = ?"
data <- dbGetQuery(con, query, params = list(123))

# Disconnect
dbDisconnect(con)
```

### MySQL/MariaDB

```{r}
library(RMySQL)

# Connect
con <- dbConnect(RMySQL::MySQL(),
                 dbname = "database_name",
                 host = "localhost",
                 port = 3306,
                 user = "username",
                 password = "password")

# Query
data <- dbGetQuery(con, "SELECT * FROM table_name")

# Disconnect
dbDisconnect(con)
```

### PostgreSQL

```{r}
library(RPostgres)

# Connect
con <- dbConnect(RPostgres::Postgres(),
                 dbname = "database_name",
                 host = "localhost",
                 port = 5432,
                 user = "username",
                 password = "password")

# Query
data <- dbGetQuery(con, "SELECT * FROM table_name")

# Disconnect
dbDisconnect(con)
```

# 7. Web Scraping and APIs

## Reading from URLs

```{r}
# Read CSV from URL
data <- read.csv("https://example.com/data.csv")

# With readr
library(readr)
data <- read_csv("https://example.com/data.csv")

# Download file first
download.file("https://example.com/data.csv", "local_data.csv")
data <- read.csv("local_data.csv")
```

## httr Package for APIs

```{r}
library(httr)

# GET request
response <- GET("https://api.example.com/data")

# Check status
status_code(response)

# Parse content
content(response, "text")
content(response, "parsed")

# POST request
response <- POST("https://api.example.com/submit",
                 body = list(param1 = "value1", param2 = "value2"),
                 encode = "json")

# Authentication
response <- GET("https://api.example.com/data",
                authenticate("username", "password"))

# API key in header
response <- GET("https://api.example.com/data",
                add_headers(Authorization = "Bearer YOUR_API_KEY"))
```

## rvest Package for Web Scraping

```{r}
library(rvest)

# Read HTML page
page <- read_html("https://example.com")

# Extract elements
titles <- page %>% html_nodes("h1") %>% html_text()
links <- page %>% html_nodes("a") %>% html_attr("href")

# Extract tables
tables <- page %>% html_table(fill = TRUE)
```

# 8. R Data Formats

## RDS (Single Object)

```{r}
# Save single object
saveRDS(data, "data.rds")

# Read single object
data <- readRDS("data.rds")

# Compressed RDS
saveRDS(data, "data.rds", compress = "xz")
```

## RData/Rda (Multiple Objects)

```{r}
# Save multiple objects
save(data1, data2, data3, file = "mydata.RData")

# Save with compression
save(data1, data2, file = "mydata.RData", compress = "xz")

# Load objects (variables will be created in environment)
load("mydata.RData")

# Save all objects in workspace
save.image("workspace.RData")
```

## feather Format (Fast, Cross-Language)

```{r}
library(feather)

# Write
write_feather(data, "data.feather")

# Read
data <- read_feather("data.feather")
```

## parquet Format (Columnar Storage)

```{r}
library(arrow)

# Write
write_parquet(data, "data.parquet")

# Read
data <- read_parquet("data.parquet")

# Read specific columns
data <- read_parquet("data.parquet", col_select = c("col1", "col2"))
```

# 9. Bioinformatics File Formats

## FASTA Files

```{r}
library(Biostrings)

# Read FASTA
sequences <- readDNAStringSet("sequences.fasta")
sequences <- readAAStringSet("proteins.fasta")

# Access sequences
sequences[[1]]
names(sequences)

# Write FASTA
writeXStringSet(sequences, "output.fasta")
```

## FASTQ Files

```{r}
library(ShortRead)

# Read FASTQ
fastq <- readFastq("reads.fastq")

# Access components
sread(fastq)  # Sequences
quality(fastq)  # Quality scores
id(fastq)  # Read IDs

# Write FASTQ
writeFastq(fastq, "output.fastq")
```

## VCF Files (Variant Call Format)

```{r}
library(VariantAnnotation)

# Read VCF
vcf <- readVcf("variants.vcf", "hg38")

# Access components
rowRanges(vcf)  # Genomic positions
info(vcf)  # Info fields
geno(vcf)  # Genotype data

# Write VCF
writeVcf(vcf, "output.vcf")
```

## BAM Files (Binary Alignment)

```{r}
library(Rsamtools)

# Index BAM file first
indexBam("aligned.bam")

# Read BAM
bam <- scanBam("aligned.bam")

# Read specific region
param <- ScanBamParam(which = GRanges("chr1", IRanges(1000000, 2000000)))
bam <- scanBam("aligned.bam", param = param)

# What fields to import
param <- ScanBamParam(what = c("qname", "flag", "rname", "pos", "mapq"))
bam <- scanBam("aligned.bam", param = param)
```

## BED Files

```{r}
library(rtracklayer)

# Read BED
bed <- import.bed("regions.bed")

# Read as GRanges
gr <- import("regions.bed")

# Write BED
export.bed(gr, "output.bed")
export(gr, "output.bed", format = "bed")
```

## GFF/GTF Files

```{r}
library(rtracklayer)

# Read GTF
gtf <- import.gff("genes.gtf")

# Read GFF3
gff <- import.gff3("genes.gff3")

# Write
export.gff(gtf, "output.gtf")
export.gff3(gff, "output.gff3")
```

## GenomicRanges Objects

```{r}
library(GenomicRanges)

# Save GRanges as RDS
saveRDS(granges_object, "granges.rds")

# Read back
granges_object <- readRDS("granges.rds")
```

# 10. Large Files and Streaming

## Reading Large Files in Chunks

```{r}
# Base R - read in chunks
chunk_size <- 10000
con <- file("large_file.csv", "r")

while(length(chunk <- readLines(con, n = chunk_size)) > 0) {
  # Process chunk
  print(paste("Processing", length(chunk), "lines"))
}
close(con)

# readr - read chunks
library(readr)
read_csv_chunked("large_file.csv",
                 callback = DataFrameCallback$new(function(x, pos) {
                   # Process chunk x
                   print(nrow(x))
                 }),
                 chunk_size = 10000)
```

## data.table for Large Files

```{r}
library(data.table)

# fread is very fast for large files
data <- fread("large_file.csv")

# Select columns to reduce memory
data <- fread("large_file.csv", select = c("col1", "col2"))
```

## arrow for Huge Datasets

```{r}
library(arrow)

# Read dataset from multiple parquet files
dataset <- open_dataset("data_directory/", format = "parquet")

# Query with dplyr
library(dplyr)
result <- dataset %>%
  filter(value > 100) %>%
  select(id, value, category) %>%
  collect()  # Brings into memory
```

# 11. Compressed Files

## Reading Compressed Files

```{r}
# Most R functions handle compression automatically

# Gzip
data <- read.csv("data.csv.gz")
data <- read_csv("data.csv.gz")

# Bzip2
data <- read.csv("data.csv.bz2")

# Xz
data <- read.csv("data.csv.xz")

# Zip (need to extract first or specify file)
data <- read.csv(unz("archive.zip", "data.csv"))
```

## Writing Compressed Files

```{r}
# Gzip
write.csv(data, gzfile("output.csv.gz"))
write_csv(data, "output.csv.gz")  # readr handles automatically

# Bzip2
write.csv(data, bzfile("output.csv.bz2"))

# Xz
write.csv(data, xzfile("output.csv.xz"))
```

# 12. Google Sheets

## googlesheets4 Package

```{r}
library(googlesheets4)

# Authenticate
gs4_auth()

# Read from Google Sheets
data <- read_sheet("spreadsheet_url_or_id")

# Specific sheet
data <- read_sheet("spreadsheet_url_or_id", sheet = "Sheet2")

# Specific range
data <- read_sheet("spreadsheet_url_or_id", range = "A1:D100")

# Write to Google Sheets
write_sheet(data, "spreadsheet_url_or_id")

# Create new sheet
sheet_write(data, ss = "spreadsheet_url_or_id", sheet = "NewSheet")
```

# 13. Cloud Storage

## AWS S3

```{r}
library(aws.s3)

# Read from S3
data <- s3read_using(read.csv, object = "s3://bucket-name/data.csv")

# Write to S3
s3write_using(data, write.csv, object = "s3://bucket-name/output.csv")

# List objects
bucket_list_df("bucket-name")
```

# 14. Best Practices

## File Path Management

```{r}
# Use here package for project-relative paths
library(here)

data <- read.csv(here("data", "raw", "data.csv"))
write.csv(result, here("output", "results.csv"))

# Use file.path for cross-platform compatibility
data_path <- file.path("data", "raw", "data.csv")
data <- read.csv(data_path)
```

## Error Handling

```{r}
# Check file exists
if (file.exists("data.csv")) {
  data <- read.csv("data.csv")
} else {
  stop("File not found")
}

# Try-catch for file operations
data <- tryCatch({
  read.csv("data.csv")
}, error = function(e) {
  message("Error reading file: ", e$message)
  NULL
})
```

## Memory Efficiency

```{r}
# Specify column types to save memory
data <- read.csv("data.csv", colClasses = c(
  "integer", "character", "numeric", "factor"
))

# Read only needed columns
library(readr)
data <- read_csv("data.csv", col_select = c(id, value, category))

# Use data.table for large files
library(data.table)
data <- fread("large_file.csv", select = c("col1", "col2"))
```

# Summary

This document covered:

- **Text files**: CSV, TSV with base R, readr, data.table
- **Excel files**: readxl, openxlsx, writexl
- **Statistical formats**: SPSS, SAS, Stata with haven
- **Structured data**: JSON, XML
- **Databases**: SQLite, MySQL, PostgreSQL
- **Web data**: APIs, web scraping
- **R formats**: RDS, RData, feather, parquet
- **Bioinformatics**: FASTA, FASTQ, VCF, BAM, BED, GTF/GFF
- **Large files**: Chunked reading, arrow
- **Compressed files**: gz, bz2, xz, zip
- **Cloud storage**: Google Sheets, AWS S3

**Key Recommendations**:
- Use `readr` for CSV/TSV (fast, consistent)
- Use `data.table::fread()` for very large files
- Use `readxl` for reading Excel, `openxlsx` for writing
- Use `arrow` and parquet for huge datasets
- Always specify `stringsAsFactors = FALSE` in base R (or use readr)
- Use compression for large files
- Use proper error handling

**Next Steps**: Proceed to Data Manipulation to learn dplyr, tidyr, and data.table.

# References

- [readr documentation](https://readr.tidyverse.org/)
- [data.table documentation](https://rdatatable.gitlab.io/data.table/)
- [Bioconductor workflows](https://bioconductor.org/packages/release/BiocViews.html#___Workflow)
- [R Data Import/Export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html)
