---
title: "Data Manipulation in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Base R Data Manipulation

## Subsetting and Indexing

```{r}
# Create sample data
df <- data.frame(
  id = 1:10,
  name = paste0("Sample_", 1:10),
  value = c(23, 45, 67, 12, 89, 34, 56, 78, 90, 23),
  group = rep(c("A", "B"), each = 5)
)

# Subset rows
df[df$value > 50, ]
df[df$group == "A", ]
df[df$value > 30 & df$group == "B", ]

# Subset columns
df[, c("name", "value")]
df[c("name", "value")]

# Subset function
subset(df, value > 50)
subset(df, value > 50, select = c(name, value))
subset(df, group == "A", select = -id)

# Order/sort
df[order(df$value), ]
df[order(-df$value), ]  # Descending
df[order(df$group, -df$value), ]  # Multiple columns
```

## Adding and Modifying Columns

```{r}
# Add new column
df$category <- ifelse(df$value > 50, "High", "Low")

# Transform column
df$value_log <- log(df$value)

# Conditional transformation
df$value_adjusted <- ifelse(df$group == "A", df$value * 1.1, df$value * 0.9)

# Remove column
df$value_log <- NULL
```

## Aggregation

```{r}
# aggregate() function
aggregate(value ~ group, data = df, FUN = mean)
aggregate(value ~ group, data = df, FUN = sum)
aggregate(cbind(value, id) ~ group, data = df, FUN = mean)

# tapply() - apply function to each group
tapply(df$value, df$group, mean)
tapply(df$value, df$group, sum)

# by() - more general grouping
by(df$value, df$group, summary)
```

## Merging/Joining

```{r}
# Create sample data
df1 <- data.frame(id = 1:5, value_x = c(10, 20, 30, 40, 50))
df2 <- data.frame(id = c(2, 3, 4, 5, 6), value_y = c(100, 200, 300, 400, 500))

# Inner join
merge(df1, df2, by = "id")

# Left join
merge(df1, df2, by = "id", all.x = TRUE)

# Right join
merge(df1, df2, by = "id", all.y = TRUE)

# Outer join
merge(df1, df2, by = "id", all = TRUE)

# Join by different column names
df1_renamed <- data.frame(id_x = 1:5, value = c(10, 20, 30, 40, 50))
df2_renamed <- data.frame(id_y = c(2, 3, 4, 5, 6), value = c(100, 200, 300, 400, 500))
merge(df1_renamed, df2_renamed, by.x = "id_x", by.y = "id_y")
```

# 2. dplyr Package (Tidyverse)

dplyr is the most popular package for data manipulation in R.

```{r}
library(dplyr)
```

## Core dplyr Verbs

### select() - Choose columns

```{r}
# Select specific columns
df %>% select(name, value)

# Select range of columns
df %>% select(id:value)

# Select by exclusion
df %>% select(-id)

# Select with helpers
df %>% select(starts_with("val"))
df %>% select(ends_with("e"))
df %>% select(contains("am"))
df %>% select(matches("^v"))

# Reorder columns
df %>% select(name, group, everything())
```

### filter() - Choose rows

```{r}
# Filter by condition
df %>% filter(value > 50)

# Multiple conditions (AND)
df %>% filter(value > 30, group == "A")
df %>% filter(value > 30 & group == "A")

# OR condition
df %>% filter(value > 80 | value < 20)

# Filter by multiple values
df %>% filter(group %in% c("A", "B"))

# Filter out NA
df %>% filter(!is.na(value))

# Filter with string operations
df %>% filter(grepl("Sample_[1-5]", name))

# Slice rows by position
df %>% slice(1:5)
df %>% slice_head(n = 3)
df %>% slice_tail(n = 3)
df %>% slice_max(value, n = 3)
df %>% slice_min(value, n = 3)
df %>% slice_sample(n = 5)
```

### mutate() - Create/modify columns

```{r}
# Add new column
df %>% mutate(value_squared = value^2)

# Multiple new columns
df %>% mutate(
  value_log = log(value),
  value_sqrt = sqrt(value),
  value_category = ifelse(value > 50, "High", "Low")
)

# Transform existing column
df %>% mutate(value = value * 2)

# Conditional mutation
df %>% mutate(
  adjusted_value = case_when(
    group == "A" & value > 50 ~ value * 1.2,
    group == "A" & value <= 50 ~ value * 1.1,
    group == "B" & value > 50 ~ value * 0.9,
    TRUE ~ value  # default case
  )
)

# Mutate with across() for multiple columns
df %>% mutate(across(where(is.numeric), ~.x * 2))
df %>% mutate(across(c(value, id), ~.x + 10))

# transmute() - create new columns and drop old ones
df %>% transmute(
  name,
  value_log = log(value)
)
```

### summarise() / summarize() - Aggregate data

```{r}
# Single summary
df %>% summarise(mean_value = mean(value))

# Multiple summaries
df %>% summarise(
  n = n(),
  mean_value = mean(value),
  sd_value = sd(value),
  min_value = min(value),
  max_value = max(value),
  median_value = median(value)
)

# With NA handling
df %>% summarise(
  mean_value = mean(value, na.rm = TRUE)
)
```

### arrange() - Sort rows

```{r}
# Sort ascending
df %>% arrange(value)

# Sort descending
df %>% arrange(desc(value))

# Sort by multiple columns
df %>% arrange(group, desc(value))
```

### group_by() - Group operations

```{r}
# Group and summarize
df %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_value = mean(value),
    sd_value = sd(value)
  )

# Group and mutate (creates within-group calculations)
df %>%
  group_by(group) %>%
  mutate(
    mean_group = mean(value),
    centered_value = value - mean(value)
  )

# Multiple grouping variables
df %>%
  group_by(group, category) %>%
  summarise(count = n())

# Ungroup
df %>%
  group_by(group) %>%
  mutate(mean_group = mean(value)) %>%
  ungroup()
```

## Advanced dplyr Operations

### Window Functions

```{r}
# Ranking
df %>%
  mutate(
    rank = row_number(value),
    min_rank = min_rank(value),
    dense_rank = dense_rank(value)
  )

# Grouped ranking
df %>%
  group_by(group) %>%
  mutate(rank_in_group = row_number(desc(value)))

# Cumulative operations
df %>%
  arrange(id) %>%
  mutate(
    cumsum_value = cumsum(value),
    cummean_value = cummean(value)
  )

# Lead and lag
df %>%
  arrange(id) %>%
  mutate(
    previous_value = lag(value),
    next_value = lead(value),
    diff_from_previous = value - lag(value)
  )
```

### Joins

```{r}
df1 <- data.frame(id = 1:5, value_x = c(10, 20, 30, 40, 50))
df2 <- data.frame(id = c(2, 3, 4, 5, 6), value_y = c(100, 200, 300, 400, 500))

# Inner join
inner_join(df1, df2, by = "id")

# Left join
left_join(df1, df2, by = "id")

# Right join
right_join(df1, df2, by = "id")

# Full join
full_join(df1, df2, by = "id")

# Semi join (keeps rows in df1 that have match in df2)
semi_join(df1, df2, by = "id")

# Anti join (keeps rows in df1 that DON'T have match in df2)
anti_join(df1, df2, by = "id")

# Join by different column names
left_join(df1, df2, by = c("id" = "id"))

# Join by multiple columns
# df1 %>% left_join(df2, by = c("id", "group"))
```

### Reshaping with dplyr

```{r}
# Row-wise operations
df %>%
  rowwise() %>%
  mutate(row_mean = mean(c(value, id)))

# Counting
df %>% count(group)
df %>% count(group, sort = TRUE)

# Distinct values
df %>% distinct(group)
df %>% distinct(group, .keep_all = TRUE)

# Add count to data
df %>% add_count(group)
```

## Piping

```{r}
# Chain multiple operations
result <- df %>%
  filter(value > 30) %>%
  mutate(value_log = log(value)) %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_log = mean(value_log)
  ) %>%
  arrange(desc(mean_log))

result
```

# 3. tidyr Package

tidyr is used for reshaping data.

```{r}
library(tidyr)
```

## Pivoting

### pivot_longer() - Wide to long

```{r}
# Create wide data
wide_df <- data.frame(
  id = 1:3,
  year_2020 = c(100, 200, 300),
  year_2021 = c(150, 250, 350),
  year_2022 = c(180, 280, 380)
)

# Pivot to long format
long_df <- wide_df %>%
  pivot_longer(
    cols = starts_with("year_"),
    names_to = "year",
    values_to = "value"
  )
long_df

# With name transformation
wide_df %>%
  pivot_longer(
    cols = starts_with("year_"),
    names_to = "year",
    names_prefix = "year_",
    names_transform = list(year = as.integer),
    values_to = "value"
  )
```

### pivot_wider() - Long to wide

```{r}
# Pivot back to wide format
long_df %>%
  pivot_wider(
    names_from = year,
    values_from = value
  )

# With name prefix
long_df %>%
  pivot_wider(
    names_from = year,
    names_prefix = "year_",
    values_from = value
  )
```

## Separating and Uniting

### separate() - Split one column into multiple

```{r}
# Create data with combined column
df_combined <- data.frame(
  id = 1:3,
  sample_info = c("Patient_A_2020", "Patient_B_2021", "Patient_C_2022")
)

# Separate into multiple columns
df_combined %>%
  separate(sample_info, into = c("type", "patient", "year"), sep = "_")

# Keep original column
df_combined %>%
  separate(sample_info, into = c("type", "patient", "year"),
           sep = "_", remove = FALSE)

# separate_rows() - separate and expand to multiple rows
df_multi <- data.frame(
  id = 1:2,
  values = c("1,2,3", "4,5,6")
)

df_multi %>%
  separate_rows(values, sep = ",", convert = TRUE)
```

### unite() - Combine multiple columns

```{r}
# Create separate columns
df_separate <- data.frame(
  id = 1:3,
  year = c(2020, 2021, 2022),
  month = c("01", "02", "03"),
  day = c("15", "20", "25")
)

# Unite into one column
df_separate %>%
  unite("date", year, month, day, sep = "-")

# Keep original columns
df_separate %>%
  unite("date", year, month, day, sep = "-", remove = FALSE)
```

## Handling Missing Values

```{r}
# Create data with NAs
df_na <- data.frame(
  id = 1:5,
  x = c(1, NA, 3, NA, 5),
  y = c(NA, 2, 3, 4, 5)
)

# Drop rows with any NA
df_na %>% drop_na()

# Drop rows with NA in specific columns
df_na %>% drop_na(x)

# Fill NA with specific value
df_na %>% replace_na(list(x = 0, y = 0))

# Fill NA with previous/next value
df_na %>% fill(x, .direction = "down")
df_na %>% fill(x, .direction = "up")

# complete() - make implicit missing values explicit
df_incomplete <- data.frame(
  group = c("A", "A", "B"),
  value = c(1, 2, 3)
)

df_incomplete %>%
  complete(group = c("A", "B", "C"))
```

## Nesting

```{r}
# Nest data
nested_df <- df %>%
  group_by(group) %>%
  nest()

nested_df

# Access nested data
nested_df$data[[1]]

# Unnest
nested_df %>%
  unnest(cols = data)
```

# 4. data.table Package

data.table is extremely fast for large datasets.

```{r}
library(data.table)

# Create data.table
dt <- data.table(
  id = 1:10,
  name = paste0("Sample_", 1:10),
  value = c(23, 45, 67, 12, 89, 34, 56, 78, 90, 23),
  group = rep(c("A", "B"), each = 5)
)

# Convert data.frame to data.table
dt <- as.data.table(df)
```

## data.table Syntax: DT[i, j, by]

- **i**: which rows (WHERE)
- **j**: what to do (SELECT/UPDATE)
- **by**: grouped by what (GROUP BY)

### Selecting Rows (i)

```{r}
# Filter rows
dt[value > 50]
dt[group == "A"]
dt[value > 30 & group == "B"]

# Select by row number
dt[1:5]

# Order
dt[order(value)]
dt[order(-value)]
```

### Selecting Columns (j)

```{r}
# Select columns
dt[, .(name, value)]
dt[, list(name, value)]

# Select single column (returns vector)
dt[, value]

# Select single column (returns data.table)
dt[, .(value)]

# Compute on columns
dt[, mean(value)]
dt[, .(mean_value = mean(value), sd_value = sd(value))]
```

### Grouping (by)

```{r}
# Group and summarize
dt[, .(mean_value = mean(value)), by = group]

# Multiple aggregations
dt[, .(
  n = .N,
  mean_value = mean(value),
  sd_value = sd(value)
), by = group]

# Group by multiple columns
dt[, .(mean_value = mean(value)), by = .(group, category)]
```

## Modify by Reference

data.table allows modification by reference (very fast, no copy).

```{r}
# Add column
dt[, new_col := value * 2]

# Multiple columns
dt[, `:=`(
  col1 = value + 10,
  col2 = value - 10
)]

# Conditional update
dt[group == "A", value := value * 1.1]

# Delete column
dt[, new_col := NULL]
```

## Advanced data.table

### Chaining

```{r}
# Chain operations
dt[value > 30][order(-value)][, .(name, value)]
```

### Special Symbols

```{r}
# .N - number of rows
dt[, .N, by = group]

# .I - row numbers
dt[, .I[1:3], by = group]

# .SD - Subset of Data
dt[, lapply(.SD, mean), by = group, .SDcols = c("value", "id")]

# .GRP - group counter
dt[, group_id := .GRP, by = group]
```

### Keys and Fast Subsetting

```{r}
# Set key
setkey(dt, group)

# Fast subsetting by key
dt["A"]
dt[c("A", "B")]

# Joins with keys
dt1 <- data.table(id = 1:5, value_x = c(10, 20, 30, 40, 50), key = "id")
dt2 <- data.table(id = c(2, 3, 4, 5, 6), value_y = c(100, 200, 300, 400, 500), key = "id")

dt1[dt2]  # Right join
merge(dt1, dt2, all.x = TRUE)  # Left join
merge(dt1, dt2, all = TRUE)  # Outer join
```

### Reshaping

```{r}
# melt (wide to long)
wide_dt <- data.table(
  id = 1:3,
  A = c(1, 2, 3),
  B = c(4, 5, 6)
)

long_dt <- melt(wide_dt, id.vars = "id",
                variable.name = "group", value.name = "value")
long_dt

# dcast (long to wide)
dcast(long_dt, id ~ group, value.var = "value")
```

# 5. stringr Package (String Manipulation)

```{r}
library(stringr)

# Create example strings
genes <- c("TP53", "BRCA1", "EGFR", "KRAS")
samples <- c("Sample_A_001", "Sample_B_002", "Sample_C_003")

# String length
str_length(genes)

# Convert case
str_to_upper(genes)
str_to_lower(genes)
str_to_title(samples)

# Detect pattern
str_detect(genes, "R")
str_detect(genes, "^BR")  # Starts with BR

# Subset strings
str_subset(genes, "R")  # Keep matches
str_subset(genes, "R", negate = TRUE)  # Keep non-matches

# Extract matches
str_extract(samples, "[A-Z]")
str_extract_all(samples, "[0-9]")

# Replace
str_replace(samples, "_", "-")
str_replace_all(samples, "_", "-")

# Remove whitespace
str_trim("  text  ")
str_squish("  too    much    space  ")

# Split strings
str_split(samples, "_")
str_split(samples, "_", simplify = TRUE)

# Combine strings
str_c(genes, "gene", sep = "_")
str_glue("Gene: {genes}")

# Pad strings
str_pad(1:5, width = 3, pad = "0")

# Truncate strings
str_trunc("Very long string", width = 10)
```

# 6. lubridate Package (Date/Time)

```{r}
library(lubridate)

# Parse dates
ymd("2024-01-15")
mdy("01/15/2024")
dmy("15-01-2024")

# Parse datetimes
ymd_hms("2024-01-15 14:30:00")

# Get components
date <- ymd("2024-01-15")
year(date)
month(date)
day(date)
wday(date)
wday(date, label = TRUE)

# Set components
year(date) <- 2025
date

# Arithmetic
date + days(7)
date + months(2)
date + years(1)

# Intervals
start <- ymd("2024-01-01")
end <- ymd("2024-12-31")
interval(start, end)
time_length(interval(start, end), "months")

# Floor/ceiling dates
floor_date(ymd("2024-01-15"), "month")
ceiling_date(ymd("2024-01-15"), "month")
round_date(ymd("2024-01-15"), "month")
```

# 7. Real-World Example Workflows

## Example 1: Gene Expression Analysis

```{r}
# Simulated gene expression data
set.seed(123)
expression_data <- data.frame(
  sample_id = rep(paste0("Sample_", 1:10), each = 100),
  gene = rep(paste0("Gene_", 1:100), times = 10),
  expression = rnorm(1000, mean = 100, sd = 20),
  condition = rep(rep(c("Control", "Treatment"), each = 5), each = 100)
)

# Analysis workflow
results <- expression_data %>%
  group_by(gene, condition) %>%
  summarise(
    mean_expr = mean(expression),
    sd_expr = sd(expression),
    n = n()
  ) %>%
  pivot_wider(
    names_from = condition,
    values_from = c(mean_expr, sd_expr)
  ) %>%
  mutate(
    fold_change = mean_expr_Treatment / mean_expr_Control,
    log2_fc = log2(fold_change)
  ) %>%
  filter(abs(log2_fc) > 0.5) %>%
  arrange(desc(abs(log2_fc)))

head(results)
```

## Example 2: Clinical Data Processing

```{r}
# Simulated clinical data
clinical <- data.frame(
  patient_id = paste0("PT", 1:20),
  age = sample(25:75, 20, replace = TRUE),
  diagnosis_date = sample(seq(ymd("2020-01-01"), ymd("2023-12-31"), by = "day"), 20),
  treatment = sample(c("A", "B", "C"), 20, replace = TRUE),
  response = sample(c("Complete", "Partial", "None"), 20, replace = TRUE)
)

# Processing workflow
clinical_processed <- clinical %>%
  mutate(
    age_group = case_when(
      age < 40 ~ "Young",
      age < 60 ~ "Middle",
      TRUE ~ "Senior"
    ),
    year = year(diagnosis_date),
    days_since = as.numeric(today() - diagnosis_date)
  ) %>%
  group_by(treatment, response) %>%
  summarise(
    n_patients = n(),
    mean_age = mean(age),
    .groups = "drop"
  )

clinical_processed
```

# Summary

This document covered comprehensive data manipulation techniques:

**Base R**:
- Subsetting, indexing, merging
- Simple but can be verbose

**dplyr**:
- select, filter, mutate, summarise, arrange, group_by
- Pipe-friendly, readable syntax
- Best for most data analysis tasks

**tidyr**:
- pivot_longer, pivot_wider
- separate, unite
- Handling missing values

**data.table**:
- DT[i, j, by] syntax
- Extremely fast for large data
- Modification by reference

**stringr**:
- Consistent string manipulation
- Pattern matching and replacement

**lubridate**:
- Easy date/time handling
- Arithmetic and parsing

**When to use what**:
- **dplyr/tidyr**: Most data analysis, readable code
- **data.table**: Very large datasets, performance critical
- **Base R**: Quick operations, no dependencies

**Next Steps**: Proceed to Data Visualization to learn ggplot2 and other visualization tools.

# References

- [dplyr documentation](https://dplyr.tidyverse.org/)
- [tidyr documentation](https://tidyr.tidyverse.org/)
- [data.table documentation](https://rdatatable.gitlab.io/data.table/)
- [stringr documentation](https://stringr.tidyverse.org/)
- [lubridate documentation](https://lubridate.tidyverse.org/)
- [R for Data Science](https://r4ds.had.co.nz/)
