---
title: "Data Visualization in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# 1. Base R Graphics

Base R provides quick plotting capabilities without additional packages.

## Basic Plots

```{r}
# Sample data
set.seed(123)
x <- rnorm(100)
y <- x + rnorm(100)
groups <- sample(c("A", "B", "C"), 100, replace = TRUE)

# Scatter plot
plot(x, y)
plot(x, y, main = "Scatter Plot", xlab = "X Variable", ylab = "Y Variable")

# With colors and points
plot(x, y, col = "blue", pch = 16, cex = 1.5)

# Line plot
plot(1:10, (1:10)^2, type = "l")
plot(1:10, (1:10)^2, type = "b")  # Both points and lines

# Multiple plots
plot(x, y, col = as.factor(groups), pch = 16)
legend("topleft", legend = unique(groups), col = 1:3, pch = 16)
```

## Histogram

```{r}
# Basic histogram
hist(x)

# Customized
hist(x, breaks = 20, col = "lightblue", main = "Distribution",
     xlab = "Value", border = "white")

# Add density line
hist(x, breaks = 20, col = "lightblue", freq = FALSE)
lines(density(x), col = "red", lwd = 2)
```

## Boxplot

```{r}
# Basic boxplot
boxplot(rnorm(100))

# Grouped boxplot
boxplot(y ~ groups, main = "Boxplot by Group",
        col = c("lightblue", "lightgreen", "lightpink"))

# Horizontal
boxplot(y ~ groups, horizontal = TRUE)
```

## Barplot

```{r}
# Simple barplot
counts <- table(groups)
barplot(counts, col = "steelblue", main = "Counts by Group")

# Grouped barplot
data_matrix <- matrix(c(3, 5, 7, 4, 6, 8), nrow = 2)
barplot(data_matrix, beside = TRUE, col = c("lightblue", "lightpink"),
        legend.text = c("Group 1", "Group 2"))
```

## Par() for Multiple Plots

```{r}
# Set up 2x2 plotting area
par(mfrow = c(2, 2))
plot(x, y, main = "Scatter")
hist(x, main = "Histogram")
boxplot(x, main = "Boxplot")
barplot(counts, main = "Barplot")

# Reset
par(mfrow = c(1, 1))
```

# 2. ggplot2 - Grammar of Graphics

ggplot2 is the most powerful and flexible visualization package in R.

```{r}
library(ggplot2)
```

## ggplot2 Basics

The grammar consists of:
- **Data**: The dataset
- **Aesthetics** (aes): Mappings (x, y, color, size, etc.)
- **Geoms**: Geometric objects (points, lines, bars, etc.)
- **Scales**: Control aesthetic mappings
- **Themes**: Overall visual appearance

## Scatter Plots

```{r}
# Basic scatter plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# With color by group
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point()

# With size and transparency
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl), size = hp)) +
  geom_point(alpha = 0.6)

# Add smooth line
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)

# Different smoothers per group
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Line Plots

```{r}
# Time series example
time_data <- data.frame(
  time = 1:100,
  value = cumsum(rnorm(100)),
  group = rep(c("A", "B"), each = 50)
)

ggplot(time_data, aes(x = time, y = value)) +
  geom_line()

# Multiple lines
ggplot(time_data, aes(x = time, y = value, color = group)) +
  geom_line(linewidth = 1)

# With points
ggplot(time_data, aes(x = time, y = value, color = group)) +
  geom_line() +
  geom_point()
```

## Bar Plots

```{r}
# Count bar plot
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()

# With fill color
ggplot(mtcars, aes(x = factor(cyl), fill = factor(gear))) +
  geom_bar()

# Stacked vs dodged
ggplot(mtcars, aes(x = factor(cyl), fill = factor(gear))) +
  geom_bar(position = "dodge")

# Bar plot with values
summary_data <- mtcars %>%
  group_by(cyl) %>%
  summarise(mean_mpg = mean(mpg))

ggplot(summary_data, aes(x = factor(cyl), y = mean_mpg)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(mean_mpg, 1)), vjust = -0.5)
```

## Histograms and Density Plots

```{r}
# Histogram
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white")

# Density plot
ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "steelblue", alpha = 0.5)

# Overlapping densities
ggplot(mtcars, aes(x = mpg, fill = factor(cyl))) +
  geom_density(alpha = 0.5)

# Histogram with density
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue") +
  geom_density(color = "red", linewidth = 1)
```

## Boxplots and Violin Plots

```{r}
# Boxplot
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()

# With color
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot()

# With points
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5)

# Violin plot
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_violin()

# Violin + boxplot
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_violin() +
  geom_boxplot(width = 0.2)
```

## Faceting

```{r}
# Facet wrap (one variable)
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl)

# Facet wrap with custom layout
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl, ncol = 2)

# Facet grid (two variables)
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_grid(gear ~ cyl)

# Free scales
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl, scales = "free")
```

## Customization

### Themes

```{r}
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3)

# Built-in themes
p + theme_bw()
p + theme_minimal()
p + theme_classic()
p + theme_dark()

# Custom theme modifications
p + theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 12)
  )
```

### Labels and Titles

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(
    title = "Fuel Efficiency by Weight",
    subtitle = "Motor Trend Car Road Tests",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    color = "Cylinders",
    caption = "Source: mtcars dataset"
  )
```

### Scales

```{r}
# Color scales
ggplot(mtcars, aes(x = wt, y = mpg, color = hp)) +
  geom_point(size = 3) +
  scale_color_gradient(low = "blue", high = "red")

# Manual colors
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_manual(values = c("red", "green", "blue"))

# Brewer color palettes
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1")

# Axis scales
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 6, 1)) +
  scale_y_continuous(limits = c(0, 40))

# Log scale
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  scale_y_log10()
```

### Coordinate Systems

```{r}
# Flip coordinates
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  coord_flip()

# Fixed aspect ratio
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  coord_fixed(ratio = 1/5)

# Polar coordinates (for pie charts)
df <- data.frame(
  category = c("A", "B", "C"),
  value = c(30, 40, 30)
)

ggplot(df, aes(x = "", y = value, fill = category)) +
  geom_col() +
  coord_polar(theta = "y")
```

## Advanced ggplot2

### Error Bars

```{r}
summary_stats <- mtcars %>%
  group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    se = sd(mpg) / sqrt(n())
  )

ggplot(summary_stats, aes(x = factor(cyl), y = mean_mpg)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_mpg - se, ymax = mean_mpg + se),
                width = 0.2)

# Point range
ggplot(summary_stats, aes(x = factor(cyl), y = mean_mpg)) +
  geom_pointrange(aes(ymin = mean_mpg - se, ymax = mean_mpg + se))
```

### Annotations

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  annotate("text", x = 4, y = 30, label = "High efficiency zone",
           color = "red") +
  annotate("rect", xmin = 1.5, xmax = 3, ymin = 25, ymax = 35,
           alpha = 0.2, fill = "blue")
```

### Statistical Transformations

```{r}
# Regression line with equation
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  stat_poly_eq(formula = y ~ x, parse = TRUE)

# Confidence intervals
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2)
```

# 3. Heatmaps

## Basic Heatmap with ggplot2

```{r}
# Create matrix data
set.seed(123)
mat_data <- matrix(rnorm(100), nrow = 10)
rownames(mat_data) <- paste0("Gene", 1:10)
colnames(mat_data) <- paste0("Sample", 1:10)

# Convert to long format for ggplot2
heatmap_data <- as.data.frame(mat_data) %>%
  tibble::rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to = "Sample", values_to = "Expression")

ggplot(heatmap_data, aes(x = Sample, y = Gene, fill = Expression)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## pheatmap Package

```{r}
library(pheatmap)

# Basic heatmap
pheatmap(mat_data)

# With clustering
pheatmap(mat_data, clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete")

# With annotations
annotation_row <- data.frame(
  GeneType = rep(c("TypeA", "TypeB"), each = 5),
  row.names = rownames(mat_data)
)

annotation_col <- data.frame(
  Condition = rep(c("Control", "Treatment"), each = 5),
  row.names = colnames(mat_data)
)

pheatmap(mat_data,
         annotation_row = annotation_row,
         annotation_col = annotation_col,
         scale = "row")
```

## ComplexHeatmap Package

```{r eval=FALSE}
library(ComplexHeatmap)

# Advanced heatmap
Heatmap(mat_data,
        name = "Expression",
        col = circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red")),
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = TRUE,
        show_column_names = TRUE)

# Multiple heatmaps
ht1 <- Heatmap(mat_data[, 1:5], name = "Control")
ht2 <- Heatmap(mat_data[, 6:10], name = "Treatment")
ht1 + ht2
```

# 4. Specialized Bioinformatics Plots

## Volcano Plot

```{r}
# Simulate differential expression data
set.seed(123)
volcano_data <- data.frame(
  gene = paste0("Gene", 1:1000),
  log2FC = rnorm(1000, mean = 0, sd = 1),
  pvalue = runif(1000, 0, 1)
) %>%
  mutate(
    padj = p.adjust(pvalue, method = "BH"),
    significant = padj < 0.05 & abs(log2FC) > 1
  )

ggplot(volcano_data, aes(x = log2FC, y = -log10(padj), color = significant)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("gray", "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 Adjusted P-value") +
  theme_minimal()
```

## MA Plot

```{r}
# MA plot (log ratio vs mean average)
ma_data <- data.frame(
  A = runif(1000, 0, 15),  # Average expression
  M = rnorm(1000, 0, 1)     # Log fold change
)

ggplot(ma_data, aes(x = A, y = M)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red") +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  labs(title = "MA Plot", x = "Average Expression", y = "Log Fold Change") +
  theme_minimal()
```

## Manhattan Plot (GWAS)

```{r}
# Simulate GWAS data
set.seed(123)
manhattan_data <- data.frame(
  chr = rep(1:22, each = 100),
  pos = rep(1:100, 22),
  pvalue = runif(2200, 1e-10, 1)
) %>%
  mutate(
    logp = -log10(pvalue)
  )

ggplot(manhattan_data, aes(x = pos, y = logp, color = factor(chr))) +
  geom_point(alpha = 0.6) +
  facet_grid(. ~ chr, scales = "free_x", space = "free_x") +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed", color = "red") +
  labs(title = "Manhattan Plot", x = "Chromosome Position", y = "-Log10 P-value") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        panel.spacing = unit(0.1, "lines"))
```

# 5. Interactive Visualizations

## plotly

```{r eval=FALSE}
library(plotly)

# Convert ggplot to plotly
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3)

ggplotly(p)

# Native plotly
plot_ly(mtcars, x = ~wt, y = ~mpg, color = ~factor(cyl),
        type = "scatter", mode = "markers")

# 3D scatter
plot_ly(mtcars, x = ~wt, y = ~mpg, z = ~hp, color = ~factor(cyl),
        type = "scatter3d", mode = "markers")
```

## ggiraph (Interactive ggplot2)

```{r eval=FALSE}
library(ggiraph)

# Create interactive plot
p <- ggplot(mtcars, aes(x = wt, y = mpg, tooltip = rownames(mtcars))) +
  geom_point_interactive(aes(data_id = rownames(mtcars)), size = 3)

girafe(ggobj = p)
```

# 6. Multi-panel Figures

## patchwork Package

```{r}
library(patchwork)

# Create plots
p1 <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point() + ggtitle("Plot 1")
p2 <- ggplot(mtcars, aes(x = factor(cyl))) + geom_bar() + ggtitle("Plot 2")
p3 <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) + geom_boxplot() + ggtitle("Plot 3")

# Combine plots
p1 + p2 + p3

# Custom layout
(p1 | p2) / p3

# With common legend
p1 + p2 + plot_layout(guides = "collect")
```

## cowplot Package

```{r}
library(cowplot)

# Arrange plots in grid
plot_grid(p1, p2, p3, labels = c("A", "B", "C"), ncol = 2)

# With different sizes
plot_grid(p1, p2, p3, labels = c("A", "B", "C"),
          ncol = 2, rel_widths = c(1, 1.5))
```

# 7. Saving Plots

```{r eval=FALSE}
# Save ggplot2 plot
p <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()

ggsave("plot.png", p, width = 8, height = 6, dpi = 300)
ggsave("plot.pdf", p, width = 8, height = 6)
ggsave("plot.svg", p, width = 8, height = 6)

# Save base R plot
png("baseplot.png", width = 800, height = 600, res = 120)
plot(mtcars$wt, mtcars$mpg)
dev.off()

# PDF for multiple pages
pdf("multipage.pdf", width = 8, height = 6)
plot(mtcars$wt, mtcars$mpg)
hist(mtcars$mpg)
dev.off()
```

# 8. Best Practices

## Color Palettes

```{r}
library(RColorBrewer)

# Display all palettes
# display.brewer.all()

# Use ColorBrewer palettes
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1")

# Viridis (colorblind-friendly)
library(viridis)
ggplot(mtcars, aes(x = wt, y = mpg, color = hp)) +
  geom_point(size = 3) +
  scale_color_viridis_c()
```

## Accessibility

```{r}
# Use colorblind-friendly palettes
# Avoid red-green combinations
# Use shapes in addition to colors

ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl), shape = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_viridis_d()
```

## Publication-Quality Figures

```{r}
# Clean, minimal theme
# High resolution (300+ dpi)
# Clear labels and titles
# Appropriate figure size

p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    title = "Fuel Efficiency vs Vehicle Weight"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

p

# ggsave("figure1.png", p, width = 6, height = 4, dpi = 300)
```

# Summary

This document covered comprehensive data visualization:

**Base R Graphics**:
- Quick and simple
- plot(), hist(), boxplot(), barplot()
- Good for exploratory analysis

**ggplot2**:
- Grammar of graphics
- Highly customizable
- Layered approach
- Best for publication-quality figures

**Specialized Plots**:
- Heatmaps (pheatmap, ComplexHeatmap)
- Volcano plots, MA plots, Manhattan plots
- Bioinformatics-specific visualizations

**Interactive**:
- plotly for web-based interaction
- ggiraph for interactive ggplot2

**Multi-panel**:
- patchwork for combining plots
- cowplot for complex layouts

**Best Practices**:
- Use appropriate plot types
- Choose colorblind-friendly palettes
- Clear labels and titles
- High resolution for publications
- Consider your audience

**Next Steps**: Proceed to Statistical Analysis to learn hypothesis testing, regression, and more.

# References

- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [R Graphics Cookbook](https://r-graphics.org/)
- [Data Visualization with ggplot2 Cheat Sheet](https://github.com/rstudio/cheatsheets)
- [Fundamentals of Data Visualization by Claus Wilke](https://clauswilke.com/dataviz/)
