---
title: "Statistical Analysis in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
```

# 1. Descriptive Statistics

## Measures of Central Tendency

```{r}
# Sample data
data <- c(12, 15, 18, 20, 22, 25, 28, 30, 35, 40)

# Mean
mean(data)
mean(data, na.rm = TRUE)  # Remove NAs

# Median
median(data)

# Mode (no built-in function)
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
get_mode(c(1, 2, 2, 3, 3, 3, 4))

# Weighted mean
weights <- c(1, 1, 2, 2, 2, 2, 1, 1, 1, 1)
weighted.mean(data, weights)
```

## Measures of Dispersion

```{r}
# Variance
var(data)

# Standard deviation
sd(data)

# Range
range(data)
diff(range(data))  # Range value

# Interquartile range
IQR(data)

# Quantiles
quantile(data)
quantile(data, probs = c(0.25, 0.5, 0.75))
quantile(data, probs = seq(0, 1, 0.1))  # Deciles

# Coefficient of variation
cv <- sd(data) / mean(data) * 100
cv
```

## Summary Statistics

```{r}
# Five-number summary
fivenum(data)

# Detailed summary
summary(data)

# For data frames
summary(mtcars)

# Custom summary
library(dplyr)
mtcars %>%
  summarise(
    n = n(),
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    median_mpg = median(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg),
    q25 = quantile(mpg, 0.25),
    q75 = quantile(mpg, 0.75)
  )

# By groups
mtcars %>%
  group_by(cyl) %>%
  summarise(
    n = n(),
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg)
  )
```

## Distribution Shape

```{r}
# Skewness
library(moments)
skewness(data)

# Kurtosis
kurtosis(data)

# Normality test
shapiro.test(data)
```

# 2. Probability Distributions

## Normal Distribution

```{r}
# Density
dnorm(0, mean = 0, sd = 1)  # PDF at x=0

# Cumulative probability
pnorm(1.96, mean = 0, sd = 1)  # P(X <= 1.96)
pnorm(1.96, mean = 0, sd = 1, lower.tail = FALSE)  # P(X > 1.96)

# Quantile function (inverse CDF)
qnorm(0.975, mean = 0, sd = 1)  # 97.5th percentile

# Random generation
set.seed(123)
rnorm(10, mean = 0, sd = 1)

# Visualize
x <- seq(-4, 4, 0.01)
plot(x, dnorm(x), type = "l", main = "Normal Distribution",
     ylab = "Density", xlab = "x")
```

## Other Common Distributions

```{r}
# Uniform
runif(10, min = 0, max = 1)

# Binomial
rbinom(10, size = 10, prob = 0.5)
dbinom(5, size = 10, prob = 0.5)  # P(X = 5)

# Poisson
rpois(10, lambda = 3)
dpois(3, lambda = 3)  # P(X = 3)

# Exponential
rexp(10, rate = 1)

# Chi-square
rchisq(10, df = 5)

# Student's t
rt(10, df = 10)

# F distribution
rf(10, df1 = 5, df2 = 10)
```

# 3. Hypothesis Testing

## One-Sample Tests

### One-Sample t-test

```{r}
# Test if mean differs from hypothesized value
data <- rnorm(30, mean = 5, sd = 2)

# Two-sided test
t.test(data, mu = 5)

# One-sided test
t.test(data, mu = 5, alternative = "greater")
t.test(data, mu = 5, alternative = "less")

# Extract components
result <- t.test(data, mu = 5)
result$statistic
result$p.value
result$conf.int
```

### Wilcoxon Signed-Rank Test (Non-parametric)

```{r}
# For non-normal data
wilcox.test(data, mu = 5)
```

## Two-Sample Tests

### Independent Samples t-test

```{r}
# Create two groups
group1 <- rnorm(30, mean = 5, sd = 2)
group2 <- rnorm(30, mean = 6, sd = 2)

# Assuming equal variances
t.test(group1, group2, var.equal = TRUE)

# Welch's t-test (unequal variances)
t.test(group1, group2, var.equal = FALSE)

# Using formula with data frame
df <- data.frame(
  value = c(group1, group2),
  group = rep(c("A", "B"), each = 30)
)
t.test(value ~ group, data = df)
```

### Paired t-test

```{r}
# For matched pairs
before <- rnorm(20, mean = 100, sd = 15)
after <- before + rnorm(20, mean = 5, sd = 10)

t.test(before, after, paired = TRUE)
t.test(after - before, mu = 0)  # Equivalent
```

### Mann-Whitney U Test (Wilcoxon Rank-Sum)

```{r}
# Non-parametric alternative to t-test
wilcox.test(group1, group2)
```

## Categorical Data Tests

### Chi-Square Test of Independence

```{r}
# Create contingency table
table_data <- matrix(c(20, 10, 15, 25), nrow = 2,
                     dimnames = list(c("Male", "Female"),
                                    c("Treatment", "Control")))
table_data

# Chi-square test
chisq.test(table_data)

# With correction
chisq.test(table_data, correct = FALSE)

# From data frame
chisq.test(table(mtcars$cyl, mtcars$gear))
```

### Fisher's Exact Test

```{r}
# For small sample sizes
fisher.test(table_data)
```

### Chi-Square Goodness of Fit

```{r}
# Test if observed frequencies match expected
observed <- c(25, 30, 20, 25)
expected_prob <- c(0.25, 0.25, 0.25, 0.25)

chisq.test(observed, p = expected_prob)
```

## Multiple Group Comparisons

### One-Way ANOVA

```{r}
# Compare means across multiple groups
group_data <- data.frame(
  value = c(rnorm(20, 5, 1), rnorm(20, 6, 1), rnorm(20, 7, 1)),
  group = rep(c("A", "B", "C"), each = 20)
)

# ANOVA
aov_result <- aov(value ~ group, data = group_data)
summary(aov_result)

# Post-hoc pairwise comparisons
TukeyHSD(aov_result)

# Plot
plot(TukeyHSD(aov_result))
```

### Kruskal-Wallis Test (Non-parametric ANOVA)

```{r}
# Non-parametric alternative
kruskal.test(value ~ group, data = group_data)

# Post-hoc test
pairwise.wilcox.test(group_data$value, group_data$group,
                     p.adjust.method = "BH")
```

### Two-Way ANOVA

```{r}
# Two factors
two_way_data <- data.frame(
  value = rnorm(60),
  factor1 = rep(c("A", "B"), each = 30),
  factor2 = rep(rep(c("X", "Y", "Z"), each = 10), 2)
)

aov_2way <- aov(value ~ factor1 * factor2, data = two_way_data)
summary(aov_2way)
```

## Multiple Testing Correction

```{r}
# Multiple p-values
p_values <- c(0.001, 0.01, 0.03, 0.05, 0.10)

# Bonferroni correction
p.adjust(p_values, method = "bonferroni")

# Benjamini-Hochberg (FDR)
p.adjust(p_values, method = "BH")

# Other methods
p.adjust(p_values, method = "holm")
p.adjust(p_values, method = "hochberg")
```

# 4. Correlation and Association

## Pearson Correlation

```{r}
# For normally distributed continuous data
cor(mtcars$mpg, mtcars$wt)

# With significance test
cor.test(mtcars$mpg, mtcars$wt)

# Correlation matrix
cor(mtcars[, c("mpg", "wt", "hp", "disp")])

# Visualize
library(corrplot)
cor_matrix <- cor(mtcars[, c("mpg", "wt", "hp", "disp")])
corrplot(cor_matrix, method = "circle")
```

## Spearman Correlation

```{r}
# For non-parametric or ordinal data
cor(mtcars$mpg, mtcars$wt, method = "spearman")
cor.test(mtcars$mpg, mtcars$wt, method = "spearman")
```

## Kendall's Tau

```{r}
# Another non-parametric correlation
cor(mtcars$mpg, mtcars$wt, method = "kendall")
cor.test(mtcars$mpg, mtcars$wt, method = "kendall")
```

# 5. Regression Analysis

## Simple Linear Regression

```{r}
# Fit model
model <- lm(mpg ~ wt, data = mtcars)

# Summary
summary(model)

# Coefficients
coef(model)
confint(model)  # Confidence intervals

# Predictions
predict(model, newdata = data.frame(wt = c(2, 3, 4)))
predict(model, newdata = data.frame(wt = c(2, 3, 4)),
        interval = "confidence")

# Residuals
residuals(model)
fitted(model)

# Plot
plot(mtcars$wt, mtcars$mpg)
abline(model, col = "red")

# Diagnostic plots
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

## Multiple Linear Regression

```{r}
# Multiple predictors
model_multi <- lm(mpg ~ wt + hp + cyl, data = mtcars)
summary(model_multi)

# With interaction
model_interact <- lm(mpg ~ wt * hp, data = mtcars)
summary(model_interact)

# Polynomial regression
model_poly <- lm(mpg ~ poly(wt, 2), data = mtcars)
summary(model_poly)
```

## Model Comparison

```{r}
model1 <- lm(mpg ~ wt, data = mtcars)
model2 <- lm(mpg ~ wt + hp, data = mtcars)
model3 <- lm(mpg ~ wt + hp + cyl, data = mtcars)

# Compare with ANOVA
anova(model1, model2, model3)

# AIC comparison
AIC(model1, model2, model3)

# BIC comparison
BIC(model1, model2, model3)
```

## Assumptions and Diagnostics

```{r}
# Normality of residuals
shapiro.test(residuals(model))

# Homoscedasticity
library(lmtest)
bptest(model)  # Breusch-Pagan test

# Multicollinearity (VIF)
library(car)
vif(model_multi)

# Outliers and influential points
# Cook's distance
cooks.distance(model)
plot(cooks.distance(model))

# Leverage
hatvalues(model)
```

## Logistic Regression

```{r}
# Binary outcome
# Create binary outcome
mtcars$high_mpg <- ifelse(mtcars$mpg > median(mtcars$mpg), 1, 0)

# Fit logistic regression
logit_model <- glm(high_mpg ~ wt + hp, data = mtcars, family = binomial)
summary(logit_model)

# Odds ratios
exp(coef(logit_model))
exp(confint(logit_model))

# Predictions (probabilities)
predict(logit_model, type = "response")

# Predictions (classes)
predicted_class <- ifelse(predict(logit_model, type = "response") > 0.5, 1, 0)

# Confusion matrix
table(Predicted = predicted_class, Actual = mtcars$high_mpg)

# Model performance
library(pROC)
roc_obj <- roc(mtcars$high_mpg, predict(logit_model, type = "response"))
auc(roc_obj)
plot(roc_obj)
```

## Generalized Linear Models (GLM)

```{r}
# Poisson regression (count data)
poisson_model <- glm(hp ~ wt + cyl, data = mtcars, family = poisson)
summary(poisson_model)

# Gamma regression (positive continuous)
gamma_model <- glm(mpg ~ wt + hp, data = mtcars, family = Gamma)
summary(gamma_model)
```

# 6. Survival Analysis

```{r}
library(survival)

# Create survival object
# Simulated survival data
surv_data <- data.frame(
  time = c(5, 8, 12, 15, 20, 25, 30, 35, 40, 45),
  status = c(1, 1, 0, 1, 1, 0, 1, 0, 1, 1),  # 1 = event, 0 = censored
  group = rep(c("A", "B"), each = 5)
)

# Survival object
surv_obj <- Surv(time = surv_data$time, event = surv_data$status)

# Kaplan-Meier survival curve
km_fit <- survfit(surv_obj ~ group, data = surv_data)
summary(km_fit)

# Plot survival curves
plot(km_fit, col = c("red", "blue"), main = "Kaplan-Meier Survival Curves")
legend("topright", legend = c("Group A", "Group B"), col = c("red", "blue"), lty = 1)

# Log-rank test
survdiff(surv_obj ~ group, data = surv_data)

# Cox proportional hazards model
cox_model <- coxph(surv_obj ~ group, data = surv_data)
summary(cox_model)
```

# 7. Non-Parametric Tests

```{r}
# Sign test
library(BSDA)
# SIGN.test(before, after)

# Friedman test (non-parametric repeated measures ANOVA)
# Create repeated measures data
repeated_data <- matrix(rnorm(60), ncol = 3)
friedman.test(repeated_data)

# Kolmogorov-Smirnov test (compare distributions)
ks.test(rnorm(100), "pnorm")  # Test against normal
ks.test(rnorm(100), rnorm(100))  # Two-sample
```

# 8. Power Analysis

```{r}
library(pwr)

# Power for t-test
# Given: effect size, alpha, power → calculate sample size
pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.8, type = "two.sample")

# Given: effect size, alpha, n → calculate power
pwr.t.test(d = 0.5, sig.level = 0.05, n = 64, type = "two.sample")

# Power for ANOVA
pwr.anova.test(k = 3, f = 0.25, sig.level = 0.05, power = 0.8)

# Power for correlation
pwr.r.test(r = 0.3, sig.level = 0.05, power = 0.8)

# Power for chi-square
pwr.chisq.test(w = 0.3, df = 2, sig.level = 0.05, power = 0.8)
```

# 9. Bayesian Statistics

```{r eval=FALSE}
library(rstan)
library(brms)

# Bayesian linear regression
bayesian_model <- brm(mpg ~ wt + hp, data = mtcars, family = gaussian())
summary(bayesian_model)

# Posterior distributions
plot(bayesian_model)

# Predictions with uncertainty
posterior_predict(bayesian_model, newdata = data.frame(wt = 3, hp = 100))
```

# 10. Mixed Models

```{r}
library(lme4)

# Linear mixed model (random intercepts)
# Example with simulated hierarchical data
set.seed(123)
hier_data <- data.frame(
  subject = rep(1:10, each = 5),
  time = rep(1:5, 10),
  value = rnorm(50) + rep(rnorm(10), each = 5)
)

lmer_model <- lmer(value ~ time + (1|subject), data = hier_data)
summary(lmer_model)

# Random slopes and intercepts
lmer_model2 <- lmer(value ~ time + (time|subject), data = hier_data)
summary(lmer_model2)

# Generalized linear mixed model
glmer_model <- glmer(cbind(value > 0, value <= 0) ~ time + (1|subject),
                     data = hier_data, family = binomial)
```

# 11. Time Series Analysis

```{r}
# Create time series
ts_data <- ts(rnorm(100), start = c(2020, 1), frequency = 12)

# Plot
plot(ts_data)

# Decomposition
decomposed <- decompose(ts_data)
plot(decomposed)

# ACF and PACF
acf(ts_data)
pacf(ts_data)

# ARIMA model
library(forecast)
arima_model <- auto.arima(ts_data)
summary(arima_model)

# Forecast
forecasted <- forecast(arima_model, h = 12)
plot(forecasted)
```

# 12. Bootstrap and Resampling

```{r}
library(boot)

# Define statistic function
mean_func <- function(data, indices) {
  return(mean(data[indices]))
}

# Bootstrap
boot_result <- boot(data = mtcars$mpg, statistic = mean_func, R = 1000)
boot_result

# Confidence interval
boot.ci(boot_result, type = "perc")

# Plot
plot(boot_result)
```

# 13. Effect Sizes

```{r}
library(effectsize)

# Cohen's d (standardized mean difference)
cohens_d(mpg ~ am, data = mtcars)

# Eta squared (ANOVA effect size)
aov_result <- aov(mpg ~ factor(cyl), data = mtcars)
eta_squared(aov_result)

# Odds ratio
# For contingency tables
```

# Summary

This document covered comprehensive statistical analysis:

**Descriptive Statistics**:
- Central tendency, dispersion, distribution shape

**Probability Distributions**:
- Normal, binomial, Poisson, etc.

**Hypothesis Testing**:
- t-tests, chi-square, ANOVA
- Parametric and non-parametric tests
- Multiple testing correction

**Correlation**:
- Pearson, Spearman, Kendall

**Regression**:
- Linear, multiple, logistic
- GLMs, mixed models

**Survival Analysis**:
- Kaplan-Meier, Cox regression

**Advanced Topics**:
- Power analysis
- Bayesian statistics
- Time series
- Bootstrap

**Key Principles**:
- Check assumptions
- Use appropriate tests
- Report effect sizes
- Correct for multiple testing
- Visualize results

**Next Steps**: Proceed to Bioconductor Ecosystem for bioinformatics-specific analysis.

# References

- [R for Data Science](https://r4ds.had.co.nz/)
- [An Introduction to Statistical Learning](https://www.statlearning.com/)
- [Modern Statistics for Modern Biology](https://www.huber.embl.de/msmb/)
- [The R Book by Michael Crawley](https://www.wiley.com/en-us/The+R+Book%2C+2nd+Edition-p-9780470973929)
