---
title: "Bioconductor Ecosystem"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Introduction to Bioconductor

Bioconductor is an open-source software project providing tools for bioinformatics and computational biology in R.

## Installation

```{r}
# Install BiocManager
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install Bioconductor packages
BiocManager::install("GenomicRanges")
BiocManager::install(c("DESeq2", "edgeR", "limma"))

# Check Bioconductor version
BiocManager::version()

# Update packages
BiocManager::install()
```

## Core Packages

```{r}
# Essential packages
BiocManager::install(c(
  "Biostrings",        # Sequence manipulation
  "GenomicRanges",     # Genomic intervals
  "GenomicFeatures",   # Gene annotations
  "SummarizedExperiment", # Data containers
  "rtracklayer",       # Import/export genomic data
  "AnnotationDbi",     # Annotation databases
  "biomaRt"            # BioMart queries
))
```

# 2. Core Data Structures

## GenomicRanges

```{r}
library(GenomicRanges)

# Create GRanges object
gr <- GRanges(
  seqnames = c("chr1", "chr1", "chr2"),
  ranges = IRanges(start = c(100, 200, 300), end = c(150, 250, 350)),
  strand = c("+", "-", "+"),
  score = c(5, 7, 3),
  gene_id = c("gene1", "gene2", "gene3")
)
gr

# Access components
seqnames(gr)
ranges(gr)
start(gr)
end(gr)
width(gr)
strand(gr)
mcols(gr)  # Metadata columns

# Subsetting
gr[1]
gr[strand(gr) == "+"]

# Extend ranges
resize(gr, width = 100)
flank(gr, width = 50)
promoters(gr, upstream = 2000, downstream = 200)

# Operations
shift(gr, 10)  # Shift positions
narrow(gr, start = 10, end = -10)  # Narrow ranges
```

## IRanges

```{r}
library(IRanges)

# Create interval ranges
ir <- IRanges(start = c(1, 5, 10), end = c(4, 8, 15))
ir

# Properties
start(ir)
end(ir)
width(ir)

# Operations
reduce(ir)  # Merge overlapping ranges
disjoin(ir)  # Create disjoint ranges
gaps(ir)  # Find gaps between ranges
coverage(ir)  # Coverage vector
```

## GRangesList

```{r}
# List of GRanges (e.g., exons per transcript)
grl <- GRangesList(
  gene1 = GRanges("chr1", IRanges(c(100, 200, 300), c(150, 250, 350))),
  gene2 = GRanges("chr2", IRanges(c(400, 500), c(450, 550)))
)
grl

# Access
grl[[1]]
grl$gene1

# Operations
unlist(grl)  # Flatten to GRanges
```

## Range Operations

```{r}
gr1 <- GRanges("chr1", IRanges(c(1, 5, 10), c(4, 8, 15)))
gr2 <- GRanges("chr1", IRanges(c(3, 7, 12), c(6, 9, 16)))

# Find overlaps
findOverlaps(gr1, gr2)
countOverlaps(gr1, gr2)
subsetByOverlaps(gr1, gr2)

# Set operations
union(gr1, gr2)
intersect(gr1, gr2)
setdiff(gr1, gr2)

# Distance
distance(gr1, gr2)
distanceToNearest(gr1, gr2)

# Nearest features
nearest(gr1, gr2)
precede(gr1, gr2)
follow(gr1, gr2)
```

## SummarizedExperiment

```{r}
library(SummarizedExperiment)

# Create SummarizedExperiment
counts <- matrix(rpois(100, lambda = 10), nrow = 10, ncol = 10)
rownames(counts) <- paste0("Gene", 1:10)
colnames(counts) <- paste0("Sample", 1:10)

rowData <- GRanges("chr1", IRanges(seq(100, 1000, length.out = 10),
                                   width = 50))
colData <- DataFrame(
  condition = rep(c("Control", "Treatment"), each = 5),
  batch = rep(1:2, 5)
)

se <- SummarizedExperiment(
  assays = list(counts = counts),
  rowRanges = rowData,
  colData = colData
)
se

# Access
assay(se)
assay(se, "counts")
rowRanges(se)
colData(se)

# Subset
se[1:5, 1:5]
se[, se$condition == "Control"]
```

# 3. Sequence Analysis

## Biostrings

```{r}
library(Biostrings)

# DNA sequences
dna <- DNAString("ATCGATCGATCG")
dna

# Multiple sequences
dna_set <- DNAStringSet(c("ATCG", "GCTA", "TTAA"))
dna_set

# Basic operations
length(dna)
width(dna_set)
reverseComplement(dna)
translate(dna)

# Pattern matching
matchPattern("ATG", dna)
countPattern("AT", dna)

# GC content
letterFrequency(dna, "GC", as.prob = TRUE)

# Read FASTA
# sequences <- readDNAStringSet("sequences.fasta")
# writeXStringSet(sequences, "output.fasta")
```

## BSgenome

```{r}
library(BSgenome)

# Available genomes
# available.genomes()

# Load human genome
library(BSgenome.Hsapiens.UCSC.hg38)
genome <- BSgenome.Hsapiens.UCSC.hg38

# Get sequence
getSeq(genome, GRanges("chr1", IRanges(1000, 2000)))

# Multiple regions
regions <- GRanges(c("chr1", "chr2"), IRanges(c(1000, 2000), c(1100, 2100)))
getSeq(genome, regions)
```

# 4. Gene Annotation

## GenomicFeatures

```{r}
library(GenomicFeatures)

# Load annotation from GTF
# txdb <- makeTxDbFromGFF("genes.gtf", format = "gtf")

# Or use pre-built annotations
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# Extract features
genes_gr <- genes(txdb)
transcripts_gr <- transcripts(txdb)
exons_gr <- exons(txdb)
cds_gr <- cds(txdb)

# Exons by gene
exonsByGene <- exonsBy(txdb, by = "gene")

# Introns
intronsByTranscript <- intronsByTranscript(txdb)

# 5' and 3' UTRs
fiveUTRsByTranscript <- fiveUTRsByTranscript(txdb)
threeUTRsByTranscript <- threeUTRsByTranscript(txdb)

# Promoters
promoters(txdb, upstream = 2000, downstream = 200)
```

## AnnotationDbi

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)

# Available keys
keytypes(org.Hs.eg.db)

# Gene symbols to Entrez IDs
symbols <- c("TP53", "BRCA1", "EGFR")
select(org.Hs.eg.db, keys = symbols, columns = c("ENTREZID", "ENSEMBL"),
       keytype = "SYMBOL")

# Map between identifiers
mapIds(org.Hs.eg.db, keys = symbols, column = "ENTREZID", keytype = "SYMBOL")

# Get gene information
genes_info <- select(org.Hs.eg.db,
                     keys = symbols,
                     columns = c("ENTREZID", "GENENAME", "CHR"),
                     keytype = "SYMBOL")
genes_info
```

## biomaRt

```{r}
library(biomaRt)

# List marts
# listMarts()

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Available attributes
# listAttributes(mart)

# Available filters
# listFilters(mart)

# Query
genes <- c("ENSG00000141510", "ENSG00000012048")
getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "chromosome_name",
                     "start_position", "end_position"),
      filters = "ensembl_gene_id",
      values = genes,
      mart = mart)

# Get sequences
# getSequence(id = genes, type = "ensembl_gene_id", seqType = "gene_exon", mart = mart)
```

# 5. RNA-seq Analysis

## DESeq2

```{r}
library(DESeq2)

# Create count matrix (genes × samples)
counts <- matrix(rpois(1000, lambda = 10), nrow = 100, ncol = 10)
rownames(counts) <- paste0("Gene", 1:100)
colnames(counts) <- paste0("Sample", 1:10)

# Sample information
coldata <- data.frame(
  condition = factor(rep(c("Control", "Treatment"), each = 5)),
  row.names = colnames(counts)
)

# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = counts,
                               colData = coldata,
                               design = ~ condition)

# Run DESeq2 pipeline
dds <- DESeq(dds)

# Get results
res <- results(dds, contrast = c("condition", "Treatment", "Control"))
res

# Order by p-value
res_ordered <- res[order(res$padj), ]

# Extract significant genes
sig_genes <- subset(res, padj < 0.05 & abs(log2FoldChange) > 1)

# Visualization
plotMA(res)
plotCounts(dds, gene = rownames(res)[1], intgroup = "condition")

# Volcano plot
library(ggplot2)
res_df <- as.data.frame(res)
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = padj < 0.05 & abs(log2FoldChange) > 1)) +
  theme_minimal()

# PCA plot
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = "condition")

# Heatmap of top genes
library(pheatmap)
top_genes <- head(order(res$padj), 20)
pheatmap(assay(vsd)[top_genes, ],
         annotation_col = as.data.frame(colData(dds)["condition"]),
         scale = "row")
```

## edgeR

```{r}
library(edgeR)

# Create DGEList object
y <- DGEList(counts = counts, group = coldata$condition)

# Filter low-count genes
keep <- filterByExpr(y)
y <- y[keep, , keep.lib.sizes = FALSE]

# Normalization
y <- calcNormFactors(y)

# Design matrix
design <- model.matrix(~ condition, data = coldata)

# Estimate dispersion
y <- estimateDisp(y, design)

# Fit model
fit <- glmQLFit(y, design)

# Differential expression test
qlf <- glmQLFTest(fit, coef = 2)

# Top genes
topTags(qlf)

# Extract results
results_edgeR <- as.data.frame(topTags(qlf, n = Inf))
```

## limma-voom

```{r}
library(limma)

# Design matrix
design <- model.matrix(~ condition, data = coldata)

# Create DGEList
y <- DGEList(counts = counts)

# Filter and normalize
keep <- filterByExpr(y, design)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)

# Voom transformation
v <- voom(y, design, plot = TRUE)

# Fit linear model
fit <- lmFit(v, design)
fit <- eBayes(fit)

# Extract results
results_limma <- topTable(fit, coef = 2, n = Inf)
```

# 6. ChIP-seq and Peak Analysis

## DiffBind

```{r}
library(DiffBind)

# Read sample sheet
# samples <- dba(sampleSheet = "samples.csv")

# Count reads in peaks
# samples <- dba.count(samples)

# Establish contrast
# samples <- dba.contrast(samples, categories = DBA_CONDITION)

# Differential binding analysis
# samples <- dba.analyze(samples)

# Get results
# results <- dba.report(samples)

# Visualization
# dba.plotPCA(samples)
# dba.plotHeatmap(samples)
```

## ChIPseeker

```{r}
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Read peaks
# peaks <- readPeakFile("peaks.bed")

# Annotate peaks
# peakAnno <- annotatePeak(peaks, tssRegion = c(-3000, 3000),
#                          TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene)

# Visualize
# plotAnnoPie(peakAnno)
# plotDistToTSS(peakAnno)
```

# 7. Single-Cell RNA-seq

## Seurat

```{r}
library(Seurat)

# Create Seurat object
# counts <- Read10X("filtered_gene_bc_matrices/hg38/")
# seurat_obj <- CreateSeuratObject(counts = counts, project = "sample")

# Standard workflow
# seurat_obj <- NormalizeData(seurat_obj)
# seurat_obj <- FindVariableFeatures(seurat_obj)
# seurat_obj <- ScaleData(seurat_obj)
# seurat_obj <- RunPCA(seurat_obj)
# seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10)
# seurat_obj <- FindClusters(seurat_obj, resolution = 0.5)
# seurat_obj <- RunUMAP(seurat_obj, dims = 1:10)

# Visualization
# DimPlot(seurat_obj, reduction = "umap")
# FeaturePlot(seurat_obj, features = c("CD3D", "CD8A"))

# Find markers
# markers <- FindAllMarkers(seurat_obj)
# top10 <- markers %>% group_by(cluster) %>% top_n(10, avg_log2FC)
# DoHeatmap(seurat_obj, features = top10$gene)
```

## SingleCellExperiment

```{r}
library(SingleCellExperiment)

# Create SCE object
counts <- matrix(rpois(1000, lambda = 5), nrow = 100, ncol = 10)
sce <- SingleCellExperiment(assays = list(counts = counts))

# Add cell metadata
colData(sce)$cellType <- sample(c("TypeA", "TypeB"), 10, replace = TRUE)

# Add gene metadata
rowData(sce)$geneSymbol <- paste0("Gene", 1:100)

# Reduced dimensions
reducedDim(sce, "PCA") <- matrix(rnorm(20), nrow = 10, ncol = 2)

# Access
assay(sce, "counts")
colData(sce)
rowData(sce)
reducedDim(sce, "PCA")
```

# 8. Variant Calling and Analysis

## VariantAnnotation

```{r}
library(VariantAnnotation)

# Read VCF
# vcf <- readVcf("variants.vcf", "hg38")

# Access components
# rowRanges(vcf)  # Genomic positions
# ref(vcf)  # Reference alleles
# alt(vcf)  # Alternate alleles
# info(vcf)  # Info fields
# geno(vcf)  # Genotype data

# Filter variants
# vcf_filtered <- vcf[qual(vcf) > 30]

# Write VCF
# writeVcf(vcf, "output.vcf")
```

# 9. Pathway and Gene Set Enrichment

## clusterProfiler

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# Gene list (Entrez IDs)
genes <- c("1", "2", "3", "4", "5")  # Example

# GO enrichment
ego <- enrichGO(gene = genes,
                OrgDb = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont = "BP",  # Biological Process
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.05)

# View results
head(ego)

# Visualization
# dotplot(ego)
# barplot(ego)
# enrichMap(ego)

# KEGG enrichment
# keg <- enrichKEGG(gene = genes,
#                   organism = "hsa",
#                   pvalueCutoff = 0.05)

# GSEA (Gene Set Enrichment Analysis)
# Ranked gene list
gene_list <- sort(rnorm(1000), decreasing = TRUE)
names(gene_list) <- paste0("Gene", 1:1000)

# gseGO(geneList = gene_list,
#       OrgDb = org.Hs.eg.db,
#       ont = "BP",
#       pvalueCutoff = 0.05)
```

## fgsea

```{r}
library(fgsea)

# Ranked gene list
ranks <- sort(rnorm(1000), decreasing = TRUE)
names(ranks) <- paste0("Gene", 1:1000)

# Gene sets (pathways)
pathways <- list(
  pathway1 = paste0("Gene", 1:50),
  pathway2 = paste0("Gene", 51:100),
  pathway3 = paste0("Gene", 101:150)
)

# Run GSEA
fgsea_results <- fgsea(pathways = pathways,
                       stats = ranks,
                       minSize = 15,
                       maxSize = 500)

fgsea_results

# Visualization
# plotEnrichment(pathways[[1]], ranks)
```

# 10. Import/Export Genomic Data

## rtracklayer

```{r}
library(rtracklayer)

# Import BED
# bed <- import.bed("regions.bed")

# Import GTF/GFF
# gtf <- import.gff("genes.gtf")

# Import BigWig
# bw <- import.bw("coverage.bw")

# Export GRanges to BED
# export.bed(gr, "output.bed")

# Export to GTF
# export.gff(gr, "output.gtf")

# Export to BigWig
# export.bw(gr, "output.bw")

# Generic import/export
# data <- import("file.bed", format = "bed")
# export(gr, "output.bed", format = "bed")
```

# 11. Genomic Visualization

## Gviz

```{r}
library(Gviz)

# Genome axis
gtrack <- GenomeAxisTrack()

# Gene model track
# data(geneModels)
# grtrack <- GeneRegionTrack(geneModels, genome = "hg19", chromosome = "chr7",
#                           name = "Genes")

# Data track
# dtrack <- DataTrack(range = gr, data = "score", chromosome = "chr1",
#                    genome = "hg38", name = "Coverage")

# Plot tracks
# plotTracks(list(gtrack, grtrack, dtrack), from = 1000, to = 10000)
```

## karyoploteR

```{r}
library(karyoploteR)

# Create ideogram
# kp <- plotKaryotype(genome = "hg38")

# Add data
# kpPoints(kp, data = gr, y = gr$score)
# kpPlotRegions(kp, data = gr)
```

# Summary

This document covered the Bioconductor ecosystem:

**Core Data Structures**:
- GRanges, IRanges, GRangesList
- SummarizedExperiment

**Sequence Analysis**:
- Biostrings, BSgenome

**Gene Annotation**:
- GenomicFeatures, AnnotationDbi, biomaRt

**RNA-seq**:
- DESeq2, edgeR, limma-voom

**ChIP-seq**:
- DiffBind, ChIPseeker

**Single-Cell**:
- Seurat, SingleCellExperiment

**Variant Analysis**:
- VariantAnnotation

**Pathway Analysis**:
- clusterProfiler, fgsea

**Visualization**:
- Gviz, karyoploteR

**Import/Export**:
- rtracklayer

**Key Principles**:
- Use appropriate data structures
- Leverage existing annotations
- Follow standard workflows
- Document analysis thoroughly

**Next Steps**: Proceed to Machine Learning for predictive modeling and classification.

# References

- [Bioconductor](https://bioconductor.org/)
- [Bioconductor Workflows](https://bioconductor.org/packages/release/BiocViews.html#___Workflow)
- [Orchestrating Single-Cell Analysis with Bioconductor](https://osca.bioconductor.org/)
- [RNA-seq analysis with DESeq2](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)
