---
title: "Machine Learning in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
```

# 1. Introduction to Machine Learning in R

Machine learning can be divided into:
- **Supervised Learning**: Classification and Regression
- **Unsupervised Learning**: Clustering and Dimensionality Reduction
- **Semi-supervised and Reinforcement Learning**

# 2. Data Preparation

## Train-Test Split

```{r}
# Load example dataset
data(iris)

# Set seed for reproducibility
set.seed(123)

# Simple random split
train_indices <- sample(1:nrow(iris), 0.7 * nrow(iris))
train_data <- iris[train_indices, ]
test_data <- iris[-train_indices, ]

# Using caret
library(caret)
train_index <- createDataPartition(iris$Species, p = 0.7, list = FALSE)
train_data <- iris[train_index, ]
test_data <- iris[-train_index, ]
```

## Feature Scaling

```{r}
# Normalization (0-1 range)
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
iris_normalized <- as.data.frame(lapply(iris[, 1:4], normalize))

# Standardization (z-score)
iris_scaled <- as.data.frame(scale(iris[, 1:4]))

# Using caret
preproc <- preProcess(iris[, 1:4], method = c("center", "scale"))
iris_processed <- predict(preproc, iris[, 1:4])
```

## Handling Missing Values

```{r}
# Create data with missing values
data_na <- iris
data_na[sample(1:150, 10), 1] <- NA

# Remove rows with NA
data_complete <- na.omit(data_na)

# Impute with mean
data_na$Sepal.Length[is.na(data_na$Sepal.Length)] <-
  mean(data_na$Sepal.Length, na.rm = TRUE)

# Impute with caret
library(caret)
preproc <- preProcess(data_na[, 1:4], method = "medianImpute")
data_imputed <- predict(preproc, data_na[, 1:4])
```

# 3. Supervised Learning - Classification

## Logistic Regression

```{r}
# Binary classification
iris_binary <- iris[iris$Species != "setosa", ]
iris_binary$Species <- droplevels(iris_binary$Species)

# Fit model
logit_model <- glm(Species ~ Sepal.Length + Sepal.Width,
                   data = iris_binary, family = binomial)
summary(logit_model)

# Predictions
predictions <- predict(logit_model, type = "response")
predicted_class <- ifelse(predictions > 0.5, "virginica", "versicolor")

# Confusion matrix
table(Predicted = predicted_class, Actual = iris_binary$Species)

# Accuracy
mean(predicted_class == iris_binary$Species)
```

## K-Nearest Neighbors (KNN)

```{r}
library(class)

# Prepare data
train_features <- train_data[, 1:4]
test_features <- test_data[, 1:4]
train_labels <- train_data$Species
test_labels <- test_data$Species

# KNN classification
knn_pred <- knn(train = train_features,
                test = test_features,
                cl = train_labels,
                k = 5)

# Confusion matrix
table(Predicted = knn_pred, Actual = test_labels)

# Accuracy
mean(knn_pred == test_labels)

# Try different k values
accuracy_k <- sapply(1:20, function(k) {
  pred <- knn(train_features, test_features, train_labels, k = k)
  mean(pred == test_labels)
})

# Plot
plot(1:20, accuracy_k, type = "b", xlab = "k", ylab = "Accuracy")
```

## Decision Trees

```{r}
library(rpart)
library(rpart.plot)

# Fit decision tree
tree_model <- rpart(Species ~ ., data = train_data, method = "class")

# Visualize tree
rpart.plot(tree_model)

# Predictions
tree_pred <- predict(tree_model, test_data, type = "class")

# Confusion matrix
confusionMatrix(tree_pred, test_data$Species)

# Variable importance
tree_model$variable.importance
```

## Random Forest

```{r}
library(randomForest)

# Fit random forest
rf_model <- randomForest(Species ~ ., data = train_data, ntree = 500)
print(rf_model)

# Predictions
rf_pred <- predict(rf_model, test_data)

# Confusion matrix
confusionMatrix(rf_pred, test_data$Species)

# Variable importance
importance(rf_model)
varImpPlot(rf_model)

# Tune mtry parameter
tuned_rf <- tuneRF(train_data[, -5], train_data$Species,
                   stepFactor = 1.5, improve = 0.01, ntree = 500)
```

## Support Vector Machines (SVM)

```{r}
library(e1071)

# Fit SVM
svm_model <- svm(Species ~ ., data = train_data, kernel = "radial")

# Predictions
svm_pred <- predict(svm_model, test_data)

# Confusion matrix
confusionMatrix(svm_pred, test_data$Species)

# Tune parameters
svm_tune <- tune(svm, Species ~ ., data = train_data,
                 ranges = list(cost = c(0.1, 1, 10),
                              gamma = c(0.1, 1, 10)))
print(svm_tune)

# Best model
best_svm <- svm_tune$best.model
svm_pred_tuned <- predict(best_svm, test_data)
confusionMatrix(svm_pred_tuned, test_data$Species)
```

## Naive Bayes

```{r}
library(e1071)

# Fit Naive Bayes
nb_model <- naiveBayes(Species ~ ., data = train_data)

# Predictions
nb_pred <- predict(nb_model, test_data)

# Confusion matrix
confusionMatrix(nb_pred, test_data$Species)
```

## Gradient Boosting (XGBoost)

```{r}
library(xgboost)

# Prepare data for xgboost (requires numeric matrix)
train_matrix <- model.matrix(Species ~ . - 1, data = train_data)
train_label <- as.numeric(train_data$Species) - 1

test_matrix <- model.matrix(Species ~ . - 1, data = test_data)
test_label <- as.numeric(test_data$Species) - 1

# Create DMatrix
dtrain <- xgb.DMatrix(data = train_matrix, label = train_label)
dtest <- xgb.DMatrix(data = test_matrix, label = test_label)

# Set parameters
params <- list(
  objective = "multi:softmax",
  num_class = 3,
  max_depth = 6,
  eta = 0.3
)

# Train model
xgb_model <- xgb.train(params = params, data = dtrain, nrounds = 100)

# Predictions
xgb_pred <- predict(xgb_model, dtest)

# Confusion matrix
table(Predicted = xgb_pred, Actual = test_label)

# Feature importance
importance_matrix <- xgb.importance(colnames(train_matrix), model = xgb_model)
xgb.plot.importance(importance_matrix)
```

# 4. Supervised Learning - Regression

## Linear Regression

```{r}
# Use mtcars dataset
data(mtcars)
set.seed(123)
train_idx <- sample(1:nrow(mtcars), 0.7 * nrow(mtcars))
train <- mtcars[train_idx, ]
test <- mtcars[-train_idx, ]

# Fit model
lm_model <- lm(mpg ~ wt + hp + cyl, data = train)
summary(lm_model)

# Predictions
lm_pred <- predict(lm_model, test)

# Evaluation
library(Metrics)
rmse(test$mpg, lm_pred)
mae(test$mpg, lm_pred)
cor(test$mpg, lm_pred)^2  # R-squared

# Plot
plot(test$mpg, lm_pred, xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")
```

## Random Forest Regression

```{r}
# Fit random forest for regression
rf_reg <- randomForest(mpg ~ ., data = train, ntree = 500)

# Predictions
rf_pred <- predict(rf_reg, test)

# Evaluation
rmse(test$mpg, rf_pred)
mae(test$mpg, rf_pred)

# Variable importance
importance(rf_reg)
```

## XGBoost Regression

```{r}
# Prepare data
train_matrix <- as.matrix(train[, -1])
train_label <- train$mpg

test_matrix <- as.matrix(test[, -1])
test_label <- test$mpg

# Create DMatrix
dtrain <- xgb.DMatrix(data = train_matrix, label = train_label)
dtest <- xgb.DMatrix(data = test_matrix, label = test_label)

# Parameters
params <- list(
  objective = "reg:squarederror",
  max_depth = 6,
  eta = 0.1
)

# Train
xgb_reg <- xgb.train(params = params, data = dtrain, nrounds = 100)

# Predictions
xgb_pred <- predict(xgb_reg, dtest)

# Evaluation
rmse(test_label, xgb_pred)
mae(test_label, xgb_pred)
```

# 5. Model Evaluation

## Cross-Validation

```{r}
library(caret)

# 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Train with cross-validation
cv_model <- train(Species ~ ., data = iris, method = "rf",
                  trControl = train_control)
print(cv_model)

# Repeated cross-validation
train_control_repeat <- trainControl(method = "repeatedcv",
                                     number = 10, repeats = 3)
cv_model_repeat <- train(Species ~ ., data = iris, method = "rf",
                         trControl = train_control_repeat)
```

## ROC Curve and AUC

```{r}
library(pROC)

# Binary classification example
iris_binary <- iris[iris$Species != "setosa", ]
iris_binary$Species <- factor(ifelse(iris_binary$Species == "virginica", 1, 0))

# Fit logistic regression
logit <- glm(Species ~ ., data = iris_binary, family = binomial)
predictions <- predict(logit, type = "response")

# ROC curve
roc_obj <- roc(iris_binary$Species, predictions)
plot(roc_obj, main = "ROC Curve")

# AUC
auc(roc_obj)

# Optimal threshold
coords(roc_obj, "best")
```

## Performance Metrics

```{r}
library(caret)

# Confusion matrix with detailed metrics
predictions <- predict(rf_model, test_data)
cm <- confusionMatrix(predictions, test_data$Species)
print(cm)

# Extract metrics
cm$overall['Accuracy']
cm$byClass['Sensitivity']
cm$byClass['Specificity']
cm$byClass['Precision']
cm$byClass['F1']
```

# 6. Unsupervised Learning - Clustering

## K-Means Clustering

```{r}
# Prepare data (remove Species column)
iris_features <- iris[, 1:4]

# K-means with k=3
set.seed(123)
kmeans_result <- kmeans(iris_features, centers = 3, nstart = 25)

# Cluster assignments
kmeans_result$cluster

# Cluster centers
kmeans_result$centers

# Visualize
library(factoextra)
fviz_cluster(kmeans_result, data = iris_features)

# Determine optimal k (Elbow method)
wss <- sapply(1:10, function(k) {
  kmeans(iris_features, k, nstart = 10)$tot.withinss
})
plot(1:10, wss, type = "b", xlab = "Number of Clusters", ylab = "Within-cluster SS")

# Silhouette method
fviz_nbclust(iris_features, kmeans, method = "silhouette")

# Gap statistic
fviz_nbclust(iris_features, kmeans, method = "gap_stat")
```

## Hierarchical Clustering

```{r}
# Calculate distance matrix
dist_matrix <- dist(iris_features, method = "euclidean")

# Hierarchical clustering
hc <- hclust(dist_matrix, method = "complete")

# Dendrogram
plot(hc, labels = iris$Species, main = "Hierarchical Clustering")
rect.hclust(hc, k = 3, border = "red")

# Cut tree to get clusters
clusters <- cutree(hc, k = 3)

# Visualize
fviz_dend(hc, k = 3, rect = TRUE)

# Different linkage methods
hc_single <- hclust(dist_matrix, method = "single")
hc_average <- hclust(dist_matrix, method = "average")
hc_ward <- hclust(dist_matrix, method = "ward.D2")
```

## DBSCAN

```{r}
library(dbscan)

# DBSCAN clustering
db <- dbscan(iris_features, eps = 0.5, minPts = 5)

# Cluster assignments (0 = noise)
db$cluster

# Visualize
fviz_cluster(db, data = iris_features, geom = "point")

# Find optimal eps
kNNdistplot(iris_features, k = 5)
abline(h = 0.5, col = "red")
```

# 7. Dimensionality Reduction

## Principal Component Analysis (PCA)

```{r}
# Perform PCA
pca_result <- prcomp(iris[, 1:4], center = TRUE, scale. = TRUE)

# Summary
summary(pca_result)

# Loadings
pca_result$rotation

# Scores
head(pca_result$x)

# Scree plot
plot(pca_result, type = "l")
fviz_eig(pca_result)

# Biplot
biplot(pca_result)
fviz_pca_biplot(pca_result, col.ind = iris$Species)

# Variance explained
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
plot(var_explained, type = "b", xlab = "PC", ylab = "Variance Explained")

# Cumulative variance
cumsum(var_explained)
```

## t-SNE

```{r}
library(Rtsne)

# Run t-SNE
set.seed(123)
tsne_result <- Rtsne(as.matrix(iris[, 1:4]), dims = 2, perplexity = 30)

# Plot
tsne_data <- data.frame(
  tsne1 = tsne_result$Y[, 1],
  tsne2 = tsne_result$Y[, 2],
  Species = iris$Species
)

ggplot(tsne_data, aes(x = tsne1, y = tsne2, color = Species)) +
  geom_point(size = 2) +
  theme_minimal()
```

## UMAP

```{r}
library(umap)

# Run UMAP
umap_result <- umap(iris[, 1:4])

# Plot
umap_data <- data.frame(
  umap1 = umap_result$layout[, 1],
  umap2 = umap_result$layout[, 2],
  Species = iris$Species
)

ggplot(umap_data, aes(x = umap1, y = umap2, color = Species)) +
  geom_point(size = 2) +
  theme_minimal()
```

# 8. Feature Selection

## Filter Methods

```{r}
library(caret)

# Correlation-based
cor_matrix <- cor(iris[, 1:4])
high_cor <- findCorrelation(cor_matrix, cutoff = 0.75)

# Variance-based (remove low variance)
nzv <- nearZeroVar(iris[, 1:4])

# Chi-square test for categorical
# chisq.test based selection
```

## Wrapper Methods

```{r}
library(caret)

# Recursive Feature Elimination (RFE)
control <- rfeControl(functions = rfFuncs, method = "cv", number = 10)
rfe_result <- rfe(iris[, 1:4], iris[, 5], sizes = c(1:4), rfeControl = control)

print(rfe_result)
predictors(rfe_result)
plot(rfe_result, type = c("g", "o"))
```

## Embedded Methods

```{r}
# LASSO for feature selection
library(glmnet)

# Prepare data
x <- as.matrix(mtcars[, -1])
y <- mtcars$mpg

# Fit LASSO
lasso_model <- cv.glmnet(x, y, alpha = 1)
plot(lasso_model)

# Coefficients at optimal lambda
coef(lasso_model, s = "lambda.min")

# Non-zero coefficients
coef(lasso_model, s = "lambda.min")[coef(lasso_model, s = "lambda.min")[, 1] != 0, ]
```

# 9. Hyperparameter Tuning

## Grid Search

```{r}
library(caret)

# Define grid
tuneGrid <- expand.grid(
  mtry = c(2, 3, 4),
  splitrule = c("gini", "extratrees"),
  min.node.size = c(1, 3, 5)
)

# Train with grid search
train_control <- trainControl(method = "cv", number = 5)
model_tuned <- train(Species ~ ., data = iris,
                     method = "ranger",
                     tuneGrid = tuneGrid,
                     trControl = train_control)

print(model_tuned)
plot(model_tuned)

# Best parameters
model_tuned$bestTune
```

## Random Search

```{r}
# Random search
train_control_random <- trainControl(method = "cv", number = 5,
                                     search = "random")
model_random <- train(Species ~ ., data = iris,
                      method = "ranger",
                      tuneLength = 15,
                      trControl = train_control_random)

print(model_random)
```

# 10. Ensemble Methods

## Bagging

```{r}
library(ipred)

# Bagging
bag_model <- bagging(Species ~ ., data = train_data, nbagg = 25)

# Predictions
bag_pred <- predict(bag_model, test_data)

# Accuracy
mean(bag_pred == test_data$Species)
```

## Boosting

```{r}
library(adabag)

# AdaBoost
ada_model <- boosting(Species ~ ., data = train_data, boos = TRUE, mfinal = 50)

# Predictions
ada_pred <- predict(ada_model, test_data)

# Accuracy
ada_pred$confusion
ada_pred$error
```

## Stacking

```{r}
library(caretEnsemble)

# Define control
train_control <- trainControl(
  method = "cv",
  number = 5,
  savePredictions = "final",
  classProbs = TRUE
)

# Train multiple models
model_list <- caretList(
  Species ~ ., data = train_data,
  trControl = train_control,
  methodList = c("rf", "rpart", "knn")
)

# Stack models
stack_model <- caretStack(model_list, method = "glm")

# Predictions
stack_pred <- predict(stack_model, test_data)

# Accuracy
mean(stack_pred == test_data$Species)
```

# Summary

This document covered machine learning in R:

**Data Preparation**:
- Train-test split, scaling, missing values

**Classification**:
- Logistic regression, KNN, decision trees
- Random forest, SVM, Naive Bayes, XGBoost

**Regression**:
- Linear regression, random forest, XGBoost

**Evaluation**:
- Cross-validation, ROC/AUC, confusion matrix

**Clustering**:
- K-means, hierarchical, DBSCAN

**Dimensionality Reduction**:
- PCA, t-SNE, UMAP

**Feature Selection**:
- Filter, wrapper, embedded methods

**Hyperparameter Tuning**:
- Grid search, random search

**Ensemble Methods**:
- Bagging, boosting, stacking

**Key Frameworks**:
- caret: Unified interface
- tidymodels: Modern framework
- Base packages: Individual algorithms

**Next Steps**: Proceed to Reproducible Research for R Markdown, version control, and project organization.

# References

- [An Introduction to Statistical Learning](https://www.statlearning.com/)
- [Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/)
- [caret documentation](https://topepo.github.io/caret/)
- [tidymodels](https://www.tidymodels.org/)
