---
title: "Advanced Programming in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Functional Programming

## apply Family

### apply()

```{r}
# Apply function to matrix rows or columns
mat <- matrix(1:12, nrow = 3)

# Apply to rows (MARGIN = 1)
apply(mat, 1, sum)

# Apply to columns (MARGIN = 2)
apply(mat, 2, mean)

# Custom function
apply(mat, 1, function(x) max(x) - min(x))
```

### lapply() and sapply()

```{r}
# lapply returns a list
my_list <- list(a = 1:5, b = 6:10, c = 11:15)
lapply(my_list, mean)

# sapply simplifies to vector/matrix
sapply(my_list, mean)

# With custom function
sapply(my_list, function(x) sum(x^2))

# On data frames
lapply(mtcars[, 1:3], mean)
sapply(mtcars[, 1:3], sd)
```

### tapply()

```{r}
# Apply function to groups
tapply(mtcars$mpg, mtcars$cyl, mean)
tapply(mtcars$hp, list(mtcars$cyl, mtcars$gear), mean)
```

### mapply()

```{r}
# Multivariate apply
mapply(function(x, y) x + y, 1:5, 6:10)

# Multiple arguments
mapply(rep, 1:4, 4:1)

# With data frames
mapply(function(x, y) cor(x, y),
       mtcars[c("mpg", "wt")],
       mtcars[c("hp", "disp")])
```

### vapply()

```{r}
# Like sapply but with pre-specified output type (safer)
vapply(mtcars[1:3], mean, numeric(1))

# Specifies exact output format
vapply(mtcars[1:3], function(x) c(min(x), max(x)), numeric(2))
```

## purrr Package (Functional Programming)

```{r}
library(purrr)

# map() - always returns list
map(mtcars[1:3], mean)

# map_dbl() - returns numeric vector
map_dbl(mtcars[1:3], mean)

# map_chr() - returns character vector
map_chr(iris[1:4], class)

# map_df() - returns data frame
map_df(mtcars[1:3], ~ data.frame(mean = mean(.), sd = sd(.)))

# map2() - iterate over two inputs
map2_dbl(mtcars$mpg, mtcars$wt, ~ .x / .y)

# pmap() - iterate over multiple inputs
pmap_dbl(list(mtcars$mpg, mtcars$wt, mtcars$hp),
         function(x, y, z) x + y + z)

# Anonymous function shortcuts
map_dbl(mtcars[1:3], ~ mean(.) * 2)

# Keep, discard, compact
keep(mtcars, is.numeric)
discard(mtcars, is.numeric)
compact(list(1, NULL, 3, NULL, 5))

# Reduce
reduce(1:10, `+`)
reduce(list(a = 1:5, b = 3:7, c = 5:9), intersect)

# Safely handle errors
safe_log <- safely(log)
safe_log(10)
safe_log("a")

# Possibly for default values
possible_log <- possibly(log, otherwise = NA)
possible_log("a")

# Quietly capture messages/warnings
quiet_log <- quietly(log)
quiet_log(-1)
```

## Anonymous Functions

```{r}
# Traditional
sapply(1:5, function(x) x^2)

# R 4.1+ shorthand
sapply(1:5, \(x) x^2)

# purrr style
map_dbl(1:5, ~ .^2)
```

## Higher-Order Functions

```{r}
# Functions that return functions
make_power <- function(n) {
  function(x) x^n
}

square <- make_power(2)
cube <- make_power(3)

square(5)
cube(5)

# Function composition
library(purrr)
compose(sqrt, mean, sum)(1:10)
```

## Closures

```{r}
# Function that captures environment
make_counter <- function() {
  count <- 0
  function() {
    count <<- count + 1
    count
  }
}

counter1 <- make_counter()
counter1()  # 1
counter1()  # 2

counter2 <- make_counter()
counter2()  # 1
```

# 2. Object-Oriented Programming

## S3 Objects

```{r}
# Create S3 object
person <- list(
  name = "Alice",
  age = 30
)
class(person) <- "person"

# Create print method
print.person <- function(x, ...) {
  cat("Person:", x$name, "\n")
  cat("Age:", x$age, "\n")
}

print(person)

# Generic function
summary.person <- function(object, ...) {
  cat("Summary of person:", object$name, "\n")
}

summary(person)

# Constructor function
new_person <- function(name, age) {
  structure(
    list(name = name, age = age),
    class = "person"
  )
}

alice <- new_person("Alice", 30)
```

## S4 Objects

```{r}
# Define S4 class
setClass("Person",
         slots = c(
           name = "character",
           age = "numeric"
         ))

# Create instance
alice <- new("Person", name = "Alice", age = 30)

# Access slots
alice@name
alice@age

# Generic function
setGeneric("getName", function(object) standardGeneric("getName"))

# Method
setMethod("getName", "Person", function(object) object@name)

getName(alice)

# Validator
setValidity("Person", function(object) {
  if (object@age < 0) {
    return("Age must be positive")
  }
  TRUE
})

# Constructor with validation
Person <- function(name, age) {
  new("Person", name = name, age = age)
}
```

## R6 Objects

```{r}
library(R6)

# Define R6 class
Person <- R6Class("Person",
  public = list(
    name = NULL,
    age = NULL,

    initialize = function(name, age) {
      self$name <- name
      self$age <- age
    },

    print = function() {
      cat("Person:", self$name, "\n")
      cat("Age:", self$age, "\n")
    },

    birthday = function() {
      self$age <- self$age + 1
      invisible(self)
    }
  ),

  private = list(
    secret = "private data"
  )
)

# Create instance
alice <- Person$new("Alice", 30)
alice$print()
alice$birthday()
alice$age

# Inheritance
Employee <- R6Class("Employee",
  inherit = Person,
  public = list(
    company = NULL,

    initialize = function(name, age, company) {
      super$initialize(name, age)
      self$company <- company
    }
  )
)

bob <- Employee$new("Bob", 25, "Acme Corp")
```

# 3. Environments and Scoping

## Environment Basics

```{r}
# Current environment
environment()

# Global environment
.GlobalEnv

# Create new environment
env <- new.env()

# Assign to environment
env$x <- 10
env$y <- 20

# List contents
ls(env)

# Get value
get("x", envir = env)

# Parent environment
parent.env(env)

# Environment chain
search()
```

## Scoping Rules

```{r}
# Lexical scoping
x <- 1

f <- function() {
  y <- 2
  g <- function() {
    z <- 3
    c(x, y, z)
  }
  g()
}

f()  # Can access x (global), y (parent), z (local)

# Dynamic scoping with environment
x <- 1
f <- function() {
  y <- 2
  environment()
}
env_f <- f()
env_f$y
```

## Non-Standard Evaluation (NSE)

```{r}
# Capture expressions without evaluating
expr <- quote(x + y)
expr

# Evaluate in environment
x <- 5
y <- 10
eval(expr)

# Substitute
substitute(x + y)

# With environment
eval(quote(mean(x)), list(x = 1:10))

# deparse and parse
code <- deparse(expr)
code
parsed <- parse(text = code)
eval(parsed)
```

# 4. Tidy Evaluation

```{r}
library(rlang)
library(dplyr)

# Enquo and !!
my_summarise <- function(data, group_var, sum_var) {
  group_var <- enquo(group_var)
  sum_var <- enquo(sum_var)

  data %>%
    group_by(!!group_var) %>%
    summarise(total = sum(!!sum_var))
}

my_summarise(mtcars, cyl, mpg)

# Multiple variables with ...
my_select <- function(data, ...) {
  data %>% select(...)
}

my_select(mtcars, mpg, cyl, hp)

# := for dynamic names
my_mutate <- function(data, var, value) {
  var <- ensym(var)
  data %>% mutate(!!var := value)
}

my_mutate(mtcars, new_col, 10)

# Curly-curly {{}}
my_plot <- function(data, x, y) {
  ggplot(data, aes({{x}}, {{y}})) +
    geom_point()
}

library(ggplot2)
my_plot(mtcars, wt, mpg)
```

# 5. Error Handling

## tryCatch

```{r}
# Basic error handling
result <- tryCatch(
  {
    log("not a number")
  },
  error = function(e) {
    message("Error occurred: ", e$message)
    NA
  },
  warning = function(w) {
    message("Warning: ", w$message)
  }
)

# With finally
tryCatch(
  {
    x <- 10
    y <- x / 2
  },
  error = function(e) NULL,
  finally = {
    message("Cleanup code")
  }
)

# Multiple error types
safe_divide <- function(x, y) {
  tryCatch(
    x / y,
    error = function(e) {
      if (grepl("non-numeric", e$message)) {
        message("Input must be numeric")
      } else {
        message("Unknown error")
      }
      NA
    }
  )
}
```

## withCallingHandlers

```{r}
# Handle without interrupting
withCallingHandlers(
  {
    warning("First warning")
    warning("Second warning")
    message("Done")
  },
  warning = function(w) {
    message("Caught warning: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

## purrr Error Handling

```{r}
library(purrr)

# safely - returns list with result and error
safe_log <- safely(log)
safe_log(10)
safe_log("a")

# Use with map
map(list(10, "a", 100), safe_log)

# possibly - returns default on error
possible_log <- possibly(log, otherwise = NA)
map_dbl(list(10, "a", 100), possible_log)

# quietly - captures output
quiet_log <- quietly(log)
quiet_log(-1)
```

# 6. Debugging

## Debugging Tools

```{r eval=FALSE}
# Print debugging
print(x)
cat("Value of x:", x, "\n")

# browser() - interactive debugging
my_function <- function(x) {
  y <- x * 2
  browser()  # Execution stops here
  z <- y + 1
  z
}

# debug() - step through function
debug(my_function)
my_function(10)
undebug(my_function)

# debugonce() - debug only once
debugonce(my_function)

# trace() - insert code at specific points
trace(mean, quote(print(x)))
untrace(mean)

# recover() - examine call stack on error
options(error = recover)
# ... run code that errors ...
options(error = NULL)

# traceback() - show call stack
# After error occurs:
traceback()
```

## Profiling

```{r eval=FALSE}
# Rprof - basic profiling
Rprof("profile.out")
# ... code to profile ...
Rprof(NULL)
summaryRprof("profile.out")

# profvis - visual profiling
library(profvis)
profvis({
  x <- rnorm(10000)
  y <- x^2
  z <- log(y + 1)
})

# microbenchmark - precise timing
library(microbenchmark)
microbenchmark(
  method1 = sum(1:1000),
  method2 = Reduce(`+`, 1:1000),
  times = 100
)

# bench - comprehensive benchmarking
library(bench)
bench::mark(
  sum(1:1000),
  Reduce(`+`, 1:1000)
)
```

# 7. Metaprogramming

## Code as Data

```{r}
# Create expression
expr <- quote(x + y)
expr

# Modify expression
expr[[1]]  # Function
expr[[2]]  # First argument
expr[[3]]  # Second argument

# Build expression
call("+", quote(x), quote(y))

# as.call and as.list
expr_list <- as.list(expr)
as.call(list(quote(mean), quote(x)))
```

## Computing on the Language

```{r}
# Walk expression tree
library(lobstr)
ast(f(x, y = g(z)))

# Modify function calls
substitute_call <- function(call, env = parent.frame()) {
  if (is.call(call)) {
    call[-1] <- lapply(call[-1], substitute_call, env)
  }
  call
}

# Unevaluated expressions
expr(x + y)
exprs(x = a + b, y = c + d)

# Quasiquotation
x <- 5
expr(y + !!x)  # Unquote x

# Unquote-splice
values <- c(1, 2, 3)
expr(sum(!!!values))
```

## Code Generation

```{r}
# Generate function
make_function <- function(args, body, env = parent.frame()) {
  args <- as.pairlist(args)
  eval(call("function", args, body), env)
}

f <- make_function(alist(x = , y = ), quote(x + y))
f(3, 4)

# rlang::new_function
library(rlang)
new_function(
  exprs(x = , y = ),
  expr({x + y})
)
```

# 8. Advanced Function Features

## Function Arguments

```{r}
# Default arguments
f <- function(x = 1, y = x * 2) {
  x + y
}

# Missing arguments
f <- function(x, y) {
  if (missing(y)) {
    y <- x * 2
  }
  x + y
}

# match.arg for restricted choices
greet <- function(name, style = c("formal", "casual")) {
  style <- match.arg(style)
  switch(style,
         formal = paste("Dear", name),
         casual = paste("Hi", name))
}

greet("Alice", "formal")
```

## do.call

```{r}
# Call function with list of arguments
args_list <- list(x = 1:10, na.rm = TRUE)
do.call(mean, args_list)

# Combine multiple data frames
df_list <- list(
  data.frame(x = 1:3),
  data.frame(x = 4:6),
  data.frame(x = 7:9)
)
do.call(rbind, df_list)
```

## Lazy Evaluation

```{r}
# Arguments evaluated only when needed
f <- function(x, y) {
  if (x > 0) {
    x
  } else {
    y
  }
}

f(5, stop("This won't be evaluated"))

# Force evaluation
f <- function(x) {
  force(x)
  function() x
}
```

# 9. Parallel Programming

## parallel Package

```{r}
library(parallel)

# Detect cores
detectCores()

# parLapply (parallel lapply)
cl <- makeCluster(4)
result <- parLapply(cl, 1:10, function(x) x^2)
stopCluster(cl)

# mclapply (multicore, Unix/Mac only)
mclapply(1:10, function(x) x^2, mc.cores = 4)

# parSapply
cl <- makeCluster(4)
result <- parSapply(cl, 1:10, function(x) x^2)
stopCluster(cl)

# Export variables to cluster
cl <- makeCluster(4)
clusterExport(cl, "my_variable")
clusterEvalQ(cl, library(dplyr))
result <- parLapply(cl, 1:10, some_function)
stopCluster(cl)
```

## foreach and doParallel

```{r}
library(foreach)
library(doParallel)

# Setup parallel backend
registerDoParallel(cores = 4)

# Parallel loop
result <- foreach(i = 1:10, .combine = c) %dopar% {
  i^2
}

# With packages
result <- foreach(i = 1:10, .packages = "dplyr") %dopar% {
  # Use dplyr functions
}

# Stop parallel backend
stopImplicitCluster()
```

## future Package

```{r}
library(future)
library(furrr)

# Set plan
plan(multisession, workers = 4)

# Future map
future_map_dbl(1:10, ~ .^2)

# Explicit futures
f <- future({
  Sys.sleep(2)
  42
})
value(f)
```

# Summary

This document covered advanced R programming:

**Functional Programming**:
- apply family (apply, lapply, sapply, etc.)
- purrr for functional programming
- Higher-order functions and closures

**Object-Oriented Programming**:
- S3 (simple, flexible)
- S4 (formal, type-safe)
- R6 (reference semantics)

**Environments and Scoping**:
- Environment manipulation
- Lexical vs dynamic scoping
- Non-standard evaluation

**Tidy Evaluation**:
- enquo, !!, {{}}
- Programming with dplyr/ggplot2

**Error Handling**:
- tryCatch, withCallingHandlers
- purrr::safely, possibly, quietly

**Debugging and Profiling**:
- browser, debug, trace
- profvis, microbenchmark

**Metaprogramming**:
- Computing on the language
- Code generation
- Quasiquotation

**Advanced Features**:
- Function arguments and evaluation
- do.call
- Lazy evaluation

**Parallel Programming**:
- parallel package
- foreach/doParallel
- future/furrr

**Key Principles**:
- Understand when to use each paradigm
- Profile before optimizing
- Handle errors gracefully
- Use functional programming for clarity

**Next Steps**: Proceed to Performance Optimization for vectorization, Rcpp, and memory management.

# References

- [Advanced R by Hadley Wickham](https://adv-r.hadley.nz/)
- [Tidy Evaluation](https://tidyeval.tidyverse.org/)
- [R6 Documentation](https://r6.r-lib.org/)
- [purrr Documentation](https://purrr.tidyverse.org/)
