---
title: "Performance Optimization in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Profiling R Code

## Why Profile?

- Identify bottlenecks
- Measure actual performance (not guesses)
- Guide optimization efforts

## system.time()

```{r}
# Basic timing
system.time({
  x <- rnorm(1e6)
  y <- x^2
})

# User time: CPU time
# System time: OS operations
# Elapsed time: Wall clock time
```

## microbenchmark

```{r}
library(microbenchmark)

# Compare methods
microbenchmark(
  loop = {
    result <- numeric(1000)
    for (i in 1:1000) result[i] <- i^2
  },
  vectorized = {
    result <- (1:1000)^2
  },
  times = 100
)

# With different units
microbenchmark(
  sqrt(1:1000),
  (1:1000)^0.5,
  unit = "us"
)
```

## bench

```{r}
library(bench)

# Comprehensive benchmarking
results <- mark(
  loop = {
    result <- numeric(1000)
    for (i in 1:1000) result[i] <- i^2
  },
  vectorized = (1:1000)^2,
  check = FALSE  # Don't check if results are identical
)

results
plot(results)
```

## profvis

```{r eval=FALSE}
library(profvis)

# Visual profiling
profvis({
  # Data generation
  n <- 1e5
  x <- rnorm(n)
  y <- rnorm(n)

  # Slow operations
  z <- numeric(n)
  for (i in 1:n) {
    z[i] <- x[i] + y[i]
  }

  # Analysis
  lm(z ~ x + y)
})
```

## Rprof

```{r eval=FALSE}
# Low-level profiling
Rprof("profile.out")

# Code to profile
for (i in 1:1000) {
  x <- rnorm(100)
  mean(x)
}

Rprof(NULL)

# Analyze
summaryRprof("profile.out")
```

# 2. Vectorization

## Avoid Loops

```{r}
# Bad: Loop
loop_version <- function(n) {
  result <- numeric(n)
  for (i in 1:n) {
    result[i] <- sqrt(i)
  }
  result
}

# Good: Vectorized
vectorized_version <- function(n) {
  sqrt(1:n)
}

microbenchmark(
  loop_version(1000),
  vectorized_version(1000),
  times = 100
)
```

## Vectorized Functions

```{r}
# Arithmetic operations
x <- 1:1000
y <- 1:1000

# All vectorized
x + y
x * y
x / y
x^y

# Comparison operators
x > 500
x == y

# Logical operations
(x > 500) & (y < 600)

# Math functions
sqrt(x)
log(x)
exp(x)
```

## Row/Column Operations

```{r}
mat <- matrix(rnorm(10000), nrow = 100)

# rowSums, colSums (much faster than apply)
microbenchmark(
  rowSums = rowSums(mat),
  apply = apply(mat, 1, sum),
  times = 100
)

# Other optimized functions
rowMeans(mat)
colMeans(mat)
rowSums(mat)
colSums(mat)
```

## ifelse vs if/else

```{r}
x <- rnorm(1000)

# Vectorized ifelse
result1 <- ifelse(x > 0, "positive", "negative")

# For single values, if/else is faster
value <- 5
result2 <- if (value > 0) "positive" else "negative"

# case_when for complex conditions
library(dplyr)
result3 <- case_when(
  x > 1 ~ "high",
  x > 0 ~ "medium",
  TRUE ~ "low"
)
```

# 3. Memory Management

## Object Size

```{r}
# Check object size
x <- rnorm(1e6)
object.size(x)
format(object.size(x), units = "MB")

# All objects
sapply(ls(), function(obj) {
  format(object.size(get(obj)), units = "MB")
})
```

## Memory Profiling

```{r eval=FALSE}
library(profmem)

# Profile memory
profmem({
  x <- numeric(1e6)
  y <- x * 2
  z <- data.frame(x, y)
})
```

## Avoid Copies

```{r}
# Bad: Creates multiple copies
slow_function <- function(df) {
  df$new_col1 <- df$x + 1
  df$new_col2 <- df$y + 1
  df$new_col3 <- df$z + 1
  df
}

# Better: All at once
fast_function <- function(df) {
  df$new_col1 <- df$x + 1
  df$new_col2 <- df$y + 1
  df$new_col3 <- df$z + 1
  df
}

# Best: data.table (modify in place)
library(data.table)
dt_function <- function(dt) {
  dt[, `:=`(
    new_col1 = x + 1,
    new_col2 = y + 1,
    new_col3 = z + 1
  )]
}
```

## Pre-allocate Vectors

```{r}
# Bad: Growing vector
slow <- function(n) {
  result <- NULL
  for (i in 1:n) {
    result <- c(result, i^2)
  }
  result
}

# Good: Pre-allocated
fast <- function(n) {
  result <- numeric(n)
  for (i in 1:n) {
    result[i] <- i^2
  }
  result
}

microbenchmark(
  slow(1000),
  fast(1000),
  times = 10
)
```

## Garbage Collection

```{r eval=FALSE}
# Manual garbage collection
gc()

# Check memory usage
gc(verbose = TRUE)

# Disable GC during critical operations
gcold <- gcinfo(TRUE)
# ... operations ...
gcinfo(gcold)
```

# 4. Efficient Data Structures

## data.table

```{r}
library(data.table)
library(dplyr)

# Create large dataset
n <- 1e6
df <- data.frame(
  id = 1:n,
  group = sample(letters[1:5], n, replace = TRUE),
  value = rnorm(n)
)
dt <- as.data.table(df)

# Compare performance
microbenchmark(
  dplyr = df %>% group_by(group) %>% summarise(mean = mean(value)),
  datatable = dt[, .(mean = mean(value)), by = group],
  times = 10
)
```

## Matrix vs Data Frame

```{r}
# For numeric data, matrices are faster
n <- 1000
df <- data.frame(matrix(rnorm(n * 100), ncol = 100))
mat <- as.matrix(df)

microbenchmark(
  df_colMeans = colMeans(df),
  mat_colMeans = colMeans(mat),
  times = 100
)
```

## Lists vs Vectors

```{r}
# Vectors for homogeneous data
v <- 1:1000

# Lists for heterogeneous data
l <- as.list(1:1000)

# Vectors are faster
microbenchmark(
  vector = sum(v),
  list = sum(unlist(l)),
  times = 100
)
```

# 5. Rcpp - C++ Integration

## Basic Rcpp

```{r}
library(Rcpp)

# Simple C++ function
cppFunction('
  int add_cpp(int x, int y) {
    return x + y;
  }
')

add_cpp(5, 10)

# Rcpp vs R loop
cppFunction('
  NumericVector cumsum_cpp(NumericVector x) {
    int n = x.size();
    NumericVector result(n);
    result[0] = x[0];
    for (int i = 1; i < n; i++) {
      result[i] = result[i-1] + x[i];
    }
    return result;
  }
')

x <- rnorm(10000)
microbenchmark(
  R = cumsum(x),
  Rcpp = cumsum_cpp(x),
  times = 100
)
```

## Rcpp with RcppArmadillo

```{r eval=FALSE}
library(RcppArmadillo)

# Matrix operations
src <- '
  // [[Rcpp::depends(RcppArmadillo)]]
  #include <RcppArmadillo.h>

  // [[Rcpp::export]]
  arma::mat multiply_arma(arma::mat A, arma::mat B) {
    return A * B;
  }
'

sourceCpp(code = src)

A <- matrix(rnorm(100), 10, 10)
B <- matrix(rnorm(100), 10, 10)

microbenchmark(
  R = A %*% B,
  Arma = multiply_arma(A, B),
  times = 100
)
```

## When to Use Rcpp

Use Rcpp when:
- Loops that can't be vectorized
- Recursive functions
- Need to call C/C++ libraries
- Working with complex data structures

Don't use Rcpp for:
- Already vectorized operations
- Simple arithmetic
- When R is fast enough

# 6. Parallel Computing

## parallel Package

```{r}
library(parallel)

# Detect cores
num_cores <- detectCores() - 1  # Leave one free

# Parallel lapply
x <- 1:1000
result <- mclapply(x, function(i) {
  Sys.sleep(0.01)
  i^2
}, mc.cores = num_cores)

# Parallel for Windows
cl <- makeCluster(num_cores)
result <- parLapply(cl, x, function(i) i^2)
stopCluster(cl)
```

## foreach

```{r}
library(foreach)
library(doParallel)

# Register parallel backend
registerDoParallel(cores = num_cores)

# Parallel foreach
result <- foreach(i = 1:1000, .combine = c) %dopar% {
  i^2
}

# Stop cluster
stopImplicitCluster()
```

## future

```{r}
library(future)
library(furrr)

# Set plan
plan(multisession, workers = num_cores)

# Parallel map
result <- future_map(1:1000, ~ .^2)

# Reset to sequential
plan(sequential)
```

# 7. Database Operations

## Use Databases for Large Data

```{r eval=FALSE}
library(DBI)
library(RSQLite)

# Connect to database
con <- dbConnect(SQLite(), "large_data.db")

# Write data in chunks
for (chunk in data_chunks) {
  dbWriteTable(con, "data", chunk, append = TRUE)
}

# Query only what you need
result <- dbGetQuery(con, "
  SELECT *
  FROM data
  WHERE value > 100
  LIMIT 1000
")

# Use indexes
dbExecute(con, "CREATE INDEX idx_value ON data(value)")

dbDisconnect(con)
```

## dbplyr

```{r eval=FALSE}
library(dplyr)
library(dbplyr)

# Connect
con <- DBI::dbConnect(RSQLite::SQLite(), "data.db")
data_tbl <- tbl(con, "data")

# Use dplyr syntax (translated to SQL)
result <- data_tbl %>%
  filter(value > 100) %>%
  group_by(group) %>%
  summarise(mean_value = mean(value)) %>%
  collect()  # Execute and bring to R
```

# 8. Caching and Memoization

## memoise Package

```{r}
library(memoise)

# Slow function
slow_function <- function(n) {
  Sys.sleep(2)
  n^2
}

# Memoized version
fast_function <- memoise(slow_function)

# First call is slow
system.time(fast_function(10))

# Second call is instant
system.time(fast_function(10))

# Clear cache
forget(fast_function)
```

## knitr Caching

````{r eval=FALSE}
```{r cache=TRUE}
# This chunk will be cached
expensive_computation <- function() {
  Sys.sleep(10)
  rnorm(1e6)
}
result <- expensive_computation()
```
````

# 9. Code Optimization Strategies

## General Principles

```{r eval=FALSE}
# 1. Use built-in functions
mean(x)  # vs manual calculation

# 2. Vectorize
x + y  # vs loop

# 3. Pre-allocate
result <- numeric(n)  # vs growing vector

# 4. Use appropriate data structures
matrix for numeric data  # vs data.frame

# 5. Avoid unnecessary copies
modify in place with data.table

# 6. Use specialized packages
data.table, Rcpp

# 7. Parallelize when appropriate
mclapply, foreach

# 8. Cache expensive computations
memoise, knitr cache
```

## Specific Optimizations

```{r}
# Use which.min/which.max
x <- rnorm(1e6)
microbenchmark(
  which = which.min(x),
  manual = which(x == min(x))[1],
  times = 100
)

# Use anyNA instead of any(is.na())
x <- rnorm(1e6)
microbenchmark(
  anyNA = anyNA(x),
  any_is.na = any(is.na(x)),
  times = 100
)

# Use && and || for scalars (short-circuit)
# Use & and | for vectors

# %in% for membership testing
x <- sample(1:1000, 100)
lookup <- 1:500
microbenchmark(
  match = x %in% lookup,
  sapply = sapply(x, function(i) i %in% lookup),
  times = 100
)
```

# 10. Best Practices

## Optimization Workflow

```{r eval=FALSE}
# 1. Make it work
# Write clear, correct code first

# 2. Make it right
# Ensure correctness with tests

# 3. Profile
# Find actual bottlenecks
profvis({your_code()})

# 4. Optimize bottlenecks
# Focus on slow parts only

# 5. Verify
# Ensure still correct
testthat::test_that("optimized version works", {
  expect_equal(slow_version(), fast_version())
})

# 6. Measure
# Confirm improvement
microbenchmark(
  slow_version(),
  fast_version()
)
```

## When to Optimize

Optimize when:
- Code is too slow for practical use
- Profiling shows clear bottleneck
- You have correct tests in place

Don't optimize:
- Prematurely (before profiling)
- Code that runs rarely
- At the cost of readability (unless necessary)

## Memory vs Speed Trade-offs

```{r eval=FALSE}
# More memory, faster
cache_results <- TRUE

# Less memory, slower
process_in_chunks <- TRUE

# Balance based on constraints
```

# Summary

This document covered performance optimization:

**Profiling**:
- system.time, microbenchmark, bench
- profvis for visual profiling
- Identify bottlenecks before optimizing

**Vectorization**:
- Avoid loops when possible
- Use built-in vectorized functions
- rowSums/colSums faster than apply

**Memory**:
- Pre-allocate vectors
- Avoid unnecessary copies
- Use appropriate data structures
- Monitor memory usage

**Rcpp**:
- C++ integration for speed
- Use for unavoidable loops
- Matrix operations with RcppArmadillo

**Parallel Computing**:
- parallel, foreach, future packages
- Utilize multiple cores
- For embarrassingly parallel tasks

**Data Structures**:
- data.table for large datasets
- Matrices for numeric data
- Databases for huge datasets

**Caching**:
- memoise for expensive functions
- knitr caching for reports

**Best Practices**:
- Profile first
- Optimize bottlenecks only
- Maintain correctness
- Document trade-offs

**Golden Rules**:
1. Correct first, fast second
2. Profile before optimizing
3. Vectorize when possible
4. Use appropriate data structures
5. Consider memory constraints

**Next Steps**: Proceed to Specialized Bioinformatics for pathway analysis, network analysis, and more.

# References

- [Efficient R Programming](https://csgillespie.github.io/efficientR/)
- [Advanced R - Performance](https://adv-r.hadley.nz/perf-improve.html)
- [Rcpp for Seamless R and C++ Integration](http://www.rcpp.org/)
- [data.table documentation](https://rdatatable.gitlab.io/data.table/)
