---
title: "Specialized Bioinformatics Topics"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Multiple Sequence Alignment

## Biostrings for Pairwise Alignment

```{r}
library(Biostrings)

# Create sequences
seq1 <- DNAString("ACGTACGTACGT")
seq2 <- DNAString("ACGTAAGTACGT")

# Pairwise alignment
alignment <- pairwiseAlignment(seq1, seq2)
alignment

# Alignment score
score(alignment)

# View alignment
writePairwiseAlignments(alignment)

# Protein alignment
prot1 <- AAString("PAWHEAE")
prot2 <- AAString("HEAGAWGHE")
pairwiseAlignment(prot1, prot2)
```

## msa for Multiple Alignment

```{r}
library(msa)

# Load sequences
sequences <- readAAStringSet(system.file("examples", "exampleAA.fasta",
                                         package = "msa"))

# Multiple alignment
alignment <- msa(sequences, method = "ClustalW")
# Or: msa(sequences, method = "ClustalOmega")
# Or: msa(sequences, method = "Muscle")

# View alignment
print(alignment, show = "complete")

# Convert to seqinr format
alignment_seqinr <- msaConvert(alignment, type = "seqinr::alignment")

# Export
msaConvert(alignment, type = "bios2mds::align")
```

# 2. Phylogenetic Analysis

## ape Package

```{r}
library(ape)

# Read tree
# tree <- read.tree("tree.nwk")

# Create random tree
tree <- rtree(10)

# Plot tree
plot(tree)

# Root tree
rooted_tree <- root(tree, outgroup = "t1")

# Prune tree
pruned_tree <- drop.tip(tree, c("t1", "t2"))

# Ultrametric tree
ultra_tree <- chronos(tree)

# Node labels
tree$node.label <- paste0("Node", 1:tree$Nnode)
plot(tree, show.node.label = TRUE)

# Branch lengths
tree$edge.length
```

## phangorn for Phylogeny Construction

```{r}
library(phangorn)

# Read alignment
# alignment <- read.phyDat("alignment.fasta", format = "fasta")

# Distance matrix
# dist_matrix <- dist.ml(alignment)

# Neighbor-joining tree
# nj_tree <- NJ(dist_matrix)

# Maximum likelihood
# fit <- pml(nj_tree, alignment)
# fit_optim <- optim.pml(fit, optNni = TRUE, optBf = TRUE,
#                        optQ = TRUE, optGamma = TRUE)

# Bootstrap
# bs <- bootstrap.pml(fit_optim, bs = 100, optNni = TRUE)

# Plot tree with bootstrap values
# plotBS(fit_optim$tree, bs)
```

## ggtree for Tree Visualization

```{r}
library(ggtree)
library(ggplot2)

# Basic tree plot
ggtree(tree) +
  geom_tiplab() +
  theme_tree2()

# Circular tree
ggtree(tree, layout = "circular") +
  geom_tiplab()

# With node labels
ggtree(tree) +
  geom_tiplab() +
  geom_nodelab(color = "red")

# Phylogram
ggtree(tree, branch.length = "none") +
  geom_tiplab()

# Time tree
ggtree(ultra_tree) +
  theme_tree2() +
  geom_tiplab()

# Add data
tip_data <- data.frame(
  label = tree$tip.label,
  category = sample(c("A", "B"), length(tree$tip.label), replace = TRUE)
)

ggtree(tree) %<+% tip_data +
  geom_tiplab(aes(color = category)) +
  geom_tippoint(aes(shape = category))
```

# 3. Network Analysis

## igraph

```{r}
library(igraph)

# Create graph
g <- graph(c(1,2, 2,3, 3,4, 4,1, 1,5), directed = FALSE)

# Plot
plot(g)

# From data frame
edges <- data.frame(
  from = c("A", "A", "B", "C"),
  to = c("B", "C", "C", "D")
)
g <- graph_from_data_frame(edges, directed = FALSE)

# Add node attributes
V(g)$color <- "lightblue"
V(g)$size <- 20

# Add edge attributes
E(g)$width <- c(1, 2, 3, 4)

# Network metrics
degree(g)
betweenness(g)
closeness(g)
transitivity(g)

# Community detection
communities <- cluster_louvain(g)
plot(communities, g)

# Shortest paths
shortest_paths(g, from = "A", to = "D")

# Centrality
degree_centrality <- degree(g)
betweenness_centrality <- betweenness(g)
```

## Gene Regulatory Networks

```{r}
library(igraph)

# Create gene network
genes <- paste0("Gene", 1:20)
edges <- data.frame(
  regulator = sample(genes, 30, replace = TRUE),
  target = sample(genes, 30, replace = TRUE)
)
grn <- graph_from_data_frame(edges, directed = TRUE)

# Find hub genes
hub_score(grn)
authority_score(grn)

# Subgraph
subgraph <- induced_subgraph(grn, vids = 1:10)

# Export
write_graph(grn, "network.graphml", format = "graphml")
```

## visNetwork for Interactive Networks

```{r}
library(visNetwork)

# Create nodes and edges
nodes <- data.frame(
  id = 1:5,
  label = paste("Node", 1:5),
  group = c("A", "A", "B", "B", "C")
)

edges <- data.frame(
  from = c(1, 2, 3, 4),
  to = c(2, 3, 4, 5)
)

# Interactive network
visNetwork(nodes, edges) %>%
  visGroups(groupname = "A", color = "red") %>%
  visGroups(groupname = "B", color = "blue") %>%
  visGroups(groupname = "C", color = "green") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE)
```

# 4. Metabolomics

## xcms for LC-MS Data

```{r}
library(xcms)

# Read files
# files <- c("sample1.mzML", "sample2.mzML")
# raw_data <- readMSData(files, mode = "onDisk")

# Peak detection
# cwp <- CentWaveParam(peakwidth = c(20, 80), ppm = 25)
# peaks <- findChromPeaks(raw_data, param = cwp)

# Retention time correction
# peaks <- adjustRtime(peaks, param = ObiwarpParam())

# Peak grouping
# peaks <- groupChromPeaks(peaks, param = PeakDensityParam())

# Fill missing peaks
# peaks <- fillChromPeaks(peaks)

# Extract feature table
# feature_table <- featureValues(peaks)
```

## CAMERA for Annotation

```{r}
library(CAMERA)

# Annotate isotopes and adducts
# xsa <- xsAnnotate(peaks)
# xsa <- groupFWHM(xsa)
# xsa <- findIsotopes(xsa)
# xsa <- groupCorr(xsa)
# xsa <- findAdducts(xsa, polarity = "positive")

# Get annotation table
# annotation <- getPeaklist(xsa)
```

# 5. Metagenomics and Microbiome

## phyloseq

```{r}
library(phyloseq)

# Create phyloseq object
# otu_table <- read.csv("otu_table.csv", row.names = 1)
# tax_table <- read.csv("taxonomy.csv", row.names = 1)
# sample_data <- read.csv("metadata.csv", row.names = 1)

# ps <- phyloseq(
#   otu_table(otu_table, taxa_are_rows = TRUE),
#   tax_table(as.matrix(tax_table)),
#   sample_data(sample_data)
# )

# Example with included data
data(GlobalPatterns)
ps <- GlobalPatterns

# Explore
sample_names(ps)
tax_table(ps)
otu_table(ps)

# Filter low abundance
ps_filtered <- filter_taxa(ps, function(x) sum(x) > 100, TRUE)

# Transform counts
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

# Ordination
ps_ord <- ordinate(ps, method = "PCoA", distance = "bray")
plot_ordination(ps, ps_ord, color = "SampleType")

# Barplot
plot_bar(ps, fill = "Phylum")

# Alpha diversity
plot_richness(ps, measures = c("Shannon", "Simpson"))
```

## vegan for Ecology

```{r}
library(vegan)

# Create community matrix
data(varespec)

# Diversity indices
diversity(varespec, index = "shannon")
diversity(varespec, index = "simpson")
specnumber(varespec)  # Species richness

# Ordination
nmds <- metaMDS(varespec, distance = "bray")
plot(nmds)

# Permanova
data(varechem)
adonis2(varespec ~ N + P + K, data = varechem)

# Distance matrix
dist_matrix <- vegdist(varespec, method = "bray")
```

# 6. Population Genetics

## adegenet

```{r}
library(adegenet)

# Example data
data(nancycats)

# Summary
summary(nancycats)

# Principal Component Analysis
pca_result <- glPca(nancycats)
scatter(pca_result)

# Discriminant Analysis
dapc_result <- dapc(nancycats, n.pca = 30, n.da = 2)
scatter(dapc_result)

# Population structure
compoplot(dapc_result)

# Find optimal K
grp <- find.clusters(nancycats, max.n.clust = 10)
```

## SNPRelate for GWAS

```{r}
library(SNPRelate)

# Open GDS file
# genofile <- snpgdsOpen("genotype.gds")

# LD-based SNP pruning
# snpset <- snpgdsLDpruning(genofile, ld.threshold = 0.2)

# PCA
# pca <- snpgdsPCA(genofile, snp.id = snpset$snp.id)
# plot(pca$eigenvect[,1], pca$eigenvect[,2])

# Identity by descent
# ibd <- snpgdsIBDMLE(genofile, snp.id = snpset$snp.id)

# Close file
# snpgdsClose(genofile)
```

# 7. Protein Structure Analysis

## bio3d

```{r}
library(bio3d)

# Read PDB file
# pdb <- read.pdb("1HEL")

# Atom selection
# ca_atoms <- atom.select(pdb, "calpha")

# Visualize
# view(pdb)

# RMSD calculation
# pdbs <- pdbaln(c("1HEL", "1LYZ"))
# rmsd(pdbs)

# Normal mode analysis
# modes <- nma(pdb)
# plot(modes)

# Calculate secondary structure
# ss <- dssp(pdb)

# Distance matrix
# dm <- dm(pdb, inds = ca_atoms)
```

## Rpdb

```{r}
library(Rpdb)

# Read PDB
# pdb <- read.pdb("structure.pdb")

# Get coordinates
# coords <- coords(pdb)

# Atom types
# atoms(pdb)

# Residues
# residues(pdb)
```

# 8. Integrative Multi-Omics

## mixOmics

```{r}
library(mixOmics)

# PLS-DA (Partial Least Squares Discriminant Analysis)
# X <- as.matrix(gene_expression)
# Y <- factor(sample_groups)
# plsda_result <- plsda(X, Y, ncomp = 3)

# Plot
# plotIndiv(plsda_result)
# plotVar(plsda_result)

# DIABLO for multi-omics integration
# data_list <- list(
#   mRNA = mrna_data,
#   protein = protein_data,
#   metabolite = metabolite_data
# )
# design <- matrix(0.1, ncol = length(data_list), nrow = length(data_list))
# diag(design) <- 0
# result <- block.plsda(X = data_list, Y = outcome, ncomp = 5, design = design)

# Circos plot
# circosPlot(result)
```

## MOFA (Multi-Omics Factor Analysis)

```{r}
library(MOFA2)

# Create MOFA object
# data_list <- list(
#   RNA = rna_matrix,
#   Protein = protein_matrix,
#   Metabolite = metabolite_matrix
# )
# MOFAobject <- create_mofa(data_list)

# Prepare MOFA
# model_opts <- get_default_model_options(MOFAobject)
# train_opts <- get_default_training_options(MOFAobject)
# MOFAobject <- prepare_mofa(MOFAobject, model_options = model_opts)

# Train model
# MOFAmodel <- run_mofa(MOFAobject, train_options = train_opts)

# Visualize
# plot_variance_explained(MOFAmodel)
# plot_factor(MOFAmodel, factor = 1)
```

# 9. Spatial Transcriptomics

## Seurat for Spatial Data

```{r}
library(Seurat)

# Load spatial data
# spatial_obj <- Load10X_Spatial("path/to/spatial/data")

# Normalize
# spatial_obj <- SCTransform(spatial_obj, assay = "Spatial")

# Dimensionality reduction
# spatial_obj <- RunPCA(spatial_obj)
# spatial_obj <- RunUMAP(spatial_obj, dims = 1:30)

# Clustering
# spatial_obj <- FindNeighbors(spatial_obj, dims = 1:30)
# spatial_obj <- FindClusters(spatial_obj)

# Spatial visualization
# SpatialDimPlot(spatial_obj)
# SpatialFeaturePlot(spatial_obj, features = "gene_name")

# Identify spatially variable features
# spatial_obj <- FindSpatiallyVariableFeatures(spatial_obj)
```

# 10. Image Analysis

## EBImage

```{r}
library(EBImage)

# Read image
# img <- readImage("cell_image.tif")

# Display
# display(img)

# Threshold
# img_thresh <- thresh(img, w = 10, h = 10, offset = 0.05)

# Morphological operations
# img_erode <- erode(img_thresh, kern = makeBrush(5, shape = "disc"))
# img_dilate <- dilate(img_thresh, kern = makeBrush(5, shape = "disc"))

# Watershed segmentation
# nmask <- watershed(distmap(img_thresh), tolerance = 2)

# Compute features
# features <- computeFeatures.shape(nmask)
# features_intensity <- computeFeatures.basic(nmask, img)

# Count objects
# max(nmask)
```

# 11. Text Mining for Biomedical Literature

## pubmed.mineR

```{r}
library(pubmed.mineR)

# Search PubMed
# results <- pumed_search("CRISPR cancer therapy")

# Download abstracts
# abstracts <- readabs(results)

# Extract information
# genes <- Gengrep(abstracts)
# diseases <- Disgrep(abstracts)

# Co-occurrence
# cooccurrence <- cooc(genes, diseases)
```

## rentrez

```{r}
library(rentrez)

# Search databases
entrez_dbs()

# Search PubMed
search_results <- entrez_search(db = "pubmed", term = "R bioconductor", retmax = 10)

# Get summaries
summaries <- entrez_summary(db = "pubmed", id = search_results$ids)

# Fetch records
records <- entrez_fetch(db = "pubmed", id = search_results$ids[1], rettype = "xml")
```

# 12. Simulation and Modeling

## deSolve for ODE Models

```{r}
library(deSolve)

# Define ODE system (SIR model)
sir_model <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS <- -beta * S * I / N
    dI <- beta * S * I / N - gamma * I
    dR <- gamma * I
    list(c(dS, dI, dR))
  })
}

# Parameters
parameters <- c(beta = 0.5, gamma = 0.1, N = 1000)
state <- c(S = 999, I = 1, R = 0)
times <- seq(0, 100, by = 0.1)

# Solve
out <- ode(y = state, times = times, func = sir_model, parms = parameters)

# Plot
library(ggplot2)
out_df <- as.data.frame(out)
ggplot(out_df) +
  geom_line(aes(x = time, y = S, color = "Susceptible")) +
  geom_line(aes(x = time, y = I, color = "Infected")) +
  geom_line(aes(x = time, y = R, color = "Recovered")) +
  labs(y = "Number of individuals", color = "Status")
```

## simmer for Discrete Event Simulation

```{r}
library(simmer)

# Create simulation environment
env <- simmer("Hospital")

# Define trajectory
patient <- trajectory() %>%
  seize("doctor", 1) %>%
  timeout(function() rnorm(1, 15, 5)) %>%
  release("doctor", 1)

# Run simulation
env %>%
  add_resource("doctor", capacity = 2) %>%
  add_generator("patient", patient, function() rexp(1, 1/10)) %>%
  run(until = 480)

# Get results
get_mon_arrivals(env)
get_mon_resources(env)
```

# Summary

This document covered specialized bioinformatics topics:

**Sequence Analysis**:
- Multiple sequence alignment (msa, Biostrings)
- Phylogenetic analysis (ape, phangorn, ggtree)

**Network Analysis**:
- igraph for network construction
- Gene regulatory networks
- Interactive visualization (visNetwork)

**Omics Technologies**:
- Metabolomics (xcms, CAMERA)
- Metagenomics (phyloseq, vegan)
- Proteomics (bio3d)

**Population Genetics**:
- adegenet for population structure
- SNPRelate for GWAS

**Multi-Omics Integration**:
- mixOmics for data integration
- MOFA for factor analysis

**Spatial Analysis**:
- Spatial transcriptomics (Seurat)

**Image Analysis**:
- EBImage for microscopy

**Literature Mining**:
- PubMed searching and text mining

**Modeling**:
- ODE models (deSolve)
- Discrete event simulation (simmer)

**Key Principles**:
- Use domain-specific packages
- Understand your data type
- Integrate multiple data sources
- Validate results biologically

**Next Steps**: Proceed to Web Applications and Dashboards for Shiny and interactive tools.

# References

- [Bioconductor Workflows](https://bioconductor.org/packages/release/BiocViews.html#___Workflow)
- [phyloseq tutorials](https://joey711.github.io/phyloseq/)
- [igraph documentation](https://igraph.org/r/)
- [mixOmics](http://mixomics.org/)
