---
title: "Web Applications and Dashboards in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Introduction to Shiny

Shiny is an R package for building interactive web applications.

## Basic Shiny App Structure

```{r}
library(shiny)

# UI (User Interface)
ui <- fluidPage(
  titlePanel("My First Shiny App"),

  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
                  "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)
    ),

    mainPanel(
      plotOutput("distPlot")
    )
  )
)

# Server
server <- function(input, output) {
  output$distPlot <- renderPlot({
    x <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    hist(x, breaks = bins, col = "darkgray", border = "white")
  })
}

# Run app
shinyApp(ui = ui, server = server)
```

## File Structure

```
app_directory/
├── app.R          # Single file app
# OR
├── ui.R           # UI definition
├── server.R       # Server logic
└── global.R       # (Optional) Global variables/functions
```

# 2. User Interface (UI)

## Layout Functions

```{r}
# Fluid page
ui <- fluidPage(
  titlePanel("Title"),
  sidebarLayout(
    sidebarPanel(),
    mainPanel()
  )
)

# Navbar page
ui <- navbarPage("App Title",
  tabPanel("Plot",
    # content
  ),
  tabPanel("Table",
    # content
  ),
  tabPanel("About",
    # content
  )
)

# Fluid row
ui <- fluidPage(
  fluidRow(
    column(4, # content),
    column(8, # content)
  )
)

# Tab panels
ui <- fluidPage(
  tabsetPanel(
    tabPanel("Plot", plotOutput("plot")),
    tabPanel("Summary", verbatimTextOutput("summary")),
    tabPanel("Table", tableOutput("table"))
  )
)
```

## Input Widgets

```{r}
# Text input
textInput("name", "Enter your name:")

# Numeric input
numericInput("age", "Enter age:", value = 25, min = 0, max = 120)

# Slider
sliderInput("slider", "Select value:", min = 0, max = 100, value = 50)

# Select dropdown
selectInput("choice", "Choose one:",
            choices = c("Option 1", "Option 2", "Option 3"))

# Multiple select
selectInput("choices", "Choose multiple:",
            choices = c("A", "B", "C"), multiple = TRUE)

# Radio buttons
radioButtons("radio", "Select one:",
             choices = c("Yes", "No"))

# Checkboxes
checkboxGroupInput("checks", "Select all that apply:",
                   choices = c("A", "B", "C"))

# Single checkbox
checkboxInput("agree", "I agree", value = FALSE)

# Date input
dateInput("date", "Select date:")

# Date range
dateRangeInput("dates", "Select date range:")

# File upload
fileInput("file", "Choose file:")

# Action button
actionButton("button", "Click me!")

# Submit button
submitButton("Submit")
```

## Output Functions

```{r}
# In UI
plotOutput("plot")
tableOutput("table")
textOutput("text")
verbatimTextOutput("summary")
uiOutput("ui")  # Dynamic UI
htmlOutput("html")

# In server
output$plot <- renderPlot({ ... })
output$table <- renderTable({ ... })
output$text <- renderText({ ... })
output$summary <- renderPrint({ ... })
output$ui <- renderUI({ ... })
output$html <- renderUI({ HTML(...) })
```

# 3. Reactive Programming

## Reactive Sources

```{r}
server <- function(input, output) {
  # Input values are reactive
  # They automatically trigger updates

  output$text <- renderText({
    paste("You entered:", input$name)
  })
}
```

## Reactive Expressions

```{r}
server <- function(input, output) {
  # Reactive expression (caches result)
  filtered_data <- reactive({
    mtcars[mtcars$cyl == input$cyl, ]
  })

  # Use reactive expression
  output$plot <- renderPlot({
    plot(filtered_data()$wt, filtered_data()$mpg)
  })

  output$table <- renderTable({
    filtered_data()
  })
}
```

## observe and observeEvent

```{r}
server <- function(input, output) {
  # observe - runs when dependencies change
  observe({
    print(paste("Input changed to:", input$value))
  })

  # observeEvent - runs when specific event occurs
  observeEvent(input$button, {
    print("Button was clicked!")
  })
}
```

## isolate

```{r}
server <- function(input, output) {
  # Prevent reactivity with isolate
  observeEvent(input$button, {
    # This only runs when button is clicked
    # Not when input$value changes
    value <- isolate(input$value)
    print(value)
  })
}
```

## eventReactive

```{r}
server <- function(input, output) {
  # Only update when button is clicked
  data <- eventReactive(input$button, {
    rnorm(input$n)
  })

  output$plot <- renderPlot({
    hist(data())
  })
}
```

## reactiveValues

```{r}
server <- function(input, output) {
  # Create reactive values object
  values <- reactiveValues(
    counter = 0,
    data = NULL
  )

  # Update values
  observeEvent(input$increment, {
    values$counter <- values$counter + 1
  })

  # Use values
  output$text <- renderText({
    paste("Count:", values$counter)
  })
}
```

# 4. Advanced Shiny Features

## Dynamic UI

```{r}
ui <- fluidPage(
  selectInput("type", "Select type:", choices = c("Plot", "Table")),
  uiOutput("dynamic_output")
)

server <- function(input, output) {
  output$dynamic_output <- renderUI({
    if (input$type == "Plot") {
      plotOutput("plot")
    } else {
      tableOutput("table")
    }
  })

  output$plot <- renderPlot({
    plot(mtcars$wt, mtcars$mpg)
  })

  output$table <- renderTable({
    mtcars
  })
}
```

## File Upload and Download

```{r}
ui <- fluidPage(
  # Upload
  fileInput("file", "Upload CSV:"),
  tableOutput("contents"),

  # Download
  downloadButton("download", "Download data")
)

server <- function(input, output) {
  # Upload
  data <- reactive({
    req(input$file)
    read.csv(input$file$datapath)
  })

  output$contents <- renderTable({
    data()
  })

  # Download
  output$download <- downloadHandler(
    filename = function() {
      paste("data-", Sys.Date(), ".csv", sep = "")
    },
    content = function(file) {
      write.csv(data(), file, row.names = FALSE)
    }
  )
}
```

## Progress Indicators

```{r}
server <- function(input, output) {
  output$plot <- renderPlot({
    # Create progress indicator
    withProgress(message = "Creating plot", value = 0, {
      n <- 10
      for (i in 1:n) {
        incProgress(1/n, detail = paste("Step", i))
        Sys.sleep(0.5)
      }
    })

    plot(mtcars$wt, mtcars$mpg)
  })
}
```

## Modals

```{r}
ui <- fluidPage(
  actionButton("show", "Show modal")
)

server <- function(input, output) {
  observeEvent(input$show, {
    showModal(modalDialog(
      title = "Important message",
      "This is an important message!",
      easyClose = TRUE,
      footer = modalButton("Close")
    ))
  })
}
```

## Notifications

```{r}
server <- function(input, output) {
  observeEvent(input$button, {
    showNotification("Process complete!", type = "message")
    # Types: "message", "warning", "error"
  })
}
```

# 5. Shiny Modules

Modules make code reusable and maintainable.

```{r}
# Module UI
histogramUI <- function(id) {
  ns <- NS(id)
  tagList(
    sliderInput(ns("bins"), "Number of bins:", 1, 50, 30),
    plotOutput(ns("plot"))
  )
}

# Module server
histogramServer <- function(id, data) {
  moduleServer(id, function(input, output, session) {
    output$plot <- renderPlot({
      hist(data(), breaks = input$bins)
    })
  })
}

# Use module in app
ui <- fluidPage(
  histogramUI("hist1"),
  histogramUI("hist2")
)

server <- function(input, output) {
  data1 <- reactive({ rnorm(100) })
  data2 <- reactive({ rnorm(100, mean = 10) })

  histogramServer("hist1", data1)
  histogramServer("hist2", data2)
}
```

# 6. Shiny Themes and Styling

## shinythemes

```{r}
library(shinythemes)

ui <- fluidPage(
  theme = shinytheme("cerulean"),
  # Or: "darkly", "flatly", "journal", "readable", "spacelab", etc.
  titlePanel("Themed App")
)
```

## bslib

```{r}
library(bslib)

ui <- fluidPage(
  theme = bs_theme(
    bg = "#FFFFFF",
    fg = "#000000",
    primary = "#0199F8",
    base_font = font_google("Inter")
  ),
  titlePanel("Custom Theme")
)
```

## Custom CSS

```{r}
ui <- fluidPage(
  tags$head(
    tags$style(HTML("
      .custom-class {
        color: blue;
        font-size: 20px;
      }
    "))
  ),
  div(class = "custom-class", "Styled text")
)
```

# 7. Interactive Visualizations

## plotly

```{r}
library(plotly)

ui <- fluidPage(
  plotlyOutput("plot")
)

server <- function(input, output) {
  output$plot <- renderPlotly({
    plot_ly(mtcars, x = ~wt, y = ~mpg, type = "scatter", mode = "markers")
  })
}
```

## DT (DataTables)

```{r}
library(DT)

ui <- fluidPage(
  DTOutput("table")
)

server <- function(input, output) {
  output$table <- renderDT({
    datatable(mtcars,
              options = list(pageLength = 10),
              filter = "top",
              selection = "multiple")
  })

  # Get selected rows
  observeEvent(input$table_rows_selected, {
    selected <- input$table_rows_selected
    print(mtcars[selected, ])
  })
}
```

## leaflet (Maps)

```{r}
library(leaflet)

ui <- fluidPage(
  leafletOutput("map")
)

server <- function(input, output) {
  output$map <- renderLeaflet({
    leaflet() %>%
      addTiles() %>%
      addMarkers(lng = -73.935242, lat = 40.730610,
                 popup = "New York City")
  })
}
```

# 8. Shiny Dashboards

## shinydashboard

```{r}
library(shinydashboard)

ui <- dashboardPage(
  dashboardHeader(title = "Dashboard"),

  dashboardSidebar(
    sidebarMenu(
      menuItem("Dashboard", tabName = "dashboard", icon = icon("dashboard")),
      menuItem("Data", tabName = "data", icon = icon("table"))
    )
  ),

  dashboardBody(
    tabItems(
      tabItem(tabName = "dashboard",
        fluidRow(
          valueBox(10 * 2, "New Orders", icon = icon("shopping-cart")),
          valueBox(100, "Approval Rating", icon = icon("thumbs-up"), color = "green"),
          infoBox("Progress", "70%", icon = icon("chart-line"), color = "purple")
        ),
        fluidRow(
          box(
            title = "Histogram",
            status = "primary",
            solidHeader = TRUE,
            plotOutput("plot1")
          ),
          box(
            title = "Controls",
            status = "warning",
            sliderInput("slider", "Number of observations:", 1, 100, 50)
          )
        )
      ),
      tabItem(tabName = "data",
        tableOutput("table")
      )
    )
  )
)

server <- function(input, output) {
  output$plot1 <- renderPlot({
    hist(rnorm(input$slider))
  })

  output$table <- renderTable({
    head(mtcars)
  })
}

shinyApp(ui, server)
```

# 9. flexdashboard

```{r}
---
title: "My Dashboard"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(ggplot2)
```

Row {data-height=200}
-----------------------------------------------------------------------

### Chart A

```{r}
valueBox(42, icon = "fa-pencil")
```

### Chart B

```{r}
gauge(91, min = 0, max = 100, symbol = '%', gaugeSectors(
  success = c(80, 100), warning = c(40, 79), danger = c(0, 39)
))
```

Row
-----------------------------------------------------------------------

### Chart C

```{r}
renderPlot({
  ggplot(mtcars, aes(wt, mpg)) + geom_point()
})
```

### Chart D

```{r}
renderTable({
  head(mtcars)
})
```
```

# 10. Deployment

## Local Deployment

```{r}
# Run from R
shiny::runApp("app_directory")

# Specify port
shiny::runApp("app_directory", port = 8080)
```

## shinyapps.io

```{r}
library(rsconnect)

# Set account
rsconnect::setAccountInfo(
  name = "your_account",
  token = "your_token",
  secret = "your_secret"
)

# Deploy
rsconnect::deployApp("app_directory")

# Update
rsconnect::deployApp("app_directory", forceUpdate = TRUE)
```

## Shiny Server (Open Source)

```bash
# Install Shiny Server on Ubuntu
sudo apt-get install gdebi-core
wget https://download3.rstudio.org/ubuntu-18.04/x86_64/shiny-server-xxx.deb
sudo gdebi shiny-server-xxx.deb

# Apps go in: /srv/shiny-server/
# Logs in: /var/log/shiny-server/
```

## Docker

```dockerfile
FROM rocker/shiny:latest

# Install R packages
RUN R -e "install.packages(c('shiny', 'ggplot2', 'dplyr'))"

# Copy app
COPY ./app /srv/shiny-server/app

# Expose port
EXPOSE 3838

# Run
CMD ["/usr/bin/shiny-server"]
```

# 11. Performance Optimization

## Caching

```{r}
server <- function(input, output) {
  # Cache expensive computations
  expensive_data <- reactive({
    # This only reruns when input$file changes
    read.csv(input$file$datapath)
  }) %>% bindCache(input$file)
}
```

## Async Processing

```{r}
library(promises)
library(future)
plan(multisession)

server <- function(input, output) {
  output$plot <- renderPlot({
    future({
      # Long computation in background
      Sys.sleep(5)
      mtcars
    }) %...>% {
      plot(.$wt, .$mpg)
    }
  })
}
```

## Debouncing and Throttling

```{r}
server <- function(input, output) {
  # Debounce - wait for pause in input
  debounced_value <- debounce(reactive(input$text), 1000)

  # Throttle - limit update frequency
  throttled_value <- throttle(reactive(input$slider), 1000)

  output$text <- renderText({
    debounced_value()
  })
}
```

# 12. Testing Shiny Apps

## shinytest

```{r}
library(shinytest)

# Record test
recordTest("app_directory")

# Run tests
testApp("app_directory")

# In testthat
library(testthat)
test_that("app works", {
  app <- ShinyDriver$new("app_directory")
  app$setInputs(bins = 30)
  output <- app$getValue(name = "distPlot")
  expect_true(!is.null(output))
})
```

# Summary

This document covered web applications and dashboards:

**Shiny Basics**:
- UI and server structure
- Inputs and outputs
- Layouts

**Reactive Programming**:
- Reactive expressions
- observe/observeEvent
- isolate, eventReactive
- reactiveValues

**Advanced Features**:
- Dynamic UI
- File upload/download
- Progress, modals, notifications
- Modules for reusability

**Styling**:
- shinythemes
- bslib
- Custom CSS

**Interactive Visualizations**:
- plotly
- DT (DataTables)
- leaflet (maps)

**Dashboards**:
- shinydashboard
- flexdashboard

**Deployment**:
- shinyapps.io
- Shiny Server
- Docker

**Performance**:
- Caching
- Async processing
- Debouncing/throttling

**Testing**:
- shinytest
- testthat integration

**Best Practices**:
- Modularize code
- Optimize reactivity
- Test thoroughly
- Document well

**Next Steps**: Proceed to Reporting and Communication for creating professional tables and figures.

# References

- [Mastering Shiny](https://mastering-shiny.org/)
- [Shiny Documentation](https://shiny.rstudio.com/)
- [shinydashboard](https://rstudio.github.io/shinydashboard/)
- [flexdashboard](https://pkgs.rstudio.com/flexdashboard/)
