---
title: "Reporting and Communication"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(knitr)
library(kableExtra)
library(ggplot2)
```

# 1. Creating Professional Tables

## knitr::kable()

```{r}
# Basic table
kable(head(mtcars))

# With caption
kable(head(iris), caption = "First 6 rows of iris dataset")

# Customize column names
kable(head(mtcars[, 1:3]),
      col.names = c("Miles/(US) gallon", "Number of cylinders", "Displacement"))

# Format numbers
kable(head(mtcars[, 1:3]), digits = 1)

# Alignment
kable(head(mtcars[, 1:3]), align = c("l", "c", "r"))
```

## kableExtra

```{r}
library(kableExtra)

# Styled table
kable(head(mtcars)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "left")

# Column specifications
kable(head(mtcars)) %>%
  kable_styling() %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "red") %>%
  column_spec(3, background = "lightblue")

# Row specifications
kable(head(mtcars)) %>%
  kable_styling() %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4F81BD") %>%
  row_spec(1:3, background = "#F0F0F0")

# Group rows
kable(mtcars[1:10, 1:4]) %>%
  kable_styling() %>%
  pack_rows("Group 1", 1, 4) %>%
  pack_rows("Group 2", 5, 10)

# Add footnotes
kable(head(mtcars)) %>%
  kable_styling() %>%
  footnote(general = "Data from mtcars dataset",
           number = c("Note 1", "Note 2"))

# Scroll box for large tables
kable(mtcars) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "400px")
```

## gt Package

```{r}
library(gt)

# Create gt table
gt_table <- head(mtcars) %>%
  gt() %>%
  tab_header(
    title = "Motor Trend Car Road Tests",
    subtitle = "First 6 cars"
  ) %>%
  tab_source_note(
    source_note = "Source: 1974 Motor Trend US magazine"
  ) %>%
  fmt_number(
    columns = c(mpg, disp, drat, wt, qsec),
    decimals = 1
  ) %>%
  cols_label(
    mpg = "MPG",
    cyl = "Cylinders",
    disp = "Displacement"
  ) %>%
  tab_style(
    style = cell_fill(color = "lightblue"),
    locations = cells_body(columns = mpg, rows = mpg > 20)
  )

gt_table
```

## flextable

```{r}
library(flextable)

# Create flextable
ft <- flextable(head(mtcars))

# Customize
ft <- ft %>%
  set_header_labels(
    mpg = "Miles/(US) gallon",
    cyl = "Number of cylinders"
  ) %>%
  color(i = ~ mpg > 20, j = "mpg", color = "red") %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  autofit()

ft
```

## DT (DataTables)

```{r}
library(DT)

# Interactive table
datatable(mtcars,
          options = list(
            pageLength = 10,
            searching = TRUE,
            ordering = TRUE
          ),
          filter = "top",
          rownames = FALSE)

# Format columns
datatable(iris) %>%
  formatRound(columns = 1:4, digits = 2) %>%
  formatStyle('Sepal.Length',
              backgroundColor = styleInterval(5, c('lightblue', 'lightgreen')))
```

# 2. Figure Best Practices

## Publication-Quality Figures

```{r}
# Clean, professional theme
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Fuel Efficiency vs Vehicle Weight",
    subtitle = "Motor Trend Car Road Tests",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    color = "Cylinders",
    caption = "Source: 1974 Motor Trend US magazine"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "gray90")
  ) +
  scale_color_brewer(palette = "Set1")

p

# Save with high resolution
ggsave("figure1.png", p, width = 8, height = 6, dpi = 300)
ggsave("figure1.pdf", p, width = 8, height = 6)
```

## Multi-Panel Figures

```{r}
library(patchwork)

p1 <- ggplot(mtcars, aes(wt, mpg)) + geom_point() + labs(tag = "A")
p2 <- ggplot(mtcars, aes(factor(cyl))) + geom_bar() + labs(tag = "B")
p3 <- ggplot(mtcars, aes(factor(cyl), mpg)) + geom_boxplot() + labs(tag = "C")

# Combine
combined <- (p1 | p2) / p3 +
  plot_annotation(
    title = "Motor Trend Analysis",
    subtitle = "Multiple aspects of car performance",
    caption = "Data from mtcars dataset"
  )

combined
```

## Colorblind-Friendly Palettes

```{r}
library(viridis)

# Viridis colors
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  scale_color_viridis_d() +
  theme_minimal()

# ColorBrewer
library(RColorBrewer)
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

# 3. Output Formats

## HTML Documents

```yaml
---
title: "My Report"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 8
    fig_height: 6
    fig_caption: true
---
```

## PDF Documents

```yaml
---
title: "My Report"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 5
    fig_caption: true
    keep_tex: true
---
```

## Word Documents

```yaml
---
title: "My Report"
output:
  word_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    reference_docx: template.docx
---
```

## beamer Presentations

```yaml
---
title: "My Presentation"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    slide_level: 2
---

# Section 1

## Slide 1

Content here

## Slide 2

More content
```

## xaringan (HTML Presentations)

```yaml
---
title: "Presentation Title"
subtitle: "Subtitle"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# First Slide

Content here

---

# Second Slide

More content
```

# 4. Document Customization

## Custom CSS

```yaml
---
output:
  html_document:
    css: custom.css
---
```

```css
/* custom.css */
body {
  font-family: 'Arial', sans-serif;
  font-size: 14px;
}

h1 {
  color: #2C3E50;
  border-bottom: 2px solid #3498DB;
}

.author, .date {
  font-style: italic;
  color: #7F8C8D;
}
```

## Custom LaTeX

```yaml
---
output:
  pdf_document:
    includes:
      in_header: header.tex
      before_body: before_body.tex
      after_body: after_body.tex
---
```

## Template Documents

```yaml
---
output:
  word_document:
    reference_docx: my_template.docx
  pdf_document:
    template: my_template.tex
---
```

# 5. Cross-Referencing

## bookdown for Cross-References

```yaml
---
output:
  bookdown::html_document2:
    number_sections: true
  bookdown::pdf_document2:
    number_sections: true
---
```

```markdown
See Figure \@ref(fig:plot1) and Table \@ref(tab:table1).

```{r plot1, fig.cap="My plot"}
plot(mtcars$wt, mtcars$mpg)
```

```{r table1}
kable(head(mtcars), caption = "My table")
```
```

# 6. Citations and Bibliography

## Setting up Bibliography

```yaml
---
title: "My Report"
bibliography: references.bib
csl: nature.csl
output: html_document
---
```

## BibTeX File (references.bib)

```bibtex
@article{wickham2014,
  title={Tidy data},
  author={Wickham, Hadley},
  journal={Journal of Statistical Software},
  volume={59},
  number={10},
  year={2014}
}
```

## Citing in Text

```markdown
According to @wickham2014, tidy data is important.

Multiple citations [@wickham2014; @author2020].

# References
```

# 7. Interactive Documents

## HTML Widgets

```{r}
library(plotly)
library(leaflet)
library(DT)

# Interactive plot
plot_ly(mtcars, x = ~wt, y = ~mpg, type = "scatter", mode = "markers")

# Interactive map
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -73.935242, lat = 40.730610, popup = "NYC")

# Interactive table
datatable(mtcars, filter = "top")
```

## runtime: shiny

```yaml
---
title: "Interactive Document"
output: html_document
runtime: shiny
---

```{r}
sliderInput("bins", "Number of bins:", 1, 50, 30)

renderPlot({
  hist(faithful$waiting, breaks = input$bins)
})
```
```

# 8. Long Documents and Books

## bookdown

```yaml
# index.Rmd
---
title: "My Book"
author: "Your Name"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    config:
      toc:
        collapse: section
      download: ["pdf", "epub"]
  bookdown::pdf_book:
    keep_tex: yes
---

# Preface {-}

This is the preface.
```

```yaml
# _bookdown.yml
book_filename: "mybook"
chapter_name: "Chapter "
rmd_files: ["index.Rmd", "01-intro.Rmd", "02-methods.Rmd"]
output_dir: "docs"
```

# 9. Scientific Writing Templates

## rticles Package

```{r eval=FALSE}
# Install
install.packages("rticles")

# Available templates
rticles::journals()

# Create new document
# File > New File > R Markdown > From Template
# Choose journal template
```

# 10. Best Practices

## Document Structure

```markdown
---
title: "Analysis Title"
author: "Author Name"
date: "`r Sys.Date()`"
output: html_document
---

# Executive Summary

Brief overview of findings

# Introduction

- Background
- Objectives
- Research questions

# Methods

## Data Collection
## Data Processing
## Statistical Analysis

# Results

## Descriptive Statistics
## Main Findings
## Subgroup Analysis

# Discussion

- Interpretation
- Limitations
- Future directions

# Conclusions

Key takeaways

# References

# Appendix

## Session Information
```

## Code Organization

```{r eval=FALSE}
# Setup chunk (at beginning)
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Don't show code
  warning = FALSE,   # Don't show warnings
  message = FALSE,   # Don't show messages
  fig.width = 8,     # Default figure width
  fig.height = 6,    # Default figure height
  dpi = 300          # High resolution
)

# Load libraries
library(tidyverse)
library(kableExtra)

# Set ggplot2 theme
theme_set(theme_minimal())
```

# Import data (hidden)
```{r}
data <- read.csv("data.csv")
```

# Show results (visible)
```{r echo=TRUE}
summary(data)
```
```

## Figure and Table Captions

```{r}
# Good caption
# ```{r fig-scatter, fig.cap="Figure 1. Relationship between vehicle weight and fuel efficiency. Points are colored by number of cylinders. Linear regression lines show negative correlation within each cylinder group. Data from 1974 Motor Trend US magazine."}
plot(mtcars$wt, mtcars$mpg)
# ```

# Table with informative caption
kable(head(mtcars), caption = "Table 1. Summary statistics for the first 6 vehicles in the Motor Trend Car Road Tests dataset. Values shown include fuel efficiency (mpg), engine configuration (cyl, hp), and vehicle dimensions (wt).")
```

## Reproducibility

```{r}
# Always include session info
sessionInfo()

# Or use devtools
library(devtools)
session_info()

# Include data sources
# Document package versions
# Include random seed if applicable
set.seed(123)
```

# 11. Accessibility

## Alt Text for Images

```yaml
```{r fig.alt="Scatter plot showing negative relationship between weight and MPG"}
plot(mtcars$wt, mtcars$mpg)
```
```

## Semantic Markup

```markdown
# Use proper heading hierarchy

## Second level

### Third level

- Use lists appropriately
- Use tables for tabular data
- Use emphasis (*italic*) and strong (**bold**) appropriately
```

# 12. Collaborative Writing

## Version Control

```bash
# Track Rmd files
git add analysis.Rmd
git commit -m "Add exploratory analysis"

# Ignore output files
# .gitignore:
*.html
*.pdf
*_cache/
*_files/
```

## Comments and TODOs

```markdown
<!-- This is a comment that won't appear in output -->

<!-- TODO: Add more detailed explanation here -->

<!-- NOTE: Check these numbers with collaborator -->
```

# Summary

This document covered reporting and communication:

**Tables**:
- kable for simple tables
- kableExtra for styled tables
- gt for complex tables
- DT for interactive tables

**Figures**:
- Publication-quality plots
- Multi-panel figures
- Colorblind-friendly palettes
- High-resolution output

**Output Formats**:
- HTML, PDF, Word
- Presentations (Beamer, xaringan)
- Books (bookdown)
- Journal articles (rticles)

**Customization**:
- Themes and styling
- Custom CSS/LaTeX
- Templates

**Advanced Features**:
- Cross-referencing
- Citations and bibliography
- Interactive documents
- HTML widgets

**Best Practices**:
- Clear structure
- Informative captions
- Reproducible
- Accessible

**Key Principles**:
- Know your audience
- Clear and concise
- Professional appearance
- Fully reproducible
- Well-documented

**Next Steps**: Proceed to Package Development for creating R packages.

# References

- [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
- [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [bookdown: Authoring Books with R Markdown](https://bookdown.org/yihui/bookdown/)
- [gt Documentation](https://gt.rstudio.com/)
- [kableExtra Documentation](https://haozhu233.github.io/kableExtra/)
