---
title: "R Package Development"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Introduction to Package Development

## Why Create Packages?

- **Organize code**: Structured, reusable functions
- **Share work**: With colleagues or the community
- **Documentation**: Force yourself to document well
- **Testing**: Built-in framework for tests
- **Reproducibility**: Versioned, installable units

## Package Structure

```
mypackage/
├── DESCRIPTION          # Package metadata
├── NAMESPACE            # Function exports/imports
├── R/                   # R code (.R files)
├── man/                 # Documentation (.Rd files)
├── tests/              # Unit tests
│   └── testthat/
├── vignettes/          # Long-form documentation
├── data/               # Data files
├── inst/               # Installed files
├── src/                # Compiled code (C/C++)
├── .Rbuildignore       # Files to ignore when building
└── README.md           # GitHub README
```

# 2. Getting Started

## Create Package

```{r}
# Using usethis
library(usethis)
create_package("~/mypackage")

# Or in RStudio: File > New Project > New Directory > R Package
```

## Essential Setup

```{r}
# Set up Git
use_git()

# Set up GitHub
use_github()

# Use roxygen2 for documentation
use_roxygen_md()

# Add license
use_mit_license("Your Name")
# Or: use_gpl3_license(), use_cc0_license()

# Add README
use_readme_md()

# Add NEWS file
use_news_md()
```

# 3. DESCRIPTION File

```yaml
Package: mypackage
Title: What The Package Does (One Line, Title Case)
Version: 0.1.0
Authors@R:
    person("First", "Last", , "email@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: MIT + file LICENSE
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.0
Imports:
    dplyr,
    ggplot2
Suggests:
    testthat (>= 3.0.0),
    knitr,
    rmarkdown
VignetteBuilder: knitr
URL: https://github.com/username/mypackage
BugReports: https://github.com/username/mypackage/issues
```

## Add Dependencies

```{r}
# Add to Imports (required)
use_package("dplyr")

# Add to Suggests (optional)
use_package("ggplot2", type = "Suggests")

# Specific version
use_package("rlang", min_version = "0.4.0")
```

# 4. Writing Functions

## Function File (R/)

```{r}
# Create new function file
use_r("my_function")

# In R/my_function.R:
#' Add Two Numbers
#'
#' This function adds two numbers together.
#'
#' @param x A numeric value
#' @param y A numeric value
#'
#' @return The sum of x and y
#' @export
#'
#' @examples
#' add_numbers(2, 3)
#' add_numbers(10, 20)
add_numbers <- function(x, y) {
  if (!is.numeric(x) || !is.numeric(y)) {
    stop("Both x and y must be numeric")
  }
  x + y
}
```

## Documentation with roxygen2

```{r}
#' Function Title
#'
#' Function description (one paragraph).
#'
#' @param param1 Description of param1
#' @param param2 Description of param2
#' @param ... Additional arguments
#'
#' @return Description of return value
#' @export
#'
#' @examples
#' my_function(10, 20)
#'
#' @seealso \code{\link{related_function}}
#' @references Author (Year). Title. Journal.
#'
#' @importFrom dplyr filter mutate
#' @importFrom ggplot2 ggplot aes
my_function <- function(param1, param2, ...) {
  # Function body
}
```

## Common roxygen2 Tags

```{r}
#' @param parameter Description
#' @return Description of return value
#' @export                    # Export function
#' @examples                  # Usage examples
#' @seealso                   # Related functions
#' @references                # Citations
#' @author Author Name
#' @keywords keyword1 keyword2
#'
#' @import package            # Import entire package
#' @importFrom package function  # Import specific function
#' @importClassesFrom package class
#' @importMethodsFrom package method
#'
#' @family function_family    # Group related functions
#' @inherit source_function   # Inherit documentation
#' @inheritParams source_function  # Inherit parameter docs
#' @inheritDotParams source_function  # Inherit ... docs
#'
#' @section Warning:          # Special sections
#' Warning text here
#' @section Details:
#' Detailed description
```

## Generate Documentation

```{r}
# Generate .Rd files from roxygen comments
devtools::document()

# Or
roxygen2::roxygenise()
```

# 5. NAMESPACE

NAMESPACE controls which functions are exported and which are imported.

```{r}
# Automatically generated by roxygen2
# Don't edit manually!

# Export functions
export(my_function)
export(another_function)

# Import from other packages
importFrom(dplyr, filter)
importFrom(dplyr, mutate)
importFrom(ggplot2, ggplot)

# Import all from package
import(magrittr)

# S3 methods
S3method(print, myclass)
```

# 6. Data in Packages

## Add Data

```{r}
# Create data object
my_data <- data.frame(
  x = 1:10,
  y = rnorm(10)
)

# Save as package data
usethis::use_data(my_data, overwrite = TRUE)

# Document data (in R/data.R)
#' Example Dataset
#'
#' A dataset containing example values.
#'
#' @format A data frame with 10 rows and 2 variables:
#' \describe{
#'   \item{x}{Integer values from 1 to 10}
#'   \item{y}{Random normal values}
#' }
#' @source Generated for package examples
"my_data"
```

## Internal Data

```{r}
# Data for internal use only
usethis::use_data(internal_data, internal = TRUE)
```

## Raw Data

```{r}
# Store raw data
usethis::use_data_raw("my_data")

# This creates data-raw/my_data.R
# Add code to process raw data and create cleaned version
```

# 7. Testing

## Setup Testing

```{r}
# Set up testthat
usethis::use_testthat()

# Create test file
usethis::use_test("my_function")
```

## Writing Tests

```{r}
# In tests/testthat/test-my_function.R
test_that("add_numbers works correctly", {
  expect_equal(add_numbers(2, 3), 5)
  expect_equal(add_numbers(-1, 1), 0)
  expect_equal(add_numbers(0, 0), 0)
})

test_that("add_numbers handles errors", {
  expect_error(add_numbers("a", 5))
  expect_error(add_numbers(5, "b"))
  expect_error(add_numbers(NA, 5))
})

test_that("add_numbers handles edge cases", {
  expect_equal(add_numbers(Inf, 1), Inf)
  expect_equal(add_numbers(1e10, 1e10), 2e10)
})
```

## Common Expectations

```{r}
# Equality
expect_equal(object, expected)
expect_identical(object, expected)

# Type
expect_type(object, "double")
expect_s3_class(object, "data.frame")
expect_s4_class(object, "GRanges")

# Truthiness
expect_true(condition)
expect_false(condition)

# Errors and warnings
expect_error(code, regexp = "error message")
expect_warning(code, regexp = "warning message")
expect_message(code, regexp = "message")

# Output
expect_output(print(x), "expected output")
expect_silent(code)

# Length
expect_length(object, n)

# NA
expect_null(object)
expect_na(object)

# Comparison
expect_gt(x, y)  # Greater than
expect_lt(x, y)  # Less than
expect_gte(x, y)
expect_lte(x, y)

# Vector comparison
expect_setequal(x, y)
expect_contains(x, y)
```

## Run Tests

```{r}
# Run all tests
devtools::test()

# Test specific file
testthat::test_file("tests/testthat/test-my_function.R")

# Test coverage
covr::package_coverage()
covr::report()
```

# 8. Vignettes

## Create Vignette

```{r}
usethis::use_vignette("my-vignette")
```

## Vignette Template

```markdown
---
title: "Getting Started with mypackage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mypackage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mypackage)
```

# Introduction

This vignette shows how to use mypackage.

# Basic Usage

```{r}
result <- my_function(10, 20)
```

# Advanced Examples

More examples here...
```

## Build Vignettes

```{r}
# Build all vignettes
devtools::build_vignettes()

# View vignette
vignette("my-vignette", package = "mypackage")
```

# 9. Package Website with pkgdown

## Setup

```{r}
# Install pkgdown
install.packages("pkgdown")

# Initialize
usethis::use_pkgdown()

# Build site
pkgdown::build_site()
```

## Customize

```yaml
# _pkgdown.yml
url: https://username.github.io/mypackage

template:
  bootstrap: 5

reference:
  - title: "Main Functions"
    contents:
    - function1
    - function2
  - title: "Helper Functions"
    contents:
    - helper1
    - helper2

articles:
  - title: "Tutorials"
    contents:
    - vignette1
    - vignette2

navbar:
  left:
    - text: "Reference"
      href: reference/index.html
    - text: "Articles"
      href: articles/index.html
  right:
    - icon: fa-github
      href: https://github.com/username/mypackage
```

# 10. Checking and Building

## Check Package

```{r}
# Full package check
devtools::check()

# Quick check
devtools::load_all()

# Check for specific issues
rcmdcheck::rcmdcheck()
```

## Common Issues

```{r}
# Fix NAMESPACE
devtools::document()

# Check documentation
devtools::check_man()

# Spell check
spelling::spell_check_package()

# Check URLs
urlchecker::url_check()

# Check for CRAN
devtools::check(cran = TRUE)
```

## Build Package

```{r}
# Build source package
devtools::build()

# Build binary package
devtools::build(binary = TRUE)

# Install package
devtools::install()

# Install from GitHub
devtools::install_github("username/mypackage")
```

# 11. Version Control and Releases

## Versioning

```{r}
# Increment version (in DESCRIPTION)
usethis::use_version()

# Or manually follow semantic versioning:
# Major.Minor.Patch
# 0.1.0 -> Initial development
# 1.0.0 -> First stable release
# 1.1.0 -> Add features
# 1.1.1 -> Bug fixes
```

## GitHub Releases

```{r}
# Create release
usethis::use_github_release()

# Add to NEWS.md
```

## Continuous Integration

```{r}
# GitHub Actions
usethis::use_github_action("check-standard")
usethis::use_github_action("test-coverage")
usethis::use_github_action("pkgdown")
```

# 12. CRAN Submission

## Prepare for CRAN

```{r}
# Run checks
devtools::check(cran = TRUE)

# Check on different platforms
devtools::check_win_devel()
devtools::check_mac_release()
rhub::check_for_cran()

# Check reverse dependencies
revdepcheck::revdep_check()

# Update cran-comments.md
usethis::use_cran_comments()
```

## Submit to CRAN

```{r}
# Submit
devtools::submit_cran()

# Or manually:
# 1. Build source package
# 2. Submit at https://cran.r-project.org/submit.html
# 3. Confirm email
```

## CRAN Comments Template

```markdown
## Test environments
* local: macOS (R 4.2.0)
* GitHub Actions: windows-latest, macOS-latest, ubuntu-latest (R-devel, R-release)
* win-builder: R-devel, R-release

## R CMD check results
0 errors | 0 warnings | 0 notes

## Downstream dependencies
No downstream dependencies currently.
```

# 13. Bioconductor Submission

## Bioconductor Requirements

```{r}
# Check Bioconductor requirements
BiocCheck::BiocCheck(".")

# Add biocViews
# In DESCRIPTION:
# biocViews: Software, Genetics, Sequencing
```

## Submit to Bioconductor

1. Create package following Bioconductor guidelines
2. Add to GitHub
3. Open issue at https://github.com/Bioconductor/Contributions
4. Follow review process

# 14. Advanced Topics

## S3 Methods

```{r}
# Create S3 class
my_class <- function(x) {
  structure(list(data = x), class = "my_class")
}

# Print method
#' @export
print.my_class <- function(x, ...) {
  cat("My Class Object\n")
  print(x$data)
}

# Summary method
#' @export
summary.my_class <- function(object, ...) {
  summary(object$data)
}

# Register S3 methods in NAMESPACE
# S3method(print, my_class)
# S3method(summary, my_class)
```

## S4 Classes

```{r}
# Define S4 class (in R/AllClasses.R)
#' My S4 Class
#' @slot data numeric vector
#' @exportClass MyClass
setClass("MyClass",
  slots = c(data = "numeric")
)

# Constructor
#' @export
MyClass <- function(data) {
  new("MyClass", data = data)
}

# Methods (in R/AllMethods.R)
#' @export
setMethod("show", "MyClass", function(object) {
  cat("MyClass object with", length(object@data), "elements\n")
})
```

## Compiled Code (C/C++)

```{r}
# Use Rcpp
usethis::use_rcpp()

# Create C++ file (src/)
# src/add.cpp:
// [[Rcpp::export]]
double add_cpp(double x, double y) {
  return x + y;
}

# Compile and reload
devtools::load_all()
```

# 15. Best Practices

## Code Style

```{r}
# Use styler to format code
styler::style_pkg()

# Use lintr to check style
lintr::lint_package()
```

## Documentation

- Write clear, concise documentation
- Include examples for all exported functions
- Create vignettes for workflows
- Keep NEWS.md updated

## Testing

- Aim for >80% code coverage
- Test edge cases
- Test error conditions
- Use continuous integration

## Maintenance

- Respond to issues and PRs
- Keep dependencies updated
- Follow semantic versioning
- Deprecate functions properly

```{r}
# Deprecate function
#' @export
old_function <- function() {
  .Deprecated("new_function")
  new_function()
}
```

# 16. Useful Commands

```{r}
# Development workflow
devtools::load_all()     # Load package
devtools::document()     # Update documentation
devtools::test()         # Run tests
devtools::check()        # Check package

# Documentation
devtools::build_readme()
pkgdown::build_site()

# Installation
devtools::install()
devtools::install_github("user/pkg")

# Checking
goodpractice::gp()
spelling::spell_check_package()
urlchecker::url_check()

# Coverage
covr::package_coverage()
```

# Summary

This document covered R package development:

**Setup**:
- Package structure
- DESCRIPTION file
- Dependencies

**Code**:
- Writing functions
- roxygen2 documentation
- NAMESPACE

**Data**:
- Package data
- Internal data
- Data documentation

**Testing**:
- testthat framework
- Test coverage
- Continuous integration

**Documentation**:
- Vignettes
- pkgdown websites
- README

**Distribution**:
- GitHub
- CRAN submission
- Bioconductor submission

**Advanced**:
- S3/S4 classes
- Compiled code
- Best practices

**Key Principles**:
- Document everything
- Test thoroughly
- Version carefully
- Follow standards
- Maintain actively

**Resources**:
- Hadley Wickham's "R Packages" book
- devtools and usethis packages
- CRAN and Bioconductor guidelines

Creating packages:
- Organizes your work
- Facilitates sharing
- Ensures reproducibility
- Builds community

**Congratulations!** You've completed all 15 R knowledge modules essential for data scientists and computational bioinformaticians.

# References

- [R Packages (2e)](https://r-pkgs.org/)
- [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
- [Bioconductor Package Guidelines](https://bioconductor.org/developers/package-guidelines/)
- [rOpenSci Packages Guide](https://devguide.ropensci.org/)
