---
title: "Essential Data Analysis Toolkit: dplyr, tidyr, ggplot2, and caret"
author: "Comprehensive Guide with Practical Examples"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Introduction

This document provides comprehensive coverage of the most essential R packages for data analysis:

- **dplyr & tidyr**: Data manipulation and reshaping
- **ggplot2**: Data visualization
- **caret**: Machine learning workflows

These form the core toolkit for data scientists and are used in nearly every analysis project.

---

# Part 1: Data Manipulation with dplyr and tidyr

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## 1.1 Core dplyr Functions

### select() - Column Selection

```{r}
# Load example data
data(mtcars)
data(iris)

# Select specific columns
mtcars %>%
  select(mpg, cyl, hp)

# Select range of columns
mtcars %>%
  select(mpg:hp)

# Select by position
mtcars %>%
  select(1:3)

# Exclude columns
mtcars %>%
  select(-mpg, -cyl)

# Select helpers
iris %>%
  select(starts_with("Sepal"))

iris %>%
  select(ends_with("Width"))

iris %>%
  select(contains("pal"))

# Select by type
iris %>%
  select(where(is.numeric))

iris %>%
  select(where(is.factor))

# Rename while selecting
mtcars %>%
  select(miles_per_gallon = mpg, cylinders = cyl)

# Reorder columns
mtcars %>%
  select(hp, mpg, everything())
```

### filter() - Row Filtering

```{r}
# Single condition
mtcars %>%
  filter(mpg > 20)

# Multiple conditions (AND)
mtcars %>%
  filter(mpg > 20, cyl == 6)

mtcars %>%
  filter(mpg > 20 & cyl == 6)

# OR conditions
mtcars %>%
  filter(mpg > 25 | hp > 200)

# Complex conditions
mtcars %>%
  filter((mpg > 20 & cyl == 4) | (mpg > 15 & cyl == 8))

# IN operator
mtcars %>%
  filter(cyl %in% c(4, 6))

# NOT IN
mtcars %>%
  filter(!cyl %in% c(4, 6))

# Between
mtcars %>%
  filter(between(mpg, 15, 20))

# Remove NAs
df_with_na <- data.frame(x = c(1, 2, NA, 4), y = c(NA, 2, 3, 4))
df_with_na %>%
  filter(!is.na(x))

# String matching
iris %>%
  filter(grepl("^set", Species))

# Filter by quantile
mtcars %>%
  filter(mpg > quantile(mpg, 0.75))

# Top n values
mtcars %>%
  slice_max(mpg, n = 5)

mtcars %>%
  slice_min(mpg, n = 5)

# Random sample
mtcars %>%
  slice_sample(n = 10)

mtcars %>%
  slice_sample(prop = 0.25)
```

### mutate() - Create/Modify Columns

```{r}
# Add new column
mtcars %>%
  mutate(gpm = 1/mpg)

# Multiple new columns
mtcars %>%
  mutate(
    gpm = 1/mpg,
    hp_per_cyl = hp/cyl,
    wt_tons = wt/2
  )

# Modify existing column
mtcars %>%
  mutate(hp = hp * 1.1)

# Conditional mutation
mtcars %>%
  mutate(
    efficiency = case_when(
      mpg > 25 ~ "High",
      mpg > 20 ~ "Medium",
      TRUE ~ "Low"
    )
  )

# Multiple conditions
mtcars %>%
  mutate(
    category = case_when(
      cyl == 4 & mpg > 25 ~ "Efficient Small",
      cyl == 4 & mpg <= 25 ~ "Average Small",
      cyl == 6 ~ "Medium",
      cyl == 8 & hp > 200 ~ "Powerful Large",
      TRUE ~ "Other Large"
    )
  )

# Use previous mutations
mtcars %>%
  mutate(
    hp_per_cyl = hp/cyl,
    hp_per_cyl_scaled = scale(hp_per_cyl)
  )

# Across multiple columns
iris %>%
  mutate(across(starts_with("Sepal"), ~ . * 10))

iris %>%
  mutate(across(where(is.numeric), ~ round(., 1)))

# Multiple functions across columns
iris %>%
  mutate(across(where(is.numeric),
                list(mean = mean, sd = sd),
                .names = "{.col}_{.fn}"))

# Conditional across
iris %>%
  mutate(across(where(is.numeric),
                ~ if_else(. > mean(.), "High", "Low")))

# Window functions
mtcars %>%
  mutate(
    rank = row_number(desc(mpg)),
    percentile = percent_rank(mpg),
    cumulative_hp = cumsum(hp)
  )

# Lead and lag
mtcars %>%
  arrange(mpg) %>%
  mutate(
    mpg_prev = lag(mpg),
    mpg_next = lead(mpg),
    mpg_diff = mpg - lag(mpg)
  )
```

### summarise() - Aggregate Data

```{r}
# Single summary
mtcars %>%
  summarise(mean_mpg = mean(mpg))

# Multiple summaries
mtcars %>%
  summarise(
    n = n(),
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    median_mpg = median(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg),
    q25 = quantile(mpg, 0.25),
    q75 = quantile(mpg, 0.75)
  )

# Count distinct values
mtcars %>%
  summarise(
    n_cars = n(),
    n_cyl_types = n_distinct(cyl)
  )

# Across in summarise
iris %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, sd = sd)))

# Custom summary function
custom_summary <- function(x) {
  c(
    mean = mean(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    n = sum(!is.na(x))
  )
}

mtcars %>%
  summarise(across(c(mpg, hp), custom_summary))
```

### arrange() - Sort Data

```{r}
# Sort ascending
mtcars %>%
  arrange(mpg)

# Sort descending
mtcars %>%
  arrange(desc(mpg))

# Multiple columns
mtcars %>%
  arrange(cyl, desc(mpg))

# Sort by expression
mtcars %>%
  arrange(desc(hp/wt))

# Put NAs first or last
df_with_na <- data.frame(x = c(3, 1, NA, 2))
df_with_na %>% arrange(x)  # NA last (default)
```

### group_by() - Group Operations

```{r}
# Basic grouping
mtcars %>%
  group_by(cyl) %>%
  summarise(mean_mpg = mean(mpg))

# Multiple grouping variables
mtcars %>%
  group_by(cyl, am) %>%
  summarise(
    n = n(),
    mean_mpg = mean(mpg),
    mean_hp = mean(hp)
  )

# Group and mutate (within-group operations)
mtcars %>%
  group_by(cyl) %>%
  mutate(
    mean_mpg_by_cyl = mean(mpg),
    mpg_centered = mpg - mean(mpg)
  )

# Group and filter
mtcars %>%
  group_by(cyl) %>%
  filter(mpg > mean(mpg))

# Count by group
mtcars %>%
  count(cyl, am)

# Add count as column
mtcars %>%
  group_by(cyl) %>%
  add_count()

# Top n per group
mtcars %>%
  group_by(cyl) %>%
  slice_max(mpg, n = 2)

# Window functions by group
mtcars %>%
  group_by(cyl) %>%
  mutate(
    rank_within_cyl = row_number(desc(mpg)),
    mpg_percentile = percent_rank(mpg)
  )
```

## 1.2 Complex dplyr Workflows

### Real-World Example 1: Customer Analysis

```{r}
# Create sample customer data
set.seed(123)
customers <- data.frame(
  customer_id = 1:1000,
  age = sample(18:80, 1000, replace = TRUE),
  gender = sample(c("M", "F"), 1000, replace = TRUE),
  region = sample(c("North", "South", "East", "West"), 1000, replace = TRUE),
  purchase_amount = round(rnorm(1000, 100, 50), 2),
  purchase_date = sample(seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day"),
                        1000, replace = TRUE),
  product_category = sample(c("Electronics", "Clothing", "Food", "Books"), 1000, replace = TRUE)
)

# Complex analysis workflow
customer_analysis <- customers %>%
  # Add derived columns
  mutate(
    age_group = case_when(
      age < 25 ~ "18-24",
      age < 35 ~ "25-34",
      age < 50 ~ "35-49",
      TRUE ~ "50+"
    ),
    month = format(purchase_date, "%Y-%m"),
    high_value = purchase_amount > 150
  ) %>%
  # Group by multiple dimensions
  group_by(region, age_group, product_category) %>%
  # Calculate metrics
  summarise(
    n_customers = n_distinct(customer_id),
    n_purchases = n(),
    total_revenue = sum(purchase_amount),
    avg_purchase = mean(purchase_amount),
    pct_high_value = mean(high_value) * 100,
    .groups = "drop"
  ) %>%
  # Filter to significant segments
  filter(n_purchases >= 5) %>%
  # Add revenue rank
  arrange(desc(total_revenue)) %>%
  mutate(revenue_rank = row_number()) %>%
  # Reorder columns
  select(revenue_rank, region, age_group, product_category,
         n_customers, n_purchases, total_revenue, avg_purchase, pct_high_value)

head(customer_analysis, 10)
```

### Real-World Example 2: Gene Expression Analysis

```{r}
# Create sample gene expression data
set.seed(123)
gene_expression <- data.frame(
  gene_id = paste0("Gene_", rep(1:100, each = 6)),
  sample_id = paste0("Sample_", rep(1:6, times = 100)),
  condition = rep(rep(c("Control", "Treatment"), each = 3), times = 100),
  replicate = rep(rep(1:3, times = 2), times = 100),
  expression = rnorm(600, mean = 100, sd = 20)
) %>%
  mutate(
    expression = ifelse(
      gene_id %in% paste0("Gene_", 1:20) & condition == "Treatment",
      expression * runif(n(), 1.5, 2.5),
      expression
    )
  )

# Differential expression analysis workflow
diff_expr_analysis <- gene_expression %>%
  # Calculate statistics by gene and condition
  group_by(gene_id, condition) %>%
  summarise(
    n = n(),
    mean_expr = mean(expression),
    sd_expr = sd(expression),
    se_expr = sd_expr / sqrt(n),
    .groups = "drop"
  ) %>%
  # Reshape to wide format for comparison
  pivot_wider(
    names_from = condition,
    values_from = c(mean_expr, sd_expr, se_expr)
  ) %>%
  # Calculate fold change and statistics
  mutate(
    fold_change = mean_expr_Treatment / mean_expr_Control,
    log2_fc = log2(fold_change),
    abs_log2_fc = abs(log2_fc),
    significant = abs_log2_fc > 1
  ) %>%
  # Add gene ranks
  arrange(desc(abs_log2_fc)) %>%
  mutate(rank = row_number()) %>%
  # Classify genes
  mutate(
    regulation = case_when(
      log2_fc > 1 ~ "Upregulated",
      log2_fc < -1 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  )

# Summary of results
diff_expr_analysis %>%
  count(regulation) %>%
  mutate(percentage = n / sum(n) * 100)

# Top upregulated genes
diff_expr_analysis %>%
  filter(regulation == "Upregulated") %>%
  arrange(desc(log2_fc)) %>%
  select(rank, gene_id, log2_fc, mean_expr_Control, mean_expr_Treatment) %>%
  head(10)
```

## 1.3 tidyr for Data Reshaping

### pivot_longer() - Wide to Long

```{r}
# Create wide data
wide_data <- data.frame(
  id = 1:5,
  name = paste("Person", 1:5),
  age = c(25, 30, 35, 28, 32),
  score_2020 = c(85, 90, 78, 92, 88),
  score_2021 = c(88, 92, 80, 95, 90),
  score_2022 = c(90, 94, 85, 97, 92)
)

# Basic pivot longer
wide_data %>%
  pivot_longer(
    cols = starts_with("score_"),
    names_to = "year",
    values_to = "score"
  )

# Extract year as numeric
wide_data %>%
  pivot_longer(
    cols = starts_with("score_"),
    names_to = "year",
    names_prefix = "score_",
    names_transform = list(year = as.integer),
    values_to = "score"
  )

# Multiple value columns
results_wide <- data.frame(
  id = 1:3,
  test1_score = c(85, 90, 78),
  test1_time = c(45, 40, 50),
  test2_score = c(88, 92, 80),
  test2_time = c(42, 38, 48)
)

results_wide %>%
  pivot_longer(
    cols = -id,
    names_to = c("test", ".value"),
    names_sep = "_"
  )
```

### pivot_wider() - Long to Wide

```{r}
# Create long data
long_data <- data.frame(
  id = rep(1:3, each = 3),
  year = rep(c(2020, 2021, 2022), times = 3),
  value = c(100, 110, 120, 200, 210, 220, 300, 310, 320)
)

# Basic pivot wider
long_data %>%
  pivot_wider(
    names_from = year,
    values_from = value
  )

# With name prefix
long_data %>%
  pivot_wider(
    names_from = year,
    names_prefix = "year_",
    values_from = value
  )

# Multiple value columns
long_multi <- data.frame(
  id = rep(1:2, each = 2),
  metric = rep(c("weight", "height"), times = 2),
  value = c(70, 170, 80, 180),
  unit = c("kg", "cm", "kg", "cm")
)

long_multi %>%
  pivot_wider(
    names_from = metric,
    values_from = c(value, unit)
  )
```

### separate() and unite()

```{r}
# Separate column
combined_data <- data.frame(
  id = 1:5,
  date_time = c("2023-01-15_14:30", "2023-02-20_16:45",
                "2023-03-10_09:15", "2023-04-05_11:00",
                "2023-05-12_15:30")
)

combined_data %>%
  separate(date_time,
           into = c("date", "time"),
           sep = "_")

# Separate into multiple columns
patient_data <- data.frame(
  id = 1:3,
  info = c("John_Doe_35_M", "Jane_Smith_28_F", "Bob_Johnson_42_M")
)

patient_data %>%
  separate(info,
           into = c("first_name", "last_name", "age", "gender"),
           sep = "_",
           convert = TRUE)  # Auto-convert types

# Unite columns
separated_data <- data.frame(
  id = 1:3,
  first = c("John", "Jane", "Bob"),
  last = c("Doe", "Smith", "Johnson")
)

separated_data %>%
  unite("full_name", first, last, sep = " ")

# Unite keeping originals
separated_data %>%
  unite("full_name", first, last, sep = " ", remove = FALSE)
```

### Advanced Reshaping Example

```{r}
# Complex clinical trial data
clinical_data <- data.frame(
  patient_id = rep(1:4, each = 3),
  visit = rep(c("Baseline", "Week_4", "Week_8"), times = 4),
  treatment = rep(c("Drug_A", "Drug_A", "Placebo", "Placebo"), each = 3),
  biomarker1 = round(rnorm(12, 100, 15), 1),
  biomarker2 = round(rnorm(12, 50, 10), 1),
  symptom_score = sample(0:10, 12, replace = TRUE)
)

# Complex transformation pipeline
clinical_transformed <- clinical_data %>%
  # Pivot biomarkers to long format
  pivot_longer(
    cols = starts_with("biomarker"),
    names_to = "biomarker",
    values_to = "value"
  ) %>%
  # Create combined identifier
  unite("measurement", biomarker, visit, sep = "_", remove = FALSE) %>%
  # Calculate baseline-adjusted values
  group_by(patient_id, biomarker) %>%
  mutate(
    baseline_value = first(value),
    change_from_baseline = value - baseline_value,
    pct_change = (value - baseline_value) / baseline_value * 100
  ) %>%
  ungroup() %>%
  # Pivot back to wide for analysis
  select(patient_id, treatment, measurement, change_from_baseline) %>%
  pivot_wider(
    names_from = measurement,
    values_from = change_from_baseline
  )

head(clinical_transformed)
```

## 1.4 Combining dplyr and tidyr

### Complete Data Transformation Workflow

```{r}
# Create messy sales data
sales_messy <- data.frame(
  transaction = paste0("TXN_", 1:20),
  customer_info = c(
    rep("Region_North_Type_Premium", 5),
    rep("Region_South_Type_Standard", 5),
    rep("Region_East_Type_Premium", 5),
    rep("Region_West_Type_Standard", 5)
  ),
  Q1_2023 = round(rnorm(20, 10000, 2000), 0),
  Q2_2023 = round(rnorm(20, 12000, 2000), 0),
  Q3_2023 = round(rnorm(20, 11000, 2000), 0),
  Q4_2023 = round(rnorm(20, 15000, 3000), 0)
)

# Complete transformation
sales_clean <- sales_messy %>%
  # Separate customer info
  separate(customer_info,
           into = c("region_label", "region", "type_label", "customer_type"),
           sep = "_") %>%
  select(-region_label, -type_label) %>%
  # Pivot quarters to long format
  pivot_longer(
    cols = matches("Q[1-4]_2023"),
    names_to = "quarter",
    values_to = "sales"
  ) %>%
  # Extract quarter number
  mutate(quarter_num = as.integer(gsub("Q([1-4])_2023", "\\1", quarter))) %>%
  # Calculate metrics by group
  group_by(region, customer_type, quarter_num) %>%
  summarise(
    n_transactions = n(),
    total_sales = sum(sales),
    avg_sales = mean(sales),
    median_sales = median(sales),
    .groups = "drop"
  ) %>%
  # Calculate growth rates
  arrange(region, customer_type, quarter_num) %>%
  group_by(region, customer_type) %>%
  mutate(
    qoq_growth = (total_sales - lag(total_sales)) / lag(total_sales) * 100,
    cumulative_sales = cumsum(total_sales)
  ) %>%
  ungroup() %>%
  # Add rankings
  group_by(quarter_num) %>%
  mutate(
    sales_rank = dense_rank(desc(total_sales))
  ) %>%
  ungroup()

# View results
sales_clean %>%
  filter(quarter_num == 4) %>%
  arrange(sales_rank)
```

---

# Part 2: Data Visualization with ggplot2

```{r}
library(ggplot2)
library(scales)
library(RColorBrewer)
```

## 2.1 ggplot2 Grammar of Graphics

### Basic Components

```{r}
# Basic structure
# ggplot(data, aes(x, y, ...)) + geom_*() + ...

# Minimal plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# With color
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point()

# With size
ggplot(mtcars, aes(x = wt, y = mpg, size = hp, color = factor(cyl))) +
  geom_point(alpha = 0.7)
```

## 2.2 Scatter Plots

```{r}
# Basic scatter plot
p1 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, alpha = 0.6) +
  labs(
    title = "Fuel Efficiency vs Vehicle Weight",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_minimal()

p1

# Scatter with trend line
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "With Linear Regression Line") +
  theme_minimal()

# Multiple groups
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_brewer(palette = "Set1", name = "Cylinders") +
  labs(
    title = "Fuel Efficiency by Cylinder Count",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Scatter with marginal distributions
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()

# Text labels
top_cars <- mtcars %>%
  arrange(desc(mpg)) %>%
  head(5) %>%
  mutate(car = rownames(.))

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(alpha = 0.5) +
  geom_text(data = top_cars, aes(label = car),
            hjust = -0.1, size = 3, check_overlap = TRUE) +
  theme_minimal()

# Bubble chart
ggplot(mtcars, aes(x = wt, y = mpg, size = hp, color = qsec)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(range = c(3, 12), name = "Horsepower") +
  scale_color_viridis_c(name = "Quarter Mile Time") +
  labs(
    title = "Multi-dimensional Car Performance",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_minimal()
```

## 2.3 Line Plots

```{r}
# Create time series data
set.seed(123)
time_data <- data.frame(
  date = seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day"),
  value = cumsum(rnorm(365, 0, 10)) + 100
) %>%
  mutate(
    month = format(date, "%Y-%m"),
    group = sample(c("A", "B", "C"), n(), replace = TRUE)
  )

# Basic line plot
ggplot(time_data, aes(x = date, y = value)) +
  geom_line(color = "steelblue", size = 1) +
  labs(
    title = "Time Series Data",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal()

# Multiple lines
monthly_data <- time_data %>%
  group_by(month, group) %>%
  summarise(mean_value = mean(value), .groups = "drop") %>%
  mutate(month_date = as.Date(paste0(month, "-01")))

ggplot(monthly_data, aes(x = month_date, y = mean_value, color = group)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Monthly Trends by Group",
    x = "Month",
    y = "Average Value"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Area plot
ggplot(time_data %>% head(100), aes(x = date, y = value)) +
  geom_area(fill = "steelblue", alpha = 0.5) +
  geom_line(color = "darkblue", size = 1) +
  labs(title = "Area Plot") +
  theme_minimal()

# Stacked area
ggplot(monthly_data, aes(x = month_date, y = mean_value, fill = group)) +
  geom_area(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Stacked Area Chart") +
  theme_minimal()

# Step plot
ggplot(time_data %>% head(50), aes(x = date, y = value)) +
  geom_step(color = "steelblue", size = 1) +
  labs(title = "Step Plot") +
  theme_minimal()
```

## 2.4 Bar Plots

```{r}
# Count bar plot
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Cylinder Counts",
    x = "Number of Cylinders",
    y = "Count"
  ) +
  theme_minimal()

# Bar plot with values
summary_data <- mtcars %>%
  group_by(cyl) %>%
  summarise(mean_mpg = mean(mpg), se = sd(mpg)/sqrt(n()))

ggplot(summary_data, aes(x = factor(cyl), y = mean_mpg)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_mpg - se, ymax = mean_mpg + se),
                width = 0.2) +
  geom_text(aes(label = round(mean_mpg, 1)), vjust = -1) +
  labs(
    title = "Mean MPG by Cylinder Count",
    x = "Cylinders",
    y = "Mean MPG"
  ) +
  theme_minimal()

# Grouped bar plot
grouped_data <- mtcars %>%
  group_by(cyl, am) %>%
  summarise(mean_mpg = mean(mpg), .groups = "drop") %>%
  mutate(transmission = factor(am, labels = c("Automatic", "Manual")))

ggplot(grouped_data, aes(x = factor(cyl), y = mean_mpg, fill = transmission)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "MPG by Cylinder Count and Transmission",
    x = "Cylinders",
    y = "Mean MPG",
    fill = "Transmission"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Stacked bar plot
ggplot(grouped_data, aes(x = factor(cyl), y = mean_mpg, fill = transmission)) +
  geom_col(position = "stack") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Stacked Bar Plot") +
  theme_minimal()

# Horizontal bar plot
top_cars <- mtcars %>%
  arrange(desc(mpg)) %>%
  head(10) %>%
  mutate(car = rownames(.))

ggplot(top_cars, aes(x = reorder(car, mpg), y = mpg)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Most Fuel Efficient Cars",
    x = "",
    y = "Miles per Gallon"
  ) +
  theme_minimal()
```

## 2.5 Distributions

```{r}
# Histogram
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of MPG",
    x = "Miles per Gallon",
    y = "Frequency"
  ) +
  theme_minimal()

# Histogram with density
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 15, fill = "lightblue", color = "white") +
  geom_density(color = "darkblue", size = 1.2) +
  labs(title = "Histogram with Density Overlay") +
  theme_minimal()

# Overlapping histograms
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 20) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Overlapping Histograms") +
  theme_minimal()

# Density plot
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Density Plot by Species",
    x = "Sepal Length",
    y = "Density"
  ) +
  theme_minimal()

# Violin plot
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.2, fill = "white", alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Violin Plot with Boxplot Overlay",
    x = "Species",
    y = "Sepal Length"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Ridge plot (requires ggridges)
library(ggridges)
ggplot(iris, aes(x = Sepal.Length, y = Species, fill = Species)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Ridge Plot") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 2.6 Box Plots

```{r}
# Basic boxplot
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "MPG Distribution by Cylinder Count",
    x = "Cylinders",
    y = "Miles per Gallon"
  ) +
  theme_minimal()

# Boxplot with points
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkred") +
  labs(title = "Boxplot with Individual Points") +
  theme_minimal()

# Grouped boxplot
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(am))) +
  geom_boxplot() +
  scale_fill_manual(
    values = c("lightblue", "lightcoral"),
    labels = c("Automatic", "Manual"),
    name = "Transmission"
  ) +
  labs(title = "Grouped Boxplot") +
  theme_minimal()

# Notched boxplot
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(notch = TRUE, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Notched Boxplot") +
  theme_minimal() +
  theme(legend.position = "none")

# Horizontal boxplot
ggplot(mtcars, aes(x = mpg, y = factor(cyl))) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Horizontal Boxplot",
    x = "Miles per Gallon",
    y = "Cylinders"
  ) +
  theme_minimal()
```

## 2.7 Heatmaps

```{r}
# Correlation heatmap
cor_matrix <- cor(mtcars[, 1:7])
cor_data <- as.data.frame(as.table(cor_matrix))
names(cor_data) <- c("Var1", "Var2", "Correlation")

ggplot(cor_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = round(Correlation, 2)), size = 3) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(title = "Correlation Heatmap") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank()
  )

# Time series heatmap
# Create sample data
heatmap_data <- expand.grid(
  day = 1:31,
  hour = 0:23
) %>%
  mutate(value = sin(day/5) * cos(hour/3) + rnorm(n(), 0, 0.5))

ggplot(heatmap_data, aes(x = hour, y = day, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(
    title = "Activity Heatmap",
    x = "Hour of Day",
    y = "Day of Month"
  ) +
  theme_minimal()
```

## 2.8 Advanced Customization

```{r}
# Professional publication-ready plot
publication_plot <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl), shape = factor(am))) +
  geom_point(size = 4, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed", size = 0.8) +
  scale_color_manual(
    values = c("4" = "#E69F00", "6" = "#56B4E9", "8" = "#009E73"),
    name = "Cylinders"
  ) +
  scale_shape_manual(
    values = c("0" = 16, "1" = 17),
    labels = c("Automatic", "Manual"),
    name = "Transmission"
  ) +
  labs(
    title = "Vehicle Performance Analysis",
    subtitle = "Relationship between weight and fuel efficiency",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon (MPG)",
    caption = "Data source: Motor Trend Car Road Tests (1974)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.major = element_line(color = "gray90", size = 0.5),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, hjust = 1, color = "gray50")
  ) +
  guides(
    color = guide_legend(order = 1, nrow = 1),
    shape = guide_legend(order = 2, nrow = 1)
  )

publication_plot

# Faceted plot
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  facet_wrap(~ Species, scales = "free") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Sepal Dimensions by Species",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  )

# Multi-panel facet grid
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_grid(am ~ cyl, labeller = labeller(
    am = c("0" = "Automatic", "1" = "Manual"),
    cyl = c("4" = "4 Cyl", "6" = "6 Cyl", "8" = "8 Cyl")
  )) +
  labs(
    title = "MPG vs Weight by Transmission and Cylinders",
    x = "Weight (1000 lbs)",
    y = "MPG"
  ) +
  theme_bw()
```

## 2.9 Specialized Plots for Bioinformatics

```{r}
# Volcano plot
set.seed(123)
volcano_data <- data.frame(
  gene = paste0("Gene_", 1:1000),
  log2FC = rnorm(1000, 0, 1.5),
  pvalue = runif(1000, 0, 1)
) %>%
  mutate(
    neglog10p = -log10(pvalue),
    significant = pvalue < 0.05 & abs(log2FC) > 1,
    regulation = case_when(
      log2FC > 1 & pvalue < 0.05 ~ "Upregulated",
      log2FC < -1 & pvalue < 0.05 ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )

ggplot(volcano_data, aes(x = log2FC, y = neglog10p, color = regulation)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(
    values = c("Upregulated" = "red", "Downregulated" = "blue", "Not Significant" = "gray"),
    name = ""
  ) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  labs(
    title = "Volcano Plot",
    subtitle = "Differential Gene Expression",
    x = "Log2 Fold Change",
    y = "-Log10 P-value"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# MA plot
ma_data <- data.frame(
  A = runif(1000, 0, 15),
  M = rnorm(1000, 0, 1.5)
) %>%
  mutate(significant = abs(M) > 1)

ggplot(ma_data, aes(x = A, y = M, color = significant)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("gray", "red"), name = "Significant") +
  geom_hline(yintercept = 0, color = "blue", linetype = "solid") +
  geom_hline(yintercept = c(-1, 1), color = "blue", linetype = "dashed") +
  labs(
    title = "MA Plot",
    x = "Average Expression",
    y = "Log Fold Change"
  ) +
  theme_minimal()
```

---

# Part 3: Machine Learning with caret

```{r}
library(caret)
library(randomForest)
library(e1071)
```

## 3.1 Data Preparation

```{r}
# Load and prepare data
data(iris)
set.seed(123)

# Train-test split
trainIndex <- createDataPartition(iris$Species, p = 0.7, list = FALSE)
train_data <- iris[trainIndex, ]
test_data <- iris[-trainIndex, ]

# Check class balance
table(train_data$Species)

# Feature scaling
preproc <- preProcess(train_data[, -5], method = c("center", "scale"))
train_scaled <- predict(preproc, train_data[, -5])
test_scaled <- predict(preproc, test_data[, -5])

# Add back the target
train_scaled$Species <- train_data$Species
test_scaled$Species <- test_data$Species

# Handle missing values
data_with_na <- iris
data_with_na[sample(1:150, 10), 1] <- NA

# Impute missing values
preproc_impute <- preProcess(data_with_na[, -5],
                              method = c("medianImpute", "center", "scale"))
data_imputed <- predict(preproc_impute, data_with_na[, -5])
```

## 3.2 Model Training - Classification

### Random Forest

```{r}
# Train random forest
set.seed(123)
rf_model <- train(
  Species ~ .,
  data = train_data,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10,
    verboseIter = FALSE
  ),
  tuneGrid = expand.grid(
    mtry = c(2, 3, 4)
  )
)

# View results
print(rf_model)
plot(rf_model)

# Variable importance
varImp(rf_model)
plot(varImp(rf_model))

# Make predictions
rf_pred <- predict(rf_model, test_data)

# Confusion matrix
confusionMatrix(rf_pred, test_data$Species)
```

### Support Vector Machine

```{r}
# Train SVM
set.seed(123)
svm_model <- train(
  Species ~ .,
  data = train_data,
  method = "svmRadial",
  trControl = trainControl(
    method = "cv",
    number = 10
  ),
  tuneGrid = expand.grid(
    C = c(0.1, 1, 10),
    sigma = c(0.01, 0.1, 1)
  )
)

print(svm_model)
plot(svm_model)

# Predictions
svm_pred <- predict(svm_model, test_data)
confusionMatrix(svm_pred, test_data$Species)
```

### K-Nearest Neighbors

```{r}
# Train KNN
set.seed(123)
knn_model <- train(
  Species ~ .,
  data = train_scaled,
  method = "knn",
  trControl = trainControl(
    method = "cv",
    number = 10
  ),
  tuneGrid = expand.grid(
    k = seq(1, 20, 2)
  )
)

print(knn_model)
plot(knn_model)

# Predictions
knn_pred <- predict(knn_model, test_scaled)
confusionMatrix(knn_pred, test_scaled$Species)
```

### Gradient Boosting Machine

```{r}
# Train GBM
set.seed(123)
gbm_model <- train(
  Species ~ .,
  data = train_data,
  method = "gbm",
  trControl = trainControl(
    method = "cv",
    number = 10,
    verboseIter = FALSE
  ),
  tuneGrid = expand.grid(
    n.trees = c(50, 100, 150),
    interaction.depth = c(1, 3, 5),
    shrinkage = 0.1,
    n.minobsinnode = 10
  ),
  verbose = FALSE
)

print(gbm_model)
plot(gbm_model)
```

## 3.3 Model Training - Regression

```{r}
# Prepare regression data
data(mtcars)
set.seed(123)
trainIndex <- createDataPartition(mtcars$mpg, p = 0.7, list = FALSE)
train_reg <- mtcars[trainIndex, ]
test_reg <- mtcars[-trainIndex, ]

# Linear Regression
lm_model <- train(
  mpg ~ .,
  data = train_reg,
  method = "lm",
  trControl = trainControl(method = "cv", number = 10)
)

# Predictions
lm_pred <- predict(lm_model, test_reg)

# Performance metrics
postResample(lm_pred, test_reg$mpg)

# Random Forest Regression
rf_reg_model <- train(
  mpg ~ .,
  data = train_reg,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10
  ),
  tuneGrid = expand.grid(
    mtry = c(2, 4, 6, 8)
  )
)

print(rf_reg_model)

# Predictions
rf_reg_pred <- predict(rf_reg_model, test_reg)
postResample(rf_reg_pred, test_reg$mpg)

# Support Vector Regression
svm_reg_model <- train(
  mpg ~ .,
  data = train_reg,
  method = "svmRadial",
  trControl = trainControl(method = "cv", number = 10),
  tuneLength = 10
)

svm_reg_pred <- predict(svm_reg_model, test_reg)
postResample(svm_reg_pred, test_reg$mpg)
```

## 3.4 Advanced Cross-Validation

```{r}
# Repeated cross-validation
set.seed(123)
train_control_repeated <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 3,
  savePredictions = "final",
  classProbs = TRUE,
  summaryFunction = multiClassSummary
)

rf_repeated <- train(
  Species ~ .,
  data = train_data,
  method = "rf",
  trControl = train_control_repeated,
  metric = "Accuracy",
  tuneLength = 5
)

print(rf_repeated)

# Leave-one-out cross-validation
train_control_loocv <- trainControl(
  method = "LOOCV"
)

# Bootstrap
train_control_boot <- trainControl(
  method = "boot",
  number = 25
)

# Custom train/test splits
custom_folds <- createFolds(train_data$Species, k = 5, returnTrain = TRUE)
train_control_custom <- trainControl(
  method = "cv",
  index = custom_folds
)
```

## 3.5 Hyperparameter Tuning

```{r}
# Grid search
tune_grid <- expand.grid(
  mtry = seq(2, 4, 1),
  splitrule = c("gini", "extratrees"),
  min.node.size = c(1, 5, 10)
)

set.seed(123)
rf_tuned <- train(
  Species ~ .,
  data = train_data,
  method = "ranger",
  trControl = trainControl(method = "cv", number = 10),
  tuneGrid = tune_grid
)

print(rf_tuned)
plot(rf_tuned)

# Random search
train_control_random <- trainControl(
  method = "cv",
  number = 10,
  search = "random"
)

set.seed(123)
rf_random <- train(
  Species ~ .,
  data = train_data,
  method = "rf",
  trControl = train_control_random,
  tuneLength = 15
)

plot(rf_random)
```

## 3.6 Model Comparison

```{r}
# Train multiple models
set.seed(123)
train_control <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final",
  classProbs = TRUE
)

# Random Forest
model_rf <- train(
  Species ~ .,
  data = train_data,
  method = "rf",
  trControl = train_control
)

# SVM
model_svm <- train(
  Species ~ .,
  data = train_data,
  method = "svmRadial",
  trControl = train_control
)

# KNN
model_knn <- train(
  Species ~ .,
  data = train_scaled,
  method = "knn",
  trControl = train_control
)

# Naive Bayes
model_nb <- train(
  Species ~ .,
  data = train_data,
  method = "naive_bayes",
  trControl = train_control
)

# Compare models
model_list <- list(
  RF = model_rf,
  SVM = model_svm,
  KNN = model_knn,
  NB = model_nb
)

# Resampling comparison
resamples_result <- resamples(model_list)
summary(resamples_result)

# Visualize comparison
dotplot(resamples_result)
bwplot(resamples_result)

# Test set performance
results <- data.frame(
  Model = names(model_list),
  Accuracy = sapply(model_list, function(m) {
    pred <- predict(m, if(deparse(substitute(m)) == "model_knn") test_scaled else test_data)
    confusionMatrix(pred, test_data$Species)$overall["Accuracy"]
  })
)

print(results)
```

## 3.7 Feature Selection

```{r}
# Recursive Feature Elimination (RFE)
set.seed(123)
control_rfe <- rfeControl(
  functions = rfFuncs,
  method = "cv",
  number = 10
)

rfe_result <- rfe(
  x = train_data[, -5],
  y = train_data$Species,
  sizes = c(1, 2, 3, 4),
  rfeControl = control_rfe
)

print(rfe_result)
plot(rfe_result, type = c("g", "o"))
predictors(rfe_result)

# Feature importance across models
importance_rf <- varImp(model_rf)
importance_svm <- varImp(model_svm)

plot(importance_rf, main = "RF Feature Importance")
plot(importance_svm, main = "SVM Feature Importance")
```

## 3.8 Complete Machine Learning Workflow

```{r}
# Complete workflow with real-world considerations
set.seed(123)

# 1. Data preparation
data(iris)

# Create train/validation/test split
trainIndex <- createDataPartition(iris$Species, p = 0.6, list = FALSE)
train_full <- iris[trainIndex, ]
temp <- iris[-trainIndex, ]

valIndex <- createDataPartition(temp$Species, p = 0.5, list = FALSE)
validation <- temp[valIndex, ]
test_final <- temp[-valIndex, ]

# 2. Preprocessing
preproc_params <- preProcess(
  train_full[, -5],
  method = c("center", "scale", "nzv")
)

train_processed <- predict(preproc_params, train_full)
validation_processed <- predict(preproc_params, validation)
test_processed <- predict(preproc_params, test_final)

# 3. Define training control
train_control <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final",
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  verboseIter = TRUE
)

# 4. Train multiple models with hyperparameter tuning
models <- list()

# Random Forest
models$rf <- train(
  Species ~ .,
  data = train_processed,
  method = "rf",
  trControl = train_control,
  tuneGrid = expand.grid(mtry = c(2, 3, 4)),
  metric = "Accuracy"
)

# SVM
models$svm <- train(
  Species ~ .,
  data = train_processed,
  method = "svmRadial",
  trControl = train_control,
  tuneLength = 10,
  metric = "Accuracy"
)

# XGBoost
models$xgb <- train(
  Species ~ .,
  data = train_processed,
  method = "xgbTree",
  trControl = train_control,
  tuneLength = 5,
  metric = "Accuracy",
  verbosity = 0
)

# 5. Model selection on validation set
validation_results <- sapply(models, function(model) {
  pred <- predict(model, validation_processed)
  confusionMatrix(pred, validation_processed$Species)$overall["Accuracy"]
})

print("Validation Results:")
print(validation_results)

# Select best model
best_model_name <- names(which.max(validation_results))
best_model <- models[[best_model_name]]

# 6. Final evaluation on test set
test_pred <- predict(best_model, test_processed)
final_cm <- confusionMatrix(test_pred, test_processed$Species)

print(paste("Best Model:", best_model_name))
print(final_cm)

# 7. Feature importance
final_importance <- varImp(best_model)
plot(final_importance, main = paste("Feature Importance -", best_model_name))

# 8. Prediction probabilities
pred_probs <- predict(best_model, test_processed, type = "prob")
print(head(pred_probs))
```

---

# Part 4: Integrated Workflow Example

## Real-World Data Analysis Pipeline

```{r}
# Complete data analysis workflow combining all tools
set.seed(123)

# 1. DATA IMPORT AND PREPARATION (dplyr/tidyr)
# Simulate realistic customer purchase data
customers <- data.frame(
  customer_id = 1:500,
  age = sample(18:80, 500, replace = TRUE),
  gender = sample(c("M", "F"), 500, replace = TRUE),
  income = round(rnorm(500, 50000, 20000)),
  region = sample(c("North", "South", "East", "West"), 500, replace = TRUE)
)

purchases <- data.frame(
  purchase_id = 1:2000,
  customer_id = sample(1:500, 2000, replace = TRUE),
  product_category = sample(c("Electronics", "Clothing", "Food", "Books"), 2000, replace = TRUE),
  amount = round(rnorm(2000, 100, 50)),
  date = sample(seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day"),
                2000, replace = TRUE)
)

# 2. DATA MANIPULATION (dplyr/tidyr)
customer_features <- purchases %>%
  # Join with customer data
  left_join(customers, by = "customer_id") %>%
  # Add time features
  mutate(
    month = format(date, "%Y-%m"),
    quarter = paste0("Q", quarter(date)),
    day_of_week = weekdays(date)
  ) %>%
  # Calculate customer-level features
  group_by(customer_id) %>%
  summarise(
    n_purchases = n(),
    total_spent = sum(amount),
    avg_purchase = mean(amount),
    purchase_range = max(amount) - min(amount),
    n_categories = n_distinct(product_category),
    days_active = n_distinct(date),
    .groups = "drop"
  ) %>%
  # Join back customer demographics
  left_join(customers, by = "customer_id") %>%
  # Create target variable (high-value customer)
  mutate(
    high_value = factor(ifelse(total_spent > median(total_spent), "High", "Low")),
    age_group = cut(age, breaks = c(0, 30, 50, 100), labels = c("Young", "Middle", "Senior"))
  ) %>%
  # Select relevant features
  select(-customer_id)

# 3. EXPLORATORY VISUALIZATION (ggplot2)
# Distribution of total spend
p1 <- ggplot(customer_features, aes(x = total_spent)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = median(customer_features$total_spent),
             color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Customer Spending",
    x = "Total Spent ($)",
    y = "Number of Customers"
  ) +
  theme_minimal()

# Spending by demographics
p2 <- ggplot(customer_features, aes(x = age_group, y = total_spent, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Spending by Age Group and Gender",
    x = "Age Group",
    y = "Total Spent ($)"
  ) +
  theme_minimal()

# Feature correlations
numeric_features <- customer_features %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_features)
cor_data <- as.data.frame(as.table(cor_matrix))
names(cor_data) <- c("Var1", "Var2", "Correlation")

p3 <- ggplot(cor_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Feature Correlation Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display plots
library(patchwork)
(p1 | p2) / p3

# 4. MACHINE LEARNING (caret)
# Prepare data for modeling
ml_data <- customer_features %>%
  select(-total_spent) %>%  # Remove target-related feature
  na.omit()

# Train-test split
trainIndex <- createDataPartition(ml_data$high_value, p = 0.7, list = FALSE)
train_ml <- ml_data[trainIndex, ]
test_ml <- ml_data[-trainIndex, ]

# Define training control
train_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Train models
models <- list()

# Logistic Regression
models$glm <- train(
  high_value ~ .,
  data = train_ml,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"
)

# Random Forest
models$rf <- train(
  high_value ~ .,
  data = train_ml,
  method = "rf",
  trControl = train_control,
  tuneGrid = expand.grid(mtry = c(2, 4, 6)),
  metric = "ROC"
)

# Gradient Boosting
models$gbm <- train(
  high_value ~ .,
  data = train_ml,
  method = "gbm",
  trControl = train_control,
  verbose = FALSE,
  metric = "ROC"
)

# Compare models
resamples_result <- resamples(models)
summary(resamples_result)

# Visualize comparison
dotplot(resamples_result, metric = "ROC")

# Select best model and evaluate
best_model <- models$rf  # Based on comparison

# Test set predictions
test_pred <- predict(best_model, test_ml)
test_prob <- predict(best_model, test_ml, type = "prob")

# Confusion matrix
cm <- confusionMatrix(test_pred, test_ml$high_value)
print(cm)

# ROC curve
library(pROC)
roc_obj <- roc(test_ml$high_value, test_prob$High)
plot(roc_obj, main = "ROC Curve - Customer Segmentation Model")
auc(roc_obj)

# Feature importance
importance <- varImp(best_model)
plot(importance, main = "Feature Importance for Customer Value Prediction")

# 5. RESULTS VISUALIZATION (ggplot2)
# Create prediction results dataframe
results_df <- test_ml %>%
  mutate(
    predicted = test_pred,
    probability = test_prob$High,
    correct = (predicted == high_value)
  )

# Prediction probability distribution
ggplot(results_df, aes(x = probability, fill = high_value)) +
  geom_density(alpha = 0.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Prediction Probability Distribution",
    x = "Predicted Probability of High Value",
    y = "Density",
    fill = "Actual Class"
  ) +
  theme_minimal()

# Confusion matrix heatmap
cm_data <- as.data.frame(cm$table)
ggplot(cm_data, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Confusion Matrix",
    x = "Actual Class",
    y = "Predicted Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# 6. BUSINESS INSIGHTS (dplyr + ggplot2)
# Analyze misclassifications
misclassified <- results_df %>%
  filter(!correct) %>%
  group_by(age_group, gender, region) %>%
  summarise(
    n_errors = n(),
    avg_n_purchases = mean(n_purchases),
    avg_purchase = mean(avg_purchase),
    .groups = "drop"
  ) %>%
  arrange(desc(n_errors))

print("Top Misclassification Patterns:")
print(head(misclassified))

# Success rate by segment
segment_performance <- results_df %>%
  group_by(region, age_group) %>%
  summarise(
    n = n(),
    accuracy = mean(correct),
    avg_prob = mean(probability),
    .groups = "drop"
  )

ggplot(segment_performance, aes(x = age_group, y = accuracy, fill = region)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = scales::percent(accuracy, accuracy = 0.1)),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Model Accuracy by Customer Segment",
    x = "Age Group",
    y = "Accuracy",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Summary

This comprehensive guide covered:

## dplyr & tidyr
- All major verbs with extensive examples
- Complex real-world workflows
- Data reshaping strategies
- Advanced chaining patterns

## ggplot2
- All major plot types
- Professional customization
- Publication-ready figures
- Bioinformatics-specific plots

## caret
- Complete ML workflows
- Multiple model comparison
- Hyperparameter tuning
- Feature selection
- Model evaluation

## Integrated Workflow
- End-to-end data analysis
- Combining all three tools
- Real-world business case
- From data to insights

# Key Takeaways

1. **dplyr/tidyr**: Master these for efficient data manipulation
2. **ggplot2**: Create publication-quality visualizations
3. **caret**: Streamline machine learning workflows
4. **Integration**: Combine all three for powerful analyses

# Next Steps

- Practice with your own datasets
- Explore additional geoms and aesthetics
- Try different ML algorithms
- Build complete analysis pipelines

# References

- [dplyr documentation](https://dplyr.tidyverse.org/)
- [tidyr documentation](https://tidyr.tidyverse.org/)
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [caret documentation](https://topepo.github.io/caret/)
- [R for Data Science](https://r4ds.had.co.nz/)
