---
title: "Advanced Topics and Practical Skills"
author: "Essential Knowledge for Data Scientists"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This document covers critical topics that are essential for real-world data analysis but often overlooked:

1. **data.table** - High-performance data manipulation
2. **Regular Expressions** - Text and sequence parsing
3. **tidymodels** - Modern machine learning framework
4. **Working with Databases** - Big data that doesn't fit in memory
5. **String Processing** - Advanced text manipulation
6. **Working with APIs** - Web data and services
7. **Common Pitfalls** - What to avoid and how to debug
8. **Best Practices** - Professional workflows

---

# Part 1: data.table - High Performance Computing

## 1.1 Why data.table?

- **Speed**: 10-100x faster than dplyr for large datasets
- **Memory efficient**: Modify by reference (no copies)
- **Concise syntax**: Powerful operations in compact code
- **Big data**: Handles datasets larger than RAM

```{r}
library(data.table)
library(dplyr)
library(microbenchmark)
```

## 1.2 data.table Syntax: DT[i, j, by]

The data.table syntax is: `DT[where, select|update, group by]`

```{r}
# Create data.table
dt <- data.table(
  id = 1:1000000,
  group = sample(LETTERS[1:5], 1000000, replace = TRUE),
  value = rnorm(1000000),
  date = sample(seq(as.Date("2020-01-01"), as.Date("2023-12-31"), by = "day"),
                1000000, replace = TRUE)
)

# Basic operations
# Filter rows (i)
dt[value > 0]
dt[group %in% c("A", "B")]

# Select columns (j)
dt[, .(id, value)]
dt[, list(id, value)]  # Same as above

# Compute in j
dt[, mean(value)]
dt[, .(mean_val = mean(value), sd_val = sd(value))]

# Group by (by)
dt[, .(mean_val = mean(value)), by = group]
dt[, .(mean_val = mean(value), count = .N), by = .(group, year = year(date))]

# Combine all three
dt[value > 0, .(mean_val = mean(value), .N), by = group]
```

## 1.3 Special Symbols in data.table

```{r}
# .N - number of rows
dt[, .N]
dt[, .N, by = group]

# .SD - Subset of Data (current group)
dt[, lapply(.SD, mean), by = group, .SDcols = c("value")]

# .SDcols - specify columns for .SD
dt[, lapply(.SD, mean), by = group, .SDcols = is.numeric]

# .I - row numbers
dt[, .I[value == max(value)], by = group]

# .GRP - group counter
dt[, group_id := .GRP, by = group]

# .BY - current group values
dt[, .(group_name = .BY[[1]], mean = mean(value)), by = group]
```

## 1.4 Modify by Reference (:=)

**Key advantage**: No copies made, extremely fast and memory efficient

```{r}
# Add column
dt[, new_col := value * 2]

# Multiple columns
dt[, `:=`(
  col1 = value + 1,
  col2 = value - 1,
  col3 = value^2
)]

# Conditional update
dt[group == "A", value := value * 1.1]

# Delete column
dt[, new_col := NULL]

# Update by group
dt[, value_centered := value - mean(value), by = group]

# Functional form
cols <- c("col1", "col2")
dt[, (cols) := NULL]
```

## 1.5 Advanced Operations

```{r}
# Rolling joins
dt1 <- data.table(id = c(1, 2, 3), val = c(10, 20, 30), key = "id")
dt2 <- data.table(id = c(1.5, 2.5, 3.5), val = c(100, 200, 300))

# Roll forward
dt2[dt1, roll = TRUE]

# Roll backward
dt2[dt1, roll = -Inf]

# Overlap joins (for genomic intervals)
dt_intervals1 <- data.table(
  chr = c("chr1", "chr1", "chr2"),
  start = c(100, 200, 150),
  end = c(150, 250, 200),
  key = c("chr", "start", "end")
)

dt_intervals2 <- data.table(
  chr = c("chr1", "chr1"),
  start = c(120, 220),
  end = c(130, 230)
)

# Find overlaps
foverlaps(dt_intervals2, dt_intervals1, type = "any")

# Reshape
# melt (wide to long)
dt_wide <- data.table(
  id = 1:3,
  A = c(1, 2, 3),
  B = c(4, 5, 6),
  C = c(7, 8, 9)
)

dt_long <- melt(dt_wide, id.vars = "id",
                variable.name = "group", value.name = "value")

# dcast (long to wide)
dcast(dt_long, id ~ group, value.var = "value")

# Chain operations
result <- dt[
  value > 0
][
  , .(mean_val = mean(value), .N), by = group
][
  order(-mean_val)
][
  1:3
]
```

## 1.6 Keys and Indices

```{r}
# Set key for fast subsetting
setkey(dt, group)

# Now subsetting is blazing fast
dt["A"]
dt[c("A", "B")]

# Multiple keys
setkey(dt, group, id)
dt[.("A", 1:100)]

# Secondary indices
setindex(dt, date)

# Check keys and indices
key(dt)
indices(dt)

# Fast aggregation with keys
dt[, .(mean_val = mean(value)), keyby = group]  # Result is keyed
```

## 1.7 Performance Comparison

```{r}
# Create large datasets
n <- 1e6
df <- data.frame(
  id = 1:n,
  group = sample(LETTERS[1:10], n, replace = TRUE),
  value = rnorm(n)
)
dt <- as.data.table(df)
setkey(dt, group)

# Benchmark: Grouping and summarizing
microbenchmark(
  dplyr = df %>% group_by(group) %>% summarise(mean = mean(value)),
  data.table = dt[, .(mean = mean(value)), by = group],
  times = 10
)

# Benchmark: Filtering
microbenchmark(
  dplyr = df %>% filter(value > 0),
  data.table = dt[value > 0],
  times = 10
)

# Benchmark: Adding column
microbenchmark(
  dplyr = df %>% mutate(new_col = value * 2),
  data.table = dt[, new_col := value * 2],
  times = 10
)
```

## 1.8 Real-World Example: Gene Expression Analysis

```{r}
# Large gene expression dataset
set.seed(123)
expr_dt <- data.table(
  gene_id = paste0("Gene_", rep(1:10000, each = 100)),
  sample_id = paste0("Sample_", rep(1:100, times = 10000)),
  condition = rep(rep(c("Control", "Treatment"), each = 50), times = 10000),
  replicate = rep(rep(1:50, times = 2), times = 10000),
  expression = rnorm(1000000, mean = 100, sd = 20)
)

# Complex analysis with data.table
diff_expr <- expr_dt[
  # Calculate mean expression by gene and condition
  , .(mean_expr = mean(expression)), by = .(gene_id, condition)
][
  # Reshape to wide format
  dcast(.SD, gene_id ~ condition, value.var = "mean_expr")
][
  # Calculate fold change
  , `:=`(
    fold_change = Treatment / Control,
    log2fc = log2(Treatment / Control)
  )
][
  # Filter significant changes
  abs(log2fc) > 1
][
  # Sort by absolute fold change
  order(-abs(log2fc))
][
  # Add rank
  , rank := .I
][
  # Classify
  , regulation := fcase(
    log2fc > 1, "Up",
    log2fc < -1, "Down",
    default = "No change"
  )
]

# Summary by regulation
diff_expr[, .N, by = regulation]
```

---

# Part 2: Regular Expressions

## 2.1 Why Regular Expressions?

Essential for:
- Parsing FASTA headers
- Extracting gene IDs
- Cleaning messy text data
- Validating formats
- Pattern matching in sequences

```{r}
library(stringr)
```

## 2.2 Basic Patterns

```{r}
# Literal matching
str_detect("ATCGATCG", "ATC")

# Character classes
str_extract_all("Sample_A1_B2", "[A-Z]")  # All uppercase letters
str_extract_all("Sample_A1_B2", "[0-9]")  # All digits
str_extract_all("Sample_A1_B2", "[A-Za-z0-9]")  # Alphanumeric

# Predefined classes
# \\d = [0-9] (digit)
# \\D = [^0-9] (non-digit)
# \\w = [A-Za-z0-9_] (word character)
# \\W = [^A-Za-z0-9_] (non-word)
# \\s = whitespace
# \\S = non-whitespace

str_extract_all("Sample_123", "\\d+")
str_extract_all("Hello World", "\\w+")

# Anchors
str_detect("ATCG", "^AT")   # Starts with AT
str_detect("ATCG", "CG$")   # Ends with CG
str_detect("ATCG", "^ATCG$")  # Exact match

# Quantifiers
# * = 0 or more
# + = 1 or more
# ? = 0 or 1
# {n} = exactly n
# {n,} = n or more
# {n,m} = between n and m

str_extract("AAATTT", "A+")     # One or more A
str_extract("AAATTT", "A{3}")   # Exactly 3 A's
str_extract("AAATTT", "A{2,}")  # 2 or more A's
```

## 2.3 Bioinformatics Examples

```{r}
# Parse FASTA headers
fasta_headers <- c(
  ">gi|123456|ref|NM_001234.5| Homo sapiens gene [Homo sapiens]",
  ">ENSG00000139618.13 BRCA2 gene [Homo sapiens]",
  ">sp|P04637|P53_HUMAN Tumor suppressor p53"
)

# Extract RefSeq ID
str_extract(fasta_headers, "NM_\\d+\\.\\d+")

# Extract Ensembl ID
str_extract(fasta_headers, "ENSG\\d+\\.\\d+")

# Extract gene symbol
str_extract(fasta_headers, "[A-Z0-9]+(?= gene)")

# Parse complex IDs
sample_ids <- c(
  "Patient_001_Tumor_RNA_Rep1",
  "Patient_002_Normal_DNA_Rep2",
  "Patient_003_Tumor_RNA_Rep3"
)

# Extract all components
data.frame(
  sample = sample_ids,
  patient = str_extract(sample_ids, "Patient_\\d+"),
  tissue = str_extract(sample_ids, "(Tumor|Normal)"),
  type = str_extract(sample_ids, "(RNA|DNA)"),
  replicate = str_extract(sample_ids, "Rep\\d+")
)

# Validate formats
# Check if sequence is valid DNA
is_valid_dna <- function(seq) {
  str_detect(seq, "^[ATCGN]+$")
}

is_valid_dna("ATCGATCG")  # TRUE
is_valid_dna("ATCGATCX")  # FALSE

# Validate gene ID format
is_ensembl <- function(id) {
  str_detect(id, "^ENS[A-Z]{0,3}G\\d{11}(\\.\\d+)?$")
}

is_ensembl("ENSG00000139618")      # TRUE
is_ensembl("ENSG00000139618.13")   # TRUE
is_ensembl("NM_001234")             # FALSE
```

## 2.4 Advanced Patterns

```{r}
# Groups and capture
gene_info <- "BRCA1 (chr17:43044295-43125483)"

# Extract with capture groups
str_match(gene_info, "([A-Z0-9]+) \\(chr(\\d+):(\\d+)-(\\d+)\\)")

# Named groups (using stringr 1.5.0+)
pattern <- "(?<gene>[A-Z0-9]+) \\(chr(?<chr>\\d+):(?<start>\\d+)-(?<end>\\d+)\\)"

# Lookahead and lookbehind
# Positive lookahead: (?=...)
# Negative lookahead: (?!...)
# Positive lookbehind: (?<=...)
# Negative lookbehind: (?<!...)

# Extract number that comes before 'bp'
str_extract("Length: 150bp", "\\d+(?=bp)")

# Extract gene name that comes after 'gene='
str_extract("feature gene=BRCA1", "(?<=gene=)[A-Z0-9]+")

# Replace using patterns
vcf_info <- "DP=50;AF=0.25;AC=2"
str_replace_all(vcf_info, "(\\w+)=(\\d+\\.?\\d*)", "\\1: \\2")

# Split complex strings
str_split("gene1,gene2,gene3;gene4|gene5", "[,;|]")
```

## 2.5 Practical Examples

```{r}
# Clean messy gene symbols
messy_genes <- c("  BRCA1  ", "tp53", "EGFR*", "kras-v1", "TP53 ")

clean_genes <- messy_genes %>%
  str_trim() %>%                    # Remove whitespace
  str_to_upper() %>%                # Uppercase
  str_remove_all("[^A-Z0-9]") %>%  # Remove special chars
  unique()

# Parse VCF INFO field
vcf_info <- "DP=100;AF=0.45;AC=2;AN=4;BaseQRankSum=0.25"

parse_vcf_info <- function(info) {
  parts <- str_split(info, ";")[[1]]
  keys <- str_extract(parts, "^[^=]+")
  values <- str_extract(parts, "(?<==).+$")
  setNames(as.list(values), keys)
}

parse_vcf_info(vcf_info)

# Extract coordinates from multiple formats
coords <- c(
  "chr1:12345-67890",
  "1:12345-67890",
  "chr1 12345 67890",
  "chromosome1:12345-67890"
)

extract_coords <- function(x) {
  data.frame(
    chr = str_extract(x, "(?:chr|chromosome)?(\\d+|[XY])", group = 1),
    start = str_extract(x, "(?::| )(\\d+)", group = 1),
    end = str_extract(x, "[-: ](\\d+)$", group = 1)
  )
}

extract_coords(coords)

# Find motifs in sequences
sequence <- "ATCGATCGATCGTAGCTAGCTAGCTAG"

# Find all ATG (start codons)
str_locate_all(sequence, "ATG")

# Find restriction sites (e.g., EcoRI: GAATTC)
str_locate_all(sequence, "GAATTC")

# Find pattern with wildcards (N = any nucleotide)
# Using . for any character
str_locate_all(sequence, "AT.G")
```

---

# Part 3: tidymodels - Modern ML Framework

## 3.1 Why tidymodels?

- **Unified interface**: Consistent API across models
- **Modular**: Separate packages for each step
- **Tidy**: Integrates with tidyverse
- **Modern**: Successor to caret
- **Flexible**: Easy to customize workflows

```{r}
library(tidymodels)
```

## 3.2 Core Components

```{r}
# rsample: Data splitting and resampling
# recipes: Preprocessing and feature engineering
# parsnip: Model specification
# workflows: Combine preprocessing + model
# tune: Hyperparameter tuning
# yardstick: Model evaluation

# Load data
data(mtcars)
```

## 3.3 Basic Workflow

```{r}
# 1. Split data
set.seed(123)
data_split <- initial_split(mtcars, prop = 0.7, strata = am)
train_data <- training(data_split)
test_data <- testing(data_split)

# 2. Create recipe (preprocessing)
rec <- recipe(mpg ~ ., data = train_data) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

# 3. Specify model
rf_spec <- rand_forest(trees = 100) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# 4. Create workflow
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_spec)

# 5. Fit model
fit <- wf %>% fit(train_data)

# 6. Make predictions
predictions <- predict(fit, test_data)

# 7. Evaluate
metrics <- metric_set(rmse, rsq, mae)
bind_cols(test_data, predictions) %>%
  metrics(truth = mpg, estimate = .pred)
```

## 3.4 Advanced Preprocessing

```{r}
# Complex recipe with many steps
advanced_rec <- recipe(mpg ~ ., data = train_data) %>%
  # Remove zero variance
  step_zv(all_predictors()) %>%
  # Remove highly correlated
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  # Impute missing
  step_impute_median(all_numeric_predictors()) %>%
  # Create interactions
  step_interact(~ hp:wt) %>%
  # Polynomial features
  step_poly(wt, degree = 2) %>%
  # Normalize
  step_normalize(all_numeric_predictors()) %>%
  # One-hot encode
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # PCA
  step_pca(all_numeric_predictors(), num_comp = 5)

# Prepare recipe
prepped_rec <- prep(advanced_rec, training = train_data)

# Apply to data
train_processed <- bake(prepped_rec, new_data = NULL)
test_processed <- bake(prepped_rec, new_data = test_data)
```

## 3.5 Cross-Validation and Resampling

```{r}
# K-fold cross-validation
set.seed(123)
folds <- vfold_cv(train_data, v = 10, strata = am)

# Fit resamples
fit_resamples <- wf %>%
  fit_resamples(
    resamples = folds,
    metrics = metric_set(rmse, rsq, mae)
  )

# View results
collect_metrics(fit_resamples)

# Bootstrap
boot_samples <- bootstraps(train_data, times = 25)

# Monte Carlo CV
mc_samples <- mc_cv(train_data, prop = 0.8, times = 20)

# Time series CV
# Assuming time_series_data has a date column
# time_slices <- rolling_origin(time_series_data, initial = 100, assess = 20, cumulative = FALSE)
```

## 3.6 Hyperparameter Tuning

```{r}
# Specify model with tunable parameters
rf_tune_spec <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Create workflow
tune_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_tune_spec)

# Grid search
rf_grid <- grid_regular(
  trees(range = c(50, 200)),
  mtry(range = c(2, 8)),
  min_n(range = c(2, 10)),
  levels = 3
)

# Tune
set.seed(123)
tune_results <- tune_wf %>%
  tune_grid(
    resamples = folds,
    grid = rf_grid,
    metrics = metric_set(rmse, rsq)
  )

# View results
collect_metrics(tune_results)
autoplot(tune_results)

# Select best
best_params <- select_best(tune_results, metric = "rmse")

# Finalize workflow
final_wf <- tune_wf %>%
  finalize_workflow(best_params)

# Fit on full training data
final_fit <- final_wf %>% fit(train_data)

# Evaluate on test
test_results <- final_fit %>%
  predict(test_data) %>%
  bind_cols(test_data) %>%
  metrics(truth = mpg, estimate = .pred)
```

## 3.7 Model Comparison

```{r}
# Define multiple models
models <- list(
  rf = rand_forest(trees = 100) %>%
    set_engine("ranger") %>%
    set_mode("regression"),

  xgb = boost_tree(trees = 100) %>%
    set_engine("xgboost") %>%
    set_mode("regression"),

  svm = svm_rbf() %>%
    set_engine("kernlab") %>%
    set_mode("regression")
)

# Fit all models
results <- map_dfr(names(models), function(model_name) {
  wf <- workflow() %>%
    add_recipe(rec) %>%
    add_model(models[[model_name]])

  wf %>%
    fit_resamples(folds) %>%
    collect_metrics() %>%
    mutate(model = model_name)
})

# Compare
results %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = model, y = mean)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.2) +
  labs(title = "Model Comparison", y = "RMSE") +
  theme_minimal()
```

## 3.8 Feature Engineering for Bioinformatics

```{r}
# Gene expression data example
gene_rec <- recipe(outcome ~ ., data = gene_data) %>%
  # Log transform expression values
  step_log(starts_with("gene_"), offset = 1) %>%
  # Normalize
  step_normalize(all_numeric_predictors()) %>%
  # Remove low variance genes
  step_zv(all_predictors()) %>%
  # Feature selection based on correlation
  step_corr(all_numeric_predictors(), threshold = 0.95) %>%
  # PCA for dimensionality reduction
  step_pca(all_numeric_predictors(), num_comp = 50) %>%
  # Or use alternative: sparse PCA
  # step_ica(all_numeric_predictors(), num_comp = 20) %>%
  # Handle batch effects
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())
```

---

# Part 4: Working with Databases

## 4.1 Why Databases?

- Data too large for memory
- Central data storage
- Query only what you need
- Concurrent access
- Data integrity

```{r}
library(DBI)
library(RSQLite)
library(dplyr)
library(dbplyr)
```

## 4.2 SQLite (Local Database)

```{r}
# Create/connect to database
con <- dbConnect(SQLite(), "mydata.db")

# Write data to database
dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)
dbWriteTable(con, "iris", iris, overwrite = TRUE)

# List tables
dbListTables(con)

# Read entire table
dbReadTable(con, "mtcars")

# Execute SQL query
result <- dbGetQuery(con, "
  SELECT cyl, COUNT(*) as count, AVG(mpg) as avg_mpg
  FROM mtcars
  GROUP BY cyl
")

# Parameterized queries (prevent SQL injection)
cyl_value <- 6
result <- dbGetQuery(con, "
  SELECT * FROM mtcars WHERE cyl = ?
", params = list(cyl_value))

# Write query results
dbExecute(con, "
  CREATE TABLE high_mpg AS
  SELECT * FROM mtcars WHERE mpg > 25
")

# Disconnect
dbDisconnect(con)
```

## 4.3 dplyr with Databases (dbplyr)

```{r}
# Connect
con <- dbConnect(SQLite(), "mydata.db")

# Create table reference
mtcars_db <- tbl(con, "mtcars")

# Use dplyr syntax (lazy evaluation)
query <- mtcars_db %>%
  filter(cyl == 6) %>%
  select(mpg, wt, hp) %>%
  mutate(efficiency = mpg / wt) %>%
  arrange(desc(efficiency))

# View SQL query
show_query(query)

# Execute and bring to R
result <- collect(query)

# Aggregation
summary <- mtcars_db %>%
  group_by(cyl) %>%
  summarise(
    n = n(),
    mean_mpg = mean(mpg, na.rm = TRUE),
    sd_mpg = sd(mpg, na.rm = TRUE)
  ) %>%
  collect()

# Joins
iris_db <- tbl(con, "iris")

# Join tables (if they have common column)
# joined <- mtcars_db %>%
#   left_join(other_table, by = "id")

dbDisconnect(con)
```

## 4.4 PostgreSQL (Production Database)

```{r}
library(RPostgres)

# Connect to PostgreSQL
con <- dbConnect(
  Postgres(),
  host = "localhost",
  port = 5432,
  dbname = "mydb",
  user = "username",
  password = "password"
)

# Create table with appropriate types
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS experiments (
    exp_id SERIAL PRIMARY KEY,
    gene_id VARCHAR(50),
    sample_id VARCHAR(50),
    expression DOUBLE PRECISION,
    pvalue DOUBLE PRECISION,
    timestamp TIMESTAMP DEFAULT CURRENT_TIMESTAMP
  )
")

# Bulk insert (efficient)
experiments <- data.frame(
  gene_id = paste0("GENE", 1:1000),
  sample_id = paste0("SAMPLE", rep(1:10, each = 100)),
  expression = rnorm(1000, 100, 20),
  pvalue = runif(1000)
)

dbWriteTable(con, "experiments", experiments, append = TRUE)

# Create index for faster queries
dbExecute(con, "CREATE INDEX idx_gene ON experiments(gene_id)")

# Complex query
result <- dbGetQuery(con, "
  SELECT
    gene_id,
    AVG(expression) as mean_expr,
    STDDEV(expression) as sd_expr,
    MIN(pvalue) as min_pval
  FROM experiments
  WHERE expression > 50
  GROUP BY gene_id
  HAVING MIN(pvalue) < 0.05
  ORDER BY mean_expr DESC
  LIMIT 100
")

dbDisconnect(con)
```

## 4.5 Working with Large Datasets

```{r}
# Read in chunks
con <- dbConnect(SQLite(), "large_data.db")

# Process in batches
process_in_batches <- function(table_name, batch_size = 10000) {
  # Get total rows
  total <- dbGetQuery(con, paste("SELECT COUNT(*) FROM", table_name))[[1]]

  results <- list()
  for (i in seq(1, total, batch_size)) {
    # Read batch
    batch <- dbGetQuery(con, paste(
      "SELECT * FROM", table_name,
      "LIMIT", batch_size, "OFFSET", i - 1
    ))

    # Process batch
    results[[length(results) + 1]] <- process_batch(batch)
  }

  bind_rows(results)
}

# Write in chunks
write_large_dataset <- function(data, table_name, chunk_size = 10000) {
  # Split into chunks
  n_chunks <- ceiling(nrow(data) / chunk_size)

  for (i in 1:n_chunks) {
    start_row <- (i - 1) * chunk_size + 1
    end_row <- min(i * chunk_size, nrow(data))
    chunk <- data[start_row:end_row, ]

    dbWriteTable(con, table_name, chunk, append = (i > 1))
  }
}
```

---

# Part 5: Common Pitfalls and Debugging

## 5.1 Common Mistakes

```{r}
# 1. Factor vs Character
df <- data.frame(gene = c("BRCA1", "TP53"), stringsAsFactors = TRUE)
df$gene[1] <- "EGFR"  # Warning! May not work as expected
# Fix: Use stringsAsFactors = FALSE or tibbles

# 2. Floating point comparison
0.1 + 0.2 == 0.3  # FALSE!
# Fix: Use all.equal() or near()
all.equal(0.1 + 0.2, 0.3)
dplyr::near(0.1 + 0.2, 0.3)

# 3. Growing objects in loops
slow <- function(n) {
  result <- c()
  for (i in 1:n) {
    result <- c(result, i^2)  # BAD: Creates copy each time
  }
}
# Fix: Preallocate
fast <- function(n) {
  result <- numeric(n)
  for (i in 1:n) {
    result[i] <- i^2
  }
}

# 4. Forgetting to return in functions
bad_function <- function(x) {
  result <- x * 2
  # Forgot to return!
}
# Fix:
good_function <- function(x) {
  result <- x * 2
  return(result)  # Or just: result
}

# 5. Using = instead of == in conditions
if (x = 5) { }  # Error! Assignment, not comparison
if (x == 5) { }  # Correct

# 6. Not handling NAs
mean(c(1, 2, NA, 4))  # Returns NA
mean(c(1, 2, NA, 4), na.rm = TRUE)  # Returns mean

# 7. Incorrect subsetting
df[1]     # Returns data frame with first column
df[[1]]   # Returns vector of first column
df[, 1]   # Returns vector (or data frame if drop=FALSE)

# 8. Mixing up & and &&
c(TRUE, FALSE) & c(TRUE, TRUE)   # Vectorized AND
TRUE && FALSE                     # Scalar AND (short-circuit)

# 9. Forgetting quotes in aes()
ggplot(mtcars, aes(x = "mpg"))  # Wrong! Treats as string
ggplot(mtcars, aes(x = mpg))    # Correct

# 10. Not checking data types
str(data)  # Always check structure first!
```

## 5.2 Debugging Techniques

```{r}
# 1. Print debugging
my_function <- function(x) {
  print(paste("Input:", x))
  result <- x * 2
  print(paste("Result:", result))
  return(result)
}

# 2. browser() - Interactive debugging
debug_function <- function(x) {
  y <- x * 2
  browser()  # Execution stops here
  z <- y + 1
  return(z)
}

# 3. debug() - Step through function
debug(my_function)
my_function(5)
undebug(my_function)

# 4. trace() - Add code at specific points
trace(mean, quote(print(x)))
mean(1:10)
untrace(mean)

# 5. traceback() - See call stack after error
error_function <- function() {
  stop("Error!")
}
# After error: traceback()

# 6. options(error = recover) - Interactive debugging on error
options(error = recover)
# Run code that errors
# options(error = NULL)  # Reset

# 7. tryCatch() - Handle errors gracefully
safe_function <- function(x) {
  tryCatch(
    {
      result <- log(x)
      return(result)
    },
    error = function(e) {
      message("Error: ", e$message)
      return(NA)
    },
    warning = function(w) {
      message("Warning: ", w$message)
    }
  )
}

# 8. Assertions
library(assertthat)
validate_input <- function(x) {
  assert_that(is.numeric(x))
  assert_that(all(x >= 0))
  assert_that(length(x) > 0)
}

# 9. Unit tests
library(testthat)
test_that("Function works correctly", {
  expect_equal(my_function(2), 4)
  expect_error(my_function("text"))
})

# 10. Profiling
library(profvis)
profvis({
  # Code to profile
  x <- rnorm(1e6)
  y <- x^2
  z <- sum(y)
})
```

## 5.3 Memory Issues

```{r}
# Check object size
object.size(large_object)
format(object.size(large_object), units = "MB")

# List all objects and sizes
sapply(ls(), function(x) {
  format(object.size(get(x)), units = "MB")
})

# Remove objects
rm(large_object)
gc()  # Garbage collection

# Use data.table for large data
# Use databases for very large data
# Process in chunks
# Use sampling for development

# Memory-efficient alternatives
# Instead of: rbind in loop
result <- lapply(files, read.csv) %>% bind_rows()

# Monitor memory
# library(pryr)
# mem_used()
# mem_change(x <- rnorm(1e6))
```

---

# Part 6: Best Practices

## 6.1 Project Organization

```
project/
├── README.md              # Project description
├── project.Rproj         # R project file
├── renv/                 # Package management
├── data/
│   ├── raw/             # Original, immutable data
│   └── processed/       # Cleaned data
├── R/                    # Function definitions
│   ├── utils.R
│   ├── analysis.R
│   └── visualization.R
├── scripts/              # Analysis scripts
│   ├── 01_data_prep.R
│   ├── 02_analysis.R
│   └── 03_visualization.R
├── notebooks/            # R Markdown analyses
├── output/
│   ├── figures/
│   ├── tables/
│   └── reports/
├── tests/                # Unit tests
└── docs/                 # Documentation
```

## 6.2 Code Style

```{r}
# Use meaningful names
# Bad:
x <- function(y) { y * 2 }

# Good:
double_value <- function(input_value) {
  result <- input_value * 2
  return(result)
}

# Use consistent naming convention
# snake_case (recommended for R)
calculate_mean_expression <- function() {}

# Use spacing
# Bad:
result<-x+y*2

# Good:
result <- x + y * 2

# Limit line length (80 characters)
# Use styler package
library(styler)
style_file("script.R")

# Use lintr for code quality
library(lintr)
lint("script.R")

# Comment your code
# Explain WHY, not WHAT
# Bad:
x <- x + 1  # Add 1 to x

# Good:
x <- x + 1  # Adjust for 1-based indexing
```

## 6.3 Function Design

```{r}
# Good function design
calculate_statistics <- function(data,
                                  columns = NULL,
                                  na.rm = TRUE,
                                  verbose = FALSE) {
  # Validate inputs
  stopifnot(is.data.frame(data))

  if (verbose) message("Calculating statistics...")

  # Select columns
  if (is.null(columns)) {
    columns <- names(data)[sapply(data, is.numeric)]
  }

  # Calculate
  results <- data %>%
    select(all_of(columns)) %>%
    summarise(across(everything(),
                     list(mean = ~mean(., na.rm = na.rm),
                          sd = ~sd(., na.rm = na.rm)),
                     .names = "{.col}_{.fn}"))

  return(results)
}

# Document functions
#' Calculate Summary Statistics
#'
#' @param data A data frame
#' @param columns Character vector of column names
#' @param na.rm Logical, remove NA values?
#' @param verbose Logical, print messages?
#' @return A data frame with summary statistics
#' @examples
#' calculate_statistics(mtcars, c("mpg", "hp"))
```

## 6.4 Error Handling

```{r}
# Defensive programming
process_data <- function(file_path) {
  # Check file exists
  if (!file.exists(file_path)) {
    stop("File not found: ", file_path)
  }

  # Try reading file
  data <- tryCatch(
    read.csv(file_path),
    error = function(e) {
      stop("Error reading file: ", e$message)
    }
  )

  # Validate data
  if (nrow(data) == 0) {
    warning("Empty dataset")
    return(NULL)
  }

  # Process...
  return(data)
}
```

## 6.5 Testing

```{r}
library(testthat)

# Write tests
test_that("calculate_mean works correctly", {
  # Test normal case
  expect_equal(calculate_mean(c(1, 2, 3)), 2)

  # Test edge cases
  expect_equal(calculate_mean(c(5)), 5)
  expect_true(is.na(calculate_mean(c())))

  # Test error handling
  expect_error(calculate_mean("text"))
  expect_error(calculate_mean(NULL))
})

# Test with NA
test_that("handles NA values", {
  expect_true(is.na(calculate_mean(c(1, NA, 3))))
  expect_equal(calculate_mean(c(1, NA, 3), na.rm = TRUE), 2)
})
```

## 6.6 Documentation

```{r}
# README.md template
# # Project Title
#
# ## Description
# Brief description of the project
#
# ## Installation
# ```r
# # Install required packages
# install.packages(c("dplyr", "ggplot2"))
# ```
#
# ## Usage
# ```r
# # Example usage
# source("R/analysis.R")
# results <- run_analysis("data/input.csv")
# ```
#
# ## Data
# Description of data sources and formats
#
# ## Results
# Summary of key findings
#
# ## Authors
# Your name
#
# ## License
# License information
```

---

# Summary

This guide covered critical advanced topics:

## 1. data.table
- High-performance alternative to dplyr
- Syntax: DT[i, j, by]
- Modify by reference
- 10-100x faster for large datasets

## 2. Regular Expressions
- Pattern matching and text parsing
- Essential for bioinformatics
- Parsing FASTA, VCF, gene IDs
- Validating formats

## 3. tidymodels
- Modern ML framework
- Modular and flexible
- Complete workflows
- Hyperparameter tuning

## 4. Databases
- SQLite for local data
- PostgreSQL for production
- dplyr interface (dbplyr)
- Handle data larger than memory

## 5. Common Pitfalls
- Floating point comparison
- Growing objects in loops
- NA handling
- Factor vs character
- Debugging techniques

## 6. Best Practices
- Project organization
- Code style
- Function design
- Error handling
- Testing
- Documentation

# Key Takeaways

1. **Choose the right tool**: data.table for big data, dplyr for readability
2. **Master regex**: Essential for text/sequence processing
3. **Use databases**: For data that doesn't fit in memory
4. **Write defensive code**: Validate inputs, handle errors
5. **Test everything**: Unit tests prevent bugs
6. **Document well**: Future you will thank you

# Next Steps

- Practice with real datasets
- Build complete pipelines
- Contribute to open source
- Share your knowledge

# References

- [data.table documentation](https://rdatatable.gitlab.io/data.table/)
- [Regular Expressions in R](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [tidymodels](https://www.tidymodels.org/)
- [DBI documentation](https://dbi.r-dbi.org/)
- [R Packages](https://r-pkgs.org/)
