---
title: "Complete Bioinformatics Pipelines"
author: "End-to-End Workflows for NGS Data Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This document provides complete, production-ready bioinformatics pipelines:

1. **Bulk RNA-seq**: FASTQ → Counts → Differential Expression → Visualization
2. **Single-Cell RNA-seq**: 10X data → QC → Clustering → Marker genes
3. **ChIP-seq**: Alignment → Peak calling → Annotation → Visualization
4. **Variant Calling**: VCF processing → Annotation → Filtering
5. **Microbiome Analysis**: 16S/Shotgun → Diversity → Differential abundance

---

# Part 1: Bulk RNA-seq Pipeline

## 1.1 Overview

**Goal**: Identify differentially expressed genes between conditions

**Steps**:
1. Quality control (FASTQ)
2. Alignment (STAR/HISAT2)
3. Quantification (featureCounts/RSEM)
4. Quality assessment
5. Differential expression (DESeq2)
6. Functional enrichment
7. Visualization

```{r}
library(DESeq2)
library(tximport)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(clusterProfiler)
library(EnhancedVolcano)
library(pheatmap)
library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
```

## 1.2 Read Quantification

```{r}
# Option 1: Import Salmon/Kallisto quantification
# List sample files
samples <- data.frame(
  sample_id = paste0("Sample_", 1:6),
  condition = rep(c("Control", "Treatment"), each = 3),
  files = paste0("quant/", paste0("Sample_", 1:6), "/quant.sf")
)

# Load transcript-to-gene mapping
tx2gene <- read.csv("annotation/tx2gene.csv")

# Import with tximport
txi <- tximport(
  files = samples$files,
  type = "salmon",
  tx2gene = tx2gene,
  ignoreTxVersion = TRUE
)

# Option 2: Import featureCounts output
counts <- read.table("counts/featureCounts.txt",
                     header = TRUE,
                     row.names = 1,
                     skip = 1)

# Remove annotation columns, keep only counts
counts <- counts[, 6:ncol(counts)]
colnames(counts) <- samples$sample_id
```

## 1.3 Create DESeq2 Object

```{r}
# Sample metadata
coldata <- data.frame(
  row.names = samples$sample_id,
  condition = factor(samples$condition, levels = c("Control", "Treatment")),
  batch = factor(rep(1:2, 3))  # If batch effects present
)

# Create DESeq2 object from tximport
dds <- DESeqDataSetFromTximport(
  txi,
  colData = coldata,
  design = ~ condition
)

# Or from count matrix
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = coldata,
  design = ~ condition
)

# Pre-filtering: remove low count genes
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# View
dds
```

## 1.4 Quality Control

```{r}
# Variance stabilizing transformation
vsd <- vst(dds, blind = FALSE)

# PCA plot
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text(aes(label = name), hjust = 0, nudge_x = 1) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("PCA - RNA-seq Samples")

# Sample distances heatmap
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         main = "Sample-to-Sample Distances")

# Count distribution
library(reshape2)
count_melt <- melt(log10(counts(dds) + 1))
colnames(count_melt) <- c("Gene", "Sample", "Count")

ggplot(count_melt, aes(x = Sample, y = Count, fill = Sample)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Distribution of Log10 Counts")
```

## 1.5 Differential Expression Analysis

```{r}
# Run DESeq2
dds <- DESeq(dds)

# Get results
res <- results(dds, contrast = c("condition", "Treatment", "Control"))

# Add gene symbols
res$symbol <- mapIds(org.Hs.eg.db,
                     keys = rownames(res),
                     column = "SYMBOL",
                     keytype = "ENSEMBL",
                     multiVals = "first")

# Summary
summary(res)

# Number of significant genes
sum(res$padj < 0.05, na.rm = TRUE)
sum(res$padj < 0.05 & abs(res$log2FoldChange) > 1, na.rm = TRUE)

# Order by adjusted p-value
res_ordered <- res[order(res$padj), ]

# Top 10 upregulated genes
head(res_ordered[res_ordered$log2FoldChange > 0, ], 10)

# Top 10 downregulated genes
head(res_ordered[res_ordered$log2FoldChange < 0, ], 10)

# Export results
write.csv(as.data.frame(res_ordered),
          file = "results/DESeq2_results.csv")

# MA plot
plotMA(res, ylim = c(-5, 5))

# Dispersion plot
plotDispEsts(dds)
```

## 1.6 Visualization

### Volcano Plot

```{r}
library(EnhancedVolcano)

EnhancedVolcano(res,
                lab = res$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Treatment vs Control',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 4.0,
                colAlpha = 0.5,
                legendPosition = 'right',
                legendLabSize = 12,
                legendIconSize = 4.0)
```

### Heatmap of Top Genes

```{r}
# Select top 50 significant genes
top_genes <- head(order(res$padj), 50)

# Get normalized counts
mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation
annotation_col <- data.frame(
  Condition = coldata$condition,
  row.names = rownames(coldata)
)

# Plot with pheatmap
pheatmap(mat,
         annotation_col = annotation_col,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_row = 8,
         main = "Top 50 Differentially Expressed Genes")

# Advanced heatmap with ComplexHeatmap
library(ComplexHeatmap)
library(circlize)

col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

ha <- HeatmapAnnotation(
  Condition = coldata$condition,
  col = list(Condition = c("Control" = "lightblue", "Treatment" = "pink"))
)

Heatmap(mat,
        name = "Z-score",
        col = col_fun,
        top_annotation = ha,
        show_row_names = TRUE,
        show_column_names = TRUE,
        row_names_gp = gpar(fontsize = 8),
        column_title = "Top 50 DE Genes")
```

### Individual Gene Plots

```{r}
# Plot counts for specific gene
plotCounts(dds, gene = "ENSG00000139618", intgroup = "condition")

# Enhanced version
d <- plotCounts(dds, gene = "ENSG00000139618",
                intgroup = "condition", returnData = TRUE)

ggplot(d, aes(x = condition, y = count, color = condition)) +
  geom_point(position = position_jitter(w = 0.1, h = 0), size = 3) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  scale_y_log10() +
  theme_minimal() +
  ggtitle("BRCA2 Expression")
```

## 1.7 Functional Enrichment Analysis

```{r}
library(clusterProfiler)

# Get significant genes
sig_genes <- res[res$padj < 0.05 & !is.na(res$padj), ]
sig_genes <- sig_genes[order(sig_genes$log2FoldChange, decreasing = TRUE), ]

# Upregulated genes
up_genes <- rownames(sig_genes[sig_genes$log2FoldChange > 1, ])

# Downregulated genes
down_genes <- rownames(sig_genes[sig_genes$log2FoldChange < -1, ])

# GO enrichment - Biological Process
ego_up <- enrichGO(gene = up_genes,
                   OrgDb = org.Hs.eg.db,
                   keyType = "ENSEMBL",
                   ont = "BP",
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.05)

# Visualize
dotplot(ego_up, showCategory = 20)
barplot(ego_up, showCategory = 20)

# KEGG pathway enrichment
# Convert to Entrez IDs
entrez_up <- mapIds(org.Hs.eg.db,
                    keys = up_genes,
                    column = "ENTREZID",
                    keytype = "ENSEMBL",
                    multiVals = "first")
entrez_up <- na.omit(entrez_up)

kegg_up <- enrichKEGG(gene = entrez_up,
                      organism = "hsa",
                      pvalueCutoff = 0.05)

dotplot(kegg_up, showCategory = 15)

# GSEA (Gene Set Enrichment Analysis)
# Create ranked gene list
gene_list <- res$log2FoldChange
names(gene_list) <- rownames(res)
gene_list <- na.omit(gene_list)
gene_list <- sort(gene_list, decreasing = TRUE)

# Convert to Entrez IDs
gene_list_entrez <- mapIds(org.Hs.eg.db,
                           keys = names(gene_list),
                           column = "ENTREZID",
                           keytype = "ENSEMBL",
                           multiVals = "first")

gene_list_entrez <- gene_list[!is.na(gene_list_entrez)]
names(gene_list_entrez) <- gene_list_entrez[!is.na(gene_list_entrez)]

# Run GSEA
gsea_result <- gseKEGG(geneList = gene_list_entrez,
                       organism = "hsa",
                       pvalueCutoff = 0.05)

# Visualize enrichment
gseaplot2(gsea_result, geneSetID = 1:3, pvalue_table = TRUE)

# Enrichment map
emapplot(ego_up, showCategory = 30)
```

## 1.8 Complete Analysis Report

```{r}
# Generate comprehensive report
generate_rnaseq_report <- function(dds, res, output_dir = "results") {
  dir.create(output_dir, showWarnings = FALSE)

  # 1. Sample information
  write.csv(colData(dds), file.path(output_dir, "sample_info.csv"))

  # 2. All results
  write.csv(as.data.frame(res), file.path(output_dir, "all_genes.csv"))

  # 3. Significant genes
  sig <- res[res$padj < 0.05 & !is.na(res$padj), ]
  write.csv(as.data.frame(sig), file.path(output_dir, "significant_genes.csv"))

  # 4. Normalized counts
  norm_counts <- counts(dds, normalized = TRUE)
  write.csv(norm_counts, file.path(output_dir, "normalized_counts.csv"))

  # 5. Top 100 genes
  top100 <- head(order(res$padj), 100)
  write.csv(norm_counts[top100, ], file.path(output_dir, "top100_genes_counts.csv"))

  # 6. Summary statistics
  summary_stats <- data.frame(
    Total_genes = nrow(res),
    Genes_tested = sum(!is.na(res$padj)),
    Significant_padj05 = sum(res$padj < 0.05, na.rm = TRUE),
    Significant_padj01 = sum(res$padj < 0.01, na.rm = TRUE),
    Upregulated_log2FC1 = sum(res$padj < 0.05 & res$log2FoldChange > 1, na.rm = TRUE),
    Downregulated_log2FC1 = sum(res$padj < 0.05 & res$log2FoldChange < -1, na.rm = TRUE)
  )
  write.csv(summary_stats, file.path(output_dir, "summary_statistics.csv"))

  message("Report generated in: ", output_dir)
}

generate_rnaseq_report(dds, res)
```

---

# Part 2: Single-Cell RNA-seq Pipeline

## 2.1 Overview

**Goal**: Identify cell types and differentially expressed genes

**Pipeline**: 10X data → QC → Normalization → Clustering → Annotation

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
library(SingleR)
library(celldex)
```

## 2.2 Load and Create Seurat Object

```{r}
# Load 10X Genomics data
data_dir <- "filtered_gene_bc_matrices/hg38/"
data <- Read10X(data.dir = data_dir)

# Create Seurat object
seurat_obj <- CreateSeuratObject(
  counts = data,
  project = "scRNAseq",
  min.cells = 3,      # Include genes detected in ≥3 cells
  min.features = 200  # Include cells with ≥200 genes
)

# Add metadata (if available)
metadata <- read.csv("metadata.csv", row.names = 1)
seurat_obj <- AddMetaData(seurat_obj, metadata)
```

## 2.3 Quality Control

```{r}
# Calculate mitochondrial percentage
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")

# Calculate ribosomal percentage
seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^RP[SL]")

# Visualize QC metrics
VlnPlot(seurat_obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3,
        pt.size = 0.1)

# Feature scatter plots
plot1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

# Filter cells
seurat_obj <- subset(seurat_obj,
                     subset = nFeature_RNA > 200 &
                             nFeature_RNA < 6000 &
                             percent.mt < 15)

# Before and after filtering
cat("Cells before filtering:", ncol(data), "\n")
cat("Cells after filtering:", ncol(seurat_obj), "\n")
```

## 2.4 Normalization and Feature Selection

```{r}
# Normalize data
seurat_obj <- NormalizeData(seurat_obj,
                            normalization.method = "LogNormalize",
                            scale.factor = 10000)

# Find variable features
seurat_obj <- FindVariableFeatures(seurat_obj,
                                   selection.method = "vst",
                                   nfeatures = 2000)

# Identify top 10 most variable genes
top10 <- head(VariableFeatures(seurat_obj), 10)

# Plot variable features
plot1 <- VariableFeaturePlot(seurat_obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

## 2.5 Scaling and PCA

```{r}
# Scale data
all.genes <- rownames(seurat_obj)
seurat_obj <- ScaleData(seurat_obj, features = all.genes)

# Or regress out unwanted sources of variation
seurat_obj <- ScaleData(seurat_obj,
                        vars.to.regress = c("percent.mt", "nCount_RNA"))

# Run PCA
seurat_obj <- RunPCA(seurat_obj,
                     features = VariableFeatures(object = seurat_obj))

# Visualize PCA results
print(seurat_obj[["pca"]], dims = 1:5, nfeatures = 5)

VizDimLoadings(seurat_obj, dims = 1:2, reduction = "pca")

DimPlot(seurat_obj, reduction = "pca")

DimHeatmap(seurat_obj, dims = 1:15, cells = 500, balanced = TRUE)

# Elbow plot to determine dimensionality
ElbowPlot(seurat_obj, ndims = 50)
```

## 2.6 Clustering

```{r}
# Cluster cells
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:30)
seurat_obj <- FindClusters(seurat_obj, resolution = 0.5)

# Try multiple resolutions
seurat_obj <- FindClusters(seurat_obj, resolution = c(0.1, 0.3, 0.5, 0.8, 1.0))

# Visualize clusters at different resolutions
library(clustree)
clustree(seurat_obj, prefix = "RNA_snn_res.")
```

## 2.7 Dimensionality Reduction

```{r}
# Run UMAP
seurat_obj <- RunUMAP(seurat_obj, dims = 1:30)

# Visualize
DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5)

# Run t-SNE (alternative)
seurat_obj <- RunTSNE(seurat_obj, dims = 1:30)
DimPlot(seurat_obj, reduction = "tsne", label = TRUE, pt.size = 0.5)

# Color by QC metrics
FeaturePlot(seurat_obj, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

# Split by condition (if applicable)
DimPlot(seurat_obj, reduction = "umap", split.by = "condition")
```

## 2.8 Find Marker Genes

```{r}
# Find markers for all clusters
all_markers <- FindAllMarkers(seurat_obj,
                              only.pos = TRUE,
                              min.pct = 0.25,
                              logfc.threshold = 0.25)

# Top markers per cluster
top_markers <- all_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

write.csv(all_markers, "results/cluster_markers.csv")
write.csv(top_markers, "results/top_cluster_markers.csv")

# Visualize top markers
top5 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(n = 5, order_by = avg_log2FC)

DoHeatmap(seurat_obj, features = top5$gene) + NoLegend()

# Violin plots for specific markers
VlnPlot(seurat_obj, features = c("CD3D", "CD8A", "CD4"))

# Feature plots
FeaturePlot(seurat_obj, features = c("CD3D", "CD8A", "CD4", "CD14"))

# Dot plot
DotPlot(seurat_obj, features = unique(top5$gene)) +
  RotatedAxis()
```

## 2.9 Cell Type Annotation

```{r}
# Manual annotation based on marker genes
cluster_names <- c(
  "0" = "CD4 T cells",
  "1" = "CD8 T cells",
  "2" = "B cells",
  "3" = "Monocytes",
  "4" = "NK cells",
  "5" = "Dendritic cells"
)

seurat_obj <- RenameIdents(seurat_obj, cluster_names)
seurat_obj$cell_type <- Idents(seurat_obj)

DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) +
  ggtitle("Cell Type Annotation")

# Automatic annotation with SingleR
library(SingleR)
library(celldex)

# Get reference dataset
ref <- celldex::HumanPrimaryCellAtlasData()

# Convert Seurat to SingleCellExperiment
sce <- as.SingleCellExperiment(seurat_obj)

# Predict cell types
predictions <- SingleR(test = sce,
                       ref = ref,
                       labels = ref$label.main)

# Add to Seurat object
seurat_obj$singler_labels <- predictions$labels

# Visualize
DimPlot(seurat_obj, reduction = "umap", group.by = "singler_labels", label = TRUE)
```

## 2.10 Differential Expression Between Conditions

```{r}
# Compare cell types between conditions
Idents(seurat_obj) <- "cell_type"

# Find DEGs between conditions for specific cell type
cd4_cells <- subset(seurat_obj, idents = "CD4 T cells")

Idents(cd4_cells) <- "condition"
de_genes <- FindMarkers(cd4_cells,
                        ident.1 = "Treatment",
                        ident.2 = "Control",
                        min.pct = 0.25)

# Volcano plot
library(EnhancedVolcano)
EnhancedVolcano(de_genes,
                lab = rownames(de_genes),
                x = 'avg_log2FC',
                y = 'p_val_adj',
                title = 'CD4 T cells: Treatment vs Control')
```

## 2.11 Trajectory Analysis (Pseudotime)

```{r}
library(monocle3)

# Convert Seurat to cell_data_set
cds <- as.cell_data_set(seurat_obj)

# Preprocess
cds <- preprocess_cds(cds, num_dim = 30)

# Dimensionality reduction
cds <- reduce_dimension(cds)

# Cluster cells
cds <- cluster_cells(cds)

# Learn trajectory
cds <- learn_graph(cds)

# Order cells in pseudotime
cds <- order_cells(cds)

# Plot trajectory
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = FALSE)

# Plot by cell type
plot_cells(cds,
           color_cells_by = "cell_type",
           label_cell_groups = TRUE)
```

---

# Part 3: ChIP-seq Pipeline

## 3.1 Overview

**Goal**: Identify protein-DNA binding sites

**Pipeline**: BAM files → Peak calling → Annotation → Visualization

```{r}
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(clusterProfiler)
library(GenomicRanges)
library(rtracklayer)
library(DiffBind)
```

## 3.2 Peak Calling (Usually done outside R)

```bash
# Using MACS2 (in bash/shell)
macs2 callpeak \
  -t treatment.bam \
  -c control.bam \
  -f BAM \
  -g hs \
  -n sample_name \
  --outdir peaks/ \
  -q 0.05

# Or MACS3 for ChIP-seq
macs3 callpeak \
  -t treatment.bam \
  -c control.bam \
  -f BAMPE \
  -g hs \
  -n sample_name \
  --outdir peaks/
```

## 3.3 Load and Process Peaks

```{r}
# Read peak files
peaks <- import.bed("peaks/sample_peaks.narrowPeak")

# Or read multiple samples
peak_files <- list.files("peaks/", pattern = "*.narrowPeak", full.names = TRUE)
peak_list <- lapply(peak_files, import.bed)
names(peak_list) <- gsub("_peaks.narrowPeak", "", basename(peak_files))

# Combine all peaks
all_peaks <- reduce(do.call(c, peak_list))

# Summary statistics
summary(width(peaks))
table(seqnames(peaks))

# Filter peaks by score
peaks_filtered <- peaks[peaks$score > 50]
```

## 3.4 Peak Annotation

```{r}
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# Annotate peaks
peakAnno <- annotatePeak(peaks,
                        tssRegion = c(-3000, 3000),
                        TxDb = txdb,
                        annoDb = "org.Hs.eg.db")

# View annotation
peakAnno

# Convert to data frame
peak_df <- as.data.frame(peakAnno)
write.csv(peak_df, "results/annotated_peaks.csv")

# Visualization
# Distribution of peaks relative to TSS
plotAnnoBar(peakAnno)
plotDistToTSS(peakAnno,
             title = "Distribution of ChIP peaks relative to TSS")

# Pie chart
plotAnnoPie(peakAnno)

# Genomic annotation
plotAnnoPie(peakAnno)

# Upset plot
upsetplot(peakAnno, vennpie = TRUE)

# Heatmap of ChIP peaks binding to TSS regions
tagMatrix <- getTagMatrix(peaks, windows = promoter)
tagHeatmap(tagMatrix, xlim = c(-3000, 3000))

# Average profile
plotAvgProf(tagMatrix, xlim = c(-3000, 3000),
            xlab = "Genomic Region (5'->3')",
            ylab = "Read Count Frequency")
```

## 3.5 Functional Enrichment

```{r}
# Get genes associated with peaks
genes <- peak_df$geneId

# GO enrichment
ego <- enrichGO(gene = genes,
                OrgDb = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05)

dotplot(ego, showCategory = 20)
barplot(ego, showCategory = 20)

# KEGG pathway
keg <- enrichKEGG(gene = genes,
                  organism = "hsa",
                  pvalueCutoff = 0.05)

dotplot(keg, showCategory = 15)

# Disease association
edo <- enrichDO(gene = genes,
                ont = "DO",
                pvalueCutoff = 0.05,
                readable = TRUE)
```

## 3.6 Differential Binding Analysis

```{r}
library(DiffBind)

# Create sample sheet
samples <- data.frame(
  SampleID = c("Sample1_Ctrl", "Sample2_Ctrl", "Sample3_Treat", "Sample4_Treat"),
  Condition = c("Control", "Control", "Treatment", "Treatment"),
  Replicate = c(1, 2, 1, 2),
  bamReads = c("bam/s1.bam", "bam/s2.bam", "bam/s3.bam", "bam/s4.bam"),
  Peaks = c("peaks/s1.bed", "peaks/s2.bed", "peaks/s3.bed", "peaks/s4.bed"),
  PeakCaller = "macs"
)

# Create DBA object
dba <- dba(sampleSheet = samples)

# Count reads
dba <- dba.count(dba)

# Establish contrast
dba <- dba.contrast(dba, categories = DBA_CONDITION)

# Analyze
dba <- dba.analyze(dba)

# Get differential binding sites
db_sites <- dba.report(dba, th = 0.05)

# Export
write.csv(as.data.frame(db_sites), "results/differential_binding.csv")

# Visualizations
dba.plotPCA(dba, attributes = DBA_CONDITION)
dba.plotHeatmap(dba)
dba.plotMA(dba)
dba.plotVolcano(dba)
```

## 3.7 Motif Analysis

```{r}
library(motifStack)
library(BSgenome.Hsapiens.UCSC.hg38)

# Extract sequences around peak summits
peak_seqs <- getSeq(BSgenome.Hsapiens.UCSC.hg38, peaks)

# Export for external motif analysis
writeXStringSet(peak_seqs, "peaks_sequences.fasta")

# Use MEME suite (external) or homer (external) for de novo motif discovery
```

---

# Part 4: Variant Calling Pipeline

## 4.1 Overview

**Goal**: Identify and annotate genetic variants

```{r}
library(VariantAnnotation)
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg38)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

## 4.2 Read and Filter VCF

```{r}
# Read VCF file
vcf <- readVcf("variants.vcf", "hg38")

# Basic information
vcf
rowRanges(vcf)
info(vcf)
geno(vcf)

# Filter by quality
vcf_filtered <- vcf[qual(vcf) > 30]

# Filter by depth
vcf_filtered <- vcf_filtered[info(vcf_filtered)$DP > 10]

# Filter by allele frequency
vcf_filtered <- vcf_filtered[info(vcf_filtered)$AF > 0.01]

# Export filtered VCF
writeVcf(vcf_filtered, "variants_filtered.vcf")
```

## 4.3 Variant Annotation

```{r}
# Predict coding consequences
library(VariantAnnotation)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
coding <- predictCoding(vcf, txdb, BSgenome.Hsapiens.UCSC.hg38)

# View consequences
table(coding$CONSEQUENCE)

# Filter for specific consequences
nonsyn <- coding[coding$CONSEQUENCE %in% c("nonsynonymous", "frameshift", "nonsense")]

# Add gene information
library(org.Hs.eg.db)
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = as.character(nonsyn$GENEID),
                       column = "SYMBOL",
                       keytype = "ENTREZID")
nonsyn$SYMBOL <- gene_symbols

# Export
write.csv(as.data.frame(nonsyn), "results/coding_variants.csv")
```

## 4.4 Variant Summary and Visualization

```{r}
# Variant types
variant_types <- table(info(vcf)$TYPE)
barplot(variant_types, main = "Variant Types", col = "steelblue")

# Variant quality distribution
hist(qual(vcf), breaks = 50, main = "Variant Quality Distribution",
     xlab = "Quality Score", col = "lightblue")

# Depth distribution
hist(info(vcf)$DP, breaks = 50, main = "Read Depth Distribution",
     xlab = "Depth", col = "lightgreen")

# Allele frequency
hist(info(vcf)$AF, breaks = 50, main = "Allele Frequency Distribution",
     xlab = "Allele Frequency", col = "lightpink")

# Variants per chromosome
var_per_chr <- table(seqnames(rowRanges(vcf)))
barplot(var_per_chr, las = 2, main = "Variants per Chromosome")

# Ti/Tv ratio (transitions/transversions)
# Calculate Ti/Tv ratio for quality check
```

---

# Part 5: Microbiome Analysis

## 5.1 Overview

**Goal**: Analyze microbial community composition and diversity

```{r}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(microbiome)
```

## 5.2 Create phyloseq Object

```{r}
# Load data
otu_table <- read.csv("otu_table.csv", row.names = 1)
taxonomy <- read.csv("taxonomy.csv", row.names = 1)
metadata <- read.csv("metadata.csv", row.names = 1)

# Create phyloseq object
otu <- otu_table(as.matrix(otu_table), taxa_are_rows = TRUE)
tax <- tax_table(as.matrix(taxonomy))
sam <- sample_data(metadata)

physeq <- phyloseq(otu, tax, sam)

# Or load from biom format
# physeq <- import_biom("table.biom")
```

## 5.3 Quality Control and Filtering

```{r}
# Remove low abundance taxa
physeq_filtered <- filter_taxa(physeq, function(x) sum(x) > 10, TRUE)

# Remove singletons
physeq_filtered <- filter_taxa(physeq_filtered, function(x) sum(x > 1) > 1, TRUE)

# Filter samples with low read count
physeq_filtered <- prune_samples(sample_sums(physeq_filtered) > 1000, physeq_filtered)

# Summary
physeq_filtered
```

## 5.4 Alpha Diversity

```{r}
# Calculate alpha diversity
alpha_div <- estimate_richness(physeq_filtered,
                               measures = c("Observed", "Shannon", "Simpson"))

# Add metadata
alpha_div$Group <- sample_data(physeq_filtered)$Group

# Plot
plot_richness(physeq_filtered,
              x = "Group",
              measures = c("Shannon", "Simpson"),
              color = "Group") +
  geom_boxplot() +
  theme_minimal()

# Statistical test
wilcox.test(Shannon ~ Group, data = alpha_div)
```

## 5.5 Beta Diversity

```{r}
# Ordination
physeq_rel <- transform_sample_counts(physeq_filtered, function(x) x/sum(x))

# PCoA with Bray-Curtis
ord <- ordinate(physeq_rel, method = "PCoA", distance = "bray")

plot_ordination(physeq_rel, ord, color = "Group") +
  geom_point(size = 3) +
  stat_ellipse() +
  theme_minimal()

# NMDS
ord_nmds <- ordinate(physeq_rel, method = "NMDS", distance = "bray")
plot_ordination(physeq_rel, ord_nmds, color = "Group")

# PERMANOVA
dist_matrix <- distance(physeq_rel, method = "bray")
metadata <- as(sample_data(physeq_rel), "data.frame")
adonis2(dist_matrix ~ Group, data = metadata)
```

## 5.6 Taxonomic Composition

```{r}
# Bar plot
plot_bar(physeq_filtered, fill = "Phylum") +
  facet_wrap(~Group, scales = "free_x")

# Relative abundance at phylum level
physeq_phylum <- tax_glom(physeq_rel, taxrank = "Phylum")

plot_bar(physeq_phylum, fill = "Phylum") +
  geom_bar(stat = "identity") +
  facet_wrap(~Group, scales = "free_x") +
  theme_minimal()

# Heatmap
plot_heatmap(physeq_filtered,
             taxa.label = "Genus",
             sample.label = "Sample") +
  facet_grid(~Group, scales = "free")
```

## 5.7 Differential Abundance

```{r}
library(DESeq2)

# Convert to DESeq2
diagdds <- phyloseq_to_deseq2(physeq_filtered, ~ Group)

# Run DESeq2
diagdds <- DESeq(diagdds, test = "Wald", fitType = "local")

# Results
res <- results(diagdds, contrast = c("Group", "Treatment", "Control"))

# Order by adjusted p-value
res <- res[order(res$padj), ]

# Get significant taxa
sig_taxa <- res[which(res$padj < 0.05), ]

# Add taxonomy
tax_table <- tax_table(physeq_filtered)[rownames(sig_taxa), ]
sig_taxa_annotated <- cbind(as(sig_taxa, "data.frame"), as(tax_table, "matrix"))

write.csv(sig_taxa_annotated, "results/differential_abundance.csv")
```

---

# Summary

This document provided complete, production-ready pipelines for:

1. **Bulk RNA-seq**: From counts to DEGs to enrichment
2. **Single-cell RNA-seq**: QC, clustering, annotation, trajectory
3. **ChIP-seq**: Peak calling, annotation, differential binding
4. **Variant calling**: VCF processing, annotation, filtering
5. **Microbiome**: Diversity, composition, differential abundance

## Key Tools Covered:
- DESeq2, tximport, clusterProfiler
- Seurat, SingleR, monocle3
- ChIPseeker, DiffBind
- VariantAnnotation
- phyloseq, vegan, microbiome

## Best Practices:
- Always perform QC
- Use appropriate normalization
- Multiple testing correction
- Visualize at every step
- Document parameters
- Save intermediate results

# References

- [Bioconductor Workflows](https://bioconductor.org/packages/release/BiocViews.html#___Workflow)
- [Seurat vignettes](https://satijalab.org/seurat/)
- [ChIP-seq analysis](https://hbctraining.github.io/Intro-to-ChIPseq/)
- [RNA-seq workflow](http://bioconductor.org/packages/release/workflows/html/rnaseqGene.html)
