---
title: "Cloud Computing and Big Data in R"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Introduction to Cloud Computing with R

## Why Cloud Computing?

- **Scalability**: Handle datasets larger than local memory
- **Compute Power**: Access to high-performance computing resources
- **Collaboration**: Share data and analysis pipelines
- **Cost-Effective**: Pay only for resources used
- **Reproducibility**: Consistent computing environments

## Cloud Platforms

- **AWS (Amazon Web Services)**: EC2, S3, EMR, Redshift
- **Google Cloud Platform (GCP)**: Compute Engine, Cloud Storage, BigQuery
- **Microsoft Azure**: Virtual Machines, Blob Storage, HDInsight
- **Cloud-native solutions**: RStudio Cloud, Posit Connect

# 2. AWS Integration

## AWS SDK for R (paws)

```{r}
# Install
install.packages("paws")

library(paws)

# Configure AWS credentials
# Set up ~/.aws/credentials or use environment variables
# AWS_ACCESS_KEY_ID
# AWS_SECRET_ACCESS_KEY
# AWS_DEFAULT_REGION
```

## Working with S3

```{r}
library(paws)

# Initialize S3 client
s3 <- s3()

# List buckets
buckets <- s3$list_buckets()
print(buckets$Buckets)

# Create bucket
s3$create_bucket(
  Bucket = "my-r-data-bucket",
  CreateBucketConfiguration = list(
    LocationConstraint = "us-west-2"
  )
)

# Upload file to S3
s3$put_object(
  Bucket = "my-r-data-bucket",
  Key = "data/my_data.csv",
  Body = charToRaw(paste(readLines("local_data.csv"), collapse = "\n"))
)

# Download file from S3
obj <- s3$get_object(
  Bucket = "my-r-data-bucket",
  Key = "data/my_data.csv"
)
data <- read.csv(text = rawToChar(obj$Body))

# List objects in bucket
objects <- s3$list_objects_v2(
  Bucket = "my-r-data-bucket",
  Prefix = "data/"
)

# Delete object
s3$delete_object(
  Bucket = "my-r-data-bucket",
  Key = "data/my_data.csv"
)
```

## Alternative: aws.s3 Package

```{r}
library(aws.s3)

# Set credentials
Sys.setenv("AWS_ACCESS_KEY_ID" = "your_key",
           "AWS_SECRET_ACCESS_KEY" = "your_secret",
           "AWS_DEFAULT_REGION" = "us-east-1")

# Save R object directly to S3
s3save(mtcars, iris, bucket = "my-bucket", object = "datasets.RData")

# Load R object from S3
s3load("datasets.RData", bucket = "my-bucket")

# Read CSV from S3
data <- s3read_using(read.csv, bucket = "my-bucket", object = "data.csv")

# Write CSV to S3
s3write_using(mtcars, FUN = write.csv, bucket = "my-bucket", object = "mtcars.csv")

# List bucket contents
bucket_list_df("my-bucket")

# Delete object
delete_object("data.csv", bucket = "my-bucket")
```

## AWS EC2 Instances

```{r}
library(paws)

ec2 <- ec2()

# List instances
instances <- ec2$describe_instances()

# Start instance
ec2$start_instances(InstanceIds = list("i-1234567890abcdef0"))

# Stop instance
ec2$stop_instances(InstanceIds = list("i-1234567890abcdef0"))

# Create instance (basic example)
ec2$run_instances(
  ImageId = "ami-0c55b159cbfafe1f0",  # Amazon Linux 2 AMI
  InstanceType = "t2.micro",
  MinCount = 1L,
  MaxCount = 1L,
  KeyName = "my-key-pair",
  SecurityGroupIds = list("sg-12345678"),
  TagSpecifications = list(
    list(
      ResourceType = "instance",
      Tags = list(
        list(Key = "Name", Value = "R-Analysis-Server")
      )
    )
  )
)
```

## AWS Redshift

```{r}
library(DBI)
library(RPostgres)

# Connect to Redshift
con <- dbConnect(
  RPostgres::Postgres(),
  host = "my-cluster.redshift.amazonaws.com",
  port = 5439,
  dbname = "mydb",
  user = "username",
  password = "password"
)

# Query data
query <- "
  SELECT date, region, sales, customers
  FROM sales_data
  WHERE date >= '2023-01-01'
  LIMIT 10000
"

data <- dbGetQuery(con, query)

# Write data to Redshift
dbWriteTable(con, "r_processed_data", processed_data,
             overwrite = TRUE, row.names = FALSE)

# Use dplyr with Redshift
library(dplyr)
sales_tbl <- tbl(con, "sales_data")

result <- sales_tbl %>%
  filter(date >= "2023-01-01") %>%
  group_by(region) %>%
  summarise(
    total_sales = sum(sales, na.rm = TRUE),
    avg_sales = mean(sales, na.rm = TRUE)
  ) %>%
  collect()

dbDisconnect(con)
```

# 3. Google Cloud Platform

## Google Cloud Storage

```{r}
library(googleCloudStorageR)

# Authenticate
gcs_auth("path/to/service-account-key.json")

# Set default bucket
gcs_global_bucket("my-gcs-bucket")

# List objects
objects <- gcs_list_objects()

# Upload file
gcs_upload("local_file.csv", name = "data/remote_file.csv")

# Download file
gcs_get_object("data/remote_file.csv", saveToDisk = "downloaded_file.csv")

# Save R object directly
gcs_save(mtcars, iris, file = "datasets.RData")

# Load R object
gcs_load("datasets.RData")

# Read CSV directly into memory
data <- gcs_get_object("data.csv") %>%
  textConnection() %>%
  read.csv()
```

## BigQuery

```{r}
library(bigrquery)

# Set project
project_id <- "my-project-id"
billing <- project_id  # Billing project

# Query public dataset
sql <- "
  SELECT name, gender, count
  FROM `bigquery-public-data.usa_names.usa_1910_2013`
  WHERE state = 'CA' AND year = 2013
  ORDER BY count DESC
  LIMIT 10
"

# Execute query
result <- bq_project_query(project_id, sql)
data <- bq_table_download(result)

# Use dplyr with BigQuery
library(dplyr)

# Connect to dataset
con <- dbConnect(
  bigrquery::bigquery(),
  project = project_id,
  dataset = "my_dataset",
  billing = billing
)

# Query with dplyr
names_tbl <- tbl(con, "usa_names")

ca_names <- names_tbl %>%
  filter(state == "CA", year == 2013) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  collect()

# Create table in BigQuery
bq_table_create(
  bq_table(project_id, "my_dataset", "new_table"),
  fields = as_bq_fields(mtcars)
)

# Upload data
bq_table_upload(
  bq_table(project_id, "my_dataset", "new_table"),
  mtcars
)

dbDisconnect(con)
```

## Google Compute Engine

```{r}
library(googleComputeEngineR)

# Authenticate
gce_auth("path/to/service-account-key.json")

# Set project and zone
gce_global_project("my-project-id")
gce_global_zone("us-central1-a")

# List instances
instances <- gce_list_instances()

# Create instance with RStudio
vm <- gce_vm(
  name = "rstudio-server",
  predefined_type = "n1-standard-4",
  template = "rstudio",
  username = "rstudio",
  password = "secure_password"
)

# SSH into instance
gce_ssh(vm, "whoami")

# Transfer file
gce_ssh_upload(vm, "local_file.R", "remote_file.R")

# Stop instance
gce_vm_stop(vm)

# Start instance
gce_vm_start(vm)

# Delete instance
gce_vm_delete(vm)
```

# 4. Apache Spark with sparklyr

## Setup and Connection

```{r}
# Install
install.packages("sparklyr")
library(sparklyr)

# Install Spark locally (for testing)
spark_install(version = "3.4.0")

# Connect to local Spark
sc <- spark_connect(master = "local")

# Connect to Spark cluster
sc <- spark_connect(
  master = "spark://master-node:7077",
  spark_home = "/opt/spark",
  config = list(
    spark.executor.memory = "4g",
    spark.executor.cores = 4,
    spark.dynamicAllocation.enabled = TRUE
  )
)

# Connect to Spark on AWS EMR
sc <- spark_connect(
  master = "yarn",
  app_name = "my-analysis",
  config = spark_config()
)
```

## Loading Data

```{r}
# Copy R data frame to Spark
mtcars_spark <- copy_to(sc, mtcars, overwrite = TRUE)

# Read from file
# CSV
sales_spark <- spark_read_csv(
  sc,
  name = "sales",
  path = "s3://my-bucket/data/sales.csv",
  header = TRUE,
  infer_schema = TRUE
)

# Parquet (recommended for big data)
sales_spark <- spark_read_parquet(
  sc,
  name = "sales",
  path = "s3://my-bucket/data/sales.parquet"
)

# JSON
sales_spark <- spark_read_json(
  sc,
  name = "sales",
  path = "s3://my-bucket/data/sales.json"
)

# From Hive table
sales_spark <- spark_read_table(sc, "sales_table")

# Multiple files
sales_spark <- spark_read_csv(
  sc,
  name = "all_sales",
  path = "s3://my-bucket/data/sales/*.csv"
)
```

## Data Manipulation with dplyr

```{r}
library(dplyr)

# Standard dplyr operations work with Spark
result <- mtcars_spark %>%
  filter(mpg > 20) %>%
  group_by(cyl) %>%
  summarise(
    n = n(),
    avg_mpg = mean(mpg),
    avg_hp = mean(hp),
    max_wt = max(wt)
  ) %>%
  arrange(desc(avg_mpg))

# Join operations
customers_spark <- spark_read_csv(sc, "customers", "customers.csv")
orders_spark <- spark_read_csv(sc, "orders", "orders.csv")

customer_orders <- customers_spark %>%
  left_join(orders_spark, by = "customer_id") %>%
  group_by(customer_name) %>%
  summarise(
    total_orders = n(),
    total_amount = sum(order_amount)
  )

# Window functions
library(dbplyr)

ranked_sales <- sales_spark %>%
  group_by(region) %>%
  mutate(
    sales_rank = row_number(desc(sales)),
    cumulative_sales = cumsum(sales),
    pct_of_total = sales / sum(sales) * 100
  )

# SQL queries
library(DBI)

result <- dbGetQuery(sc, "
  SELECT region,
         AVG(sales) as avg_sales,
         COUNT(*) as n_transactions
  FROM sales
  WHERE date >= '2023-01-01'
  GROUP BY region
  ORDER BY avg_sales DESC
")
```

## Machine Learning with Spark

```{r}
library(sparklyr)

# Prepare data
iris_spark <- copy_to(sc, iris, overwrite = TRUE)

# Split data
partitions <- iris_spark %>%
  sdf_random_split(training = 0.7, testing = 0.3, seed = 123)

train_data <- partitions$training
test_data <- partitions$testing

# Linear Regression
lr_model <- train_data %>%
  ml_linear_regression(Sepal_Length ~ Sepal_Width + Petal_Length + Petal_Width)

# Predictions
predictions <- ml_predict(lr_model, test_data)

# Model metrics
ml_regression_evaluator(predictions, label_col = "Sepal_Length")

# Random Forest
rf_model <- train_data %>%
  ml_random_forest_classifier(
    Species ~ Sepal_Length + Sepal_Width + Petal_Length + Petal_Width,
    num_trees = 100,
    max_depth = 5
  )

# Evaluate
rf_predictions <- ml_predict(rf_model, test_data)
ml_multiclass_classification_evaluator(rf_predictions)

# Gradient Boosted Trees
gbt_model <- train_data %>%
  ml_gbt_classifier(
    Species ~ .,
    max_iter = 20,
    max_depth = 5
  )

# K-means clustering
kmeans_model <- iris_spark %>%
  ml_kmeans(~ Sepal_Length + Sepal_Width + Petal_Length + Petal_Width,
            k = 3,
            max_iter = 100)

# Get cluster assignments
clustered_data <- ml_predict(kmeans_model, iris_spark)

# PCA
pca_model <- iris_spark %>%
  ml_pca(~ Sepal_Length + Sepal_Width + Petal_Length + Petal_Width,
         k = 2)

pca_transformed <- ml_transform(pca_model, iris_spark)
```

## Feature Engineering

```{r}
# Create feature transformers pipeline
pipeline <- ml_pipeline(sc) %>%
  ft_string_indexer("Species", "species_idx") %>%
  ft_vector_assembler(
    c("Sepal_Length", "Sepal_Width", "Petal_Length", "Petal_Width"),
    "features"
  ) %>%
  ft_standard_scaler("features", "scaled_features")

# Fit pipeline
fitted_pipeline <- ml_fit(pipeline, iris_spark)

# Transform data
transformed_data <- ml_transform(fitted_pipeline, iris_spark)

# More transformers
pipeline_advanced <- ml_pipeline(sc) %>%
  # Bucketize continuous variable
  ft_bucketizer("Sepal_Length", "sepal_length_bucket",
                splits = c(-Inf, 5, 6, 7, Inf)) %>%
  # One-hot encoding
  ft_one_hot_encoder("species_idx", "species_vec") %>%
  # Polynomial expansion
  ft_polynomial_expansion("features", "poly_features", degree = 2) %>%
  # Normalize
  ft_normalizer("poly_features", "normalized_features")
```

## Performance Optimization

```{r}
# Cache frequently used tables
cached_data <- sales_spark %>%
  sdf_register("sales_cached") %>%
  tbl_cache(sc, "sales_cached")

# Repartition for better parallelism
repartitioned <- sales_spark %>%
  sdf_repartition(partitions = 200)

# Coalesce to reduce partitions
coalesced <- sales_spark %>%
  sdf_coalesce(partitions = 10)

# Persist with storage level
library(sparklyr)
spark_dataframe(sales_spark) %>%
  invoke("persist", "MEMORY_AND_DISK")

# Broadcast small tables for joins
small_table_broadcast <- sdf_broadcast(small_table_spark)

result <- large_table_spark %>%
  left_join(small_table_broadcast, by = "key")

# Monitor Spark UI
spark_web(sc)

# View execution plan
explain(result)
```

## Writing Results

```{r}
# Write to CSV
result %>%
  spark_write_csv("output/results.csv", mode = "overwrite")

# Write to Parquet (recommended)
result %>%
  spark_write_parquet("output/results.parquet", mode = "overwrite")

# Write partitioned data
result %>%
  spark_write_parquet(
    "output/results_partitioned.parquet",
    mode = "overwrite",
    partition_by = c("year", "month")
  )

# Write to Hive table
result %>%
  spark_write_table("result_table", mode = "overwrite")

# Collect to R (use with caution on large data)
result_local <- collect(result)

# Collect sample
sample_data <- result %>%
  sdf_sample(fraction = 0.01) %>%
  collect()
```

## Disconnect

```{r}
# Close connection
spark_disconnect(sc)

# Disconnect all
spark_disconnect_all()
```

# 5. Distributed Computing with future

## Basic future Usage

```{r}
library(future)
library(future.apply)

# Set parallel backend
plan(multisession, workers = 4)

# Or use cluster
plan(cluster, workers = c("node1", "node2", "node3", "node4"))

# Simple future
f <- future({
  # Long computation
  result <- expensive_function()
  result
})

# Continue with other work...

# Get result when needed (blocks if not ready)
result <- value(f)

# Check if ready
if (resolved(f)) {
  result <- value(f)
}
```

## Parallel Apply Functions

```{r}
library(future.apply)

# Set parallel backend
plan(multisession, workers = 8)

# Parallel lapply
result <- future_lapply(1:100, function(i) {
  # Simulate computation
  Sys.sleep(0.1)
  i^2
})

# Parallel sapply
result <- future_sapply(1:100, function(i) {
  complex_calculation(i)
})

# Parallel Map
result <- future_Map(function(x, y) {
  x * y + runif(1)
}, x = 1:100, y = 101:200)

# Progress bar
library(progressr)

with_progress({
  p <- progressor(steps = 100)
  result <- future_lapply(1:100, function(i) {
    p()
    Sys.sleep(0.1)
    i^2
  })
})
```

## Real Bioinformatics Example

```{r}
library(future.apply)
library(Biostrings)

# Process multiple BAM files in parallel
plan(multisession, workers = 8)

bam_files <- list.files("bam_files/", pattern = "\\.bam$", full.names = TRUE)

# Count reads in parallel
read_counts <- future_lapply(bam_files, function(bam_file) {
  library(Rsamtools)

  # Count mapped reads
  param <- ScanBamParam(what = "seq")
  reads <- scanBam(bam_file, param = param)

  data.frame(
    file = basename(bam_file),
    total_reads = length(reads[[1]]$seq),
    stringsAsFactors = FALSE
  )
})

read_counts_df <- do.call(rbind, read_counts)

# Parallel sequence analysis
sequences <- readDNAStringSet("sequences.fasta")

gc_content <- future_sapply(sequences, function(seq) {
  letterFrequency(seq, letters = "GC", as.prob = TRUE)[1]
})

# Parallel BLAST searches
blast_results <- future_lapply(sequences, function(seq) {
  # Run BLAST
  blast_result <- blast_sequence(seq)
  blast_result
})
```

# 6. Memory Management for Big Data

## Chunked Processing

```{r}
# Process large CSV in chunks
library(readr)

process_chunk <- function(chunk, pos) {
  # Process each chunk
  chunk %>%
    filter(value > 0) %>%
    group_by(category) %>%
    summarise(
      total = sum(value),
      count = n()
    )
}

# Read and process in chunks
result <- read_csv_chunked(
  "large_file.csv",
  callback = DataFrameCallback$new(process_chunk),
  chunk_size = 10000
)

# Aggregate results
final_result <- result %>%
  group_by(category) %>%
  summarise(
    total = sum(total),
    count = sum(count)
  )
```

## ff Package for Out-of-Memory Data

```{r}
library(ff)

# Create ff object
x <- ff(vmode = "double", length = 1e9)

# Read large file into ff
data_ff <- read.csv.ffdf(
  file = "large_file.csv",
  header = TRUE,
  first.rows = 10000,
  next.rows = 50000
)

# Subset
subset_data <- data_ff[data_ff$value > 0, ]

# Compute statistics
mean_value <- mean(data_ff$value)

# Use with biglm for large-scale linear models
library(biglm)
model <- biglm(y ~ x1 + x2 + x3, data = data_ff)
```

## bigmemory Package

```{r}
library(bigmemory)

# Create big.matrix
x <- big.matrix(nrow = 1e6, ncol = 100, type = "double")

# Attach to existing file
x <- attach.big.matrix("data.desc")

# Read from file
x <- read.big.matrix(
  "large_file.csv",
  header = TRUE,
  type = "double",
  backingfile = "data.bin",
  descriptorfile = "data.desc"
)

# Operations
mean_col1 <- mean(x[, 1])
subset_data <- x[x[, 1] > 0, ]

# Use with biglm
library(biganalytics)
model <- biglm.big.matrix(y ~ x1 + x2, data = x)
```

## data.table for Efficient Processing

```{r}
library(data.table)

# Read large CSV efficiently
dt <- fread("large_file.csv")

# Memory-efficient operations (modify by reference)
dt[, new_col := value * 2]
dt[value > 0, filtered_col := "high"]

# Efficient grouping
result <- dt[, .(
  mean_val = mean(value),
  sum_val = sum(value),
  .N
), by = .(category, subcategory)]

# Rolling joins for time series
setkey(dt, date)
dt[other_dt, roll = "nearest"]

# Fast file writing
fwrite(dt, "output.csv")
```

# 7. Complete Cloud Pipeline Example

## RNA-seq Analysis on AWS

```{r}
library(aws.s3)
library(DESeq2)
library(future.apply)

# Setup
Sys.setenv("AWS_DEFAULT_REGION" = "us-east-1")
plan(multisession, workers = 8)

# 1. Download count matrices from S3
count_files <- get_bucket_df("rnaseq-data", prefix = "counts/") %>%
  filter(grepl("\\.txt$", Key)) %>%
  pull(Key)

# Download in parallel
local_files <- future_lapply(count_files, function(file) {
  local_path <- file.path("temp", basename(file))
  save_object(file, bucket = "rnaseq-data", file = local_path)
  local_path
})

# 2. Read and combine count matrices
count_data <- future_lapply(local_files, function(file) {
  read.table(file, header = TRUE, row.names = 1)
}) %>%
  do.call(cbind, .)

# 3. Read metadata from S3
metadata <- s3read_using(
  read.csv,
  bucket = "rnaseq-data",
  object = "metadata/sample_info.csv"
)

# 4. Run DESeq2
dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData = metadata,
  design = ~ condition
)

dds <- DESeq(dds)
results <- results(dds)

# 5. Generate plots
library(ggplot2)

# Volcano plot
volcano_plot <- ggplot(as.data.frame(results),
                       aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = padj < 0.05 & abs(log2FoldChange) > 1)) +
  theme_minimal() +
  labs(title = "Volcano Plot")

# 6. Upload results to S3
s3write_using(
  as.data.frame(results),
  FUN = write.csv,
  bucket = "rnaseq-data",
  object = "results/deseq2_results.csv"
)

ggsave("volcano_plot.png", volcano_plot)
put_object(
  file = "volcano_plot.png",
  bucket = "rnaseq-data",
  object = "results/volcano_plot.png"
)

# 7. Save R session
s3save(dds, results,
       bucket = "rnaseq-data",
       object = "results/workspace.RData")

# Cleanup local files
unlink("temp", recursive = TRUE)
```

## Spark-based Variant Analysis

```{r}
library(sparklyr)
library(dplyr)

# Connect to Spark on EMR
sc <- spark_connect(master = "yarn", app_name = "variant-analysis")

# Read VCF data (converted to Parquet)
variants <- spark_read_parquet(
  sc,
  name = "variants",
  path = "s3://genomics-data/vcf/variants.parquet"
)

# Filter and annotate variants
filtered_variants <- variants %>%
  filter(
    QUAL > 30,
    DP > 10,
    !is.na(CHROM)
  ) %>%
  mutate(
    is_snp = nchar(REF) == 1 & nchar(ALT) == 1,
    is_indel = nchar(REF) != nchar(ALT)
  ) %>%
  group_by(CHROM, is_snp) %>%
  summarise(
    n_variants = n(),
    avg_qual = mean(QUAL),
    avg_depth = mean(DP)
  )

# Annotate with gene information
genes <- spark_read_parquet(sc, "genes", "s3://genomics-data/genes.parquet")

annotated_variants <- variants %>%
  spark_apply(function(df) {
    # Custom annotation function
    df$gene <- annotate_with_genes(df$CHROM, df$POS)
    df
  }, columns = c(colnames(variants), "gene" = "character"))

# Calculate per-sample statistics
sample_stats <- variants %>%
  group_by(SAMPLE) %>%
  summarise(
    total_variants = n(),
    novel_variants = sum(is.na(RSID)),
    avg_qual = mean(QUAL),
    ts_tv_ratio = sum(is_transition) / sum(is_transversion)
  )

# Write results
sample_stats %>%
  spark_write_parquet(
    "s3://genomics-data/results/sample_stats.parquet",
    mode = "overwrite"
  )

# Collect small summary for reporting
summary <- sample_stats %>%
  collect()

# Create report and upload
library(rmarkdown)
render("variant_report.Rmd", output_file = "report.html")

library(aws.s3)
put_object(
  file = "report.html",
  bucket = "genomics-data",
  object = "reports/variant_report.html"
)

spark_disconnect(sc)
```

# 8. Best Practices

## Cost Optimization

```{r}
# 1. Use appropriate instance types
# - Use Spot instances for non-critical workloads
# - Right-size instances based on profiling

# 2. Store data efficiently
# - Use Parquet instead of CSV
# - Compress data
# - Partition large datasets

# 3. Clean up resources
library(aws.s3)

# Delete temporary objects
delete_object("temp/data.csv", bucket = "my-bucket")

# List and delete old files
old_files <- get_bucket_df("my-bucket") %>%
  filter(LastModified < Sys.Date() - 30)

for (file in old_files$Key) {
  delete_object(file, bucket = "my-bucket")
}

# 4. Use lifecycle policies
# Set up S3 lifecycle rules to transition to cheaper storage
```

## Security

```{r}
# 1. Use IAM roles instead of hardcoded credentials

# 2. Encrypt sensitive data
library(aws.s3)

# Upload with server-side encryption
put_object(
  file = "sensitive_data.csv",
  bucket = "secure-bucket",
  object = "data.csv",
  headers = list(
    "x-amz-server-side-encryption" = "AES256"
  )
)

# 3. Use VPC for Spark clusters

# 4. Rotate credentials regularly

# 5. Use least privilege principle for IAM policies
```

## Monitoring and Logging

```{r}
# Log operations
library(logger)

log_info("Starting data processing pipeline")

tryCatch({
  # Pipeline code
  data <- process_data()
  log_info("Data processing completed successfully")
}, error = function(e) {
  log_error("Pipeline failed: {e$message}")
  # Send alert
  send_alert(e$message)
})

# Monitor Spark applications
library(sparklyr)

sc <- spark_connect(master = "yarn")

# Check Spark UI
spark_web(sc)

# Get application logs
spark_log(sc, n = 100)

# CloudWatch integration for AWS
library(paws)
cw <- cloudwatch()

cw$put_metric_data(
  Namespace = "RAnalysis",
  MetricData = list(
    list(
      MetricName = "ProcessingTime",
      Value = processing_time,
      Unit = "Seconds",
      Timestamp = Sys.time()
    )
  )
)
```

# Summary

This document covered cloud computing and big data in R:

**AWS Integration**:
- S3 for object storage
- EC2 for compute instances
- Redshift for data warehousing
- EMR for Spark clusters

**Google Cloud Platform**:
- Cloud Storage
- BigQuery for analytics
- Compute Engine

**Apache Spark**:
- sparklyr for distributed computing
- Data manipulation with dplyr
- Machine learning at scale
- Feature engineering pipelines

**Distributed Computing**:
- future for parallel processing
- Chunked data processing
- Out-of-memory solutions (ff, bigmemory)

**Real-World Pipelines**:
- RNA-seq analysis on cloud
- Large-scale variant analysis
- Automated workflows

**Best Practices**:
- Cost optimization
- Security considerations
- Monitoring and logging

**Key Principles**:
- Scale compute and storage independently
- Use appropriate tools for data size
- Optimize for cost and performance
- Implement proper security
- Monitor resource usage

**Next Steps**: Advanced Visualization Techniques

# References

- [AWS SDK for R (paws)](https://paws-r.github.io/)
- [sparklyr Documentation](https://spark.rstudio.com/)
- [Google Cloud R Packages](https://cloudyr.github.io/googleComputeEngineR/)
- [future Package](https://future.futureverse.org/)
- [Apache Spark Documentation](https://spark.apache.org/docs/latest/)
