---
title: "Advanced Visualization Techniques"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Advanced Interactive Visualizations

## plotly for Interactive Plots

```{r}
library(plotly)
library(ggplot2)

# Convert ggplot to plotly
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  theme_minimal()

ggplotly(p)

# Native plotly
plot_ly(data = mtcars,
        x = ~wt,
        y = ~mpg,
        color = ~factor(cyl),
        type = "scatter",
        mode = "markers",
        marker = list(size = 10),
        text = ~paste("Car:", rownames(mtcars),
                     "<br>MPG:", mpg,
                     "<br>Weight:", wt),
        hoverinfo = "text")

# 3D scatter plot
plot_ly(iris,
        x = ~Sepal.Length,
        y = ~Sepal.Width,
        z = ~Petal.Length,
        color = ~Species,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 5))

# Interactive heatmap
plot_ly(z = ~volcano,
        type = "heatmap",
        colorscale = "Viridis")

# Animated plot
library(gapminder)

p <- plot_ly(gapminder,
             x = ~gdpPercap,
             y = ~lifeExp,
             size = ~pop,
             color = ~continent,
             frame = ~year,
             text = ~country,
             hoverinfo = "text",
             type = 'scatter',
             mode = 'markers') %>%
  layout(xaxis = list(type = "log"))

p

# Surface plot
plot_ly(z = ~volcano, type = "surface")
```

## highcharter for Advanced Charts

```{r}
library(highcharter)

# Basic chart
hchart(mtcars, "scatter",
       hcaes(x = wt, y = mpg, group = factor(cyl)))

# Time series
library(quantmod)
getSymbols("AAPL", from = "2020-01-01")

highchart(type = "stock") %>%
  hc_add_series(AAPL$AAPL.Close, name = "Apple Stock")

# Complex multi-series chart
data <- data.frame(
  month = month.abb,
  sales = runif(12, 100, 500),
  costs = runif(12, 50, 300),
  profit = runif(12, 20, 200)
)

highchart() %>%
  hc_xAxis(categories = data$month) %>%
  hc_add_series(name = "Sales", data = data$sales, type = "column") %>%
  hc_add_series(name = "Costs", data = data$costs, type = "column") %>%
  hc_add_series(name = "Profit", data = data$profit, type = "line") %>%
  hc_title(text = "Monthly Performance") %>%
  hc_yAxis(title = list(text = "Amount ($)"))

# Biological data: gene expression heatmap
hchart(cor(mtcars), type = "heatmap") %>%
  hc_colorAxis(stops = color_stops(colors = viridisLite::inferno(10)))
```

## ggiraph for Interactive ggplot2

```{r}
library(ggiraph)

# Interactive scatter plot
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point_interactive(
    aes(tooltip = paste0("Car: ", rownames(mtcars),
                        "\nMPG: ", mpg,
                        "\nWeight: ", wt),
        data_id = rownames(mtcars)),
    size = 3
  ) +
  theme_minimal()

girafe(ggobj = p,
       options = list(
         opts_hover(css = "fill:red;stroke:black;"),
         opts_selection(type = "single", css = "fill:orange;")
       ))

# Interactive bar chart
p <- ggplot(mpg, aes(x = class, fill = class)) +
  geom_bar_interactive(aes(tooltip = after_stat(count),
                           data_id = class)) +
  theme_minimal() +
  theme(legend.position = "none")

girafe(ggobj = p)

# Interactive heatmap
library(reshape2)
cor_mat <- cor(mtcars)
cor_melt <- melt(cor_mat)

p <- ggplot(cor_melt, aes(Var1, Var2, fill = value)) +
  geom_tile_interactive(
    aes(tooltip = paste0(Var1, " vs ", Var2, "\nCorr: ", round(value, 3)),
        data_id = paste(Var1, Var2))
  ) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

girafe(ggobj = p)
```

# 2. Network Visualization

## igraph for Network Analysis

```{r}
library(igraph)

# Create network from edge list
edges <- data.frame(
  from = c("A", "A", "B", "B", "C", "D"),
  to = c("B", "C", "C", "D", "D", "E"),
  weight = c(1, 2, 3, 1, 2, 1)
)

g <- graph_from_data_frame(edges, directed = TRUE)

# Basic plot
plot(g,
     vertex.color = "lightblue",
     vertex.size = 30,
     vertex.label.color = "black",
     edge.arrow.size = 0.5,
     edge.width = E(g)$weight,
     layout = layout_with_fr)

# Network metrics
degree(g)
betweenness(g)
closeness(g)
page_rank(g)$vector

# Community detection
communities <- cluster_louvain(as.undirected(g))
plot(communities, as.undirected(g))

# Advanced visualization
library(ggraph)

ggraph(g, layout = 'fr') +
  geom_edge_link(aes(width = weight),
                 arrow = arrow(length = unit(3, 'mm')),
                 end_cap = circle(3, 'mm'),
                 alpha = 0.7) +
  geom_node_point(size = 10, color = "lightblue") +
  geom_node_text(aes(label = name), repel = TRUE) +
  scale_edge_width(range = c(0.5, 3)) +
  theme_graph()
```

## visNetwork for Interactive Networks

```{r}
library(visNetwork)

# Create nodes and edges
nodes <- data.frame(
  id = 1:10,
  label = paste("Node", 1:10),
  group = sample(c("A", "B", "C"), 10, replace = TRUE),
  value = runif(10, 10, 30),
  title = paste("Node", 1:10, "info")
)

edges <- data.frame(
  from = sample(1:10, 20, replace = TRUE),
  to = sample(1:10, 20, replace = TRUE),
  value = runif(20, 1, 10),
  title = paste("Edge", 1:20)
)

# Interactive network
visNetwork(nodes, edges) %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visPhysics(stabilization = TRUE) %>%
  visEdges(arrows = "to") %>%
  visGroups(groupname = "A", color = "lightblue") %>%
  visGroups(groupname = "B", color = "lightgreen") %>%
  visGroups(groupname = "C", color = "pink") %>%
  visLegend()

# Hierarchical layout
visNetwork(nodes, edges) %>%
  visHierarchicalLayout(direction = "UD",
                       sortMethod = "directed")

# Gene regulatory network example
genes <- data.frame(
  id = 1:50,
  label = paste0("Gene", 1:50),
  group = sample(c("TF", "Target"), 50, replace = TRUE, prob = c(0.2, 0.8)),
  expression = rnorm(50, mean = 5, sd = 2)
)

# Color by expression
library(RColorBrewer)
colors <- colorRampPalette(brewer.pal(9, "RdYlBu"))(50)
genes$color <- colors[rank(genes$expression)]

regulations <- data.frame(
  from = sample(which(genes$group == "TF"), 100, replace = TRUE),
  to = sample(1:50, 100, replace = TRUE),
  type = sample(c("activation", "repression"), 100, replace = TRUE)
)

regulations$color <- ifelse(regulations$type == "activation", "green", "red")

visNetwork(genes, regulations) %>%
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1)) %>%
  visPhysics(solver = "forceAtlas2Based")
```

## Biological Networks

```{r}
# Protein-protein interaction network
library(STRINGdb)
library(igraph)

string_db <- STRINGdb$new(version = "11.5", species = 9606, score_threshold = 400)

# Get proteins
genes <- c("TP53", "BRCA1", "BRCA2", "ATM", "CHEK2")
mapped <- string_db$map(data.frame(gene = genes), "gene")

# Get interactions
interactions <- string_db$get_interactions(mapped$STRING_id)

# Create network
g <- graph_from_data_frame(interactions[, c("from", "to", "combined_score")],
                          directed = FALSE)

# Plot
library(ggraph)

ggraph(g, layout = 'fr') +
  geom_edge_link(aes(alpha = combined_score / 1000)) +
  geom_node_point(size = 8, color = "lightblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_graph() +
  labs(title = "Protein-Protein Interaction Network")

# Pathway network
library(graphite)
library(org.Hs.eg.db)

# Get KEGG pathway
pathways <- pathways("hsapiens", "kegg")
p53_pathway <- pathways[["p53 signaling pathway"]]

# Convert to igraph
g <- pathwayGraph(p53_pathway)

plot(g,
     vertex.size = 5,
     vertex.label.cex = 0.7,
     vertex.color = "lightblue",
     layout = layout_with_kk)
```

# 3. Genome Browsers and Genomic Visualizations

## Gviz for Genomic Tracks

```{r}
library(Gviz)
library(GenomicRanges)

# Define genomic region
chr <- "chr7"
start <- 26700000
end <- 27000000
genome <- "hg38"

# Ideogram track
itrack <- IdeogramTrack(genome = genome, chromosome = chr)

# Genome axis
gtrack <- GenomeAxisTrack()

# Gene track
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

grtrack <- GeneRegionTrack(txdb,
                          genome = genome,
                          chromosome = chr,
                          start = start,
                          end = end,
                          name = "Genes",
                          transcriptAnnotation = "symbol")

# Data track (coverage)
coverage_data <- data.frame(
  start = seq(start, end, length.out = 100),
  end = seq(start, end, length.out = 100) + 1000,
  score = rnorm(100, mean = 50, sd = 10)
)

coverage_gr <- GRanges(seqnames = chr,
                      ranges = IRanges(start = coverage_data$start,
                                      end = coverage_data$end),
                      score = coverage_data$score)

dtrack <- DataTrack(coverage_gr,
                   name = "Coverage",
                   type = "h")

# SNP track
snps <- data.frame(
  pos = sample(start:end, 50),
  id = paste0("rs", 1:50)
)

snp_gr <- GRanges(seqnames = chr,
                 ranges = IRanges(start = snps$pos, width = 1),
                 id = snps$id)

atrack <- AnnotationTrack(snp_gr,
                         name = "SNPs",
                         stacking = "dense")

# Plot all tracks
plotTracks(list(itrack, gtrack, grtrack, dtrack, atrack),
          from = start,
          to = end,
          chromosome = chr)

# Heatmap track for multiple samples
library(GenomicRanges)

samples <- 10
positions <- seq(start, end, length.out = 1000)

expr_matrix <- matrix(rnorm(length(positions) * samples),
                     nrow = length(positions),
                     ncol = samples)

expr_gr <- GRanges(seqnames = chr,
                  ranges = IRanges(start = positions, width = 1000))

htrack <- DataTrack(data = expr_matrix,
                   start = positions,
                   end = positions + 1000,
                   chromosome = chr,
                   genome = genome,
                   name = "Expression",
                   type = "heatmap")

plotTracks(list(itrack, gtrack, grtrack, htrack),
          from = start,
          to = end)
```

## karyoploteR for Whole Genome Visualization

```{r}
library(karyoploteR)

# Basic karyotype plot
kp <- plotKaryotype(genome = "hg38")

# Add data
data <- data.frame(
  chr = paste0("chr", sample(1:22, 1000, replace = TRUE)),
  start = sample(1:100000000, 1000),
  end = sample(1:100000000, 1000) + 10000,
  value = rnorm(1000)
)

data_gr <- makeGRangesFromDataFrame(data, keep.extra.columns = TRUE)

kp <- plotKaryotype(genome = "hg38")
kpPoints(kp, data = data_gr, y = data_gr$value)

# CNV visualization
cnv_data <- data.frame(
  chr = paste0("chr", c(1, 2, 3, 5, 7, 8, 12, 17)),
  start = c(1e6, 5e6, 10e6, 20e6, 15e6, 30e6, 40e6, 25e6),
  end = c(5e6, 15e6, 30e6, 40e6, 35e6, 50e6, 60e6, 45e6),
  cn = c(3, 1, 4, 2, 3, 1, 2, 4)
)

cnv_gr <- makeGRangesFromDataFrame(cnv_data, keep.extra.columns = TRUE)

kp <- plotKaryotype(genome = "hg38", plot.type = 2)

kpPlotRegions(kp,
             data = cnv_gr,
             col = ifelse(cnv_gr$cn > 2, "red", "blue"),
             data.panel = 1)

# Multiple data tracks
kp <- plotKaryotype(genome = "hg38", plot.type = 4)

# Track 1: CNV
kpPlotRegions(kp, data = cnv_gr, col = "red", data.panel = 1)

# Track 2: Gene expression
kpPoints(kp, data = data_gr, y = data_gr$value, data.panel = 2)

# Track 3: Coverage
kpBars(kp, data = data_gr, y1 = abs(data_gr$value), data.panel = 3)
```

## circlize for Circular Plots

```{r}
library(circlize)

# Basic circular plot
circos.par("start.degree" = 90)
circos.initialize(factors = paste0("chr", 1:22),
                 xlim = matrix(c(rep(0, 22), runif(22, 1e7, 2e8)),
                              ncol = 2))

circos.track(ylim = c(0, 1),
            panel.fun = function(x, y) {
              circos.text(CELL_META$xcenter,
                         CELL_META$ycenter,
                         CELL_META$sector.index)
            })

circos.clear()

# Genomic data circular plot
library(GenomicRanges)

# Simulated genomic data
chrom_info <- data.frame(
  chr = paste0("chr", 1:22),
  start = 1,
  end = runif(22, 1e8, 2e8)
)

circos.genomicInitialize(chrom_info)

# Add CNV track
cnv_data <- generateRandomBed(nr = 100, nc = 1)
circos.genomicTrack(cnv_data,
                   panel.fun = function(region, value, ...) {
                     circos.genomicPoints(region, value, pch = 16, cex = 0.5)
                   })

# Add links (translocations)
link_data <- data.frame(
  chr1 = sample(paste0("chr", 1:22), 50, replace = TRUE),
  pos1 = runif(50, 0, 1e8),
  chr2 = sample(paste0("chr", 1:22), 50, replace = TRUE),
  pos2 = runif(50, 0, 1e8)
)

for (i in 1:nrow(link_data)) {
  circos.link(link_data$chr1[i], link_data$pos1[i],
             link_data$chr2[i], link_data$pos2[i],
             col = rgb(1, 0, 0, 0.3))
}

circos.clear()

# Heatmap on circular plot
mat <- matrix(rnorm(100 * 10), nrow = 100)
col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

circos.heatmap(mat,
              split = sample(c("A", "B", "C", "D"), 100, replace = TRUE),
              col = col_fun,
              track.height = 0.4)

circos.clear()
```

# 4. Bioinformatics-Specific Visualizations

## ComplexHeatmap for Advanced Heatmaps

```{r}
library(ComplexHeatmap)
library(circlize)

# Generate example data
set.seed(123)
mat <- matrix(rnorm(1000), nrow = 100)
rownames(mat) <- paste0("Gene", 1:100)
colnames(mat) <- paste0("Sample", 1:10)

# Basic heatmap with clustering
Heatmap(mat,
        name = "Expression",
        show_row_names = FALSE,
        column_title = "Gene Expression Heatmap")

# Advanced heatmap with annotations
# Row annotations
row_anno <- data.frame(
  cluster = sample(c("C1", "C2", "C3"), 100, replace = TRUE),
  pathway = sample(c("P1", "P2", "P3", "P4"), 100, replace = TRUE)
)

ha_row <- rowAnnotation(
  Cluster = row_anno$cluster,
  Pathway = row_anno$pathway,
  col = list(
    Cluster = c("C1" = "red", "C2" = "blue", "C3" = "green"),
    Pathway = c("P1" = "purple", "P2" = "orange", "P3" = "pink", "P4" = "yellow")
  )
)

# Column annotations
col_anno <- data.frame(
  condition = rep(c("Control", "Treatment"), each = 5),
  batch = rep(c("B1", "B2"), 5)
)

ha_col <- HeatmapAnnotation(
  Condition = col_anno$condition,
  Batch = col_anno$batch,
  col = list(
    Condition = c("Control" = "lightblue", "Treatment" = "pink"),
    Batch = c("B1" = "gray", "B2" = "white")
  )
)

# Complex heatmap
Heatmap(mat,
        name = "Expression",
        top_annotation = ha_col,
        right_annotation = ha_row,
        show_row_names = FALSE,
        column_title = "Gene Expression with Annotations",
        row_km = 3,  # K-means clustering with 3 groups
        column_km = 2)

# Multiple heatmaps
mat1 <- mat
mat2 <- matrix(sample(c(0, 1), 100 * 10, replace = TRUE), nrow = 100)
mat3 <- matrix(runif(100 * 10), nrow = 100)

ht1 <- Heatmap(mat1, name = "Expression", show_row_names = FALSE, width = unit(6, "cm"))
ht2 <- Heatmap(mat2, name = "Mutation", col = c("0" = "white", "1" = "black"),
              show_row_names = FALSE, width = unit(3, "cm"))
ht3 <- Heatmap(mat3, name = "CNV", col = colorRamp2(c(0, 0.5, 1), c("blue", "white", "red")),
              show_row_names = FALSE, width = unit(3, "cm"))

ht1 + ht2 + ht3

# Onchoprint for cancer genomics
library(ComplexHeatmap)

mat_onco <- matrix(sample(c("MUT", "AMP", "DEL", ""), 100, replace = TRUE),
                  nrow = 10, ncol = 10)
rownames(mat_onco) <- paste0("Gene", 1:10)
colnames(mat_onco) <- paste0("Sample", 1:10)

col <- c("MUT" = "blue", "AMP" = "red", "DEL" = "green")
alter_fun <- list(
  background = function(x, y, w, h) grid.rect(x, y, w, h, gp = gpar(fill = "#CCCCCC", col = NA)),
  MUT = function(x, y, w, h) grid.rect(x, y, w, h, gp = gpar(fill = "blue", col = NA)),
  AMP = function(x, y, w, h) grid.rect(x, y, w*0.9, h*0.9, gp = gpar(fill = "red", col = NA)),
  DEL = function(x, y, w, h) grid.rect(x, y, w*0.9, h*0.9, gp = gpar(fill = "green", col = NA))
)

oncoPrint(mat_onco,
         alter_fun = alter_fun,
         col = col,
         column_title = "OncoPrint Example")
```

## EnhancedVolcano for Volcano Plots

```{r}
library(EnhancedVolcano)

# Generate DESeq2-like results
set.seed(123)
n_genes <- 5000

results <- data.frame(
  Gene = paste0("Gene", 1:n_genes),
  log2FoldChange = rnorm(n_genes, mean = 0, sd = 1.5),
  pvalue = runif(n_genes, 0, 1),
  padj = runif(n_genes, 0, 1)
)

results$padj[sample(1:n_genes, 100)] <- runif(100, 0, 0.001)
rownames(results) <- results$Gene

# Basic volcano plot
EnhancedVolcano(results,
               lab = rownames(results),
               x = 'log2FoldChange',
               y = 'padj',
               title = 'DEGs Volcano Plot',
               pCutoff = 0.05,
               FCcutoff = 1.0,
               pointSize = 2.0,
               labSize = 3.0)

# Advanced volcano with custom colors
EnhancedVolcano(results,
               lab = rownames(results),
               x = 'log2FoldChange',
               y = 'padj',
               title = 'Advanced Volcano Plot',
               pCutoff = 0.05,
               FCcutoff = 1.5,
               pointSize = 3.0,
               labSize = 4.0,
               col = c('grey30', 'forestgreen', 'royalblue', 'red2'),
               colAlpha = 0.7,
               legendPosition = 'right',
               legendLabels = c('NS', 'Log2 FC', 'p-value', 'p-value & Log2 FC'),
               drawConnectors = TRUE,
               widthConnectors = 0.5,
               selectLab = rownames(results)[1:10])
```

## ggbio for Genomic Data

```{r}
library(ggbio)
library(GenomicRanges)

# Basic track
data(genesymbol, package = "biovizBase")

autoplot(genesymbol, which = GRanges("chr1", IRanges(1e6, 5e6)))

# Multiple tracks
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# SNP data
snps <- GRanges(seqnames = "chr1",
               ranges = IRanges(start = sample(1e6:5e6, 100),
                               width = 1),
               SNP = paste0("rs", 1:100))

# Coverage data
coverage <- GRanges(seqnames = "chr1",
                   ranges = IRanges(start = seq(1e6, 5e6, length.out = 100),
                                   width = 10000),
                   score = rnorm(100, mean = 50, sd = 10))

# Create tracks
p1 <- autoplot(txdb, which = GRanges("chr1", IRanges(1e6, 5e6)))
p2 <- autoplot(snps, geom = "point")
p3 <- autoplot(coverage, geom = "line", aes(y = score))

# Combine
tracks(Gene = p1, SNP = p2, Coverage = p3, heights = c(3, 1, 2))

# Manhattan plot
snp_data <- data.frame(
  CHR = sample(1:22, 10000, replace = TRUE),
  BP = sample(1:200000000, 10000),
  P = 10^-runif(10000, 0, 8),
  SNP = paste0("rs", 1:10000)
)

plotGrandLinear(snp_data, aes(y = -log10(P)))
```

# 5. Spatial Visualizations

## Spatial Transcriptomics

```{r}
library(Seurat)
library(ggplot2)

# Example: 10X Visium data
# Load spatial data
spatial_obj <- Load10X_Spatial(
  data.dir = "path/to/spaceranger/outs",
  filename = "filtered_feature_bc_matrix.h5"
)

# Preprocessing
spatial_obj <- NormalizeData(spatial_obj)
spatial_obj <- FindVariableFeatures(spatial_obj)
spatial_obj <- ScaleData(spatial_obj)
spatial_obj <- RunPCA(spatial_obj)
spatial_obj <- FindNeighbors(spatial_obj, dims = 1:30)
spatial_obj <- FindClusters(spatial_obj, resolution = 0.5)
spatial_obj <- RunUMAP(spatial_obj, dims = 1:30)

# Spatial visualization
SpatialFeaturePlot(spatial_obj, features = "CD3E")

# Multiple features
SpatialFeaturePlot(spatial_obj,
                  features = c("CD3E", "CD8A", "CD4", "CD19"),
                  ncol = 2)

# Clusters on tissue
SpatialDimPlot(spatial_obj, label = TRUE, label.size = 3)

# Interactive spatial plot
SpatialFeaturePlot(spatial_obj,
                  features = "CD3E",
                  interactive = TRUE)

# Custom spatial visualization
coords <- GetTissueCoordinates(spatial_obj)
coords$cluster <- spatial_obj$seurat_clusters

ggplot(coords, aes(x = imagerow, y = imagecol, color = cluster)) +
  geom_point(size = 2) +
  scale_y_reverse() +
  theme_minimal() +
  labs(title = "Spatial Clusters")
```

## Geographic Visualization

```{r}
library(leaflet)

# Basic map
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -73.935242, lat = 40.730610,
            popup = "New York, NY")

# Multiple locations with data
locations <- data.frame(
  lat = runif(100, 40, 45),
  lng = runif(100, -75, -70),
  value = rnorm(100, mean = 50, sd = 10)
)

pal <- colorNumeric(palette = "viridis", domain = locations$value)

leaflet(locations) %>%
  addTiles() %>%
  addCircleMarkers(
    ~lng, ~lat,
    radius = ~value/5,
    color = ~pal(value),
    stroke = FALSE,
    fillOpacity = 0.7,
    popup = ~paste("Value:", round(value, 2))
  ) %>%
  addLegend("bottomright",
           pal = pal,
           values = ~value,
           title = "Value")

# Heatmap overlay
leaflet(locations) %>%
  addTiles() %>%
  addHeatmap(
    lng = ~lng,
    lat = ~lat,
    intensity = ~value,
    blur = 20,
    max = 0.05,
    radius = 15
  )
```

# 6. Animated Visualizations

## gganimate for Animations

```{r}
library(gganimate)
library(ggplot2)

# Simple animation
library(gapminder)

p <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = 'GDP per capita',
       y = 'Life expectancy') +
  transition_time(year) +
  ease_aes('linear')

animate(p, nframes = 100, fps = 10)

# Save animation
anim_save("gapminder.gif", animation = p)

# Biological example: gene expression over time
time_series <- data.frame(
  time = rep(0:24, each = 100),
  gene = rep(paste0("Gene", 1:100), 25),
  expression = rnorm(2500, mean = 5, sd = 2) +
               rep(sin(seq(0, 2*pi, length.out = 25)), each = 100)
)

p <- ggplot(time_series, aes(x = gene, y = expression, fill = expression)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(title = 'Time: {frame_time} hours') +
  transition_time(time)

animate(p, nframes = 50, fps = 5)

# Animated scatter
p <- ggplot(time_series, aes(time, expression, color = gene)) +
  geom_line(alpha = 0.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  transition_reveal(time)

animate(p)
```

## plotly Animations

```{r}
library(plotly)

# Animated scatter
plot_ly(gapminder,
        x = ~gdpPercap,
        y = ~lifeExp,
        size = ~pop,
        color = ~continent,
        frame = ~year,
        text = ~country,
        hoverinfo = "text",
        type = 'scatter',
        mode = 'markers') %>%
  layout(xaxis = list(type = "log"))

# Animated bar chart
df <- data.frame(
  year = rep(2015:2020, each = 5),
  category = rep(LETTERS[1:5], 6),
  value = runif(30, 10, 100)
)

plot_ly(df,
        x = ~category,
        y = ~value,
        frame = ~year,
        type = 'bar') %>%
  animation_opts(frame = 1000, transition = 300)
```

# 7. Dashboard Integration

## flexdashboard Layout

```{r eval=FALSE}
# Save as .Rmd file
---
title: "Genomics Dashboard"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
library(plotly)
library(DT)
```

# Page 1

## Row {data-height=650}

### Chart A
```{r}
plot_ly(mtcars, x = ~wt, y = ~mpg, type = "scatter", mode = "markers")
```

### Chart B
```{r}
plot_ly(iris, x = ~Sepal.Length, y = ~Sepal.Width, color = ~Species,
        type = "scatter", mode = "markers")
```

## Row {data-height=350}

### Table
```{r}
datatable(mtcars, options = list(pageLength = 5))
```

# Page 2

## Column {data-width=600}

### Gene Expression Heatmap
```{r}
mat <- matrix(rnorm(1000), nrow = 100)
plot_ly(z = mat, type = "heatmap")
```

## Column {data-width=400}

### Value Boxes
```{r}
valueBox(value = 1234, caption = "Total Genes", icon = "fa-dna", color = "primary")
```

```{r}
valueBox(value = 567, caption = "DEGs", icon = "fa-chart-line", color = "warning")
```

### Gauge
```{r}
gauge(value = 75, min = 0, max = 100, symbol = '%',
      gaugeSectors(success = c(80, 100), warning = c(40, 79), danger = c(0, 39)))
```
```

## shinyWidgets for Enhanced Interactivity

```{r}
library(shiny)
library(shinyWidgets)

ui <- fluidPage(
  titlePanel("Enhanced Visualization Dashboard"),

  sidebarLayout(
    sidebarPanel(
      pickerInput("dataset",
                 "Choose a dataset:",
                 choices = c("mtcars", "iris", "airquality"),
                 selected = "mtcars"),

      sliderTextInput("range",
                     "Select range:",
                     choices = 1:10,
                     selected = c(3, 7)),

      prettyRadioButtons("plot_type",
                        "Plot type:",
                        choices = c("scatter", "box", "violin"),
                        inline = TRUE,
                        icon = icon("check")),

      switchInput("show_legend",
                 "Show legend",
                 value = TRUE)
    ),

    mainPanel(
      tabsetPanel(
        tabPanel("Plot", plotlyOutput("main_plot")),
        tabPanel("Data", DTOutput("data_table")),
        tabPanel("Summary", verbatimTextOutput("summary"))
      )
    )
  )
)

server <- function(input, output) {
  data <- reactive({
    get(input$dataset)
  })

  output$main_plot <- renderPlotly({
    plot_ly(data(), x = ~get(names(data())[1]),
           y = ~get(names(data())[2]),
           type = "scatter", mode = "markers")
  })

  output$data_table <- renderDT({
    datatable(data())
  })

  output$summary <- renderPrint({
    summary(data())
  })
}

shinyApp(ui, server)
```

# Summary

This document covered advanced visualization techniques:

**Interactive Visualizations**:
- plotly for interactive charts
- highcharter for complex visualizations
- ggiraph for interactive ggplot2

**Network Visualization**:
- igraph and ggraph for network analysis
- visNetwork for interactive networks
- Biological network examples (PPI, pathways)

**Genomic Visualizations**:
- Gviz for genomic tracks
- karyoploteR for whole-genome views
- circlize for circular plots
- ggbio for genomic data

**Specialized Plots**:
- ComplexHeatmap for advanced heatmaps
- EnhancedVolcano for publication-quality volcano plots
- Spatial transcriptomics visualization

**Animated Visualizations**:
- gganimate for ggplot2 animations
- plotly animations
- Time-series biological data

**Dashboard Integration**:
- flexdashboard layouts
- shinyWidgets for enhanced interactivity
- Complete dashboard examples

**Key Principles**:
- Choose appropriate visualization for data type
- Make plots interactive when beneficial
- Use color schemes appropriately
- Ensure accessibility
- Publication-quality output

**Next Steps**: Specialized Statistical Methods

# References

- [plotly for R](https://plotly-r.com/)
- [Gviz User Guide](https://bioconductor.org/packages/release/bioc/vignettes/Gviz/inst/doc/Gviz.html)
- [ComplexHeatmap Complete Reference](https://jokergoo.github.io/ComplexHeatmap-reference/book/)
- [gganimate Documentation](https://gganimate.com/)
- [visNetwork Documentation](https://datastorm-open.github.io/visNetwork/)
