---
title: "Specialized Statistical Methods"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Bayesian Analysis with Stan

## Introduction to Stan

```{r}
# Install
install.packages("rstan")
install.packages("bayesplot")
install.packages("loo")

library(rstan)
library(bayesplot)

# Enable parallel computation
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Simple Linear Regression

```{r}
# Stan model code
stan_model_code <- "
data {
  int<lower=0> N;              // number of observations
  vector[N] x;                  // predictor
  vector[N] y;                  // outcome
}
parameters {
  real alpha;                   // intercept
  real beta;                    // slope
  real<lower=0> sigma;          // error sd
}
model {
  // Priors
  alpha ~ normal(0, 10);
  beta ~ normal(0, 10);
  sigma ~ cauchy(0, 5);

  // Likelihood
  y ~ normal(alpha + beta * x, sigma);
}
generated quantities {
  vector[N] y_pred;             // posterior predictions
  vector[N] log_lik;            // log-likelihood for LOO

  for (i in 1:N) {
    y_pred[i] = normal_rng(alpha + beta * x[i], sigma);
    log_lik[i] = normal_lpdf(y[i] | alpha + beta * x[i], sigma);
  }
}
"

# Prepare data
N <- nrow(mtcars)
stan_data <- list(
  N = N,
  x = mtcars$wt,
  y = mtcars$mpg
)

# Compile and fit model
fit <- stan(
  model_code = stan_model_code,
  data = stan_data,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  thin = 1
)

# Check results
print(fit)
plot(fit)

# Diagnostics
check_hmc_diagnostics(fit)
traceplot(fit, pars = c("alpha", "beta", "sigma"))

# Posterior distributions
posterior <- extract(fit)
str(posterior)

# Plot posterior
library(bayesplot)
mcmc_areas(fit, pars = c("alpha", "beta", "sigma"), prob = 0.95)

# Posterior predictive checks
y_pred <- posterior$y_pred
ppc_dens_overlay(stan_data$y, y_pred[1:100, ])
```

## Hierarchical Model

```{r}
# Hierarchical linear model (varying intercepts)
hierarchical_model <- "
data {
  int<lower=0> N;                    // number of observations
  int<lower=0> J;                    // number of groups
  int<lower=1,upper=J> group[N];     // group indicator
  vector[N] x;                       // predictor
  vector[N] y;                       // outcome
}
parameters {
  vector[J] alpha;                   // group intercepts
  real beta;                         // common slope
  real<lower=0> sigma;               // individual-level error
  real mu_alpha;                     // mean intercept
  real<lower=0> sigma_alpha;         // sd of intercepts
}
model {
  // Hyperpriors
  mu_alpha ~ normal(0, 10);
  sigma_alpha ~ cauchy(0, 5);

  // Priors
  alpha ~ normal(mu_alpha, sigma_alpha);
  beta ~ normal(0, 10);
  sigma ~ cauchy(0, 5);

  // Likelihood
  for (i in 1:N) {
    y[i] ~ normal(alpha[group[i]] + beta * x[i], sigma);
  }
}
"

# Prepare data
stan_data <- list(
  N = nrow(mtcars),
  J = length(unique(mtcars$cyl)),
  group = as.integer(factor(mtcars$cyl)),
  x = mtcars$wt,
  y = mtcars$mpg
)

# Fit model
fit_hier <- stan(
  model_code = hierarchical_model,
  data = stan_data,
  chains = 4,
  iter = 2000
)

# Extract group-level effects
alpha_samples <- extract(fit_hier, "alpha")$alpha
colMeans(alpha_samples)

# Plot group intercepts
mcmc_intervals(fit_hier, regex_pars = "alpha")
```

## Logistic Regression

```{r}
logistic_model <- "
data {
  int<lower=0> N;
  vector[N] x;
  int<lower=0,upper=1> y[N];
}
parameters {
  real alpha;
  real beta;
}
model {
  alpha ~ normal(0, 10);
  beta ~ normal(0, 10);
  y ~ bernoulli_logit(alpha + beta * x);
}
generated quantities {
  vector[N] y_pred;
  for (i in 1:N) {
    y_pred[i] = bernoulli_rng(inv_logit(alpha + beta * x[i]));
  }
}
"

# Example data
mtcars$vs_binary <- mtcars$vs

stan_data <- list(
  N = nrow(mtcars),
  x = mtcars$wt,
  y = mtcars$vs_binary
)

fit_logistic <- stan(
  model_code = logistic_model,
  data = stan_data,
  chains = 4,
  iter = 2000
)

print(fit_logistic)
```

## Model Comparison with LOO

```{r}
library(loo)

# Extract log-likelihood
log_lik_1 <- extract_log_lik(fit, parameter_name = "log_lik")

# Compute LOO
loo_1 <- loo(log_lik_1)
print(loo_1)

# Compare models
loo_compare(loo_1, loo_2)

# WAIC
waic_1 <- waic(log_lik_1)
print(waic_1)
```

## brms: Bayesian Regression Models

```{r}
library(brms)

# Simpler interface to Stan
# Linear regression
fit_brm <- brm(
  mpg ~ wt + hp,
  data = mtcars,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "Intercept"),
    prior(normal(0, 5), class = "b"),
    prior(cauchy(0, 5), class = "sigma")
  ),
  chains = 4,
  iter = 2000
)

summary(fit_brm)
plot(fit_brm)

# Posterior predictive checks
pp_check(fit_brm)

# Hierarchical model
fit_hier_brm <- brm(
  mpg ~ wt + (1 | cyl),
  data = mtcars,
  family = gaussian(),
  chains = 4,
  iter = 2000
)

summary(fit_hier_brm)

# Generalized linear models
fit_logistic_brm <- brm(
  vs ~ wt + hp,
  data = mtcars,
  family = bernoulli(),
  chains = 4,
  iter = 2000
)

# Zero-inflated Poisson
fit_zip <- brm(
  count ~ predictor + (1 | group),
  data = my_data,
  family = zero_inflated_poisson(),
  chains = 4
)

# Robust regression with Student-t errors
fit_robust <- brm(
  y ~ x,
  data = my_data,
  family = student(),
  chains = 4
)
```

# 2. Causal Inference

## Propensity Score Matching

```{r}
library(MatchIt)
library(optmatch)

# Example data
set.seed(123)
n <- 1000
data <- data.frame(
  treat = rbinom(n, 1, 0.5),
  x1 = rnorm(n),
  x2 = rnorm(n),
  x3 = rnorm(n)
)
data$y <- data$treat * 2 + data$x1 + data$x2 + rnorm(n)

# Propensity score matching
m.out <- matchit(
  treat ~ x1 + x2 + x3,
  data = data,
  method = "nearest",
  distance = "logit",
  caliper = 0.1
)

# Check balance
summary(m.out)
plot(m.out, type = "jitter")
plot(m.out, type = "hist")

# Extract matched data
matched_data <- match.data(m.out)

# Estimate treatment effect
library(lmtest)
library(sandwich)

fit <- lm(y ~ treat + x1 + x2 + x3, data = matched_data, weights = weights)
coeftest(fit, vcov. = vcovHC(fit, type = "HC3"))

# Alternative: full matching
m.out_full <- matchit(
  treat ~ x1 + x2 + x3,
  data = data,
  method = "full"
)

# Optimal matching
m.out_optimal <- matchit(
  treat ~ x1 + x2 + x3,
  data = data,
  method = "optimal",
  ratio = 1
)
```

## Inverse Probability Weighting

```{r}
library(ipw)
library(survey)

# Estimate propensity scores
ps_model <- glm(treat ~ x1 + x2 + x3,
               data = data,
               family = binomial())

data$ps <- predict(ps_model, type = "response")

# Calculate IPW weights
data$ipw <- ifelse(data$treat == 1,
                   1 / data$ps,
                   1 / (1 - data$ps))

# Stabilized weights
data$ipw_stab <- ifelse(data$treat == 1,
                        mean(data$treat) / data$ps,
                        (1 - mean(data$treat)) / (1 - data$ps))

# Check weight distribution
summary(data$ipw_stab)
hist(data$ipw_stab)

# Trim extreme weights
data$ipw_trim <- pmin(data$ipw_stab, quantile(data$ipw_stab, 0.99))

# Estimate treatment effect with IPW
design <- svydesign(ids = ~1, weights = ~ipw_trim, data = data)
svyglm(y ~ treat, design = design)

# Check balance
library(cobalt)
bal.tab(treat ~ x1 + x2 + x3,
       data = data,
       weights = "ipw_trim",
       method = "weighting")

love.plot(treat ~ x1 + x2 + x3,
         data = data,
         weights = data$ipw_trim,
         binary = "std")
```

## Difference-in-Differences

```{r}
# Panel data example
panel_data <- data.frame(
  id = rep(1:100, each = 10),
  time = rep(1:10, 100),
  treat = rep(c(rep(0, 50), rep(1, 50)), each = 10),
  post = rep(c(rep(0, 5), rep(1, 5)), 100)
)

panel_data$y <- panel_data$treat * panel_data$post * 3 +
                panel_data$time * 0.5 +
                rnorm(1000)

# Difference-in-differences
did_model <- lm(y ~ treat * post + factor(time), data = panel_data)
summary(did_model)

# With fixed effects
library(fixest)

did_fe <- feols(y ~ treat:post | id + time, data = panel_data)
summary(did_fe)

# Parallel trends test (pre-treatment)
pre_data <- panel_data[panel_data$post == 0, ]
pre_trends <- lm(y ~ treat * time, data = pre_data)
summary(pre_trends)

# Event study
panel_data$rel_time <- panel_data$time - 5  # Treatment at time 5

event_study <- feols(
  y ~ i(rel_time, treat, ref = -1) | id + time,
  data = panel_data
)

iplot(event_study)
```

## Instrumental Variables

```{r}
library(AER)
library(ivreg)

# Example: endogenous treatment
n <- 1000
z <- rnorm(n)  # Instrument
u <- rnorm(n)  # Unobserved confounder

x <- z + u + rnorm(n)  # Endogenous predictor
y <- x + u + rnorm(n)  # Outcome

data_iv <- data.frame(y = y, x = x, z = z)

# Naive OLS (biased)
ols <- lm(y ~ x, data = data_iv)
summary(ols)

# Two-stage least squares
tsls <- ivreg(y ~ x | z, data = data_iv)
summary(tsls, diagnostics = TRUE)

# Multiple instruments
z2 <- rnorm(n)
x <- z + z2 + u + rnorm(n)
y <- x + u + rnorm(n)

data_iv2 <- data.frame(y = y, x = x, z1 = z, z2 = z2)

tsls2 <- ivreg(y ~ x | z1 + z2, data = data_iv2)
summary(tsls2, diagnostics = TRUE)

# Test for weak instruments
# F-statistic should be > 10
first_stage <- lm(x ~ z1 + z2, data = data_iv2)
anova(first_stage)
```

## Regression Discontinuity Design

```{r}
library(rdd)
library(rdrobust)

# Sharp RD design
n <- 1000
x <- rnorm(n)  # Running variable
cutoff <- 0
treat <- ifelse(x >= cutoff, 1, 0)
y <- treat * 2 + x + rnorm(n)

data_rd <- data.frame(x = x, y = y, treat = treat)

# Visualize discontinuity
library(ggplot2)
ggplot(data_rd, aes(x = x, y = y, color = factor(treat))) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  geom_smooth(method = "lm") +
  theme_minimal()

# RD estimation
rd_est <- RDestimate(y ~ x, data = data_rd, cutpoint = cutoff)
summary(rd_est)
plot(rd_est)

# Robust RD with rdrobust
rd_robust <- rdrobust(y = data_rd$y, x = data_rd$x, c = cutoff)
summary(rd_robust)

# Bandwidth selection
rdbwselect(y = data_rd$y, x = data_rd$x, c = cutoff)

# Fuzzy RD (imperfect compliance)
# Treatment is not perfectly determined by cutoff
treat_fuzzy <- rbinom(n, 1, plogis((x - cutoff) * 2))
y_fuzzy <- treat_fuzzy * 2 + x + rnorm(n)

data_rd_fuzzy <- data.frame(x = x, y = y_fuzzy, treat = treat_fuzzy)

rd_fuzzy <- rdrobust(y = data_rd_fuzzy$y,
                    x = data_rd_fuzzy$x,
                    c = cutoff,
                    fuzzy = data_rd_fuzzy$treat)
summary(rd_fuzzy)
```

# 3. Spatial Statistics

## Spatial Autocorrelation

```{r}
library(sp)
library(spdep)

# Example spatial data
set.seed(123)
coords <- cbind(runif(100, 0, 10), runif(100, 0, 10))
colnames(coords) <- c("x", "y")

# Create spatial points
sp_data <- SpatialPointsDataFrame(
  coords = coords,
  data = data.frame(
    value = rnorm(100) + coords[, 1] * 0.5  # Spatial trend
  )
)

# Create spatial weights matrix
# K-nearest neighbors
nb_k <- knn2nb(knearneigh(coords, k = 5))
w_k <- nb2listw(nb_k)

# Distance-based
nb_d <- dnearneigh(coords, 0, 2)
w_d <- nb2listw(nb_d)

# Moran's I test
moran.test(sp_data$value, w_k)

# Moran's I plot
moran.plot(sp_data$value, w_k)

# Local Moran's I (LISA)
local_moran <- localmoran(sp_data$value, w_k)
sp_data$local_i <- local_moran[, 1]
sp_data$local_p <- local_moran[, 5]

# Plot significant clusters
library(ggplot2)
library(sf)

sp_sf <- st_as_sf(sp_data)

ggplot(sp_sf) +
  geom_sf(aes(color = local_i, size = ifelse(local_p < 0.05, 2, 1))) +
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  theme_minimal()

# Geary's C
geary.test(sp_data$value, w_k)
```

## Spatial Regression Models

```{r}
library(spatialreg)

# Generate data with spatial lag
set.seed(123)
n <- 100
coords <- cbind(runif(n, 0, 10), runif(n, 0, 10))
nb <- knn2nb(knearneigh(coords, k = 5))
w <- nb2listw(nb)

x <- rnorm(n)
y_spatial <- 0.5 * lag.listw(w, rnorm(n)) + x + rnorm(n, sd = 0.5)

data_spatial <- data.frame(y = y_spatial, x = x)

# Spatial lag model (SAR)
sar_model <- lagsarlm(y ~ x, data = data_spatial, listw = w)
summary(sar_model)

# Spatial error model (SEM)
sem_model <- errorsarlm(y ~ x, data = data_spatial, listw = w)
summary(sem_model)

# Spatial Durbin model
sdm_model <- lagsarlm(y ~ x, data = data_spatial, listw = w, type = "mixed")
summary(sdm_model)

# Model comparison
AIC(sar_model, sem_model, sdm_model)

# Impacts (direct, indirect, total)
impacts(sar_model, listw = w)
```

## Geostatistics and Kriging

```{r}
library(gstat)
library(sp)

# Generate spatial data
set.seed(123)
coords <- expand.grid(x = seq(0, 10, 0.5), y = seq(0, 10, 0.5))
n <- nrow(coords)

# Simulate spatial process with known variogram
library(geoR)
sim_data <- grf(n = n, grid = coords, cov.pars = c(1, 2))

data_geo <- data.frame(
  x = coords$x,
  y = coords$y,
  z = sim_data$data
)

coordinates(data_geo) <- ~x + y

# Empirical variogram
v <- variogram(z ~ 1, data_geo)
plot(v)

# Fit variogram model
v.fit <- fit.variogram(v, vgm(psill = 1, model = "Exp", range = 2, nugget = 0.1))
plot(v, v.fit)

# Ordinary kriging
grid <- expand.grid(
  x = seq(0, 10, 0.2),
  y = seq(0, 10, 0.2)
)
coordinates(grid) <- ~x + y
gridded(grid) <- TRUE

k <- krige(z ~ 1, data_geo, grid, model = v.fit)

# Plot predictions
library(ggplot2)
k_df <- as.data.frame(k)

ggplot(k_df, aes(x = x, y = y, fill = var1.pred)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Kriging Predictions", fill = "Predicted")

# Prediction variance
ggplot(k_df, aes(x = x, y = y, fill = var1.var)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Prediction Variance", fill = "Variance")

# Universal kriging (with trend)
v.uk <- variogram(z ~ x + y, data_geo)
v.uk.fit <- fit.variogram(v.uk, vgm(1, "Exp", 2, 0.1))

k.uk <- krige(z ~ x + y, data_geo, grid, model = v.uk.fit)
```

## Spatial Point Patterns

```{r}
library(spatstat)

# Create point pattern
set.seed(123)
pp <- rpoispp(lambda = 100)  # Homogeneous Poisson process

plot(pp, main = "Point Pattern")

# Summary statistics
summary(pp)

# Intensity
intensity(pp)

# Quadrat test for homogeneity
quadrat.test(pp, nx = 4, ny = 4)

# K-function (Ripley's K)
K <- Kest(pp)
plot(K)

# L-function (variance-stabilized K)
L <- Lest(pp)
plot(L)

# Pair correlation function
g <- pcf(pp)
plot(g)

# Inhomogeneous point process
# Intensity depends on location
lambda_func <- function(x, y) {
  100 * exp(-((x - 0.5)^2 + (y - 0.5)^2) / 0.1)
}

pp_inhom <- rpoispp(lambda_func)
plot(pp_inhom)

# Estimate intensity
lambda <- density(pp_inhom)
plot(lambda)

# Inhomogeneous K-function
K_inhom <- Kinhom(pp_inhom, lambda = lambda)
plot(K_inhom)

# Fit point process model
# Poisson process
fit_pois <- ppm(pp ~ 1)
summary(fit_pois)

# Inhomogeneous Poisson
fit_inhom <- ppm(pp_inhom ~ polynom(x, y, 2))
summary(fit_inhom)

# Cluster process (Thomas process)
fit_thomas <- kppm(pp ~ 1, clusters = "Thomas")
summary(fit_thomas)
```

# 4. Survival Analysis (Advanced)

## Parametric Survival Models

```{r}
library(survival)
library(flexsurv)

# Generate survival data
set.seed(123)
n <- 200
time <- rweibull(n, shape = 1.5, scale = 100)
status <- rbinom(n, 1, 0.7)
x <- rnorm(n)
group <- sample(c("A", "B"), n, replace = TRUE)

surv_data <- data.frame(time = time, status = status, x = x, group = group)

# Weibull regression
weibull_model <- flexsurvreg(Surv(time, status) ~ x + group,
                             data = surv_data,
                             dist = "weibull")

summary(weibull_model)
plot(weibull_model)

# Log-normal
lnorm_model <- flexsurvreg(Surv(time, status) ~ x + group,
                           data = surv_data,
                           dist = "lognormal")

# Gamma
gamma_model <- flexsurvreg(Surv(time, status) ~ x + group,
                          data = surv_data,
                          dist = "gamma")

# Generalized gamma (most flexible)
gengamma_model <- flexsurvreg(Surv(time, status) ~ x + group,
                             data = surv_data,
                             dist = "gengamma")

# Model comparison
AIC(weibull_model, lnorm_model, gamma_model, gengamma_model)
```

## Frailty Models

```{r}
library(survival)

# Generate clustered survival data
set.seed(123)
n_clusters <- 50
n_per_cluster <- 4

cluster_id <- rep(1:n_clusters, each = n_per_cluster)
frailty <- rep(rgamma(n_clusters, shape = 2, scale = 0.5), each = n_per_cluster)

time <- rexp(n_clusters * n_per_cluster, rate = frailty * 0.01)
status <- rbinom(n_clusters * n_per_cluster, 1, 0.7)
x <- rnorm(n_clusters * n_per_cluster)

frailty_data <- data.frame(
  time = time,
  status = status,
  x = x,
  cluster = cluster_id
)

# Cox model with frailty
frailty_model <- coxph(Surv(time, status) ~ x + frailty(cluster),
                      data = frailty_data)

summary(frailty_model)

# Gamma frailty
frailty_gamma <- coxph(Surv(time, status) ~ x + frailty(cluster, distribution = "gamma"),
                      data = frailty_data)

# Gaussian frailty
frailty_gauss <- coxph(Surv(time, status) ~ x + frailty(cluster, distribution = "gaussian"),
                      data = frailty_data)
```

## Competing Risks

```{r}
library(cmprsk)
library(survival)

# Simulate competing risks data
set.seed(123)
n <- 500
time <- rexp(n, rate = 0.01)
cause <- sample(1:3, n, replace = TRUE, prob = c(0.5, 0.3, 0.2))
status <- ifelse(runif(n) < 0.7, cause, 0)  # 0 = censored

x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)

cr_data <- data.frame(
  time = time,
  status = status,
  x1 = x1,
  x2 = x2
)

# Cumulative incidence function
cif <- cuminc(ftime = cr_data$time,
             fstatus = cr_data$status,
             group = cr_data$x2)

plot(cif, main = "Cumulative Incidence")

# Fine-Gray model for subdistribution hazards
fg_model <- crr(ftime = cr_data$time,
               fstatus = cr_data$status,
               cov1 = cbind(cr_data$x1, cr_data$x2),
               failcode = 1)  # Model for event type 1

summary(fg_model)

# Cause-specific Cox models
# Model for cause 1
cs_model1 <- coxph(Surv(time, status == 1) ~ x1 + x2, data = cr_data)

# Model for cause 2
cs_model2 <- coxph(Surv(time, status == 2) ~ x1 + x2, data = cr_data)

summary(cs_model1)
summary(cs_model2)
```

## Multi-state Models

```{r}
library(mstate)

# Example: illness-death model
# States: 1 = healthy, 2 = ill, 3 = dead

# Transition matrix
tmat <- transMat(x = list(
  c(2, 3),    # From state 1 to states 2, 3
  c(3)        # From state 2 to state 3
), names = c("Healthy", "Ill", "Dead"))

print(tmat)

# Simulate data
set.seed(123)
n <- 200

# Transition 1->2
time_12 <- rexp(n, rate = 0.02)
status_12 <- rbinom(n, 1, 0.6)

# Transition 1->3
time_13 <- rexp(n, rate = 0.01)
status_13 <- rbinom(n, 1, 0.4)

# Transition 2->3
time_23 <- rexp(n, rate = 0.03)
status_23 <- rbinom(n, 1, 0.7)

# Prepare data in long format
ms_data <- data.frame(
  id = rep(1:n, each = 2),
  from = rep(c(1, 1), n),
  to = rep(c(2, 3), n),
  trans = rep(c(1, 2), n),
  Tstart = 0,
  Tstop = c(rbind(time_12, time_13)),
  status = c(rbind(status_12, status_13)),
  x = rep(rnorm(n), each = 2)
)

# Fit multi-state model
msfit <- coxph(Surv(Tstart, Tstop, status) ~ x + strata(trans),
              data = ms_data)

summary(msfit)

# Predict transition probabilities
# New covariate pattern
newdata <- data.frame(x = 0, trans = 1:2, strata = 1:2)

# Cumulative hazards
cumhaz <- msfit(msfit, trans = tmat)

# State probabilities
pstate <- probtrans(cumhaz, predt = 0)
plot(pstate)
```

# 5. Time Series Analysis (Advanced)

## GARCH Models

```{r}
library(rugarch)

# Simulate GARCH(1,1) data
set.seed(123)
spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0))
)

sim_garch <- ugarchpath(spec, n.sim = 1000)
returns <- as.numeric(fitted(sim_garch))

# Fit GARCH model
garch_spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(1, 1))
)

garch_fit <- ugarchfit(spec = garch_spec, data = returns)

# Results
summary(garch_fit)
plot(garch_fit)

# Forecast
garch_forecast <- ugarchforecast(garch_fit, n.ahead = 10)
plot(garch_forecast)

# GJR-GARCH (asymmetric)
gjr_spec <- ugarchspec(
  variance.model = list(model = "gjrGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(1, 1))
)

gjr_fit <- ugarchfit(spec = gjr_spec, data = returns)

# EGARCH
egarch_spec <- ugarchspec(
  variance.model = list(model = "eGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(1, 1))
)

egarch_fit <- ugarchfit(spec = egarch_spec, data = returns)

# Model comparison
infocriteria(garch_fit)
infocriteria(gjr_fit)
infocriteria(egarch_fit)
```

## State Space Models

```{r}
library(KFAS)

# Local level model
set.seed(123)
n <- 100
true_level <- cumsum(rnorm(n, 0, 0.1))
y <- true_level + rnorm(n, 0, 0.5)

# Define state space model
model <- SSModel(
  y ~ SSMtrend(degree = 1, Q = NA),
  H = NA
)

# Fit model
fit <- fitSSM(model, inits = c(0, 0))

# Kalman filter
kf <- KFS(fit$model)

# Plot
plot(y, type = "l", col = "gray")
lines(kf$alphahat, col = "red")
lines(true_level, col = "blue", lty = 2)
legend("topleft",
       legend = c("Observed", "Filtered", "True"),
       col = c("gray", "red", "blue"),
       lty = c(1, 1, 2))

# Forecasting
pred <- predict(fit$model, n.ahead = 10)

# Local linear trend
model_trend <- SSModel(
  y ~ SSMtrend(degree = 2, Q = list(NA, NA)),
  H = NA
)

fit_trend <- fitSSM(model_trend, inits = c(0, 0, 0))
kf_trend <- KFS(fit_trend$model)

# Seasonal component
y_seasonal <- true_level + 5 * sin(2 * pi * (1:n) / 12) + rnorm(n, 0, 0.5)

model_seasonal <- SSModel(
  y_seasonal ~ SSMtrend(degree = 1, Q = NA) +
               SSMseasonal(period = 12, sea.type = "dummy", Q = NA),
  H = NA
)

fit_seasonal <- fitSSM(model_seasonal, inits = rep(0, 3))
kf_seasonal <- KFS(fit_seasonal$model)

plot(y_seasonal, type = "l")
lines(kf_seasonal$alphahat[, 1], col = "red")
```

## Vector Autoregression (VAR)

```{r}
library(vars)

# Multivariate time series
data(Canada)

# Select VAR order
VARselect(Canada, lag.max = 10, type = "const")

# Fit VAR model
var_model <- VAR(Canada, p = 2, type = "const")
summary(var_model)

# Diagnostics
serial.test(var_model, lags.pt = 10, type = "PT.asymptotic")
arch.test(var_model, lags.multi = 5)
normality.test(var_model)

# Stability
roots <- roots(var_model)
all(abs(roots) < 1)  # Should be TRUE for stability

# Granger causality
causality(var_model, cause = "e")

# Impulse response functions
irf_result <- irf(var_model, impulse = "e", response = "prod", n.ahead = 20)
plot(irf_result)

# Forecast error variance decomposition
fevd_result <- fevd(var_model, n.ahead = 20)
plot(fevd_result)

# Forecasting
var_forecast <- predict(var_model, n.ahead = 10, ci = 0.95)
plot(var_forecast)

# Structural VAR
amat <- diag(4)
amat[2, 1] <- NA
amat[3, 1] <- NA
amat[4, 1] <- NA

svar_model <- SVAR(var_model, Amat = amat, Bmat = NULL, max.iter = 100)
summary(svar_model)
```

# Summary

This document covered specialized statistical methods:

**Bayesian Analysis**:
- Stan for custom Bayesian models
- Hierarchical models
- brms for simplified interface
- Model comparison with LOO/WAIC

**Causal Inference**:
- Propensity score matching
- Inverse probability weighting
- Difference-in-differences
- Instrumental variables
- Regression discontinuity

**Spatial Statistics**:
- Spatial autocorrelation (Moran's I, Geary's C)
- Spatial regression models
- Geostatistics and kriging
- Point pattern analysis

**Advanced Survival Analysis**:
- Parametric survival models
- Frailty models
- Competing risks
- Multi-state models

**Advanced Time Series**:
- GARCH models
- State space models
- Vector autoregression

**Key Principles**:
- Choose appropriate method for causal question
- Account for spatial/temporal dependence
- Use Bayesian methods when appropriate
- Validate model assumptions
- Consider uncertainty in estimates

**Next Steps**: Production and Deployment

# References

- [Stan User's Guide](https://mc-stan.org/docs/stan-users-guide/index.html)
- [Causal Inference: The Mixtape](https://mixtape.scunning.com/)
- [Applied Spatial Data Analysis with R](http://www.asdar-book.org/)
- [Survival Analysis: A Self-Learning Text](https://www.springer.com/gp/book/9781441966452)
- [Introduction to Time Series and Forecasting](https://www.springer.com/gp/book/9783319298528)
