---
title: "Production and Deployment"
author: "Computational Bioinformatics Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# 1. Building REST APIs with plumber

## Basic API Setup

```{r}
# Install
install.packages("plumber")

library(plumber)

# Create API file: api.R
#* @apiTitle Gene Expression API
#* @apiDescription API for gene expression analysis

#* Get gene expression
#* @param gene_id Gene identifier
#* @get /expression/<gene_id>
function(gene_id) {
  # Load data
  expr_data <- read.csv("expression_data.csv")

  # Filter for gene
  result <- expr_data[expr_data$gene == gene_id, ]

  if (nrow(result) == 0) {
    stop("Gene not found")
  }

  return(result)
}

#* Calculate differential expression
#* @param group1 Sample IDs for group 1 (comma-separated)
#* @param group2 Sample IDs for group 2 (comma-separated)
#* @post /deseq
function(group1, group2) {
  library(DESeq2)

  g1 <- strsplit(group1, ",")[[1]]
  g2 <- strsplit(group2, ",")[[1]]

  # Load count matrix
  counts <- read.csv("counts.csv", row.names = 1)

  # Create metadata
  coldata <- data.frame(
    condition = c(rep("group1", length(g1)), rep("group2", length(g2))),
    row.names = c(g1, g2)
  )

  # Run DESeq2
  dds <- DESeqDataSetFromMatrix(
    countData = counts[, c(g1, g2)],
    colData = coldata,
    design = ~ condition
  )

  dds <- DESeq(dds)
  results <- results(dds)

  # Return top genes
  top_genes <- head(results[order(results$padj), ], 100)

  return(as.data.frame(top_genes))
}

#* Upload file
#* @param file File to upload
#* @parser multi
#* @post /upload
function(file) {
  # Save file
  write.csv(file, "uploaded_file.csv")

  return(list(
    status = "success",
    filename = file$name,
    size = file$size
  ))
}

# Run API
# pr("api.R") %>% pr_run(port = 8000)
```

## Advanced API Features

```{r}
# api_advanced.R

#* @apiTitle Advanced Bioinformatics API

# Global setup
#* @preempt cors
function(req, res) {
  res$setHeader("Access-Control-Allow-Origin", "*")
  res$setHeader("Access-Control-Allow-Methods", "GET, POST, PUT, DELETE")
  res$setHeader("Access-Control-Allow-Headers", "Content-Type, Authorization")

  if (req$REQUEST_METHOD == "OPTIONS") {
    res$status <- 200
    return(list())
  }
  plumber::forward()
}

# Authentication
#* @filter auth
function(req, res) {
  token <- req$HTTP_AUTHORIZATION

  if (is.null(token) || token != "Bearer secret-token") {
    res$status <- 401
    return(list(error = "Unauthorized"))
  }

  plumber::forward()
}

# Logging
#* @filter logger
function(req) {
  cat(
    format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    req$REQUEST_METHOD,
    req$PATH_INFO,
    "\n"
  )
  plumber::forward()
}

# Error handling
#* @get /error
function() {
  stop("This is an intentional error")
}

# Custom error handler
#* @errorHandler
function(req, res, err) {
  res$status <- 500
  list(
    error = "Internal Server Error",
    message = err$message
  )
}

# Async endpoint with promises
library(promises)
library(future)
plan(multisession)

#* @get /async
function() {
  promise(function(resolve, reject) {
    # Long-running task
    result <- future({
      Sys.sleep(5)
      "Task completed"
    })
    resolve(result)
  })
}

# Streaming response
#* @get /stream
#* @serializer unboxedJSON
function(res) {
  res$setHeader("Content-Type", "application/x-ndjson")

  for (i in 1:10) {
    res$body <- paste0(res$body, jsonlite::toJSON(list(count = i)), "\n")
    Sys.sleep(1)
  }

  res
}

# Caching
library(memoise)

expensive_computation <- memoise(function(x) {
  Sys.sleep(2)
  x^2
})

#* @get /compute/<x:int>
function(x) {
  result <- expensive_computation(x)
  list(input = x, result = result)
}

# Run with swagger UI
# pr("api_advanced.R") %>%
#   pr_run(port = 8000, swagger = TRUE)
```

## API Testing

```{r}
library(httr)
library(testthat)

# Test basic endpoint
test_that("GET /expression works", {
  response <- GET("http://localhost:8000/expression/BRCA1")

  expect_equal(status_code(response), 200)
  expect_true("gene" %in% names(content(response)))
})

# Test POST endpoint
test_that("POST /deseq works", {
  response <- POST(
    "http://localhost:8000/deseq",
    body = list(
      group1 = "sample1,sample2,sample3",
      group2 = "sample4,sample5,sample6"
    ),
    encode = "json"
  )

  expect_equal(status_code(response), 200)
  result <- content(response)
  expect_true(length(result) > 0)
})

# Test authentication
test_that("Auth required", {
  response <- GET(
    "http://localhost:8000/protected",
    add_headers(Authorization = "Bearer wrong-token")
  )

  expect_equal(status_code(response), 401)
})

# Load testing
library(bench)

bench::mark(
  GET("http://localhost:8000/expression/BRCA1"),
  iterations = 100
)
```

## Deployment Options

```{r}
# 1. RStudio Connect
# Deploy directly from RStudio
# rsconnect::deployAPI("api.R", server = "connect.example.com")

# 2. DigitalOcean with plumber
# Install on server:
# sudo apt-get install r-base
# R -e "install.packages('plumber')"

# Create systemd service: /etc/systemd/system/plumber-api.service
# [Unit]
# Description=Plumber API
# After=network.target
#
# [Service]
# Type=simple
# User=rstudio
# WorkingDirectory=/home/rstudio/api
# ExecStart=/usr/bin/Rscript -e "plumber::pr('api.R') %>% plumber::pr_run(host='0.0.0.0', port=8000)"
# Restart=on-failure
#
# [Install]
# WantedBy=multi-user.target

# Start service:
# sudo systemctl start plumber-api
# sudo systemctl enable plumber-api

# 3. Using Docker (see Docker section below)
```

# 2. Docker Containerization

## Basic Dockerfile for R

```dockerfile
# Dockerfile
FROM rocker/r-ver:4.3.0

# Install system dependencies
RUN apt-get update && apt-get install -y \
    libcurl4-openssl-dev \
    libssl-dev \
    libxml2-dev \
    libhdf5-dev \
    libpng-dev \
    && rm -rf /var/lib/apt/lists/*

# Install R packages
RUN R -e "install.packages(c('plumber', 'jsonlite', 'dplyr'))"

# Copy application files
WORKDIR /app
COPY api.R .
COPY data/ ./data/

# Expose port
EXPOSE 8000

# Run API
CMD ["R", "-e", "plumber::pr('api.R') %>% plumber::pr_run(host='0.0.0.0', port=8000)"]
```

## Building and Running

```bash
# Build image
docker build -t r-api:latest .

# Run container
docker run -d -p 8000:8000 --name r-api-container r-api:latest

# View logs
docker logs r-api-container

# Stop container
docker stop r-api-container

# Remove container
docker rm r-api-container
```

## Bioinformatics Pipeline Dockerfile

```dockerfile
# Dockerfile for RNA-seq pipeline
FROM bioconductor/bioconductor_docker:RELEASE_3_17

# Install additional tools
RUN apt-get update && apt-get install -y \
    samtools \
    bcftools \
    bedtools \
    && rm -rf /var/lib/apt/lists/*

# Install Bioconductor packages
RUN R -e "BiocManager::install(c( \
    'DESeq2', \
    'edgeR', \
    'limma', \
    'tximport', \
    'GenomicRanges', \
    'rtracklayer', \
    'Rsubread' \
))"

# Install additional R packages
RUN R -e "install.packages(c( \
    'optparse', \
    'data.table', \
    'ggplot2', \
    'pheatmap' \
))"

# Copy pipeline scripts
WORKDIR /pipeline
COPY scripts/ ./scripts/

# Default command
CMD ["/bin/bash"]
```

## Docker Compose for Multi-Container

```yaml
# docker-compose.yml
version: '3.8'

services:
  api:
    build: ./api
    ports:
      - "8000:8000"
    volumes:
      - ./data:/app/data
      - ./logs:/app/logs
    environment:
      - DATABASE_URL=postgresql://user:pass@db:5432/mydb
    depends_on:
      - db
    restart: unless-stopped

  db:
    image: postgres:15
    environment:
      - POSTGRES_USER=user
      - POSTGRES_PASSWORD=pass
      - POSTGRES_DB=mydb
    volumes:
      - postgres_data:/var/lib/postgresql/data
    ports:
      - "5432:5432"

  redis:
    image: redis:7
    ports:
      - "6379:6379"

  nginx:
    image: nginx:latest
    ports:
      - "80:80"
      - "443:443"
    volumes:
      - ./nginx.conf:/etc/nginx/nginx.conf
      - ./certs:/etc/nginx/certs
    depends_on:
      - api

volumes:
  postgres_data:
```

```bash
# Run all services
docker-compose up -d

# View logs
docker-compose logs -f api

# Stop all services
docker-compose down

# Rebuild
docker-compose up -d --build
```

## Advanced Docker Features

```dockerfile
# Multi-stage build for smaller images
FROM rocker/r-ver:4.3.0 AS builder

RUN apt-get update && apt-get install -y \
    libcurl4-openssl-dev \
    libssl-dev

RUN R -e "install.packages(c('plumber', 'dplyr'))"

# Final stage
FROM rocker/r-ver:4.3.0

COPY --from=builder /usr/local/lib/R/site-library /usr/local/lib/R/site-library

WORKDIR /app
COPY api.R .

EXPOSE 8000
CMD ["R", "-e", "plumber::pr('api.R') %>% plumber::pr_run(host='0.0.0.0', port=8000)"]
```

# 3. Continuous Integration/Continuous Deployment

## GitHub Actions for R Package

```yaml
# .github/workflows/R-CMD-check.yaml
name: R-CMD-check

on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]

jobs:
  R-CMD-check:
    runs-on: ${{ matrix.config.os }}

    name: ${{ matrix.config.os }} (${{ matrix.config.r }})

    strategy:
      fail-fast: false
      matrix:
        config:
          - {os: ubuntu-latest, r: 'release'}
          - {os: ubuntu-latest, r: 'devel'}
          - {os: macOS-latest, r: 'release'}
          - {os: windows-latest, r: 'release'}

    steps:
      - uses: actions/checkout@v3

      - uses: r-lib/actions/setup-r@v2
        with:
          r-version: ${{ matrix.config.r }}

      - uses: r-lib/actions/setup-pandoc@v2

      - name: Query dependencies
        run: |
          install.packages('remotes')
          saveRDS(remotes::dev_package_deps(dependencies = TRUE), ".github/depends.Rds", version = 2)
        shell: Rscript {0}

      - name: Cache R packages
        uses: actions/cache@v3
        with:
          path: ${{ env.R_LIBS_USER }}
          key: ${{ runner.os }}-${{ hashFiles('.github/R-version') }}-1-${{ hashFiles('.github/depends.Rds') }}
          restore-keys: ${{ runner.os }}-${{ hashFiles('.github/R-version') }}-1-

      - name: Install system dependencies
        if: runner.os == 'Linux'
        run: |
          sudo apt-get update
          sudo apt-get install -y libcurl4-openssl-dev

      - name: Install dependencies
        run: |
          remotes::install_deps(dependencies = TRUE)
          remotes::install_cran("rcmdcheck")
        shell: Rscript {0}

      - name: Check
        run: |
          rcmdcheck::rcmdcheck(args = c("--no-manual", "--as-cran"), error_on = "warning", check_dir = "check")
        shell: Rscript {0}

      - name: Upload check results
        if: failure()
        uses: actions/upload-artifact@v3
        with:
          name: ${{ runner.os }}-r${{ matrix.config.r }}-results
          path: check
```

## Test Coverage

```yaml
# .github/workflows/test-coverage.yaml
name: test-coverage

on:
  push:
    branches: [main, master]

jobs:
  test-coverage:
    runs-on: ubuntu-latest

    steps:
      - uses: actions/checkout@v3

      - uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: any::covr
          needs: coverage

      - name: Test coverage
        run: covr::codecov()
        shell: Rscript {0}
```

## Docker Build and Push

```yaml
# .github/workflows/docker.yaml
name: Docker Build and Push

on:
  push:
    branches: [main]
    tags:
      - 'v*'

jobs:
  docker:
    runs-on: ubuntu-latest

    steps:
      - name: Checkout
        uses: actions/checkout@v3

      - name: Docker meta
        id: meta
        uses: docker/metadata-action@v4
        with:
          images: |
            myusername/r-api
          tags: |
            type=ref,event=branch
            type=semver,pattern={{version}}
            type=semver,pattern={{major}}.{{minor}}

      - name: Set up Docker Buildx
        uses: docker/setup-buildx-action@v2

      - name: Login to Docker Hub
        uses: docker/login-action@v2
        with:
          username: ${{ secrets.DOCKERHUB_USERNAME }}
          password: ${{ secrets.DOCKERHUB_TOKEN }}

      - name: Build and push
        uses: docker/build-push-action@v4
        with:
          context: .
          push: true
          tags: ${{ steps.meta.outputs.tags }}
          labels: ${{ steps.meta.outputs.labels }}
          cache-from: type=gha
          cache-to: type=gha,mode=max
```

## Deploy to Cloud

```yaml
# .github/workflows/deploy.yaml
name: Deploy to Cloud

on:
  push:
    branches: [main]

jobs:
  deploy:
    runs-on: ubuntu-latest

    steps:
      - uses: actions/checkout@v3

      - name: Configure AWS credentials
        uses: aws-actions/configure-aws-credentials@v2
        with:
          aws-access-key-id: ${{ secrets.AWS_ACCESS_KEY_ID }}
          aws-secret-access-key: ${{ secrets.AWS_SECRET_ACCESS_KEY }}
          aws-region: us-east-1

      - name: Login to Amazon ECR
        id: login-ecr
        uses: aws-actions/amazon-ecr-login@v1

      - name: Build, tag, and push image to Amazon ECR
        env:
          ECR_REGISTRY: ${{ steps.login-ecr.outputs.registry }}
          ECR_REPOSITORY: r-api
          IMAGE_TAG: ${{ github.sha }}
        run: |
          docker build -t $ECR_REGISTRY/$ECR_REPOSITORY:$IMAGE_TAG .
          docker push $ECR_REGISTRY/$ECR_REPOSITORY:$IMAGE_TAG

      - name: Deploy to ECS
        run: |
          aws ecs update-service --cluster my-cluster --service r-api-service --force-new-deployment
```

# 4. Monitoring and Logging

## Logging Setup

```{r}
library(logger)
library(glue)

# Configure logger
log_threshold(INFO)
log_appender(appender_file("app.log"))

# Custom log layout
log_layout(layout_glue_generator(
  format = '{time} [{level}] {msg}'
))

# Logging in application
log_info("Starting application")

process_data <- function(data) {
  log_debug("Processing {nrow(data)} rows")

  tryCatch({
    result <- expensive_computation(data)
    log_info("Processing completed successfully")
    return(result)
  }, error = function(e) {
    log_error("Processing failed: {e$message}")
    stop(e)
  })
}

# Request logging middleware for plumber
#* @filter logger
function(req) {
  log_info(glue(
    "{req$REQUEST_METHOD} {req$PATH_INFO} - {req$REMOTE_ADDR}"
  ))
  plumber::forward()
}
```

## Performance Monitoring

```{r}
library(profvis)
library(bench)

# Profile code
profvis({
  # Your code here
  data <- read.csv("large_file.csv")
  result <- complex_analysis(data)
})

# Benchmark
bench::mark(
  method1 = {
    result <- method1(data)
  },
  method2 = {
    result <- method2(data)
  },
  iterations = 100
)

# Memory usage
library(pryr)

mem_used()
object_size(large_object)

mem_change({
  large_computation()
})
```

## Health Checks

```{r}
# plumber API with health check
#* @get /health
function() {
  # Check database connection
  db_healthy <- tryCatch({
    con <- DBI::dbConnect(...)
    DBI::dbGetQuery(con, "SELECT 1")
    DBI::dbDisconnect(con)
    TRUE
  }, error = function(e) FALSE)

  # Check data availability
  data_healthy <- file.exists("data/required_file.csv")

  # Overall status
  healthy <- db_healthy && data_healthy

  list(
    status = if (healthy) "healthy" else "unhealthy",
    checks = list(
      database = db_healthy,
      data = data_healthy
    ),
    timestamp = Sys.time()
  )
}

# Readiness check
#* @get /ready
function() {
  # Check if application is ready to serve requests
  ready <- exists("cached_data") && !is.null(cached_data)

  if (ready) {
    return(list(status = "ready"))
  } else {
    res$status <- 503
    return(list(status = "not ready"))
  }
}
```

## Metrics Collection

```{r}
# Prometheus-compatible metrics endpoint
library(openmetrics)

# Counter
requests_total <- counter(
  "api_requests_total",
  "Total number of API requests"
)

# Histogram
request_duration <- histogram(
  "api_request_duration_seconds",
  "Request duration in seconds"
)

# Gauge
active_connections <- gauge(
  "api_active_connections",
  "Number of active connections"
)

# Middleware to collect metrics
#* @filter metrics
function(req, res) {
  start_time <- Sys.time()

  requests_total$inc()
  active_connections$inc()

  result <- plumber::forward()

  duration <- as.numeric(Sys.time() - start_time)
  request_duration$observe(duration)
  active_connections$dec()

  result
}

# Metrics endpoint
#* @get /metrics
#* @serializer text
function() {
  openmetrics::render_metrics()
}
```

# 5. Production Best Practices

## Configuration Management

```{r}
library(config)

# config.yml
# default:
#   database:
#     host: localhost
#     port: 5432
#   api:
#     port: 8000
#
# production:
#   database:
#     host: prod-db.example.com
#     port: 5432
#   api:
#     port: 80

# Load config
cfg <- config::get(config = Sys.getenv("R_CONFIG_ACTIVE", "default"))

# Use configuration
db_host <- cfg$database$host
api_port <- cfg$api$port
```

## Environment Variables

```{r}
# .Renviron file
# DATABASE_URL=postgresql://user:pass@localhost:5432/mydb
# API_KEY=secret-key
# LOG_LEVEL=INFO

# Access in code
db_url <- Sys.getenv("DATABASE_URL")
api_key <- Sys.getenv("API_KEY")
log_level <- Sys.getenv("LOG_LEVEL", "WARN")

# Check required variables
required_vars <- c("DATABASE_URL", "API_KEY")
missing_vars <- required_vars[!nzchar(Sys.getenv(required_vars))]

if (length(missing_vars) > 0) {
  stop(paste("Missing required environment variables:",
            paste(missing_vars, collapse = ", ")))
}
```

## Error Handling and Retries

```{r}
library(purrr)

# Retry with exponential backoff
retry_with_backoff <- function(expr, max_attempts = 3, base_delay = 1) {
  attempt <- 1

  while (attempt <= max_attempts) {
    result <- tryCatch({
      eval(expr)
    }, error = function(e) {
      if (attempt < max_attempts) {
        delay <- base_delay * 2^(attempt - 1)
        log_warn("Attempt {attempt} failed, retrying in {delay}s: {e$message}")
        Sys.sleep(delay)
        NULL
      } else {
        log_error("All {max_attempts} attempts failed: {e$message}")
        stop(e)
      }
    })

    if (!is.null(result)) {
      return(result)
    }

    attempt <- attempt + 1
  }
}

# Usage
result <- retry_with_backoff({
  fetch_data_from_api()
})

# Graceful degradation
get_data <- function() {
  tryCatch({
    # Try primary data source
    fetch_from_database()
  }, error = function(e) {
    log_warn("Primary source failed, using cache: {e$message}")
    # Fallback to cache
    fetch_from_cache()
  })
}
```

## Database Connection Pooling

```{r}
library(pool)
library(DBI)

# Create connection pool
pool <- dbPool(
  drv = RPostgres::Postgres(),
  dbname = "mydb",
  host = "localhost",
  user = "user",
  password = "password",
  minSize = 1,
  maxSize = 10
)

# Use connection from pool
get_data <- function(query) {
  conn <- poolCheckout(pool)
  on.exit(poolReturn(conn))

  DBI::dbGetQuery(conn, query)
}

# Or use pool directly
result <- pool %>%
  dbGetQuery("SELECT * FROM users")

# Close pool when done
onStop(function() {
  poolClose(pool)
})
```

## Rate Limiting

```{r}
library(ratelimitr)

# Limit function calls
limited_api_call <- limit_rate(
  api_call,
  rate(n = 10, period = "1 minute")
)

# Token bucket for plumber
library(R6)

TokenBucket <- R6Class("TokenBucket",
  private = list(
    capacity = NULL,
    tokens = NULL,
    fill_rate = NULL,
    last_fill = NULL
  ),
  public = list(
    initialize = function(capacity, fill_rate) {
      private$capacity <- capacity
      private$tokens <- capacity
      private$fill_rate <- fill_rate
      private$last_fill <- Sys.time()
    },

    consume = function(tokens = 1) {
      # Refill tokens
      now <- Sys.time()
      elapsed <- as.numeric(now - private$last_fill)
      private$tokens <- min(
        private$capacity,
        private$tokens + elapsed * private$fill_rate
      )
      private$last_fill <- now

      # Check if enough tokens
      if (private$tokens >= tokens) {
        private$tokens <- private$tokens - tokens
        return(TRUE)
      } else {
        return(FALSE)
      }
    }
  )
)

# Use in plumber filter
bucket <- TokenBucket$new(capacity = 100, fill_rate = 10)

#* @filter rate_limit
function(req, res) {
  if (!bucket$consume(1)) {
    res$status <- 429
    return(list(error = "Rate limit exceeded"))
  }
  plumber::forward()
}
```

## Caching Strategies

```{r}
library(memoise)
library(cachem)

# Simple memoization
expensive_function <- memoise(function(x) {
  Sys.sleep(2)
  x^2
})

# Cache with expiration
cache <- cache_mem(max_size = 100 * 1024^2)  # 100 MB

cached_function <- function(key) {
  # Check cache
  cached <- cache$get(key)
  if (!is.null(cached)) {
    return(cached)
  }

  # Compute
  result <- expensive_computation(key)

  # Store in cache (expire after 1 hour)
  cache$set(key, result, expire = 3600)

  return(result)
}

# Redis cache
library(redux)

redis <- hiredis()

get_or_compute <- function(key) {
  # Try to get from Redis
  cached <- redis$GET(key)

  if (!is.null(cached)) {
    return(unserialize(cached))
  }

  # Compute
  result <- expensive_computation(key)

  # Store in Redis (expire after 1 hour)
  redis$SETEX(key, 3600, serialize(result, NULL))

  return(result)
}
```

## Background Jobs

```{r}
library(future)
library(promises)

# Async processing
plan(multisession, workers = 4)

process_async <- function(data) {
  future({
    # Long-running computation
    result <- analyze_data(data)
    result
  })
}

# Use in plumber
#* @post /analyze/async
function(req, res) {
  data <- req$body

  # Start async processing
  job_id <- uuid::UUIDgenerate()

  future({
    result <- analyze_data(data)
    # Store result
    saveRDS(result, file.path("results", paste0(job_id, ".rds")))
  })

  # Return job ID immediately
  list(
    job_id = job_id,
    status = "processing"
  )
}

#* @get /results/<job_id>
function(job_id) {
  result_file <- file.path("results", paste0(job_id, ".rds"))

  if (file.exists(result_file)) {
    result <- readRDS(result_file)
    list(
      job_id = job_id,
      status = "completed",
      result = result
    )
  } else {
    list(
      job_id = job_id,
      status = "processing"
    )
  }
}
```

# 6. Security Best Practices

## Input Validation

```{r}
library(checkmate)

validate_input <- function(data) {
  # Check types
  assert_data_frame(data)
  assert_names(colnames(data), must.include = c("id", "value"))

  # Check ranges
  assert_numeric(data$value, lower = 0, upper = 100)

  # Check patterns
  assert_character(data$id, pattern = "^[A-Z]{2}[0-9]{4}$")

  return(TRUE)
}

# Use in API
#* @post /process
function(req) {
  data <- req$body

  tryCatch({
    validate_input(data)
    result <- process_data(data)
    return(result)
  }, error = function(e) {
    res$status <- 400
    return(list(error = e$message))
  })
}
```

## SQL Injection Prevention

```{r}
library(DBI)

# Always use parameterized queries
get_user_safe <- function(user_id) {
  query <- "SELECT * FROM users WHERE id = ?"
  dbGetQuery(con, query, params = list(user_id))
}

# Never do this:
# query <- paste0("SELECT * FROM users WHERE id = ", user_id)
# dbGetQuery(con, query)
```

## API Authentication

```{r}
library(jose)
library(sodium)

# JWT authentication
#* @filter jwt_auth
function(req, res) {
  auth_header <- req$HTTP_AUTHORIZATION

  if (is.null(auth_header)) {
    res$status <- 401
    return(list(error = "Missing authorization header"))
  }

  token <- sub("Bearer ", "", auth_header)

  tryCatch({
    # Verify JWT
    payload <- jwt_decode_sig(token, secret = Sys.getenv("JWT_SECRET"))

    # Add user info to request
    req$user <- payload

    plumber::forward()
  }, error = function(e) {
    res$status <- 401
    return(list(error = "Invalid token"))
  })
}

# API key authentication
valid_api_keys <- c("key1", "key2", "key3")

#* @filter api_key_auth
function(req, res) {
  api_key <- req$HTTP_X_API_KEY

  if (is.null(api_key) || !api_key %in% valid_api_keys) {
    res$status <- 401
    return(list(error = "Invalid API key"))
  }

  plumber::forward()
}
```

## Secrets Management

```{r}
# Never commit secrets to version control!

# Use environment variables
db_password <- Sys.getenv("DB_PASSWORD")

# Or use secrets management service
library(httr)

get_secret <- function(secret_name) {
  response <- GET(
    "https://vault.example.com/v1/secret/data/myapp",
    add_headers(
      "X-Vault-Token" = Sys.getenv("VAULT_TOKEN")
    )
  )

  secrets <- content(response)
  return(secrets$data$data[[secret_name]])
}

db_password <- get_secret("database_password")
```

# Summary

This document covered production deployment and best practices:

**API Development**:
- plumber for REST APIs
- Authentication and authorization
- Error handling and validation
- Testing and documentation

**Containerization**:
- Docker for R applications
- Multi-stage builds
- Docker Compose for multi-container apps
- Kubernetes deployment

**CI/CD**:
- GitHub Actions workflows
- Automated testing and checks
- Docker builds and deployment
- Cloud deployment automation

**Monitoring**:
- Logging with logger
- Performance profiling
- Health checks
- Metrics collection

**Production Best Practices**:
- Configuration management
- Connection pooling
- Rate limiting
- Caching strategies
- Background jobs
- Error handling and retries

**Security**:
- Input validation
- SQL injection prevention
- Authentication strategies
- Secrets management

**Key Principles**:
- Automate everything
- Monitor and log extensively
- Handle errors gracefully
- Secure by default
- Scale horizontally
- Test in production-like environments

**Congratulations!** You've completed all 22 comprehensive R knowledge modules, covering everything from fundamentals to production deployment for data scientists and computational bioinformaticians.

# References

- [plumber Documentation](https://www.rplumber.io/)
- [Docker Documentation](https://docs.docker.com/)
- [GitHub Actions for R](https://github.com/r-lib/actions)
- [R in Production](https://solutions.rstudio.com/)
- [The Twelve-Factor App](https://12factor.net/)
