---
title: "r_Data_types"
author: "Ruichao Wang"
date: "2024-05-01"
output: html_document
---

## In this document, notes of common used R objects will be provided for handling the basics of R.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

### There are many common used types of data in R.

#### Numeric: 1, 3.1415, -5.276, 1e6, 3.14e5
```{r}
# add two number
7.89 + 2.34
-5 + 10
print(3.15e8)
```
```{r}
# scientific form of large numbers
x <- 18989898
format(x, scientific=TRUE)
```
#### Character: used to represent the text string, enclosed in single ('') or double ("") quotation marks
```{r}
print('hello')
"this is R"
print("123")
```
```{r}
# Paste two string of characters
x <- c("Hello", "world")
y <- c("R", "programming")
z <- paste(x, y)
print(z)
```
```{r}
# Upper case and lower case
x <- 'Hello, world'
toupper(x)
tolower(x)
```
```{r}
# using grep 
x <- c('apple', 'banana', 'red', 'orange')
grep("a", x) # return the index of element which have 'a'
```





- Vectors, one dimension, holding elements 
- 



``` {r}
a = 1.0
b = 2.6
c = a + b
print(c)
```

## Including Plots
You can also embed plots, for example:

```{r pressure}
plot(pressure)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
